### 2025-03-15 future_challenges_for_RSEng.tex
use ECEAST template

\documentclass{eceasst}
% This is an empty ECEASST article that can be used as a template
% by authors.

% Required packages
% =================
% Your \usepackage commands go here.

% Article frontmatter
% ===================
\title{Future challenges for Research Software Engineering} % Title of the article
%\short{} % Short title of the article (optional)
\author{
Florian Goth\autref{1},
Leyla Jael Castro\autref{1},
Gerasimos Chourdakis\autref{1},
Simon Christ\autref{1},
Jeremy Cohen\autref{1},
Jean-Noël Grad\autref{1},
Magnus Hagdorn\autref{1},
Toby Hodges\autref{1},
Jan Linxweiler\autref{1},
Frank Löffler\autref{1},
Michele Martone\autref{1},
Jan Philipp Thiele\autref{1},
Nick del Grosso\autref{1}
} % Authors and references to addresses
\institute{Fantasy University\autlabel{1}} % Institutes with labels
\abstract{Admittedly, it’s still only 2025 and we are actually stillworking on establishing Research Software Engineering
as a new domain with the hope of ensuring that it gets the recognition it deserves.
And while it is certainly too early to already herald the end of RSEng,
we should acknowledge that the future is in constant flux and
it is worthwhile to discuss how RSEng will/could change in response to these upcoming challenges in an open discussion.} % Abstract of the article
\keywords{Keywords go here.} % Keywords for the article

\begin{document}
\maketitle

% Main part of your article
% =========================
\section{Introduction}
This opinion article stems from a Bof session at deRSE25 in Karsruhe. Therefore it stems from a discussion
among the participants of this workshop and hence cannot claim that it represents the entire deRSE community.
Nevertheless we hope, that we succeeded in writing up something that
\begin{itemize}
\item you like to read,
\item gives you something to think about,
\item motivates you to work on creating this better future.
\end{itemize}

- Why do we do this

\section{The Situation}
In order to give a more explicit setting, imagine a world ten years from now.
Not too much, to be overwhelmingly far in the future, but distant enough for some changes to occur.
Over this timeframe Kim and Kay
were able to ride the waves of their career in Research Software Engineering and are now pondering early retirement on a sunset blessed beach.
But the world has moved on.
When they look back, what will they see as problems which are now solved?
Which tasks are persisting, where they are now happy that a younger generation is now taking up the baton and carries forward their work.
What issues have newly emerged, where they are now just happy to say: “Oh well… My successor takes care of that”.

\section{Aspects of this future}
It is diffcult to structure the different aspects in describing the future. We decided for the following

\subsection{What do we want to have achieved by 2035}

\subsection{Definition and Evolution of RSEs}
How will the Definition of an RSE evolve in ten years?
What will not change in the essence of an RSE ten years from now?
Here we consider how the topics of RSEng will change in the face of the digitalization of society.
Also here: What will our topics be after version control doesn’t need to be taught anymore? Digitalisation will also feature more heavily in the domain curricula.

\subsection{Education, and changes in the educational system}
FG-NOTE: I wonder if we need a separation between education of RSEs and changes in the education of the broader society.

\subsection{Impact of AI and automation}
Is it better than cheap PhDs?

\subsection{How have academic Institutions transformed by then? ( Point 4 in the issue )}
- Emerging structures:
  - Universities/Institutes with central pools of RSEs, each specializing in data management, visualization, HPC, etc.
  - Potential for closer collaboration between domain scientists and RSEs to share workload effectively.

- Challenges:
  - Budget and administrative hurdles to creating permanent, well-funded RSE teams.
  - Aligning expectations: researchers might not realize that specialized RSE services exist (and are beneficial).

- Networking:
  - Need to share best practices among institutions with successful RSE teams.
  - Encourage knowledge exchange to find models that suit different organizational constraints.

\subsection{Ethics and social consideration}
FG-NOTE: Given what ChatGPT has put in there, I wonder if we need to separate these two items
MM-NOTE: (Flo believes it was from Michele) Are we assuming that progress will bring prosperity? Will we remain constrained by needing to produce products/artifacts?

- Military or other sensitive applications:
  - RSEs (like physicists, mathematicians, etc.) might face ethical dilemmas if their work can be repurposed for weaponry.
  - Importance of having personal “red lines” and awareness of how software can be used.

- Corporate dominance:
  - AI and coding platforms might be increasingly controlled by a small number of large corporations.
  - RSE community might need to develop or advocate for open-source alternatives.

\subsection{What are new tasks of RTPs?}
FG-NOTE: RSEs are one type of specialization of RTPs ins science. Do we see any other?

\subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}
FG-NOTE: By Nick. I've put it here, due to the realtion with the RTP question.

\subsection{Where has abstraction brought us, and which complexities are now hidden behind it?}
FG-NOTE: What would be great, if we can put it in a black box in our daily work and not worry about it any longer?
FG-NOTE: Here we would need a catchy word, that I think should exist out there. Is it Interface? Encapsulation?
Is there a word for "abstracting sth. away?"

\subsection{Impact of Increased Usability in Research Software}

\subsection{What is in this future for RSEs?}
FG-NOTE: I think we can use this part to wrap things a bit up, and describe how this future then looks like.

\section{REALLY Long-term(10+ years) aspects}
Technically this would mean taking everything from the previous section for granted and look
at what becomes now possible.

\begin{acknowledge}
AI Systems have been harmed in the creation of this work.
\end{acknowledge}

% Bibliography with BibTeX
% ========================
\bibliographystyle{eceasst}
%\bibliography{}

\end{document}

### 2025-03-17 future_challenges_for_RSEng.tex
Add bibliography

@@
 \author{
-Florian Goth\autref{1},
-Leyla Jael Castro\autref{1},
-Gerasimos Chourdakis\autref{1},
-Simon Christ\autref{1},
-Jeremy Cohen\autref{1},
-Jean-Noël Grad\autref{1},
-Magnus Hagdorn\autref{1},
-Toby Hodges\autref{1},
-Jan Linxweiler\autref{1},
-Frank Löffler\autref{1},
-Michele Martone\autref{1},
-Jan Philipp Thiele\autref{1},
-Nick del Grosso\autref{1}
+Florian Goth\texorpdfstring{\autref{1}}{},
+Leyla Jael Castro\texorpdfstring{\autref{1}}{},
+Gerasimos Chourdakis\texorpdfstring{\autref{1}}{},
+Simon Christ\texorpdfstring{\autref{1}}{},
+Jeremy Cohen\texorpdfstring{\autref{1}}{},
+Jean-Noël Grad\texorpdfstring{\autref{1}}{},
+Magnus Hagdorn\texorpdfstring{\autref{1}}{},
+Toby Hodges\texorpdfstring{\autref{1}}{},
+Jan Linxweiler\texorpdfstring{\autref{1}}{},
+Frank Löffler\texorpdfstring{\autref{1}}{},
+Michele Martone\texorpdfstring{\autref{1}}{},
+Jan Philipp Thiele\texorpdfstring{\autref{1}}{},
+Nick del Grosso\texorpdfstring{\autref{1}}{}
 } % Authors and references to addresses
@@
 \section{Introduction}
-This opinion article stems from a Bof session at deRSE25 in Karsruhe. Therefore it stems from a discussion
+This opinion article stems from a Bof session at deRSE25 in Karlsruhe\cite{Goth2025EndRSEng}. Therefore it stems from a discussion
 among the participants of this workshop and hence cannot claim that it represents the entire deRSE community.
@@
 \bibliographystyle{eceasst}
-%\bibliography{}
+\bibliography{bibliography/bibliography,extra}
 

### 2025-03-25 future_challenges_for_RSEng.tex
let git ignore a few files
Fix footnotes and GitHub Actions
Summarise workshop output

@@
 % Your \usepackage commands go here.
+\usepackage{fancybox,framed}
 
@@
 } % Authors and references to addresses
-\institute{Fantasy University\autlabel{1}} % Institutes with labels
+\institute{\autlabel{1} Fantasy University} % Institutes with labels
 \abstract{Admittedly, it’s still only 2025 and we are actually stillworking on establishing Research Software Engineering
@@
 \section{Introduction}
-This opinion article stems from a Bof session at deRSE25 in Karlsruhe\cite{Goth2025EndRSEng}. Therefore it stems from a discussion
-among the participants of this workshop and hence cannot claim that it represents the entire deRSE community.
+This opinion article stems from a Bof session at deRSE25 in Karlsruhe\cite{Goth2025EndRSEng}.
+Therefore it stems from a discussion among the participants of this workshop and hence cannot claim that it represents the entire deRSE community.
 Nevertheless we hope, that we succeeded in writing up something that
@@
 \subsection{Definition and Evolution of RSEs}
-How will the Definition of an RSE evolve in ten years?
+\begin{framed}
+How will the definition of an RSE evolve in ten years?
+
 What will not change in the essence of an RSE ten years from now?
-Here we consider how the topics of RSEng will change in the face of the digitalization of society.
-Also here: What will our topics be after version control doesn’t need to be taught anymore? Digitalisation will also feature more heavily in the domain curricula.
+\end{framed}
+% Here we consider how the topics of RSEng will change in the face of the digitalization of society.
+% Also here: What will our topics be after version control doesn't need to be taught anymore?
+% Digitalisation will also feature more heavily in the domain curricula.
+
+The digital skills of domain scientists will increase,
+and software will become easier to interact with,
+reducing reliance on RSEs to train newcomers on the most basic tasks
+(JN: for example, versioning and archiving?).
+User training sessions might focus on more advanced topics
+that cater to a smaller audience, or involve mentoring more frequently.
+
+Regarding the definition of RSEs, we will still have a spectrum of competencies,
+from full professional RSEs to domain scientists who develop research software.
+While the RSE core competencies\cite{Goth2024} might remain unchanged,
+there might be a shift in the responsibilities of RSEs,
+in the form of a partial reallocation of the training budget
+to the other core competencies.
+
+What will \emph{not} change over that timeframe is the RSE's ability to remain
+flexible and embrace change. Keep an eye open and pick up tools and skills
+as needed as we go along.
+Communication and collaboration skills will remain essential to engage
+with domain scientists and help them find technical solutions.
 

### 2025-03-26 future_challenges_for_RSEng.tex

@@
 \usepackage{fancybox,framed}
+\usepackage{footnote}
+\makesavenoteenv[FramedParagraphWithFootnotes]{framed} % for author notes
 
@@
 
+\begin{FramedParagraphWithFootnotes}
+JN-NOTE:
+The Covid-19 pandemic acted as a catalyst for the digital transformation
+of higher education\cite{Bygstad2022}. Students and early-career researchers
+are now more familiar with digital learning tools, which lead to a shift
+in the responsibilities of teachers. To quote from the original study,
+``with so many digital resources at hand,
+the task of the lecturer will be fewer lectures,
+and to act more as a facilitator of resources,
+and to monitor activities and results over time.''\cite{Bygstad2022}
+
+RSE-relevant open educational resources and master's programmes are tracked
+in the Learning and Teaching RSE database\footnote{\url{https://de-rse.org/learn-and-teach/}}
+and UK SSI resources hub\footnote{\url{https://www.software.ac.uk/resource-hub}}.
+The Carpentries now offer specialised software workshops
+for data scientists\footnote{\url{https://datacarpentry.org}},
+librarians\footnote{\url{https://librarycarpentry.org}},
+and soon HPC practitioners\footnote{\url{https://www.hpc-carpentry.org}}.
+\end{FramedParagraphWithFootnotes}
+
+\begin{FramedParagraphWithFootnotes}
+JN-NOTE:
+In recent years, we saw the formation of RSE institutes and advocacy groups, such as
+the UK Software Sustainability Institute\footnote{\url{https://www.software.ac.uk}},
+the US Research Software Sustainability Institute\footnote{\url{https://urssi.us}},
+the European Virtual Institute for Research Software Excellence\footnote{\url{https://everse.software}},
+and the EOSC Opportunity Area Expert Group 7 on Research Software%
+\footnote{\url{https://eosc.eu/opportunity-area-exp/oa7-research-software}}.
+We expect these initiatives to help increase the visibility and accelerate
+the professionalisation of the RSE role.
+In addition, a multitude of RSE communities self-organised at the national
+and regional level to support more viable RSE career paths in academia.
+We can expect RSEs to become more easily accessible,
+possibly through centrally-funded long-term positions at research institutions.
+\end{FramedParagraphWithFootnotes}
+
+\begin{FramedParagraphWithFootnotes}
+JN-NOTE:
+New RSE specialisations might arise.
+HPC-RSE and AI-RSE are quite recent developments,
+whose appeal is stimulated by large public and private
+investments in HPC facilities and AI technologies.
+We could imagine new RSE specialisations to emerge in the fields
+of quantum computing and federated online platforms.
+% European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories
+% European federation platform for HPC: https://eurohpc-ju.europa.eu/paving-way-eurohpc-federation-platform-2024-12-19_en
+% European quantum computing: https://digital-strategy.ec.europa.eu/en/policies/quantum
+\end{FramedParagraphWithFootnotes}
+
 \subsection{Education, and changes in the educational system}

### 2025-03-27 future_challenges_for_RSEng.tex
Add section "Impact of Complexity and Usability in Research Software"
Co-authored-by: Jan Linxweiler <jlinx@users.noreply.github.com>

--- 2025-03-17 future_challenges_for_RSEng.tex
@@
 
-\subsection{Where has abstraction brought us, and which complexities are now hidden behind it?}
-FG-NOTE: What would be great, if we can put it in a black box in our daily work and not worry about it any longer?
-FG-NOTE: Here we would need a catchy word, that I think should exist out there. Is it Interface? Encapsulation?
-Is there a word for "abstracting sth. away?"
+\subsection{Impact of Complexity and Usability in Research Software}
 
-\subsection{Impact of Increased Usability in Research Software}
+The workshop participants found that in earlier learning stages, details may be intentionally hidden, fearing a diminishing capability to handle technical complexity in research environments.
+As a result, RSEs invest more effort in making individual research software components less complex with better user experience.
+This made participants worry about growing mismatch of peoples' expectations.
+Participants identified the over-reliance on generative AI tools as an extreme manifestation of this issue.
+No concrete examples were given on which complexities are now hidden behind abstractions.
+
+We acknowledge but would also like to challenge the observations of the participants.
+We believe that ``we are standing on the shoulders of giants'': By reducing the complexity of individual software components, we are now able to develop systems of complexity and capability not previously possible, allowing the scientific community to address questions not previously in grasp.
+When it comes to teaching, it is true that students need to learn how to handle complexity, but it is also important to learn how to abstract from details (create black boxes) and build hierarchical systems from those abstractions and thereby manage complexity.
 

### 2025-03-31 future_challenges_for_RSEng.tex
Break lines
Extend with a sentence to address both roles and suggest a future

@@
 
-The workshop participants found that in earlier learning stages, details may be intentionally hidden, fearing a diminishing capability to handle technical complexity in research environments.
-As a result, RSEs invest more effort in making individual research software components less complex with better user experience.
+The workshop participants found that in earlier learning stages,
+details may be intentionally hidden,
+fearing a diminishing capability to handle technical complexity in research environments.
+As a result, RSEs invest more effort in making individual research software components
+less complex with better user experience.
 This made participants worry about growing mismatch of peoples' expectations.
@@
 We acknowledge but would also like to challenge the observations of the participants.
-We believe that ``we are standing on the shoulders of giants'': By reducing the complexity of individual software components, we are now able to develop systems of complexity and capability not previously possible, allowing the scientific community to address questions not previously in grasp.
-When it comes to teaching, it is true that students need to learn how to handle complexity, but it is also important to learn how to abstract from details (create black boxes) and build hierarchical systems from those abstractions and thereby manage complexity.
+We believe that ``we are standing on the shoulders of giants'':
+By reducing the complexity of individual software components,
+we are now able to develop systems of complexity and capability not previously possible,
+allowing the scientific community to address questions not previously in grasp.
+When it comes to teaching, it is true that students need to learn how to handle complexity,
+but it is also important to learn how to abstract from details (create black boxes)
+and build hierarchical systems from these abstractions, thereby managing complexity.
+It is clear that we need both RSEs that can dive deep into single projects of significant complexity,
+as well as RSEs that can manage complex systems of almost black-box components,
+and the future of RSE education should aim to prepare students for both roles.
 

### 2025-04-02 future_challenges_for_RSEng.tex

--- 2025-03-26 future_challenges_for_RSEng.tex
@@
 
-\subsection{Where has abstraction brought us, and which complexities are now hidden behind it?}
-FG-NOTE: What would be great, if we can put it in a black box in our daily work and not worry about it any longer?
-FG-NOTE: Here we would need a catchy word, that I think should exist out there. Is it Interface? Encapsulation?
-Is there a word for "abstracting sth. away?"
+\subsection{Impact of Complexity and Usability in Research Software}
 
-\subsection{Impact of Increased Usability in Research Software}
+The workshop participants found that in earlier learning stages,
+details may be intentionally hidden,
+fearing a diminishing capability to handle technical complexity in research environments.
+As a result, RSEs invest more effort in making individual research software components
+less complex with better user experience.
+This made participants worry about growing mismatch of peoples' expectations.
+Participants identified the over-reliance on generative AI tools as an extreme manifestation of this issue.
+No concrete examples were given on which complexities are now hidden behind abstractions.
+
+We acknowledge but would also like to challenge the observations of the participants.
+We believe that ``we are standing on the shoulders of giants'':
+By reducing the complexity of individual software components,
+we are now able to develop systems of complexity and capability not previously possible,
+allowing the scientific community to address questions not previously in grasp.
+When it comes to teaching, it is true that students need to learn how to handle complexity,
+but it is also important to learn how to abstract from details (create black boxes)
+and build hierarchical systems from those abstractions and thereby manage complexity.
 

### 2025-04-02 future_challenges_for_RSEng.tex

@@
 but it is also important to learn how to abstract from details (create black boxes)
-and build hierarchical systems from those abstractions and thereby manage complexity.
+and build hierarchical systems from these abstractions, thereby managing complexity.
+It is clear that we need both RSEs that can dive deep into single projects of significant complexity,
+as well as RSEs that can manage complex systems of almost black-box components,
+and the future of RSE education should aim to prepare students for both roles.
 

### 2025-04-03 future_challenges_for_RSEng.tex

@@
 \institute{\autlabel{1} Fantasy University} % Institutes with labels
-\abstract{Admittedly, it’s still only 2025 and we are actually stillworking on establishing Research Software Engineering
-as a new domain with the hope of ensuring that it gets the recognition it deserves.
-And while it is certainly too early to already herald the end of RSEng,
-we should acknowledge that the future is in constant flux and
-it is worthwhile to discuss how RSEng will/could change in response to these upcoming challenges in an open discussion.} % Abstract of the article
+\abstract{It's 2025 and we are, admittedly, still working on establishing and growing Research Software Engineering (RSE)
+as a domain in its own right, with the hope of ensuring that it gets the recognition it deserves.
+Nonetheless, as research becomes ever more computational in nature, there are already people suggesting that RSE will
+cease to exist as a separate field, and the specialist skills that RSEs provide to support and undertake research will
+become a core part of the tooling and skill set of researchers in almost all domains.
+Are we already approaching "the end of Research Software Engineering"?! While we believe that it is certainly too early to
+already herald the demise of RSE as a separate field,
+we also recognise the need to acknowledge that the future is in constant flux and
+it is worthwhile discussing how RSE will/could change in response to these upcoming challenges in an open discussion.} % Abstract of the article
 \keywords{Keywords go here.} % Keywords for the article

### 2025-04-03 future_challenges_for_RSEng.tex
Adding some content on where we want to be in 2035.

--- 2025-04-02 future_challenges_for_RSEng.tex
@@
 \subsection{What do we want to have achieved by 2035}
+
+Looking ahead over a 10 year timeframe, we hope that the RSE community will have
+supported a number of changes and developments that help to make research software outputs
+more sustainable, robust and maintainable. We see this happening in a number of ways. However,
+we want to ensure that one of the core achievements made by the RSE community in this time
+frame is putting the infrastructure and training opportunities in place, alongside the necessary
+advocacy, to ensure that researchers receive core software skills as part of their scientific training.
+With development of software tools or scripts becoming an increasingly important part of everyday
+research activities, this is vital and it will allow RSEs to focus on other more advanced and specialist
+areas. In this context, we do not foresee the demise of RSE as an independent role. However, we
+do see the work of RSEs developing and shifting as the research community changes.
+
+The specialist skills that RSEs provide take a significant amount of time to develop and maintain.
+We would, therefore, not expect domain researchers to also become experts in a range of
+advanced RSE topics. 
+At the same time, we do recognise that the application of core software development best practices
+to enhance and address problems with research codebases, that can often make up the bulk of
+current RSE work, will change. As noted above, these are skills that we
+would expect any researcher who writes code to have in 10 years time. 
+The rapid development of AI means that researchers are likely to have access to tooling that can assist
+them in writing robust, sustainable code, lowering the barrier to developing code, or at least changing the
+required skill set to some extent.
+They can also be expected to have an understanding of frameworks to support testing, packaging and
+deployment of their code. Again, all these skills will, we anticipate, be gained as part of training
+infrastructure that will provide domain scientists with software development expertise as part of their
+core scientific training.
+
+So where do RSEs fit into this picture? We expect to see RSEs providing technical skills and input
+to research activities in range of different areas, including the following:
+\begin{itemize}
+  \item High-quality, robust architecture and design for research software
+  \item Development of software for novel and emerging architectures
+  \item Application of specialist numerical and statistical methods within software
+  \item Optimisation of codes for use on high-performance computing infrastructure
+  \item Green computing - efficient implementations of algorithms/methods and use of hardware
+  \item ...
+\end{itemize}
 

### 2025-04-05 future_challenges_for_RSEng.tex

--- 2025-04-02 future_challenges_for_RSEng.tex
@@
 \subsection{Ethics and social consideration}
+The ethics and values of the Research Software Engineer are grounded in their practice.
+They commit to the health, safety and welfare of the public and act in the interest of society,
+their employer and their clients \cite{Goth2023}.
+Sometimes these values are in direct conflict with the aims of their institution. 
+For example, their PI might insist on quickly developing some software without adhering to current
+best practices. 
+Another example might be that they are developing code that can also be used for military purposes,
+such as numerical solvers or computer vision systems.
+Like engineers, RSEs take the responsibility for the software they develop.
+RSEs need to be aware of these issues and develop their own ``red lines''.
+
+RSEs are publicly funded.
+Like data collected by scientists should adhere to the FAIR principles \cite{FAIR}, research software developed by RSEs
+should adhere to the FAIR4RS principles \cite{FAIR4RS}.
+
+A code of ethics and professional conduct for RSEs should be developed by institutions together with RSEs themselves.
+This can then form the basis for institutional policies and guidelines.
+
+\begin{FramedParagraphWithFootnotes}  
 FG-NOTE: Given what ChatGPT has put in there, I wonder if we need to separate these two items
@@
   - RSE community might need to develop or advocate for open-source alternatives.
+\end{FramedParagraphWithFootnotes}
 

### 2025-04-07 future_challenges_for_RSEng.tex
use semantic line breaks, no content change
Update future_challenges_for_RSEng.tex
Co-authored-by: Gerasimos Chourdakis <gerasimos.chourdakis@ipvs.uni-stuttgart.de>

@@
 The ethics and values of the Research Software Engineer are grounded in their practice.
-They commit to the health, safety and welfare of the public and act in the interest of society,
-their employer and their clients \cite{Goth2023}.
+They commit to the health, safety and welfare of the public and act in the interest of society, their employer and their clients \cite{Goth2023}.
 Sometimes these values are in direct conflict with the aims of their institution.
-For example, their PI might insist on quickly developing some software without adhering to current
-best practices. 
-Another example might be that they are developing code that can also be used for military purposes,
-such as numerical solvers or computer vision systems.
+For example, their PI might insist on quickly developing some software without adhering to current best practices.
+Another example might be that they are developing code that can also be used for military purposes, such as numerical solvers or computer vision systems.
 Like engineers, RSEs take the responsibility for the software they develop.
@@
 RSEs are publicly funded.
-Like data collected by scientists should adhere to the FAIR principles \cite{FAIR}, research software developed by RSEs
+Similarly to how data collected by scientists should adhere to the FAIR principles \cite{FAIR}, research software developed by RSEs
 should adhere to the FAIR4RS principles \cite{FAIR4RS}.

### 2025-04-07 future_challenges_for_RSEng.tex
use semantic line breaks, no content change
avoid potential line break before citation

--- 2025-04-05 future_challenges_for_RSEng.tex
@@
 RSEs are publicly funded.
-Like data collected by scientists should adhere to the FAIR principles \cite{FAIR}, research software developed by RSEs
-should adhere to the FAIR4RS principles \cite{FAIR4RS}.
+Similarly to how data collected by scientists should adhere to the FAIR principles \cite{FAIR}, research software developed by RSEs should adhere to the FAIR4RS principles~\cite{FAIR4RS}.
 

### 2025-04-08 future_challenges_for_RSEng.tex
Add workshop contributions. Add some ideas.

--- 2025-04-05 future_challenges_for_RSEng.tex
@@
 
-\subsection{How have academic Institutions transformed by then? ( Point 4 in the issue )}
-- Emerging structures:
-  - Universities/Institutes with central pools of RSEs, each specializing in data management, visualization, HPC, etc.
-  - Potential for closer collaboration between domain scientists and RSEs to share workload effectively.
-
-- Challenges:
-  - Budget and administrative hurdles to creating permanent, well-funded RSE teams.
-  - Aligning expectations: researchers might not realize that specialized RSE services exist (and are beneficial).
-
-- Networking:
-  - Need to share best practices among institutions with successful RSE teams.
-  - Encourage knowledge exchange to find models that suit different organizational constraints.
-
 \subsection{Ethics and social consideration}
-The ethics and values of the Research Software Engineer are grounded in their practice.
-They commit to the health, safety and welfare of the public and act in the interest of society,
-their employer and their clients \cite{Goth2023}.
-Sometimes these values are in direct conflict with the aims of their institution. 
-For example, their PI might insist on quickly developing some software without adhering to current
-best practices. 
-Another example might be that they are developing code that can also be used for military purposes,
-such as numerical solvers or computer vision systems.
-Like engineers, RSEs take the responsibility for the software they develop.
-RSEs need to be aware of these issues and develop their own ``red lines''.
-
-RSEs are publicly funded.
-Like data collected by scientists should adhere to the FAIR principles \cite{FAIR}, research software developed by RSEs
-should adhere to the FAIR4RS principles \cite{FAIR4RS}.
-
-A code of ethics and professional conduct for RSEs should be developed by institutions together with RSEs themselves.
-This can then form the basis for institutional policies and guidelines.
-
-\begin{FramedParagraphWithFootnotes}  
 FG-NOTE: Given what ChatGPT has put in there, I wonder if we need to separate these two items
@@
   - RSE community might need to develop or advocate for open-source alternatives.
-\end{FramedParagraphWithFootnotes}
+
+\subsection{How have academic Institutions transformed by then? ( Point 4 in the issue )}
+\subsubsection{What do we want to have achieved}
+We hope to see that universities use the presence of an RSE Department actively as a means to differentiate themselves from other universities.
+We also hope, that specialized clusters form, with certain RSE units specializing in certain tasks like data management, HPC, or visualization.
+At universities which have them, we want to see, that researchers naturally accept them as part of their used services.
+
+\subsubsection{Which challenges will still persist}
+The workshop participants expect that there will still be budgeting and administrative hurdles to creating
+permanent and well-funded RSE teams.
+While we hope that the barrier to researchers diminishes, the workshop participants
+still expect some work to do. We hope that the compartmentalization that occurs will help researchers reprioritize their workload.
+An upcoming challenge, will be that with the emergence of more dedicated RSE units,
+networking between them will come into the foreground, because there should be a strong need to share best practices
+among institutions with succesful RSE Teams.
+On a more organisational level, we expect the sharing of models, to fit the different organizational constraints.
+
+What are open questions here?
+Organizations will have to come up with models for hiring people. While
+a standardized RSE masters makes it easier, it will be interesting to see, if
+RSE are still often hired inhouse or if it changes.
 
@@
 \subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}
-FG-NOTE: By Nick. I've put it here, due to the realtion with the RTP question.
+FG-TEXT-START: The RSE is probably one of the first professionalizations in the academic midlevel that is able
+to collectively voice their concerns. This enables the possibility to reshape universities.
+If more people are employed at the midlevel, this makes a steadier pool of people
+available that can integrate new people that are regularly coming into departments and know the local facilities.
+On the other hand an upcoming challenge can be, how universities decide to design the contracts of people employed
+in RSE units. will they be permanent, or will there also be an up-or-out mentality?
+FG-TEXT-END
+
+FG-NOTE: By Nick. I've put it here, due to the relation with the RTP question.
 

### 2025-04-14 future_challenges_for_RSEng.tex
fix up sentence.

--- 2025-04-07 future_challenges_for_RSEng.tex
@@
 \institute{\autlabel{1} Fantasy University} % Institutes with labels
-\abstract{Admittedly, it’s still only 2025 and we are actually stillworking on establishing Research Software Engineering
-as a new domain with the hope of ensuring that it gets the recognition it deserves.
-And while it is certainly too early to already herald the end of RSEng,
-we should acknowledge that the future is in constant flux and
-it is worthwhile to discuss how RSEng will/could change in response to these upcoming challenges in an open discussion.} % Abstract of the article
+\abstract{It's 2025 and we are, admittedly, still working on establishing and growing Research Software Engineering (RSE)
+as a domain in its own right, with the hope of ensuring that it gets the recognition it deserves.
+Nonetheless, as the use of digital tools becomes ever more pervasive in a researchers day-to-day tasks, there are already people suggesting that RSE will
+cease to exist as a separate field, and the specialist skills that RSEs provide to support and undertake research will
+become a core part of the tooling and skill set of researchers in almost all domains.
+Are we already approaching "the end of Research Software Engineering"?! While we believe that it is certainly too early to
+already herald the demise of RSE as a separate field,
+we also recognise the need to acknowledge that the future is in constant flux and
+it is worthwhile discussing how RSE will/could change in response to these upcoming challenges in an open discussion.} % Abstract of the article
 \keywords{Keywords go here.} % Keywords for the article
@@
 \subsection{Ethics and social consideration}
-The ethics and values of the Research Software Engineer are grounded in their practice.
-They commit to the health, safety and welfare of the public and act in the interest of society, their employer and their clients \cite{Goth2023}.
-Sometimes these values are in direct conflict with the aims of their institution.
-For example, their PI might insist on quickly developing some software without adhering to current best practices.
-Another example might be that they are developing code that can also be used for military purposes, such as numerical solvers or computer vision systems.
-Like engineers, RSEs take the responsibility for the software they develop.
-RSEs need to be aware of these issues and develop their own ``red lines''.
-
-RSEs are publicly funded.
-Similarly to how data collected by scientists should adhere to the FAIR principles \cite{FAIR}, research software developed by RSEs should adhere to the FAIR4RS principles~\cite{FAIR4RS}.
-
-A code of ethics and professional conduct for RSEs should be developed by institutions together with RSEs themselves.
-This can then form the basis for institutional policies and guidelines.
-
-\begin{FramedParagraphWithFootnotes}  
 FG-NOTE: Given what ChatGPT has put in there, I wonder if we need to separate these two items
@@
   - RSE community might need to develop or advocate for open-source alternatives.
-\end{FramedParagraphWithFootnotes}
 

### 2025-04-14 future_challenges_for_RSEng.tex
elaborate intro parts

@@
 \end{itemize}
-
-- Why do we do this
+In order to structure the discussion, we decided on a imaginable setting ten years from now. 
+In this world we will consider the dimensions of 
+\begin{itemize}
+\item Evolution of the term RSEs
+\item The development of academic institutions
+\item How will academic education have transformed by then
+\item Ethics and social considerations
+\end{itemize}
+In each aspect we will quickly summarise on what we want to have achieved by then, and what will be the challenges that arise from that.
 
@@
 \section{Aspects of this future}
-It is diffcult to structure the different aspects in describing the future. We decided for the following
+It is difficult to structure the different aspects in describing the future. We decided to consider
+\begin{itemize}
+\item The evolution of the term RSE
+\item How is education reshaped by the digital transformation
+\item What will the impact of AI be to the work of RSEs?
+\item Changes to the Academic Institutions
+\item Ethics and social considerations
+\item Impact of alternate career paths for RTPs
+\end{itemize}
 
-\subsection{What do we want to have achieved by 2035}
+%\subsection{What do we want to have achieved by 2035}
 

### 2025-04-14 future_challenges_for_RSEng.tex
Update future_challenges_for_RSEng.tex
Co-authored-by: Gerasimos Chourdakis <gerasimos.chourdakis@ipvs.uni-stuttgart.de>

--- 2025-04-08 future_challenges_for_RSEng.tex
@@
 still expect some work to do. We hope that the compartmentalization that occurs will help researchers reprioritize their workload.
-An upcoming challenge, will be that with the emergence of more dedicated RSE units,
+An upcoming challenge will be that, with the emergence of more dedicated RSE units,
 networking between them will come into the foreground, because there should be a strong need to share best practices

### 2025-04-14 future_challenges_for_RSEng.tex
Update future_challenges_for_RSEng.tex
Co-authored-by: Jan Philipp Thiele <54978337+jpthiele@users.noreply.github.com>
work in feedback

@@
 We hope to see that universities use the presence of an RSE Department actively as a means to differentiate themselves from other universities.
+We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department to oversee specific projects[FIXME: cite group paper].
 We also hope, that specialized clusters form, with certain RSE units specializing in certain tasks like data management, HPC, or visualization.
@@
 a standardized RSE masters makes it easier, it will be interesting to see, if
-RSE are still often hired inhouse or if it changes.
+RSE are still often hired by word of mouth or if the master's really makes it more comparable.
 

### 2025-04-14 future_challenges_for_RSEng.tex
add a note about outsourcing

@@
 among institutions with succesful RSE Teams.
-On a more organisational level, we expect the sharing of models, to fit the different organizational constraints.
+On a more organisational level, we expect the sharing of models, to fit the different organisational constraints.
 
@@
 FG-TEXT-END
+
+FG-TEXT-START:
+As outlined in [cite RSE group paper] not all universities need to opt for strengthening
+the academic midlevel. Some might have chosen to set up external companies to outsource typical RSE tasks,
+thereby providing career paths outside of academia.
+FG-TEXT-END:
 

### 2025-04-16 future_challenges_for_RSEng.tex

--- 2025-04-14 future_challenges_for_RSEng.tex
@@
 
+\subsection{How have academic Institutions transformed by then? ( Point 4 in the issue )}
+- Emerging structures:
+  - Universities/Institutes with central pools of RSEs, each specializing in data management, visualization, HPC, etc.
+  - Potential for closer collaboration between domain scientists and RSEs to share workload effectively.
+
+- Challenges:
+  - Budget and administrative hurdles to creating permanent, well-funded RSE teams.
+  - Aligning expectations: researchers might not realize that specialized RSE services exist (and are beneficial).
+
+- Networking:
+  - Need to share best practices among institutions with successful RSE teams.
+  - Encourage knowledge exchange to find models that suit different organizational constraints.
+
 \subsection{Ethics and social consideration}
+The ethics and values of the Research Software Engineer are grounded in their practice.
+They commit to the health, safety and welfare of the public and act in the interest of society, their employer and their clients \cite{Goth2023}.
+Sometimes these values are in direct conflict with the aims of their institution.
+For example, their PI might insist on quickly developing some software without adhering to current best practices.
+Another example might be that they are developing code that can also be used for military purposes, such as numerical solvers or computer vision systems.
+Like engineers, RSEs take the responsibility for the software they develop.
+RSEs need to be aware of these issues and develop their own ``red lines''.
+
+RSEs are publicly funded.
+Similarly to how data collected by scientists should adhere to the FAIR principles \cite{FAIR}, research software developed by RSEs should adhere to the FAIR4RS principles~\cite{FAIR4RS}.
+
+A code of ethics and professional conduct for RSEs should be developed by institutions together with RSEs themselves.
+This can then form the basis for institutional policies and guidelines.
+
+\begin{FramedParagraphWithFootnotes}  
 FG-NOTE: Given what ChatGPT has put in there, I wonder if we need to separate these two items
@@
   - RSE community might need to develop or advocate for open-source alternatives.
-
-\subsection{How have academic Institutions transformed by then? ( Point 4 in the issue )}
-\subsubsection{What do we want to have achieved}
-We hope to see that universities use the presence of an RSE Department actively as a means to differentiate themselves from other universities.
-We also hope, that specialized clusters form, with certain RSE units specializing in certain tasks like data management, HPC, or visualization.
-At universities which have them, we want to see, that researchers naturally accept them as part of their used services.
-
-\subsubsection{Which challenges will still persist}
-The workshop participants expect that there will still be budgeting and administrative hurdles to creating
-permanent and well-funded RSE teams.
-While we hope that the barrier to researchers diminishes, the workshop participants
-still expect some work to do. We hope that the compartmentalization that occurs will help researchers reprioritize their workload.
-An upcoming challenge will be that, with the emergence of more dedicated RSE units,
-networking between them will come into the foreground, because there should be a strong need to share best practices
-among institutions with succesful RSE Teams.
-On a more organisational level, we expect the sharing of models, to fit the different organizational constraints.
-
-What are open questions here?
-Organizations will have to come up with models for hiring people. While
-a standardized RSE masters makes it easier, it will be interesting to see, if
-RSE are still often hired inhouse or if it changes.
+\end{FramedParagraphWithFootnotes}
 
@@
 \subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}
-FG-TEXT-START: The RSE is probably one of the first professionalizations in the academic midlevel that is able
-to collectively voice their concerns. This enables the possibility to reshape universities.
-If more people are employed at the midlevel, this makes a steadier pool of people
-available that can integrate new people that are regularly coming into departments and know the local facilities.
-On the other hand an upcoming challenge can be, how universities decide to design the contracts of people employed
-in RSE units. will they be permanent, or will there also be an up-or-out mentality?
-FG-TEXT-END
-
-FG-NOTE: By Nick. I've put it here, due to the relation with the RTP question.
+FG-NOTE: By Nick. I've put it here, due to the realtion with the RTP question.
 

### 2025-04-16 future_challenges_for_RSEng.tex
some color and structuring boxes

@@
 \makesavenoteenv[FramedParagraphWithFootnotes]{framed} % for author notes
+\usepackage{color}
+\usepackage[dvipsnames]{xcolor}
+\usepackage[most]{tcolorbox}
+
+\newtcbtheorem[number within=section]{whatis}{What is 2035}{
+  enhanced,
+  segmentation engine=empty,
+%  beamer,
+fuzzy shadow={2mm}{-1mm}{0mm}{0.1mm}%
+{black!50!white},
+  sharp corners,
+  attach boxed title to top left={
+    yshifttext=-1mm
+  },
+%  halign lower=flush center,
+  colback=white,
+  colframe=Maroon!75!white,
+  fonttitle=\bfseries,
+  attach boxed title to top left={yshift=-2mm,xshift=3mm},
+  fontlower=\itshape,
+  boxed title style={
+    sharp corners,
+    size=small,
+    colback=Maroon!75!black,
+    colframe=Maroon!75!black,
+  } 
+}{ist}
+
+\newtcbtheorem[number within=section]{story}{Story}{
+  enhanced,
+  segmentation engine=empty,
+%  beamer,
+fuzzy shadow={2mm}{-1mm}{0mm}{0.1mm}%
+{black!50!white},
+  sharp corners,
+  attach boxed title to top right={
+    yshifttext=-1mm
+  },
+%  halign lower=flush center,
+  colback=white,
+  colframe=ForestGreen!75!white,
+  fonttitle=\bfseries,
+  attach boxed title to top right={yshift=-2mm,xshift=3mm},
+  fontlower=\itshape,
+  boxed title style={
+    sharp corners,
+    size=small,
+    colback=ForestGreen!75!black,
+    colframe=ForestGreen!75!black,
+  }
+}{str}
 
@@
 \subsection{What are new tasks of RTPs?}
+\begin{story}{text 1 }{text 2}
  FG-NOTE: RSEs are one type of specialization of RTPs ins science. Do we see any other?
+\end{story}
 
 \subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}
+\begin{whatis}{text 1 }{text 2}
 FG-NOTE: By Nick. I've put it here, due to the realtion with the RTP question.
+\end{whatis}
 

### 2025-04-16 future_challenges_for_RSEng.tex

@@
 \makesavenoteenv[FramedParagraphWithFootnotes]{framed} % for author notes
-\usepackage{color}
-\usepackage[dvipsnames]{xcolor}
-\usepackage[most]{tcolorbox}
-
-\newtcbtheorem[number within=section]{whatis}{What is 2035}{
-  enhanced,
-  segmentation engine=empty,
-%  beamer,
-fuzzy shadow={2mm}{-1mm}{0mm}{0.1mm}%
-{black!50!white},
-  sharp corners,
-  attach boxed title to top left={
-    yshifttext=-1mm
-  },
-%  halign lower=flush center,
-  colback=white,
-  colframe=Maroon!75!white,
-  fonttitle=\bfseries,
-  attach boxed title to top left={yshift=-2mm,xshift=3mm},
-  fontlower=\itshape,
-  boxed title style={
-    sharp corners,
-    size=small,
-    colback=Maroon!75!black,
-    colframe=Maroon!75!black,
-  } 
-}{ist}
-
-\newtcbtheorem[number within=section]{story}{Story}{
-  enhanced,
-  segmentation engine=empty,
-%  beamer,
-fuzzy shadow={2mm}{-1mm}{0mm}{0.1mm}%
-{black!50!white},
-  sharp corners,
-  attach boxed title to top right={
-    yshifttext=-1mm
-  },
-%  halign lower=flush center,
-  colback=white,
-  colframe=ForestGreen!75!white,
-  fonttitle=\bfseries,
-  attach boxed title to top right={yshift=-2mm,xshift=3mm},
-  fontlower=\itshape,
-  boxed title style={
-    sharp corners,
-    size=small,
-    colback=ForestGreen!75!black,
-    colframe=ForestGreen!75!black,
-  }
-}{str}
 
@@
 
-\subsection{How have academic Institutions transformed by then? ( Point 4 in the issue )}
-- Emerging structures:
-  - Universities/Institutes with central pools of RSEs, each specializing in data management, visualization, HPC, etc.
-  - Potential for closer collaboration between domain scientists and RSEs to share workload effectively.
-
-- Challenges:
-  - Budget and administrative hurdles to creating permanent, well-funded RSE teams.
-  - Aligning expectations: researchers might not realize that specialized RSE services exist (and are beneficial).
-
-- Networking:
-  - Need to share best practices among institutions with successful RSE teams.
-  - Encourage knowledge exchange to find models that suit different organizational constraints.
-
 \subsection{Ethics and social consideration}
@@
 
+\subsection{How have academic Institutions transformed by then? ( Point 4 in the issue )}
+\subsubsection{What do we want to have achieved}
+We hope to see that universities use the presence of an RSE Department actively as a means to differentiate themselves from other universities.
+We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department to oversee specific projects[FIXME: cite group paper].
+We also hope, that specialized clusters form, with certain RSE units specializing in certain tasks like data management, HPC, or visualization.
+At universities which have them, we want to see, that researchers naturally accept them as part of their used services.
+
+\subsubsection{Which challenges will still persist}
+The workshop participants expect that there will still be budgeting and administrative hurdles to creating
+permanent and well-funded RSE teams.
+While we hope that the barrier to researchers diminishes, the workshop participants
+still expect some work to do. We hope that the compartmentalization that occurs will help researchers reprioritize their workload.
+An upcoming challenge will be that, with the emergence of more dedicated RSE units,
+networking between them will come into the foreground, because there should be a strong need to share best practices
+among institutions with succesful RSE Teams.
+On a more organisational level, we expect the sharing of models, to fit the different organisational constraints.
+
+What are open questions here?
+Organizations will have to come up with models for hiring people. While
+a standardized RSE masters makes it easier, it will be interesting to see, if
+RSE are still often hired by word of mouth or if the master's really makes it more comparable.
+
 \subsection{What are new tasks of RTPs?}
-\begin{story}{text 1 }{text 2}
 FG-NOTE: RSEs are one type of specialization of RTPs ins science. Do we see any other?
-\end{story}
 
 \subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}
-\begin{whatis}{text 1 }{text 2}
-FG-NOTE: By Nick. I've put it here, due to the realtion with the RTP question.
-\end{whatis}
+FG-TEXT-START: The RSE is probably one of the first professionalizations in the academic midlevel that is able
+to collectively voice their concerns. This enables the possibility to reshape universities.
+If more people are employed at the midlevel, this makes a steadier pool of people
+available that can integrate new people that are regularly coming into departments and know the local facilities.
+On the other hand an upcoming challenge can be, how universities decide to design the contracts of people employed
+in RSE units. will they be permanent, or will there also be an up-or-out mentality?
+FG-TEXT-END
+
+FG-TEXT-START:
+As outlined in [cite RSE group paper] not all universities need to opt for strengthening
+the academic midlevel. Some might have chosen to set up external companies to outsource typical RSE tasks,
+thereby providing career paths outside of academia.
+FG-TEXT-END:
+
+FG-NOTE: By Nick. I've put it here, due to the relation with the RTP question.
 

### 2025-04-16 future_challenges_for_RSEng.tex

@@
 \subsection{How have academic Institutions transformed by then? ( Point 4 in the issue )}
-\subsubsection{What do we want to have achieved}
+\begin{whatis}{What do we want to have achieved}{acedemic}
 We hope to see that universities use the presence of an RSE Department actively as a means to differentiate themselves from other universities.
@@
 At universities which have them, we want to see, that researchers naturally accept them as part of their used services.
+\end{whatis}
 
@@
 \subsection{What are new tasks of RTPs?}
+\begin{story}{text 1 }{text 2}
  FG-NOTE: RSEs are one type of specialization of RTPs ins science. Do we see any other?
+\end{story}
 
@@
 
-FG-NOTE: By Nick. I've put it here, due to the relation with the RTP question.
+\begin{story}{text 1 }{text 2}
+FG-NOTE: By Nick. I've put it here, due to the realtion with the RTP question.
+\end{story}
 

### 2025-04-16 future_challenges_for_RSEng.tex
start

--- 2025-04-16 future_challenges_for_RSEng.tex
@@
 \subsection{Ethics and social consideration}
+The ethics and values of the Research Software Engineer are grounded in their practice.
+They commit to the health, safety and welfare of the public and act in the interest of society, their employer and their clients \cite{Goth2023}.
+Sometimes these values are in direct conflict with the aims of their institution.
+For example, their PI might insist on quickly developing some software without adhering to current best practices.
+Another example might be that they are developing code that can also be used for military purposes, such as numerical solvers or computer vision systems.
+Like engineers, RSEs take the responsibility for the software they develop.
+RSEs need to be aware of these issues and develop their own ``red lines''.
+
+RSEs are publicly funded.
+Similarly to how data collected by scientists should adhere to the FAIR principles \cite{FAIR}, research software developed by RSEs should adhere to the FAIR4RS principles~\cite{FAIR4RS}.
+
+A code of ethics and professional conduct for RSEs should be developed by institutions together with RSEs themselves.
+This can then form the basis for institutional policies and guidelines.
+
+\begin{story}{The greater world of Kim and Kay}{ethics}
+
+\end{story}
+
+\begin{FramedParagraphWithFootnotes}  
 FG-NOTE: Given what ChatGPT has put in there, I wonder if we need to separate these two items
@@
   - RSE community might need to develop or advocate for open-source alternatives.
+\end{FramedParagraphWithFootnotes}
 
 \subsection{What are new tasks of RTPs?}
-\begin{story}{text 1 }{text 2}
  FG-NOTE: RSEs are one type of specialization of RTPs ins science. Do we see any other?
-\end{story}
 

### 2025-04-16 future_challenges_for_RSEng.tex

@@
 
+Looking ahead over a 10 year timeframe, we hope that the RSE community will have
+supported a number of changes and developments that help to make research software outputs
+more sustainable, robust and maintainable. We see this happening in a number of ways. However,
+we want to ensure that one of the core achievements made by the RSE community in this time
+frame is putting the infrastructure and training opportunities in place, alongside the necessary
+advocacy, to ensure that researchers receive core software skills as part of their scientific training.
+With development of software tools or scripts becoming an increasingly important part of everyday
+research activities, this is vital and it will allow RSEs to focus on other more advanced and specialist
+areas. In this context, we do not foresee the demise of RSE as an independent role. However, we
+do see the work of RSEs developing and shifting as the research community changes.
+
+The specialist skills that RSEs provide take a significant amount of time to develop and maintain.
+We would, therefore, not expect domain researchers to also become experts in a range of
+advanced RSE topics. 
+At the same time, we do recognise that the application of core software development best practices
+to enhance and address problems with research codebases, that can often make up the bulk of
+current RSE work, will change. As noted above, these are skills that we
+would expect any researcher who writes code to have in 10 years time. 
+The rapid development of AI means that researchers are likely to have access to tooling that can assist
+them in writing robust, sustainable code, lowering the barrier to developing code, or at least changing the
+required skill set to some extent.
+They can also be expected to have an understanding of frameworks to support testing, packaging and
+deployment of their code. Again, all these skills will, we anticipate, be gained as part of training
+infrastructure that will provide domain scientists with software development expertise as part of their
+core scientific training.
+
+So where do RSEs fit into this picture? We expect to see RSEs providing technical skills and input
+to research activities in range of different areas, including the following:
+\begin{itemize}
+  \item High-quality, robust architecture and design for research software
+  \item Development of software for novel and emerging architectures
+  \item Application of specialist numerical and statistical methods within software
+  \item Optimisation of codes for use on high-performance computing infrastructure
+  \item Green computing - efficient implementations of algorithms/methods and use of hardware
+  \item ...
+\end{itemize}
+
 \subsection{Definition and Evolution of RSEs}
@@
 \subsection{Ethics and social consideration}
-The ethics and values of the Research Software Engineer are grounded in their practice.
-They commit to the health, safety and welfare of the public and act in the interest of society, their employer and their clients \cite{Goth2023}.
-Sometimes these values are in direct conflict with the aims of their institution.
-For example, their PI might insist on quickly developing some software without adhering to current best practices.
-Another example might be that they are developing code that can also be used for military purposes, such as numerical solvers or computer vision systems.
-Like engineers, RSEs take the responsibility for the software they develop.
-RSEs need to be aware of these issues and develop their own ``red lines''.
-
-RSEs are publicly funded.
-Similarly to how data collected by scientists should adhere to the FAIR principles \cite{FAIR}, research software developed by RSEs should adhere to the FAIR4RS principles~\cite{FAIR4RS}.
-
-A code of ethics and professional conduct for RSEs should be developed by institutions together with RSEs themselves.
-This can then form the basis for institutional policies and guidelines.
-
-\begin{story}{The greater world of Kim and Kay}{ethics}
-
-\end{story}
-
-\begin{FramedParagraphWithFootnotes}  
 FG-NOTE: Given what ChatGPT has put in there, I wonder if we need to separate these two items
@@
   - RSE community might need to develop or advocate for open-source alternatives.
-\end{FramedParagraphWithFootnotes}
 
 \subsection{What are new tasks of RTPs?}
+\begin{story}{text 1 }{text 2}
  FG-NOTE: RSEs are one type of specialization of RTPs ins science. Do we see any other?
+\end{story}
 

### 2025-04-16 future_challenges_for_RSEng.tex
work in feedback from PR

@@
 advocacy, to ensure that researchers receive core software skills as part of their scientific training.
+
 With development of software tools or scripts becoming an increasingly important part of everyday
@@
 They can also be expected to have an understanding of frameworks to support testing, packaging and
-deployment of their code. Again, all these skills will, we anticipate, be gained as part of training
-infrastructure that will provide domain scientists with software development expertise as part of their
-core scientific training.
+deployment of their code as well as integration with central domain specific infrastructures.
+Again, all these skills will, we anticipate, be gained as part of training infrastructure that will
+provide domain scientists with software development expertise as part of their core scientific training.
 
 So where do RSEs fit into this picture? We expect to see RSEs providing technical skills and input
-to research activities in range of different areas, including the following:
+to research activities in a range of different areas, including the following:
 \begin{itemize}
   \item High-quality, robust architecture and design for research software
+  \item Seamless integration with research and data infrastructures
   \item Development of software for novel and emerging architectures
@@
 \begin{framed}
-How will the definition of an RSE evolve in ten years?
+\hfill How will the definition of an RSE evolve in ten years?\\
 
-What will not change in the essence of an RSE ten years from now?
+\hfill What will not change in the essence of an RSE ten years from now?
 \end{framed}

### 2025-04-17 future_challenges_for_RSEng.tex
fixes

@@
 \item The development of academic institutions
+\item What will the impact of AI be to the work of RSEs?
 \item How will academic education have transformed by then
@@
 Not too much, to be overwhelmingly far in the future, but distant enough for some changes to occur.
-Over this timeframe Kim and Kay
+Over this timeframe Kim\cite{Anzt2021} and Kay\cite{competencies-F1000}
 were able to ride the waves of their career in Research Software Engineering and are now pondering early retirement on a sunset blessed beach.
@@
 What issues have newly emerged, where they are now just happy to say: “Oh well… My successor takes care of that”.
+\begin{story}{Introduction}{intro}
+In order to bring this future to life these two personas will accompany us and we will shed light on certain aspects
+of their future in respective small story boxes.
+ \end{story}
 

### 2025-04-17 future_challenges_for_RSEng.tex

@@
 
+\subsection{How have academic Institutions transformed by then? ( Point 4 in the issue )}
+\begin{whatis}{What do we want to have achieved}{acedemic}
+We hope to see that universities use the presence of an RSE Department actively as a means to differentiate themselves from other universities.
+We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department to oversee specific projects[FIXME: cite group paper].
+We also hope, that specialized clusters form, with certain RSE units specializing in certain tasks like data management, HPC, or visualization.
+At universities which have them, we want to see, that researchers naturally accept them as part of their used services.
+\end{whatis}
+
+\subsubsection{Which challenges will still persist}
+The workshop participants expect that there will still be budgeting and administrative hurdles to creating
+permanent and well-funded RSE teams.
+While we hope that the barrier to researchers diminishes, the workshop participants
+still expect some work to do. We hope that the compartmentalization that occurs will help researchers reprioritize their workload.
+An upcoming challenge will be that, with the emergence of more dedicated RSE units,
+networking between them will come into the foreground, because there should be a strong need to share best practices
+among institutions with succesful RSE Teams.
+On a more organisational level, we expect the sharing of models, to fit the different organisational constraints.
+
+What are open questions here?
+Organizations will have to come up with models for hiring people. While
+a standardized RSE masters makes it easier, it will be interesting to see, if
+RSE are still often hired by word of mouth or if the master's really makes it more comparable.
+
 \subsection{What are new tasks of RTPs?}
@@
 \subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}
-\begin{whatis}{text 1 }{text 2}
+FG-TEXT-START: The RSE is probably one of the first professionalizations in the academic midlevel that is able
+to collectively voice their concerns. This enables the possibility to reshape universities.
+If more people are employed at the midlevel, this makes a steadier pool of people
+available that can integrate new people that are regularly coming into departments and know the local facilities.
+On the other hand an upcoming challenge can be, how universities decide to design the contracts of people employed
+in RSE units. will they be permanent, or will there also be an up-or-out mentality?
+FG-TEXT-END
+
+FG-TEXT-START:
+As outlined in [cite RSE group paper] not all universities need to opt for strengthening
+the academic midlevel. Some might have chosen to set up external companies to outsource typical RSE tasks,
+thereby providing career paths outside of academia.
+FG-TEXT-END:
+
+\begin{story}{text 1 }{text 2}
 FG-NOTE: By Nick. I've put it here, due to the realtion with the RTP question.
-\end{whatis}
+\end{story}
 

### 2025-04-17 future_challenges_for_RSEng.tex
Add a story part

@@
 
+\begin{story}{Changes to the Universities}{changes}
+ Kim found work years ago at one of the first RSE units in Arcadia. After a couple of years he moved to the university of
+Avalon which was interested in setting up their own department for RSEs in order to not fall behind the university of Arcadia.
+Kim applies his experiences and improves upon the model set forth by the University of Arcadia. Since the university
+of Avalon has to follow different rules not everything can be straightforwardly adapted. Life would be so much easier,
+if at least university regulations would be homogenized, no?
+\end{story}
+
 \subsection{What are new tasks of RTPs?}
-\begin{story}{text 1 }{text 2}
  FG-NOTE: RSEs are one type of specialization of RTPs ins science. Do we see any other?
-\end{story}
 
@@
 
-\begin{story}{text 1 }{text 2}
 FG-NOTE: By Nick. I've put it here, due to the realtion with the RTP question.
-\end{story}
 

### 2025-04-17 future_challenges_for_RSEng.tex
fixups

@@
 \subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}
-FG-TEXT-START: The RSE is probably one of the first professionalizations in the academic midlevel that is able
+The RSE is probably one of the first professionalizations in the academic midlevel that is able
 to collectively voice their concerns. This enables the possibility to reshape universities.
@@
 in RSE units. will they be permanent, or will there also be an up-or-out mentality?
-FG-TEXT-END
 
-FG-TEXT-START:
 As outlined in [cite RSE group paper] not all universities need to opt for strengthening
@@
 thereby providing career paths outside of academia.
-FG-TEXT-END:
 
@@
 AI Systems have been harmed in the creation of this work.
+We thank all the participants of our workshop at deRSE25 in Karlsruhe for their valuable input!
 \end{acknowledge}

### 2025-04-17 future_challenges_for_RSEng.tex
fixes

@@
 Not too much, to be overwhelmingly far in the future, but distant enough for some changes to occur.
-Over this timeframe Kim\cite{Anzt2021} and Kay\cite{competencies-F1000}
+Over this timeframe Kim\cite{Anzt2021} and Kay\cite{Goth2024}
 were able to ride the waves of their career in Research Software Engineering and are now pondering early retirement on a sunset blessed beach.

### 2025-04-17 future_challenges_for_RSEng.tex

@@
 \subsection{Ethics and social consideration}
-FG-NOTE: Given what ChatGPT has put in there, I wonder if we need to separate these two items
+The ethics and values of the Research Software Engineer are grounded in their practice.
+They commit to the health, safety and welfare of the public and act in the interest of society, their employer and their clients \cite{Goth2023}.
+Sometimes these values are in direct conflict with the aims of their institution.
+For example, their PI might insist on quickly developing some software without adhering to current best practices.
+Another example might be that they are developing code that can also be used for military purposes, such as numerical solvers or computer vision systems.
+Like engineers, RSEs take the responsibility for the software they develop.
+RSEs need to be aware of these issues and develop their own ``red lines''.
+
+RSEs are publicly funded.
+Similarly to how data collected by scientists should adhere to the FAIR principles \cite{FAIR}, research software developed by RSEs should adhere to the FAIR4RS principles~\cite{FAIR4RS}.
+
+A code of ethics and professional conduct for RSEs should be developed by institutions together with RSEs themselves.
+This can then form the basis for institutional policies and guidelines.
+
 MM-NOTE: (Flo believes it was from Michele) Are we assuming that progress will bring prosperity? Will we remain constrained by needing to produce products/artifacts?
 
-- Military or other sensitive applications:
-  - RSEs (like physicists, mathematicians, etc.) might face ethical dilemmas if their work can be repurposed for weaponry.
   - Importance of having personal “red lines” and awareness of how software can be used.
 
-- Corporate dominance:
-  - AI and coding platforms might be increasingly controlled by a small number of large corporations.
-  - RSE community might need to develop or advocate for open-source alternatives.
+\begin{story}{The greater world of Kim and Kay}{ethics}
+10 years have passed since Kim and Kay have started on their journey of being RSEs. Back then,
+at the universities of Eden and Utopia, life was more orderly and harmonious. Kay is working in a project that develops
+software for cheap water monitoring devices in the global south. Kim is currently designing the software service stack
+at her university, and increasingly struggles to find compelling Open Source alternatives that let's them keep their data
+to software that is backed by large corporate entities. So far she still has the support of her university board, since it values
+strategic autonomy.
+Her son Kyle on the other hand started out with a degree in Physics and then took one of the newly created RSE Master courses,
+but quickly decided that work in academia with its time limited contracts is not attractive.
+Governments around the world are struggling to reposition themselves in this emerging deglobalized world with its new blocks
+and therefore have deprioritized publicly funded research. Kyle decided to go to a defence company since they offer a
+more secure employment in this environment.
+\end{story}
+
+\subsection{What are new tasks of RTPs?}
+ FG-NOTE: RSEs are one type of specialization of RTPs in science. Do we see any other?
+
+\subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}
+
+FG-NOTE: By Nick. I've put it here, due to the realtion with the RTP question.
 

### 2025-04-17 future_challenges_for_RSEng.tex
add some sentences to ethics section (#12)
* add some sentences to ethics section

Co-authored-by: Gerasimos Chourdakis <gerasimos.chourdakis@ipvs.uni-stuttgart.de>

* some color and structuring boxes

* start

---------

Co-authored-by: Frank Löffler <frank.loeffler@uni-jena.de>
Co-authored-by: Gerasimos Chourdakis <gerasimos.chourdakis@ipvs.uni-stuttgart.de>
Co-authored-by: Florian Goth <florian.goth@physik.uni-wuerzburg.de>
Co-authored-by: Florian Goth <florian.goth@uni-wuerzburg.de>

@@
 
-\newtcbtheorem[number within=section]{whatis}{What is 2035}{
+\newcounter{whatiscounter}
+\newtcbtheorem[use counter=whatiscounter, number format=\Alph]{whatis}{What is 2035}{
   enhanced,
@@
 
-\newtcbtheorem[number within=section]{story}{Story}{
+\newcounter{storycount}
+\newtcbtheorem[use counter=storycount]{story}{Story}{
   enhanced,
@@
 and therefore have deprioritized publicly funded research. Kyle decided to go to a defence company since they offer a
-more secure employment in this environment.
+more secure employment in this polarized world.
 \end{story}
 
-\subsection{What are new tasks of RTPs?}
- FG-NOTE: RSEs are one type of specialization of RTPs in science. Do we see any other?
-
-\subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}
-
-FG-NOTE: By Nick. I've put it here, due to the realtion with the RTP question.
-
-\subsection{How have academic Institutions transformed by then? ( Point 4 in the issue )}
+\subsection{How have academic Institutions transformed by then?}
 \begin{whatis}{What do we want to have achieved}{acedemic}

### 2025-04-17 future_challenges_for_RSEng.tex
fix typo

@@
 as a domain in its own right, with the hope of ensuring that it gets the recognition it deserves.
-Nonetheless, as the use of digital tools becomes ever more pervasive in a researchers day-to-day tasks, there are already people suggesting that RSE will
+Nonetheless, as the use of digital tools becomes ever more pervasive in a researcher's day-to-day tasks, there are already people suggesting that RSE will
 cease to exist as a separate field, and the specialist skills that RSEs provide to support and undertake research will
-become a core part of the tooling and skill set of researchers in almost all domains.
-Are we already approaching "the end of Research Software Engineering"?! While we believe that it is certainly too early to
+become a core part of the tooling and skill set of researchers in almost all other domains.
++Are we already approaching ``the end of Research Software Engineering''?! While we believe that it is certainly too early to
 already herald the demise of RSE as a separate field,
@@
 \begin{itemize}
-\item you like to read,
-\item gives you something to think about,
-\item motivates you to work on creating this better future.
+\item is pleasant to read,
+\item gives something to think about,
+\item motivates engaging in work towards a better future.
 \end{itemize}
-In order to structure the discussion, we decided on a imaginable setting ten years from now. 
+In order to structure the discussion, we decided on an imaginable setting ten years from now.
 In this world we will consider the dimensions of 
@@
 In order to give a more explicit setting, imagine a world ten years from now.
-Not too much, to be overwhelmingly far in the future, but distant enough for some changes to occur.
+Not too much to be overwhelmingly far in the future, but distant enough for some changes to occur.
 Over this timeframe Kim\cite{Anzt2021} and Kay\cite{Goth2024}
 were able to ride the waves of their career in Research Software Engineering and are now pondering early retirement on a sunset blessed beach.
+%MM-NOTE: the above scenario (early retirement) is contrary to all trends in retirement age development of public employees across the "global north".
 But the world has moved on.
@@
 \item Ethics and social considerations
-\item Impact of alternate career paths for RTPs
+\item Impact of alternate career paths for RTPs (\emph{Research Technical Professional}s)
 \end{itemize}
@@
 
-Looking ahead over a 10 year timeframe, we hope that the RSE community will have
-supported a number of changes and developments that help to make research software outputs
+Looking ahead over a 10-year timeframe, we hope that the RSE community will have
+contributed to make research software and its outputs
 more sustainable, robust and maintainable. We see this happening in a number of ways. However,
@@
 
-With development of software tools or scripts becoming an increasingly important part of everyday
+With development of software tools and scripted workflows becoming an increasingly important part of everyday
 research activities, this is vital and it will allow RSEs to focus on other more advanced and specialist
+% MM-NOTE: if "this is vital" is referring to "software skills" it's not clear -- different paragraphs
 areas. In this context, we do not foresee the demise of RSE as an independent role. However, we
@@
 We expect these initiatives to help increase the visibility and accelerate
-the professionalisation of the RSE role.
+the professionalisation%
+\footnote{MM: do we want to decide for UK or US English? Or keep it mixed? Not being a native English speaker I don't know how it reads.}
+of the RSE role.
 In addition, a multitude of RSE communities self-organised at the national
@@
 \subsection{How have academic Institutions transformed by then?}
-\begin{whatis}{What do we want to have achieved}{acedemic}
+\begin{whatis}{What do we want to have achieved}{academic}
 We hope to see that universities use the presence of an RSE Department actively as a means to differentiate themselves from other universities.
@@
 networking between them will come into the foreground, because there should be a strong need to share best practices
-among institutions with succesful RSE Teams.
+among institutions with successful RSE Teams.
 On a more organisational level, we expect the sharing of models, to fit the different organisational constraints.
@@
 \subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}
-The RSE is probably one of the first professionalizations in the academic midlevel that is able
+The RSE is probably one of the first professionalizations in the academic mid-level that is able
 to collectively voice their concerns. This enables the possibility to reshape universities.
-If more people are employed at the midlevel, this makes a steadier pool of people
+If more people are employed as mid-level, this makes a steadier pool of people
 available that can integrate new people that are regularly coming into departments and know the local facilities.
-On the other hand an upcoming challenge can be, how universities decide to design the contracts of people employed
-in RSE units. will they be permanent, or will there also be an up-or-out mentality?
+On the other hand, an upcoming challenge how universities decide to design the contracts of people employed
+in RSE units. Will they be permanent, or will there also be an up-or-out mentality?
 
 As outlined in [cite RSE group paper] not all universities need to opt for strengthening
-the academic midlevel. Some might have chosen to set up external companies to outsource typical RSE tasks,
+the academic mid-level. Some might have chosen to set up external companies to outsource typical RSE tasks,
 thereby providing career paths outside of academia.
 
-FG-NOTE: By Nick. I've put it here, due to the realtion with the RTP question.
+FG-NOTE: By Nick. I've put it here, due to the relation with the RTP question.
 
@@
 This made participants worry about growing mismatch of peoples' expectations.
-Participants identified the over-reliance on generative AI tools as an extreme manifestation of this issue.
+Participants expected an over-reliance on generative AI tools as an extreme manifestation of this issue.
 No concrete examples were given on which complexities are now hidden behind abstractions.
@@
 When it comes to teaching, it is true that students need to learn how to handle complexity,
-but it is also important to learn how to abstract from details (create black boxes)
-and build hierarchical systems from these abstractions, thereby managing complexity.
+but it is also important to learn how to abstract from details
+by building hierarchical systems from these abstractions encapsulated in reusable components, thereby managing complexity.
 It is clear that we need both RSEs that can dive deep into single projects of significant complexity,
@@
 \begin{acknowledge}
-AI Systems have been harmed in the creation of this work.
+% MM-NOTE: I'd keep the following sentence out, out of several reasons.
+% AI Systems have been harmed in the creation of this work.
 We thank all the participants of our workshop at deRSE25 in Karlsruhe for their valuable input!

### 2025-05-05 future_challenges_for_RSEng.tex
an idea.

@@
 Is it better than cheap PhDs?
+FG-NOTE: AI will change the interaction with researchers at a fundamental level.
+Whereas previously RSEs have been points of knowledge that could translate human language formulated problems into technical descriptions and then provide solutions to it,
+I foresee that AI drives this low-key interaction to a highly specialized level of knowledge, since the availability of AI is far better.
 

### 2025-05-06 future_challenges_for_RSEng.tex

@@
 \begin{itemize}
-\item The evolution of the term RSE
-\item How is education reshaped by the digital transformation
-\item What will the impact of AI be to the work of RSEs?
-\item Changes to the Academic Institutions
-\item Ethics and social considerations
-\item Impact of alternate career paths for RTPs (\emph{Research Technical Professional}s)
+\item The evolution of the term RSE,
+\item How education is reshaped by the digital transformation,
+\item What the impact of AI to the work of RSEs will look like,
+\item Changes to the Academic Institutions,
+\item Ethics and social considerations,
+\item Impact of alternate career paths for RTPs (\emph{Research Technical Professional}s).
 \end{itemize}

### 2025-05-06 future_challenges_for_RSEng.tex
small rephrase formulated mostly by Julian
Co-authored-by: Julian Dehne <julian.dehne@rz.gi.de>

@@
 \subsection{Ethics and social consideration}
+
 The ethics and values of the Research Software Engineer are grounded in their practice.
 They commit to the health, safety and welfare of the public and act in the interest of society, their employer and their clients \cite{Goth2023}.
-Sometimes these values are in direct conflict with the aims of their institution.
-For example, their PI might insist on quickly developing some software without adhering to current best practices.
+
+% institutional bonds and ethics
+Sometimes these values are in direct conflict with the aims of their institution. For example, their PI might insist on quickly developing some software without adhering to current best practices.
+A code of ethics and professional conduct for RSEs should be developed by institutions together with RSEs themselves. This can then form the basis for institutional policies and guidelines.
+
+% society implications
+The technology people working as RSE might have, well directly or indirectly, ethical implications in areas such as social engineering [TODO find source], climate [TODO find sourc] and more.
 Another example might be that they are developing code that can also be used for military purposes, such as numerical solvers or computer vision systems.
@@
 
-RSEs are publicly funded.
+% data ethics
 Similarly to how data collected by scientists should adhere to the FAIR principles \cite{FAIR}, research software developed by RSEs should adhere to the FAIR4RS principles~\cite{FAIR4RS}.

### 2025-05-09 future_challenges_for_RSEng.tex
reuse parts of the template

@@
 
+\definecolor{easstblue}{rgb}{.05,.32,.66}
+
 \newcounter{whatiscounter}
@@
   colback=white,
-  colframe=Maroon!75!white,
+  colframe=BurntOrange!75!white,
   fonttitle=\bfseries,
@@
     size=small,
-    colback=Maroon!75!black,
-    colframe=Maroon!75!black,
+    colback=BurntOrange!75!black,
+    colframe=BurntOrange!75!black,
   } 
@@
   colback=white,
-  colframe=ForestGreen!75!white,
+  colframe=easstblue,
   fonttitle=\bfseries,
@@
     size=small,
-    colback=ForestGreen!75!black,
-    colframe=ForestGreen!75!black,
+    colback=easstblue!70!black,
+    colframe=easstblue!70!black,
   }

### 2025-05-09 future_challenges_for_RSEng.tex
Impact on structures

--- 2025-05-06 future_challenges_for_RSEng.tex
@@
 \subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}
-The RSE is probably one of the first professionalizations in the academic mid-level that is able
-to collectively voice their concerns. This enables the possibility to reshape universities.
-If more people are employed as mid-level, this makes a steadier pool of people
-available that can integrate new people that are regularly coming into departments and know the local facilities.
-On the other hand, an upcoming challenge how universities decide to design the contracts of people employed
-in RSE units. Will they be permanent, or will there also be an up-or-out mentality?
 
-As outlined in [cite RSE group paper] not all universities need to opt for strengthening
-the academic mid-level. Some might have chosen to set up external companies to outsource typical RSE tasks,
-thereby providing career paths outside of academia.
-
-FG-NOTE: By Nick. I've put it here, due to the relation with the RTP question.
+RSEs can either be embedded in research groups (belonging to the mid-level university staff),
+or be part of larger RSE units.
+The participants had the opinion that focusing on embedded RSEs would make onboarding new RSEs to existing projects easier.
+This, however, comes with additional funding challenges,
+as long-term funding is more difficult to acquire in lower levels of an organization.
+Third-party funding can mitigate this issue in the mid-term.
+However, the participants considered central RSE groups more likely in the future,
+due to the fact that there are already examples and activities from various universities in this direction
+(examples include the univerisities of Jena, Heidelberg, Göttingen).
+We expect funding in central groups to evolve from initially temporary positions to permanent positions,
+as the groups themselves prove themselves valuable.
+Embedded RSEs and central groups will need to network with each other,
+and potentially request and coordinate support from external RSEs.
 

### 2025-05-09 future_challenges_for_RSEng.tex
Complexity: Clarify
Co-authored-by: Jan Linxweiler <jlinx@users.noreply.github.com>

--- 2025-05-09 future_challenges_for_RSEng.tex
@@
 The workshop participants found that in earlier learning stages,
-details may be intentionally hidden,
+details of systems may be intentionally hidden,
 fearing a diminishing capability to handle technical complexity in research environments.
@@
 When it comes to teaching, it is true that students need to learn how to handle complexity,
-but it is also important to learn how to abstract from details
+but it is also important to learn how to abstract from details,
 by building hierarchical systems from these abstractions encapsulated in reusable components, thereby managing complexity.

### 2025-05-09 future_challenges_for_RSEng.tex
Update future_challenges_for_RSEng.tex
Co-authored-by: Jan Philipp Thiele <54978337+jpthiele@users.noreply.github.com>
Add reminder for citing groups paper

--- 2025-05-09 future_challenges_for_RSEng.tex
@@
 RSEs can either be embedded in research groups (belonging to the mid-level university staff),
-or be part of larger RSE units.
+or be members of larger RSE units. (TODO: Cite RSE groups paper)
 The participants had the opinion that focusing on embedded RSEs would make onboarding new RSEs to existing projects easier.
@@
 due to the fact that there are already examples and activities from various universities in this direction
-(examples include the univerisities of Jena, Heidelberg, Göttingen).
+(examples include the universities of Jena, Heidelberg, Göttingen).
 We expect funding in central groups to evolve from initially temporary positions to permanent positions,

### 2025-05-09 future_challenges_for_RSEng.tex
Add ideas GC

--- 2025-05-09 future_challenges_for_RSEng.tex
@@
 reducing reliance on RSEs to train newcomers on the most basic tasks
-(JN: for example, versioning and archiving?).
+(JN: for example, versioning and archiving?)
+(GC: Higher-level and more intuitive workflows that require less training).
 User training sessions might focus on more advanced topics
@@
 
+\begin{FramedParagraphWithFootnotes}
+GC-NOTE: With longer-term funding and more trust to pick up longer term projects,
+some RSEs can focus on code modernization, cleanup, and modularization,
+which can pay off in faster and larger development in the future.
+This is not possible by short-term funded solo RSEs.
+\end{FramedParagraphWithFootnotes}
+
 \subsection{Education, and changes in the educational system}
+
+\begin{FramedParagraphWithFootnotes}
 FG-NOTE: I wonder if we need a separation between education of RSEs and changes in the education of the broader society.
+\end{FramedParagraphWithFootnotes}
+
+\begin{FramedParagraphWithFootnotes}
+GC-NOTE: Digital competencies in the general population continue to increase, making jumping to software engineering easier than before.
+
+GC-NOTE: By that point, we have an RSE Master's.
+\end{FramedParagraphWithFootnotes}
 
@@
 Is it better than cheap PhDs?
+
+\begin{FramedParagraphWithFootnotes}
 FG-NOTE: AI will change the interaction with researchers at a fundamental level.
@@
 I foresee that AI drives this low-key interaction to a highly specialized level of knowledge, since the availability of AI is far better.
+\end{FramedParagraphWithFootnotes}
+
+\begin{FramedParagraphWithFootnotes}
+GC-NOTE: As we continue and expand documenting our work in the open and making our code available,
+there will be more structured material to train GenAI on.
+
+However, as people get quicker answers from GenAI, the community resources (Stack Overflow, Wikipedia) get poorer.
+\end{FramedParagraphWithFootnotes}
 
@@
 
-MM-NOTE: (Flo believes it was from Michele) Are we assuming that progress will bring prosperity? Will we remain constrained by needing to produce products/artifacts?
+MM-NOTE: (Flo believes it was from Michele) Are we assuming that progress will bring prosperity?
+Will we remain constrained by needing to produce products/artifacts?
+(GC: Where "product/artifacts", maybe also "publications"?)
 
@@
 
+\begin{FramedParagraphWithFootnotes}
+  GC-NOTE: New section on "New challenges":
+
+  \begin{itemize}
+    \item Networking of embedded RSEs and RSE units
+    \item Everyone needs infrastructure: does everyone need to host their own infrastructure?
+    \item Everyone hosts their infrastructure, but then everyone stays in silos? We need federated infrastructure.
+  \end{itemize}
+\end{FramedParagraphWithFootnotes}
+
 \begin{story}{Changes to the Universities}{changes}
@@
 The workshop participants found that in earlier learning stages,
-details of systems may be intentionally hidden,
+details may be intentionally hidden,
 fearing a diminishing capability to handle technical complexity in research environments.
@@
 When it comes to teaching, it is true that students need to learn how to handle complexity,
-but it is also important to learn how to abstract from details,
+but it is also important to learn how to abstract from details
 by building hierarchical systems from these abstractions encapsulated in reusable components, thereby managing complexity.
@@
 
+GC-NOTE: Shall we add another story box here?
+
 \section{REALLY Long-term(10+ years) aspects}
@@
 at what becomes now possible.
+
+GC-NOTE: RSE develop together with and for scientists.
+Better software leads to better research,
+better research leads to more better software.
 

### 2025-05-09 future_challenges_for_RSEng.tex
Fix some typos

--- 2025-05-09 future_challenges_for_RSEng.tex
@@
 software for cheap water monitoring devices in the global south. Kim is currently designing the software service stack
-at her university, and increasingly struggles to find compelling Open Source alternatives that let's them keep their data
+at her university, and increasingly struggles to find compelling Open Source alternatives that lets them keep their data
 to software that is backed by large corporate entities. So far she still has the support of her university board, since it values
 strategic autonomy.
-Her son Kyle on the other hand started out with a degree in Physics and then took one of the newly created RSE Master courses,
+Her son Kyle on the other hand started out with a degree in Physics and then took one of the newly created RSE Master's courses,
 but quickly decided that work in academia with its time limited contracts is not attractive.
 Governments around the world are struggling to reposition themselves in this emerging deglobalized world with its new blocks
-and therefore have deprioritized publicly funded research. Kyle decided to go to a defence company since they offer a
+and therefore have deprioritized publicly funded research. Kyle decided to go to a defense company since they offer a
 more secure employment in this polarized world.
@@
 We also hope, that specialized clusters form, with certain RSE units specializing in certain tasks like data management, HPC, or visualization.
-At universities which have them, we want to see, that researchers naturally accept them as part of their used services.
+At universities which have them, we want to see that researchers naturally accept them as part of their used services.
 \end{whatis}
@@
 Organizations will have to come up with models for hiring people. While
-a standardized RSE masters makes it easier, it will be interesting to see, if
+a standardized RSE Master's makes it easier, it will be interesting to see if
 RSE are still often hired by word of mouth or if the master's really makes it more comparable.

### 2025-05-09 future_challenges_for_RSEng.tex
Convert introduction to running text

@@
 \section{Introduction}
-This opinion article stems from a Bof session at deRSE25 in Karlsruhe\cite{Goth2025EndRSEng}.
-Therefore it stems from a discussion among the participants of this workshop and hence cannot claim that it represents the entire deRSE community.
-Nevertheless we hope, that we succeeded in writing up something that
-\begin{itemize}
-\item is pleasant to read,
-\item gives something to think about,
-\item motivates engaging in work towards a better future.
-\end{itemize}
-In order to structure the discussion, we decided on an imaginable setting ten years from now.
-In this world we will consider the dimensions of 
-\begin{itemize}
-\item Evolution of the term RSEs
-\item The development of academic institutions
-\item What will the impact of AI be to the work of RSEs?
-\item How will academic education have transformed by then
-\item Ethics and social considerations
-\end{itemize}
-In each aspect we will quickly summarise on what we want to have achieved by then, and what will be the challenges that arise from that.
+
+Research Software Engineering is currently an emerging field,
+with an increasing integration in various aspects of research.
+In this opinion paper, we try to imagine the state of our field
+and community in the mid-term future, focusing on future challenges.
+We present here the results of a workshop discussion at the
+deRSE25 conference in Karlsruhe\cite{Goth2025EndRSEng}, without
+claiming to represent the entire deRSE community.
+Nevertheless, by disseminating the discussed ideas,
+we aim to engage the community in imagining the future.
+In the following sections, we consider the evolution of the RSE definition,
+the development of RSE education,
+the impact of automation and generative artificial intelligence in RSE work,
+the emerging ethical considerations,
+and the transformations expected in the research organizations.
+We then attempt to make long-term predictions
+and we conclude the paper with our wishes and recommendations.
 

### 2025-05-10 future_challenges_for_RSEng.tex

--- 2025-05-09 future_challenges_for_RSEng.tex
@@
 \subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}
-The RSE is probably one of the first professionalizations in the academic mid-level that is able
-to collectively voice their concerns. This enables the possibility to reshape universities.
-If more people are employed as mid-level, this makes a steadier pool of people
-available that can integrate new people that are regularly coming into departments and know the local facilities.
-On the other hand, an upcoming challenge how universities decide to design the contracts of people employed
-in RSE units. Will they be permanent, or will there also be an up-or-out mentality?
 
-As outlined in [cite RSE group paper] not all universities need to opt for strengthening
-the academic mid-level. Some might have chosen to set up external companies to outsource typical RSE tasks,
-thereby providing career paths outside of academia.
-
-FG-NOTE: By Nick. I've put it here, due to the relation with the RTP question.
+RSEs can either be embedded in research groups (belonging to the mid-level university staff),
+or be members of larger RSE units. (TODO: Cite RSE groups paper)
+The participants had the opinion that focusing on embedded RSEs would make onboarding new RSEs to existing projects easier.
+This, however, comes with additional funding challenges,
+as long-term funding is more difficult to acquire in lower levels of an organization.
+Third-party funding can mitigate this issue in the mid-term.
+However, the participants considered central RSE groups more likely in the future,
+due to the fact that there are already examples and activities from various universities in this direction
+(examples include the universities of Jena, Heidelberg, Göttingen).
+We expect funding in central groups to evolve from initially temporary positions to permanent positions,
+as the groups themselves prove themselves valuable.
+Embedded RSEs and central groups will need to network with each other,
+and potentially request and coordinate support from external RSEs.
 

### 2025-05-10 future_challenges_for_RSEng.tex

@@
 RSE are still often hired by word of mouth or if the master's really makes it more comparable.
+
+\begin{FramedParagraphWithFootnotes}
+  GC-NOTE: New section on "New challenges":
+
+  \begin{itemize}
+    \item Networking of embedded RSEs and RSE units
+    \item Everyone needs infrastructure: does everyone need to host their own infrastructure?
+    \item Everyone hosts their infrastructure, but then everyone stays in silos? We need federated infrastructure.
+  \end{itemize}
+\end{FramedParagraphWithFootnotes}
 

### 2025-05-11 future_challenges_for_RSEng.tex
Update future_challenges_for_RSEng.tex
Co-authored-by: Magnus Hagdorn <Magnus.Hagdorn@marsupium.org>
some restructuring

@@
 
-\section{The Situation}
-In order to give a more explicit setting, imagine a world ten years from now.
-Not too much to be overwhelmingly far in the future, but distant enough for some changes to occur.
-Over this timeframe Kim\cite{Anzt2021} and Kay\cite{Goth2024}
-were able to ride the waves of their career in Research Software Engineering and are now pondering early retirement on a sunset blessed beach.
-%MM-NOTE: the above scenario (early retirement) is contrary to all trends in retirement age development of public employees across the "global north".
-But the world has moved on.
-When they look back, what will they see as problems which are now solved?
-Which tasks are persisting, where they are now happy that a younger generation is now taking up the baton and carries forward their work.
-What issues have newly emerged, where they are now just happy to say: “Oh well… My successor takes care of that”.
-\begin{story}{Introduction}{intro}
-In order to bring this future to life these two personas will accompany us and we will shed light on certain aspects
-of their future in respective small story boxes.
- \end{story}
-
 \section{Aspects of this future}
-It is difficult to structure the different aspects in describing the future. We decided to consider
-\begin{itemize}
-\item The evolution of the term RSE,
-\item How education is reshaped by the digital transformation,
-\item What the impact of AI to the work of RSEs will look like,
-\item Changes to the Academic Institutions,
-\item Ethics and social considerations,
-\item Impact of alternate career paths for RTPs (\emph{Research Technical Professional}s).
-\end{itemize}
-
 %\subsection{What do we want to have achieved by 2035}
-
 Looking ahead over a 10-year timeframe, we hope that the RSE community will have
@@
 advocacy, to ensure that researchers receive core software skills as part of their scientific training.
-
 With development of software tools and scripted workflows becoming an increasingly important part of everyday
 research activities, this is vital and it will allow RSEs to focus on other more advanced and specialist
-% MM-NOTE: if "this is vital" is referring to "software skills" it's not clear -- different paragraphs
 areas. In this context, we do not foresee the demise of RSE as an independent role. However, we
@@
 
-So where do RSEs fit into this picture? We expect to see RSEs providing technical skills and input
-to research activities in a range of different areas, including the following:
+So where do RSEs fit into this picture? Even today we expect to see RSEs providing technical skills and input
+to research activities in a range of different areas, such as
+High-quality, robust architecture and design for research software,
+Seamless integration with research and data infrastructures,
+Development of software for novel and emerging architectures,
+Application of specialist numerical and statistical methods within software,
+Optimisation of codes for use on high-performance computing infrastructure,
+and Green computing - efficient implementations of algorithms/methods and use of hardware
+
+It is already difficult to structure the spectrum of RSE tasks and responsibilities today (FIXME: Willi Hasselbrings classification paper?)
+and it is even more hopeless to structure the aspects that will become relavant for the RSE community in the future.
+Nevertheless, we decided to consider
 \begin{itemize}
-  \item High-quality, robust architecture and design for research software
-  \item Seamless integration with research and data infrastructures
-  \item Development of software for novel and emerging architectures
-  \item Application of specialist numerical and statistical methods within software
-  \item Optimisation of codes for use on high-performance computing infrastructure
-  \item Green computing - efficient implementations of algorithms/methods and use of hardware
-  \item ...
+\item The evolution of the term RSE,
+\item How education is reshaped by the digital transformation,
+\item What the impact of AI to the work of RSEs will look like,
+\item Changes to the Academic Institutions,
+\item Impact of alternate career paths for RSEs and other RTPs (\emph{Research Technical Professional}s).
+\item Ethics and social considerations,
 \end{itemize}
+
+\begin{story}{Introduction}{intro}
+In order to give a more explicit setting, imagine a world ten years from now.
+Not too much to be overwhelmingly far in the future, but distant enough for some changes to occur.
+Over this timeframe Kim\cite{Anzt2021} and Kay\cite{Goth2024}
+were able to ride the waves of their career in Research Software Engineering and are now pondering early retirement on a sunset blessed beach.
+%MM-NOTE: the above scenario (early retirement) is contrary to all trends in retirement age development of public employees across the "global north".
+But the world has moved on.
+When they look back, what will they see as problems which are now solved?
+Which tasks are persisting, where they are now happy that a younger generation is now taking up the baton and carries forward their work.
+What issues have newly emerged, where they are now just happy to say: “Oh well… My successor takes care of that”.\\
+In order to bring this future to life these two personas will accompany us and we will shed light on certain aspects
+of their future in respective small story boxes.
+ \end{story}
 
@@
 
-\subsection{What is in this future for RSEs?}
+\section{What is in this future for RSEs and where to go from here?}
 FG-NOTE: I think we can use this part to wrap things a bit up, and describe how this future then looks like.
@@
 GC-NOTE: Shall we add another story box here?
-
-\section{REALLY Long-term(10+ years) aspects}
-Technically this would mean taking everything from the previous section for granted and look
-at what becomes now possible.
 

### 2025-05-11 future_challenges_for_RSEng.tex
current status

@@
 
+\newcommand{\Eg}{E.\,g.}
+\newcommand{\eg}{e.\,g.}
+
 % Article frontmatter
@@
 \item What will the impact of AI be to the work of RSEs?
-\item How will academic education have transformed by then
+\item How will academic education have transformed by then,
+\item The Development of the RSE community,
 \item Ethics and social considerations
@@
 \item Impact of alternate career paths for RSEs and other RTPs (\emph{Research Technical Professional}s).
+\item Development of the RSE community,
 \item Ethics and social considerations,
@@
 \subsection{Definition and Evolution of RSEs}
-\begin{framed}
-\hfill How will the definition of an RSE evolve in ten years?\\
-
-\hfill What will not change in the essence of an RSE ten years from now?
-\end{framed}
 % Here we consider how the topics of RSEng will change in the face of the digitalization of society.
@@
 % Digitalisation will also feature more heavily in the domain curricula.
-
-The digital skills of domain scientists will increase,
+In this discussions the workshop participants found that
+the digital skills of domain scientists will increase,
 and software will become easier to interact with,
@@
 
-Regarding the definition of RSEs, we will still have a spectrum of competencies,
-from full professional RSEs to domain scientists who develop research software.
 While the RSE core competencies\cite{Goth2024} might remain unchanged,
@@
 to the other core competencies.
+Regarding the definition of RSEs, we will still have a spectrum of competencies,
+from full professional RSEs to domain scientists who develop research software.
 
@@
 Communication and collaboration skills will remain essential to engage
-with domain scientists and help them find technical solutions.
+with domain scientists and help them find technical solutions.\\
 
-\begin{FramedParagraphWithFootnotes}
-JN-NOTE:
+Looking at that on a broader scale, especially the education aspect deserves some attention.\\
 The Covid-19 pandemic acted as a catalyst for the digital transformation
@@
 and soon HPC practitioners\footnote{\url{https://www.hpc-carpentry.org}}.
-\end{FramedParagraphWithFootnotes}
 
-\begin{FramedParagraphWithFootnotes}
-JN-NOTE:
+\subsection{Development of the RSE community}
 In recent years, we saw the formation of RSE institutes and advocacy groups, such as
@@
 possibly through centrally-funded long-term positions at research institutions.
-\end{FramedParagraphWithFootnotes}
-
-\begin{FramedParagraphWithFootnotes}
-JN-NOTE:
-New RSE specialisations might arise.
-HPC-RSE and AI-RSE are quite recent developments,
-whose appeal is stimulated by large public and private
-investments in HPC facilities and AI technologies.
-We could imagine new RSE specialisations to emerge in the fields
-of quantum computing and federated online platforms.
-% European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories
-% European federation platform for HPC: https://eurohpc-ju.europa.eu/paving-way-eurohpc-federation-platform-2024-12-19_en
-% European quantum computing: https://digital-strategy.ec.europa.eu/en/policies/quantum
-\end{FramedParagraphWithFootnotes}
-
-\begin{FramedParagraphWithFootnotes}
-GC-NOTE: With longer-term funding and more trust to pick up longer term projects,
-some RSEs can focus on code modernization, cleanup, and modularization,
-which can pay off in faster and larger development in the future.
-This is not possible by short-term funded solo RSEs.
-\end{FramedParagraphWithFootnotes}
-
-\subsection{Education, and changes in the educational system}
-
-\begin{FramedParagraphWithFootnotes}
-FG-NOTE: I wonder if we need a separation between education of RSEs and changes in the education of the broader society.
-\end{FramedParagraphWithFootnotes}
-
-\begin{FramedParagraphWithFootnotes}
-GC-NOTE: Digital competencies in the general population continue to increase, making jumping to software engineering easier than before.
-
-GC-NOTE: By that point, we have an RSE Master's.
-\end{FramedParagraphWithFootnotes}
+The community of RSE's will get a significant boost in their acceptance if RSE Masters programmes
+become available at various universities.
 
 \subsection{Impact of AI and automation}
-Is it better than cheap PhDs?
+AI is already transforming research processes by enabling rapid analysis of massive datasets, uncovering patterns and insights that would be impossible or prohibitively time-consuming for humans alone to find.
+In research and development, AI accelerates scientific breakthroughs in fields like healthcare, climate science, and materials engineering by automating complex modeling and predictive tasks.
+For software development, AI-powered tools automate code generation, bug detection, and testing, significantly increasing productivity and reducing human error.
+Developers are shifting from manual coding to orchestrating AI-driven workflows, focusing more on high-level design, integration, and creative problem-solving while AI handles repetitive or routine tasks.
+The integration of AI is democratizing access to advanced technologies and analytics, enabling non-experts to participate in research and software creation, and fostering greater collaboration across disciplines.
+As AI continues to evolve, it will further blur the lines between human and machine roles, making research and software development faster, more efficient, and more innovative than ever before.
+With the general availability of AI the interaction with researchers will change at a fundamental level.
+Whereas previously RSEs have been points of knowledge that could translate human language formulated problems into technical descriptions and then provide solutions to it,
+It can be foreseen, that AI drives this low-key interaction to a highly specialized level of knowledge,
+since AI copilots are available 24/7.
 
-\begin{FramedParagraphWithFootnotes}
-FG-NOTE: AI will change the interaction with researchers at a fundamental level.
-Whereas previously RSEs have been points of knowledge that could translate human language formulated problems into technical descriptions and then provide solutions to it,
-I foresee that AI drives this low-key interaction to a highly specialized level of knowledge, since the availability of AI is far better.
-\end{FramedParagraphWithFootnotes}
+Currently we are expanding and documenting our work in the open and making our code available,
+therefore there will be more structured material to train GenAI on.
+However, as people get quicker answers from GenAI, the community resources (\eg Stack Overflow, Wikipedia)
+get poorer.
 
-\begin{FramedParagraphWithFootnotes}
-GC-NOTE: As we continue and expand documenting our work in the open and making our code available,
-there will be more structured material to train GenAI on.
+\subsection{Impact of Complexity and Usability in Research Software}
 
-However, as people get quicker answers from GenAI, the community resources (Stack Overflow, Wikipedia) get poorer.
-\end{FramedParagraphWithFootnotes}
+The workshop participants found that in earlier learning stages,
+details may be intentionally hidden,
+fearing a diminishing capability to handle technical complexity in research environments.
+As a result, RSEs invest more effort in making individual research software components
+less complex with better user experience.
+This made participants worry about growing mismatch of peoples' expectations.
+Participants expected an over-reliance on generative AI tools as an extreme manifestation of this issue.
+No concrete examples were given on which complexities are now hidden behind abstractions.
+
+We acknowledge but would also like to challenge the observations of the participants.
+We believe that ``we are standing on the shoulders of giants'':
+By reducing the complexity of individual software components,
+we are now able to develop systems of complexity and capability not previously possible,
+allowing the scientific community to address questions not previously in grasp.
+When it comes to teaching, it is true that students need to learn how to handle complexity,
+but it is also important to learn how to abstract from details
+by building hierarchical systems from these abstractions encapsulated in reusable components, thereby managing complexity.
+It is clear that we need both RSEs that can dive deep into single projects of significant complexity,
+as well as RSEs that can manage complex systems of almost black-box components,
+and the future of RSE education should aim to prepare students for both roles.
 
@@
 % society implications
-The technology people working as RSE might have, well directly or indirectly, ethical implications in areas such as social engineering [TODO find source], climate [TODO find sourc] and more.
+The technology people working as RSE might have, well directly or indirectly, ethical implications in areas such as social engineering [TODO find source], climate [TODO find source] and more.
 Another example might be that they are developing code that can also be used for military purposes, such as numerical solvers or computer vision systems.
@@
 
+\subsection{Education, and changes in the educational system}
+Digital competencies in the general population continue to increase, making jumping to software engineering easier than before.
+
+\begin{whatis}{What do we want to have achieved}{academic}
+School curricula have embraced the digital transformation, and have included basics of computer interaction
+in every schoolyear.
+In Universities, the first generations of RSE-Master graduates find their ay into research departments,
+and accelerate the pace of science.
+\end{whatis}
+
 \subsection{How have academic Institutions transformed by then?}
+GC-NOTE: With longer-term funding and more trust to pick up longer term projects,
+some RSEs can focus on code modernization, cleanup, and modularization,
+which can pay off in faster and larger development in the future.
+This is not possible by short-term funded solo RSEs.
+
 \begin{whatis}{What do we want to have achieved}{academic}
@@
  FG-NOTE: RSEs are one type of specialization of RTPs ins science. Do we see any other?
+\begin{FramedParagraphWithFootnotes}
+JN-NOTE:
+New RSE specialisations might arise.
+HPC-RSE and AI-RSE are quite recent developments,
+whose appeal is stimulated by large public and private
+investments in HPC facilities and AI technologies.
+We could imagine new RSE specialisations to emerge in the fields
+of quantum computing and federated online platforms.
+% European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories
+% European federation platform for HPC: https://eurohpc-ju.europa.eu/paving-way-eurohpc-federation-platform-2024-12-19_en
+% European quantum computing: https://digital-strategy.ec.europa.eu/en/policies/quantum
+\end{FramedParagraphWithFootnotes}
 
@@
 and potentially request and coordinate support from external RSEs.
-
-\subsection{Impact of Complexity and Usability in Research Software}
-
-The workshop participants found that in earlier learning stages,
-details may be intentionally hidden,
-fearing a diminishing capability to handle technical complexity in research environments.
-As a result, RSEs invest more effort in making individual research software components
-less complex with better user experience.
-This made participants worry about growing mismatch of peoples' expectations.
-Participants expected an over-reliance on generative AI tools as an extreme manifestation of this issue.
-No concrete examples were given on which complexities are now hidden behind abstractions.
-
-We acknowledge but would also like to challenge the observations of the participants.
-We believe that ``we are standing on the shoulders of giants'':
-By reducing the complexity of individual software components,
-we are now able to develop systems of complexity and capability not previously possible,
-allowing the scientific community to address questions not previously in grasp.
-When it comes to teaching, it is true that students need to learn how to handle complexity,
-but it is also important to learn how to abstract from details
-by building hierarchical systems from these abstractions encapsulated in reusable components, thereby managing complexity.
-It is clear that we need both RSEs that can dive deep into single projects of significant complexity,
-as well as RSEs that can manage complex systems of almost black-box components,
-and the future of RSE education should aim to prepare students for both roles.
 
@@
 % ========================
-\bibliographystyle{eceasst}
-\bibliography{bibliography/bibliography,extra}
+%\bibliographystyle{eceasst}
+%\bibliography{bibliography/bibliography,extra}
 

### 2025-05-12 future_challenges_for_RSEng.tex
Update future_challenges_for_RSEng.tex
Co-authored-by: Magnus Hagdorn <Magnus.Hagdorn@marsupium.org>

--- 2025-05-10 future_challenges_for_RSEng.tex
@@
 reducing reliance on RSEs to train newcomers on the most basic tasks
-(JN: for example, versioning and archiving?)
-(GC: Higher-level and more intuitive workflows that require less training).
+(JN: for example, versioning and archiving?).
 User training sessions might focus on more advanced topics
@@
 
-\begin{FramedParagraphWithFootnotes}
-GC-NOTE: With longer-term funding and more trust to pick up longer term projects,
-some RSEs can focus on code modernization, cleanup, and modularization,
-which can pay off in faster and larger development in the future.
-This is not possible by short-term funded solo RSEs.
-\end{FramedParagraphWithFootnotes}
-
 \subsection{Education, and changes in the educational system}
-
-\begin{FramedParagraphWithFootnotes}
 FG-NOTE: I wonder if we need a separation between education of RSEs and changes in the education of the broader society.
-\end{FramedParagraphWithFootnotes}
-
-\begin{FramedParagraphWithFootnotes}
-GC-NOTE: Digital competencies in the general population continue to increase, making jumping to software engineering easier than before.
-
-GC-NOTE: By that point, we have an RSE Master's.
-\end{FramedParagraphWithFootnotes}
 
@@
 Is it better than cheap PhDs?
-
-\begin{FramedParagraphWithFootnotes}
 FG-NOTE: AI will change the interaction with researchers at a fundamental level.
@@
 I foresee that AI drives this low-key interaction to a highly specialized level of knowledge, since the availability of AI is far better.
-\end{FramedParagraphWithFootnotes}
-
-\begin{FramedParagraphWithFootnotes}
-GC-NOTE: As we continue and expand documenting our work in the open and making our code available,
-there will be more structured material to train GenAI on.
-
-However, as people get quicker answers from GenAI, the community resources (Stack Overflow, Wikipedia) get poorer.
-\end{FramedParagraphWithFootnotes}
 
@@
 
-MM-NOTE: (Flo believes it was from Michele) Are we assuming that progress will bring prosperity?
-Will we remain constrained by needing to produce products/artifacts?
-(GC: Where "product/artifacts", maybe also "publications"?)
+MM-NOTE: (Flo believes it was from Michele) Are we assuming that progress will bring prosperity? Will we remain constrained by needing to produce products/artifacts?
 
@@
 software for cheap water monitoring devices in the global south. Kim is currently designing the software service stack
-at her university, and increasingly struggles to find compelling Open Source alternatives that lets them keep their data
+at her university, and increasingly struggles to find compelling Open Source alternatives that let's them keep their data
 to software that is backed by large corporate entities. So far she still has the support of her university board, since it values
 strategic autonomy.
-Her son Kyle on the other hand started out with a degree in Physics and then took one of the newly created RSE Master's courses,
+Her son Kyle on the other hand started out with a degree in Physics and then took one of the newly created RSE Master courses,
 but quickly decided that work in academia with its time limited contracts is not attractive.
 Governments around the world are struggling to reposition themselves in this emerging deglobalized world with its new blocks
-and therefore have deprioritized publicly funded research. Kyle decided to go to a defense company since they offer a
+and therefore have deprioritized publicly funded research. Kyle decided to go to a defence company since they offer a
 more secure employment in this polarized world.
@@
 We also hope, that specialized clusters form, with certain RSE units specializing in certain tasks like data management, HPC, or visualization.
-At universities which have them, we want to see that researchers naturally accept them as part of their used services.
+At universities which have them, we want to see, that researchers naturally accept them as part of their used services.
 \end{whatis}
@@
 Organizations will have to come up with models for hiring people. While
-a standardized RSE Master's makes it easier, it will be interesting to see if
+a standardized RSE masters makes it easier, it will be interesting to see, if
 RSE are still often hired by word of mouth or if the master's really makes it more comparable.
-
-\begin{FramedParagraphWithFootnotes}
-  GC-NOTE: New section on "New challenges":
-
-  \begin{itemize}
-    \item Networking of embedded RSEs and RSE units
-    \item Everyone needs infrastructure: does everyone need to host their own infrastructure?
-    \item Everyone hosts their infrastructure, but then everyone stays in silos? We need federated infrastructure.
-  \end{itemize}
-\end{FramedParagraphWithFootnotes}
 
@@
 \subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}
+The RSE is probably one of the first professionalizations in the academic mid-level that is able
+to collectively voice their concerns. This enables the possibility to reshape universities.
+If more people are employed as mid-level, this makes a steadier pool of people
+available that can integrate new people that are regularly coming into departments and know the local facilities.
+On the other hand, an upcoming challenge how universities decide to design the contracts of people employed
+in RSE units. Will they be permanent, or will there also be an up-or-out mentality?
 
-RSEs can either be embedded in research groups (belonging to the mid-level university staff),
-or be members of larger RSE units. (TODO: Cite RSE groups paper)
-The participants had the opinion that focusing on embedded RSEs would make onboarding new RSEs to existing projects easier.
-This, however, comes with additional funding challenges,
-as long-term funding is more difficult to acquire in lower levels of an organization.
-Third-party funding can mitigate this issue in the mid-term.
-However, the participants considered central RSE groups more likely in the future,
-due to the fact that there are already examples and activities from various universities in this direction
-(examples include the universities of Jena, Heidelberg, Göttingen).
-We expect funding in central groups to evolve from initially temporary positions to permanent positions,
-as the groups themselves prove themselves valuable.
-Embedded RSEs and central groups will need to network with each other,
-and potentially request and coordinate support from external RSEs.
+As outlined in [cite RSE group paper] not all universities need to opt for strengthening
+the academic mid-level. Some might have chosen to set up external companies to outsource typical RSE tasks,
+thereby providing career paths outside of academia.
+
+FG-NOTE: By Nick. I've put it here, due to the relation with the RTP question.
 
@@
 The workshop participants found that in earlier learning stages,
-details may be intentionally hidden,
+details of software frameworks and libraries may be intentionally hidden,
 fearing a diminishing capability to handle technical complexity in research environments.
@@
 
-GC-NOTE: Shall we add another story box here?
-
 \section{REALLY Long-term(10+ years) aspects}
@@
 at what becomes now possible.
-
-GC-NOTE: RSE develop together with and for scientists.
-Better software leads to better research,
-better research leads to more better software.
 

### 2025-05-12 future_challenges_for_RSEng.tex
Update future_challenges_for_RSEng.tex
Co-authored-by: Magnus Hagdorn <Magnus.Hagdorn@marsupium.org>

@@
 details of software frameworks and libraries may be intentionally hidden,
-fearing a diminishing capability to handle technical complexity in research environments.
+fearing a reduced capability to handle technical complexity in research environments.
 As a result, RSEs invest more effort in making individual research software components

### 2025-05-12 future_challenges_for_RSEng.tex

--- 2025-05-11 future_challenges_for_RSEng.tex
@@
 \section{Introduction}
-This opinion article stems from a Bof session at deRSE25 in Karlsruhe\cite{Goth2025EndRSEng}.
-Therefore it stems from a discussion among the participants of this workshop and hence cannot claim that it represents the entire deRSE community.
-Nevertheless we hope, that we succeeded in writing up something that
-\begin{itemize}
-\item is pleasant to read,
-\item gives something to think about,
-\item motivates engaging in work towards a better future.
-\end{itemize}
-In order to structure the discussion, we decided on an imaginable setting ten years from now.
-In this world we will consider the dimensions of 
-\begin{itemize}
-\item Evolution of the term RSEs
-\item The development of academic institutions
-\item What will the impact of AI be to the work of RSEs?
-\item How will academic education have transformed by then,
-\item The Development of the RSE community,
-\item Ethics and social considerations
-\end{itemize}
-In each aspect we will quickly summarise on what we want to have achieved by then, and what will be the challenges that arise from that.
+
+Research Software Engineering is currently an emerging field,
+with an increasing integration into various aspects of research.
+In this opinion paper, we try to imagine the state of our field
+and community in the mid-term future, focusing on future challenges.
+We present here the results of a workshop discussion at the
+deRSE25 conference in Karlsruhe\cite{Goth2025EndRSEng}, without
+claiming to represent the entire deRSE community.
+Nevertheless, by disseminating the discussed ideas,
+we aim to engage the community in imagining the future.
+In the following sections, we consider the evolution of the RSE definition,
+the development of RSE education,
+the impact of automation and generative artificial intelligence in RSE work,
+the emerging ethical considerations,
+and the transformations expected in the research organizations.
+We then attempt to make long-term predictions
+and we conclude the paper with our wishes and recommendations.
+
+\section{The Situation}
+In order to give a more explicit setting, imagine a world ten years from now.
+Not too much to be overwhelmingly far in the future, but distant enough for some changes to occur.
+Over this timeframe Kim\cite{Anzt2021} and Kay\cite{Goth2024}
+were able to ride the waves of their career in Research Software Engineering and are now pondering early retirement on a sunset blessed beach.
+%MM-NOTE: the above scenario (early retirement) is contrary to all trends in retirement age development of public employees across the "global north".
+But the world has moved on.
+When they look back, what will they see as problems which are now solved?
+Which tasks are persisting, where they are now happy that a younger generation is now taking up the baton and carries forward their work.
+What issues have newly emerged, where they are now just happy to say: “Oh well… My successor takes care of that”.
+\begin{story}{Introduction}{intro}
+In order to bring this future to life these two personas will accompany us and we will shed light on certain aspects
+of their future in respective small story boxes.
+ \end{story}
 

### 2025-05-12 future_challenges_for_RSEng.tex
Follow-up in resolving conflict

@@
 and we conclude the paper with our wishes and recommendations.
-
-\section{The Situation}
-In order to give a more explicit setting, imagine a world ten years from now.
-Not too much to be overwhelmingly far in the future, but distant enough for some changes to occur.
-Over this timeframe Kim\cite{Anzt2021} and Kay\cite{Goth2024}
-were able to ride the waves of their career in Research Software Engineering and are now pondering early retirement on a sunset blessed beach.
-%MM-NOTE: the above scenario (early retirement) is contrary to all trends in retirement age development of public employees across the "global north".
-But the world has moved on.
-When they look back, what will they see as problems which are now solved?
-Which tasks are persisting, where they are now happy that a younger generation is now taking up the baton and carries forward their work.
-What issues have newly emerged, where they are now just happy to say: “Oh well… My successor takes care of that”.
-\begin{story}{Introduction}{intro}
-In order to bring this future to life these two personas will accompany us and we will shed light on certain aspects
-of their future in respective small story boxes.
- \end{story}
 

### 2025-05-12 future_challenges_for_RSEng.tex
today's state

@@
 \section{Introduction}
-
-Research Software Engineering is currently an emerging field,
-with an increasing integration into various aspects of research.
-In this opinion paper, we try to imagine the state of our field
-and community in the mid-term future, focusing on future challenges.
-We present here the results of a workshop discussion at the
-deRSE25 conference in Karlsruhe\cite{Goth2025EndRSEng}, without
-claiming to represent the entire deRSE community.
-Nevertheless, by disseminating the discussed ideas,
-we aim to engage the community in imagining the future.
-In the following sections, we consider the evolution of the RSE definition,
-the development of RSE education,
-the impact of automation and generative artificial intelligence in RSE work,
-the emerging ethical considerations,
-and the transformations expected in the research organizations.
-We then attempt to make long-term predictions
-and we conclude the paper with our wishes and recommendations.
+This opinion article stems from a Bof session at deRSE25 in Karlsruhe\cite{Goth2025EndRSEng}.
+Therefore it stems from a discussion among the participants of this workshop and hence cannot claim that it represents the entire deRSE community.
+Nevertheless we hope, that we succeeded in writing up something that
+\begin{itemize}
+\item is pleasant to read,
+\item gives something to think about,
+\item motivates engaging in work towards a better future.
+\end{itemize}
+In order to structure the discussion, we decided on an imaginable setting ten years from now.
+In this world we will consider the dimensions of 
+\begin{itemize}
+\item Evolution of the term RSEs
+\item The development of academic institutions
+\item What will the impact of AI be to the work of RSEs?
+\item How will academic education have transformed by then,
+\item The Development of the RSE community,
+\item Ethics and social considerations
+\end{itemize}
+In each aspect we will quickly summarise on what we want to have achieved by then, and what will be the challenges that arise from that.
 
@@
 \subsection{How have academic Institutions transformed by then?}
-GC-NOTE: With longer-term funding and more trust to pick up longer term projects,
-some RSEs can focus on code modernization, cleanup, and modularization,
-which can pay off in faster and larger development in the future.
-This is not possible by short-term funded solo RSEs.
+Academia, while claiming to be innovative (cite WissZeitVG here), has organisationally often adopted time-scales for change that are more reminiscient to civil services like pension funds.
+Nevertheless, the presence, even today, of initiatives like the HiddenRef hints at the potential for specialisation of research tasks.
+This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.
+We have alreadyseen some specialisation trends in recent years such as
+the HPC-RSE and AI-RSE which are quite recent developments, whose appeal is stimulated by large public and private investments in HPC facilities and AI technologies.
+But we could imagine new RSE specialisations to emerge in the fields
+of quantum computing, such as the Quantum-RSE, and in federated online platforms.
+% European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories
+% European federation platform for HPC: https://eurohpc-ju.europa.eu/paving-way-eurohpc-federation-platform-2024-12-19_en
+% European quantum computing: https://digital-strategy.ec.europa.eu/en/policies/quantum
 
+On the other hand the existence of people with a certain specilised skillset opens up possiblities for pooling.
+Universities will benefit from pooling these specialists into RSE units, and offer their services to researchers.
 \begin{whatis}{What do we want to have achieved}{academic}
@@
 We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department to oversee specific projects[FIXME: cite group paper].
-We also hope, that specialized clusters form, with certain RSE units specializing in certain tasks like data management, HPC, or visualization.
+We also hope, that specialized clusters form that pool the skills of certain specialised RSEs, with certain RSE units specializing in certain tasks like data management, HPC, or visualization.
 At universities which have them, we want to see that researchers naturally accept them as part of their used services.
 \end{whatis}
+Hopefully, these RSE-units are endowed with enough trust by the university boards that translates into stable funding.
+With longer-term funding and more trust to pick up longer term projects,
+some RSEs can focus on code modernization, cleanup, and modularization,
+which can pay off in faster and larger development in the future.
+This is an opportunity that would not possible by short-term funded solo RSEs.
 
-\subsubsection{Which challenges will still persist}
+\emph{Which challenges will still persist}\\
 The workshop participants expect that there will still be budgeting and administrative hurdles to creating
@@
 among institutions with successful RSE Teams.
-On a more organisational level, we expect the sharing of models, to fit the different organisational constraints.
-
-What are open questions here?
-Organizations will have to come up with models for hiring people. While
-a standardized RSE Master's makes it easier, it will be interesting to see if
-RSE are still often hired by word of mouth or if the master's really makes it more comparable.
-
-\begin{FramedParagraphWithFootnotes}
-  GC-NOTE: New section on "New challenges":
-
-  \begin{itemize}
-    \item Networking of embedded RSEs and RSE units
-    \item Everyone needs infrastructure: does everyone need to host their own infrastructure?
-    \item Everyone hosts their infrastructure, but then everyone stays in silos? We need federated infrastructure.
-  \end{itemize}
-\end{FramedParagraphWithFootnotes}
+On a more organisational level, we expect the sharing of models on how how to organise diiferent RSEs, to fit the different organisational constraints.
+After that Organisations will have to come up with models for hiring people.
+While a standardized RSE Master's makes it easier, it will be interesting to see if
+RSE are still often hired by word of mouth within an organisation(``I have this PhD here, before their PostDoc they would be a good fit for the RSE department...'' ) or if the RSE master's degree ensures a hiring process on a more standardised level.
 
@@
 
-\subsection{What are new tasks of RTPs?}
- FG-NOTE: RSEs are one type of specialization of RTPs ins science. Do we see any other?
-\begin{FramedParagraphWithFootnotes}
-JN-NOTE:
-New RSE specialisations might arise.
-HPC-RSE and AI-RSE are quite recent developments,
-whose appeal is stimulated by large public and private
-investments in HPC facilities and AI technologies.
-We could imagine new RSE specialisations to emerge in the fields
-of quantum computing and federated online platforms.
-% European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories
-% European federation platform for HPC: https://eurohpc-ju.europa.eu/paving-way-eurohpc-federation-platform-2024-12-19_en
-% European quantum computing: https://digital-strategy.ec.europa.eu/en/policies/quantum
-\end{FramedParagraphWithFootnotes}
+\emph{New challenges}
+If we follow the ideas laid out in (FIXME: cite RSE group paper), then in addition to university wide RSE units,
+there will be embedded RSEs in the departments. Shaping and developing these network bonds will be a task for the future.
+While networking embedded RSEs and RSE units is essential for building robust, sustainable research software infrastructure, but the question arises whether every group truly needs to host and manage its own infrastructure(FIXME: maybe cite this here: https://www.dfg.de/en/research-funding/funding-opportunities/programmes/infrastructure/lis/funding-opportunities/research-software-infrastructures).
+When each team or institution hosts its own systems, this often leads to siloed environments where resources, expertise, and data are isolated, limiting collaboration and creating inefficiencies.
+Such silos can result in duplicated efforts, higher costs, and challenges in maintaining data integrity and governance, as each group independently manages similar infrastructure and workflows.
+While local control offers flexibility, it also increases complexity and makes it harder to share innovations or best practices across the wider research community.
+To overcome these challenges, federated infrastructure models are needed, enabling shared access, interoperability, and coordinated development while still allowing for some local autonomy(FIXME: maybe cite this here: https://eurise-network.github.io/).
+Federated approaches break down silos, promote knowledge exchange, and ensure that infrastructure investments benefit the broader ecosystem rather than just isolated groups
 
 \subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}
-
 RSEs can either be embedded in research groups (belonging to the mid-level university staff),
@@
 % ========================
-%\bibliographystyle{eceasst}
-%\bibliography{bibliography/bibliography,extra}
+\bibliographystyle{eceasst}
+\bibliography{bibliography/bibliography,extra}
 

### 2025-05-12 future_challenges_for_RSEng.tex

@@
 \section{Introduction}
-This opinion article stems from a Bof session at deRSE25 in Karlsruhe\cite{Goth2025EndRSEng}.
-Therefore it stems from a discussion among the participants of this workshop and hence cannot claim that it represents the entire deRSE community.
-Nevertheless we hope, that we succeeded in writing up something that
-\begin{itemize}
-\item is pleasant to read,
-\item gives something to think about,
-\item motivates engaging in work towards a better future.
-\end{itemize}
-In order to structure the discussion, we decided on an imaginable setting ten years from now.
-In this world we will consider the dimensions of 
-\begin{itemize}
-\item Evolution of the term RSEs
-\item The development of academic institutions
-\item What will the impact of AI be to the work of RSEs?
-\item How will academic education have transformed by then,
-\item The Development of the RSE community,
-\item Ethics and social considerations
-\end{itemize}
-In each aspect we will quickly summarise on what we want to have achieved by then, and what will be the challenges that arise from that.
+
+Research Software Engineering is currently an emerging field,
+with an increasing integration into various aspects of research.
+In this opinion paper, we try to imagine the state of our field
+and community in the mid-term future, focusing on future challenges.
+We present here the results of a workshop discussion at the
+deRSE25 conference in Karlsruhe\cite{Goth2025EndRSEng}, without
+claiming to represent the entire deRSE community.
+Nevertheless, by disseminating the discussed ideas,
+we aim to engage the community in imagining the future.
+In the following sections, we consider the evolution of the RSE definition,
+the development of RSE education,
+the impact of automation and generative artificial intelligence in RSE work,
+the emerging ethical considerations,
+and the transformations expected in the research organizations.
+We then attempt to make long-term predictions
+and we conclude the paper with our wishes and recommendations.
 

### 2025-05-12 future_challenges_for_RSEng.tex
now most ideas boxes worked in

@@
 FG-NOTE: I think we can use this part to wrap things a bit up, and describe how this future then looks like.
+GC-NOTE: Shall we add another story box here? Ideas what we write here?
 
-GC-NOTE: Shall we add another story box here?
-
-GC-NOTE: RSE develop together with and for scientists.
-Better software leads to better research,
+Digitalisation will lead to changes in the way societies interact and therefore will translate
+into changes in how academia perceives itself. The differentiation of job profiles in academia,
+that has started with profiles such as the RSE will continue. Societies will be facing new questions, that arise
+from technologies such as generally available AI systems. RSEs will very likely be doing different work,
+than what they are doing today, but their core task of enhancing the scientific output of researchers, other human
+beings, will stay the same.
+They will still develop software artifacts together with scientists,
+since Better software leads to better research,
 better research leads to more better software.

### 2025-05-12 future_challenges_for_RSEng.tex
syncronize structure

@@
 the emerging ethical considerations,
-and the transformations expected in the research organizations.
-We then attempt to make long-term predictions
-and we conclude the paper with our wishes and recommendations.
+and the transformations expected in the research organizations,
+and if there will be other in academia covered by new RTPs (\emph{Research Technical Professional}s).
+We conclude the paper with a summary and with our wishes and recommendations.
 
@@
 \item The evolution of the term RSE,
+\item Development of the RSE community,
+\item What the impact of AI to the work of RSEs will look like,
+\item the impact of Complexity and Usability in Research Software
+\item Ethics and social considerations,
 \item How education is reshaped by the digital transformation,
-\item What the impact of AI to the work of RSEs will look like,
 \item Changes to the Academic Institutions,
-\item Impact of alternate career paths for RSEs and other RTPs (\emph{Research Technical Professional}s).
-\item Development of the RSE community,
-\item Ethics and social considerations,
+\item Impact of alternate career paths for RSEs and other RTPs.
 \end{itemize}

### 2025-05-12 future_challenges_for_RSEng.tex

@@
 The workshop participants found that in earlier learning stages,
-details may be intentionally hidden,
-fearing a diminishing capability to handle technical complexity in research environments.
+details of software frameworks and libraries may be intentionally hidden,
+fearing a reduced capability to handle technical complexity in research environments.
 As a result, RSEs invest more effort in making individual research software components

### 2025-05-12 future_challenges_for_RSEng.tex
Rework for readability

--- 2025-05-12 future_challenges_for_RSEng.tex
@@
 
-The workshop participants found that in earlier learning stages,
-details may be intentionally hidden,
-fearing a diminishing capability to handle technical complexity in research environments.
-As a result, RSEs invest more effort in making individual research software components
-less complex with better user experience.
-This made participants worry about growing mismatch of peoples' expectations.
-Participants expected an over-reliance on generative AI tools as an extreme manifestation of this issue.
+The workshop participants found that, in earlier learning stages,
+training may abstract away complexities that will always be present in research.
+For this reason, the participants fear that the capability of researchers to deal with complexity
+of software frameworks and libraries will reduce over time.
+As a result, RSEs will need to invest more effort in making the interfaces of individual research software components
+less complex, as the user experience expectations of researchers will continue to grow.
+Participants expect an over-reliance on generative AI tools as an extreme manifestation of this issue.
 No concrete examples were given on which complexities are now hidden behind abstractions.
@@
 By reducing the complexity of individual software components,
-we are now able to develop systems of complexity and capability not previously possible,
-allowing the scientific community to address questions not previously in grasp.
+we are now able to develop systems of complexity and capability not previously possible.
+This divide-and-conquer approach allows the scientific community to address questions not previously in grasp.
 When it comes to teaching, it is true that students need to learn how to handle complexity,
-but it is also important to learn how to abstract from details
-by building hierarchical systems from these abstractions encapsulated in reusable components, thereby managing complexity.
+by being able to develop deep understanding of individual components.
+However, it is also important to learn how to manage complexity
+by building hierarchical systems of well-encapsulated, reusable components.
 It is clear that we need both RSEs that can dive deep into single projects of significant complexity,

### 2025-05-12 future_challenges_for_RSEng.tex
Add templates for the story and whatis boxes

--- 2025-05-12 future_challenges_for_RSEng.tex
@@
 \subsection{Definition and Evolution of RSEs}
+\begin{whatis}{What do we want to have achieved}{ethics}
+The term RSE is established in the minds of university boards, as a valuable member in the academic framework.
+RSEs are developed using specialized master's program, and their support is valued by researchers.
+\end{whatis}
 % Here we consider how the topics of RSEng will change in the face of the digitalization of society.
@@
 
+\begin{story}{The Definition of RSEs}{defrse}
+A story, about how Kim and Kay started out with a certain job description that now totally
+has changed meaning.
+\end{story}
+
 \subsection{Development of the RSE community}
+\begin{whatis}{What do we want to have achieved}{ethics}
+We have more national RSE communities that are interacting with other partners in their nation.
+There are regular international conferences.
+\end{whatis}
 In recent years, we saw the formation of RSE institutes and advocacy groups, such as
@@
 
+\begin{story}{Development of RSE community}{RSEcommunity}
+A story about Kim participating in the first international RSE conference in a country on the southern hemisphere.
+\end{story}
+
 \subsection{Impact of AI and automation}
+\begin{whatis}{What do we want to have achieved}{ai}
+RSEs are educated about AI models, and can help researchers reap their benefits, while
+clearly knowing their limits.
+\end{whatis}
 AI is already transforming research processes by enabling rapid analysis of massive datasets, uncovering patterns and insights that would be impossible or prohibitively time-consuming for humans alone to find.
@@
 
+\begin{story}{The interaction of RSEs with AI}{RSEs_with_AI}
+A story about Kay co-developing with AI.
+\end{story}
+
 \subsection{Impact of Complexity and Usability in Research Software}
+\begin{whatis}{What do we want to have achieved}{abstraction}
+Gerasimos? Jan?
+\end{whatis}
 
@@
 
+\begin{story}{The benefits of abstraction}{abstraction}
+A story about Kim and Kay effortlessly progressing where years ago they would be caught up in details.
+\end{story}
+
 \subsection{Ethics and social consideration}
+\begin{whatis}{What do we want to have achieved}{ethics}
+Oh well..., is there really something that we could have achieved here?
+maybe open Science? Data awareness? Data Protection?
+\end{whatis}
 
@@
 \subsection{How have academic Institutions transformed by then?}
+\begin{whatis}{What do we want to have achieved}{academic}
+Similar to RSEs that are known by 2035 in university boards, we expect to be the concept of RSE units to be in the minds
+of academic staff at any level.
+We hope to see that universities use the presence of an RSE Department actively as a means to differentiate themselves from other universities.
+We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department to oversee specific projects[FIXME: cite group paper].
+We also hope, that specialized clusters form that pool the skills of certain specialised RSEs, with certain RSE units specializing in certain tasks like data management, HPC, or visualization.
+At universities which have them, we want to see that researchers naturally accept them as part of their used services.
+\end{whatis}
 Academia, while claiming to be innovative (cite WissZeitVG here), has organisationally often adopted time-scales for change that are more reminiscient to civil services like pension funds.
@@
 Universities will benefit from pooling these specialists into RSE units, and offer their services to researchers.
-\begin{whatis}{What do we want to have achieved}{academic}
-We hope to see that universities use the presence of an RSE Department actively as a means to differentiate themselves from other universities.
-We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department to oversee specific projects[FIXME: cite group paper].
-We also hope, that specialized clusters form that pool the skills of certain specialised RSEs, with certain RSE units specializing in certain tasks like data management, HPC, or visualization.
-At universities which have them, we want to see that researchers naturally accept them as part of their used services.
-\end{whatis}
 Hopefully, these RSE-units are endowed with enough trust by the university boards that translates into stable funding.
@@
 \subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}
+\begin{whatis}{What do we want to have achieved}{academic}
+We would like to see the emergence of more dedicated personel, i.e. other RTPs next to the RSE.
+\end{whatis}
 RSEs can either be embedded in research groups (belonging to the mid-level university staff),
@@
 and potentially request and coordinate support from external RSEs.
+
+\begin{story}{New career paths}{New career paths}
+A story, about Ken delving into a new direction as RTP.
+\end{story}
 

### 2025-05-12 future_challenges_for_RSEng.tex
finalise whatis boxes

@@
 RSEs are developed using specialized master's program, and their support is valued by researchers.
+We still teach researchers basic software skills, but the content of this lecture has changed drastically.
 \end{whatis}
@@
 We have more national RSE communities that are interacting with other partners in their nation.
-There are regular international conferences.
+There are regular international conferences. The installation of RSE Master's programmes led to
+increased numbers of the members of the national societies.
 \end{whatis}
@@
 School curricula have embraced the digital transformation, and have included basics of computer interaction
-in every schoolyear.
-In Universities, the first generations of RSE-Master graduates find their ay into research departments,
+in every class level.
+In universities, the first generations of RSE-Master graduates find their way into research departments,
 and accelerate the pace of science.

### 2025-05-12 future_challenges_for_RSEng.tex
Add side boxes for complexity section

@@
 \begin{whatis}{What do we want to have achieved}{abstraction}
-Gerasimos? Jan?
+With a palette of readily reusable software components,
+and with their skills in managing hierarchical systems of complex software components with clear interfaces,
+RSEs can now build systems of scale and capability not previously possible.
 \end{whatis}
@@
 \begin{story}{The benefits of abstraction}{abstraction}
-A story about Kim and Kay effortlessly progressing where years ago they would be caught up in details.
+Years ago, Kim had to distribute the software library libkim as a source code package with a custom build system and detailed instructions on how to build it on some common systems,
+investing significant time in developing and maintaining this auxiliary toolkit and documentation.
+Around the same time, Kay was developing the research analysis tool kay-bio,
+which however only provided an application programming interface that
+required writing complex code even for simple and common analysis tasks.
+Nowadays, Kim could have started directly from a template based on established toolkits and workflows,
+while kay-bio accepts configuration in a simpler, standardized format and generative AI assistants
+can guide new users in preparing and debugging their analyses.
 \end{story}
@@
 software for cheap water monitoring devices in the global south. Kim is currently designing the software service stack
-at her university, and increasingly struggles to find compelling Open Source alternatives that lets them keep their data
-to software that is backed by large corporate entities. So far she still has the support of her university board, since it values
+at the university, and increasingly struggles to find compelling Open Source alternatives that lets them keep their data
+to software that is backed by large corporate entities. So far she still has the support of the university board, since it values
 strategic autonomy.
@@
 Avalon which was interested in setting up their own department for RSEs in order to not fall behind the university of Arcadia.
-Kim applies his experiences and improves upon the model set forth by the University of Arcadia. Since the university
+Kim improves upon the model set forth by the University of Arcadia based on valuable experience. Since the university
 of Avalon has to follow different rules not everything can be straightforwardly adapted. Life would be so much easier,

### 2025-05-12 future_challenges_for_RSEng.tex
Extend
Add affiliations

--- 2025-05-12 future_challenges_for_RSEng.tex
@@
 %\short{} % Short title of the article (optional)
+\newcommand{\authorRef}[1]{\texorpdfstring{\autref{#1}}{}}
 \author{
-Florian Goth\texorpdfstring{\autref{1}}{},
-Leyla Jael Castro\texorpdfstring{\autref{1}}{},
-Gerasimos Chourdakis\texorpdfstring{\autref{1}}{},
-Simon Christ\texorpdfstring{\autref{1}}{},
-Jeremy Cohen\texorpdfstring{\autref{1}}{},
-Jean-Noël Grad\texorpdfstring{\autref{1}}{},
-Magnus Hagdorn\texorpdfstring{\autref{1}}{},
-Toby Hodges\texorpdfstring{\autref{1}}{},
-Jan Linxweiler\texorpdfstring{\autref{1}}{},
-Frank Löffler\texorpdfstring{\autref{1}}{},
-Michele Martone\texorpdfstring{\autref{1}}{},
-Jan Philipp Thiele\texorpdfstring{\autref{1}}{},
-Nick del Grosso\texorpdfstring{\autref{1}}{}
+Florian Goth\authorRef{1},
+Leyla Jael Castro\authorRef{2},
+Gerasimos Chourdakis\authorRef{3},
+Simon Christ\authorRef{4},
+Jeremy Cohen\authorRef{5},
+Jean-Noël Grad\authorRef{6},
+Magnus Hagdorn\authorRef{7},
+Toby Hodges\authorRef{8},
+Jan Linxweiler\authorRef{9},
+Frank Löffler\authorRef{10},
+Michele Martone\authorRef{11},
+Jan Philipp Thiele\authorRef{12},
+Nick del Grosso\authorRef{13}
 } % Authors and references to addresses
-\institute{\autlabel{1} Fantasy University} % Institutes with labels
+\institute{% Institutes with labels
+\autlabel{1} Würzburg-Dresden Cluster of Excellence ct.qmat, University of Würzburg, 97074, Würzburg, Germany\par
+\autlabel{2} ZB MED Information Centre for Life Sciences, Cologne, Germany\par
+\autlabel{3} School of Computation, Information and Technology, Technical University of Munich, Garching, Germany;
+             Institute for Parallel and Distributed Systems, University of Stuttgart, Stuttgart, Germany\par
+\autlabel{4} Leibniz University Hannover, Department of Cell Biology and Biophysics, Computational Biology, Germany\par
+\autlabel{5} Imperial College London, London, UK\par
+\autlabel{6} Institute for Computational Physics, University of Stuttgart, Germany\par
+\autlabel{7} Geschäftsbereich IT, Charité Universitätsmedizin Berlin, Germany\par
+\autlabel{8} The Carpentries, USA\par
+\autlabel{9} Technische Universität Braunschweig, Germany\par
+\autlabel{10} Michael Stifel Center Jena {\&} Friedrich Schiller University Jena, Germany\par
+\autlabel{11} Leibniz Supercomputing Centre, Garching, Germany\par
+\autlabel{12} Weierstrass Institute, Berlin, Germany;
+              Leibniz University Hannover, Institute of Applied Mathematics, Scientific Computing, Hannover, Germany\par
+\autlabel{13} Institute for Experimental Epileptology and Cognition Research, Uniklinikum Bonn, Germany
+}
 \abstract{It's 2025 and we are, admittedly, still working on establishing and growing Research Software Engineering (RSE)

### 2025-05-12 future_challenges_for_RSEng.tex
Technical fixes

@@
 \subsection{Definition and Evolution of RSEs}
-\begin{whatis}{What do we want to have achieved}{ethics}
+\begin{whatis}{What do we want to have achieved}{definition-evolution-rse}
 The term RSE is established in the minds of university boards, as a valuable member in the academic framework.
@@
 \subsection{Development of the RSE community}
-\begin{whatis}{What do we want to have achieved}{ethics}
+\begin{whatis}{What do we want to have achieved}{development-rse-community}
 We have more national RSE communities that are interacting with other partners in their nation.
@@
 \subsection{Impact of AI and automation}
-\begin{whatis}{What do we want to have achieved}{ai}
+\begin{whatis}{What do we want to have achieved}{impact-ai-automation}
 RSEs are educated about AI models, and can help researchers reap their benefits, while
@@
 The ethics and values of the Research Software Engineer are grounded in their practice.
-They commit to the health, safety and welfare of the public and act in the interest of society, their employer and their clients \cite{Goth2023}.
+They commit to the health, safety and welfare of the public and act in the interest of society, their employer and their clients \cite{Goth2024}.
 
@@
 
-\begin{whatis}{What do we want to have achieved}{academic}
+\begin{whatis}{What do we want to have achieved}{education}
 School curricula have embraced the digital transformation, and have included basics of computer interaction
@@
 \subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}
-\begin{whatis}{What do we want to have achieved}{academic}
+\begin{whatis}{What do we want to have achieved}{career-paths}
 We would like to see the emergence of more dedicated personel, i.e. other RTPs next to the RSE.
@@
 RSEs can either be embedded in research groups (belonging to the mid-level university staff),
-or be members of larger RSE units. (TODO: Cite RSE groups paper)
+or be members of larger RSE units.\cite{Kempf2025-draft}
 The participants had the opinion that focusing on embedded RSEs would make onboarding new RSEs to existing projects easier.

### 2025-05-12 future_challenges_for_RSEng.tex
Technical fixes

@@
 %\short{} % Short title of the article (optional)
-\newcommand{\authorRef}[1]{\texorpdfstring{\autref{#1}}{}}
 \author{
-Florian Goth\authorRef{1},
-Leyla Jael Castro\authorRef{2},
-Gerasimos Chourdakis\authorRef{3},
-Simon Christ\authorRef{4},
-Jeremy Cohen\authorRef{5},
-Jean-Noël Grad\authorRef{6},
-Magnus Hagdorn\authorRef{7},
-Toby Hodges\authorRef{8},
-Jan Linxweiler\authorRef{9},
-Frank Löffler\authorRef{10},
-Michele Martone\authorRef{11},
-Jan Philipp Thiele\authorRef{12},
-Nick del Grosso\authorRef{13}
+Florian Goth\texorpdfstring{\autref{1}}{},
+Leyla Jael Castro\texorpdfstring{\autref{1}}{},
+Gerasimos Chourdakis\texorpdfstring{\autref{1}}{},
+Simon Christ\texorpdfstring{\autref{1}}{},
+Jeremy Cohen\texorpdfstring{\autref{1}}{},
+Jean-Noël Grad\texorpdfstring{\autref{1}}{},
+Magnus Hagdorn\texorpdfstring{\autref{1}}{},
+Toby Hodges\texorpdfstring{\autref{1}}{},
+Jan Linxweiler\texorpdfstring{\autref{1}}{},
+Frank Löffler\texorpdfstring{\autref{1}}{},
+Michele Martone\texorpdfstring{\autref{1}}{},
+Jan Philipp Thiele\texorpdfstring{\autref{1}}{},
+Nick del Grosso\texorpdfstring{\autref{1}}{}
 } % Authors and references to addresses
-\institute{% Institutes with labels
-\autlabel{1} Würzburg-Dresden Cluster of Excellence ct.qmat, University of Würzburg, 97074, Würzburg, Germany\par
-\autlabel{2} ZB MED Information Centre for Life Sciences, Cologne, Germany\par
-\autlabel{3} School of Computation, Information and Technology, Technical University of Munich, Garching, Germany;
-             Institute for Parallel and Distributed Systems, University of Stuttgart, Stuttgart, Germany\par
-\autlabel{4} Leibniz University Hannover, Department of Cell Biology and Biophysics, Computational Biology, Germany\par
-\autlabel{5} Imperial College London, London, UK\par
-\autlabel{6} Institute for Computational Physics, University of Stuttgart, Germany\par
-\autlabel{7} Geschäftsbereich IT, Charité Universitätsmedizin Berlin, Germany\par
-\autlabel{8} The Carpentries, USA\par
-\autlabel{9} Technische Universität Braunschweig, Germany\par
-\autlabel{10} Michael Stifel Center Jena {\&} Friedrich Schiller University Jena, Germany\par
-\autlabel{11} Leibniz Supercomputing Centre, Garching, Germany\par
-\autlabel{12} Weierstrass Institute, Berlin, Germany;
-              Leibniz University Hannover, Institute of Applied Mathematics, Scientific Computing, Hannover, Germany\par
-\autlabel{13} Institute for Experimental Epileptology and Cognition Research, Uniklinikum Bonn, Germany
-}
+\institute{\autlabel{1} Fantasy University} % Institutes with labels
 \abstract{It's 2025 and we are, admittedly, still working on establishing and growing Research Software Engineering (RSE)
@@
 become a core part of the tooling and skill set of researchers in almost all other domains.
-+Are we already approaching ``the end of Research Software Engineering''?! While we believe that it is certainly too early to
+Are we already approaching ``the end of Research Software Engineering''?! While we believe that it is certainly too early to
 already herald the demise of RSE as a separate field,

### 2025-05-13 future_challenges_for_RSEng.tex
Technical fixes (#22)
Address redefined TeX labels and bibliography TODOs.

@@
 \begin{story}{The benefits of abstraction}{abstraction}
-A story about Kim and Kay effortlessly progressing where years ago they would be caught up in details.
+Years ago, Kim had to distribute the software library libkim as a source code package with a custom build system and detailed instructions on how to build it on some common systems,
+investing significant time in developing and maintaining this auxiliary toolkit and documentation.
+Around the same time, Kay was developing the research analysis tool kay-bio,
+which however only provided an application programming interface that
+required writing complex code even for simple and common analysis tasks.
+Nowadays, Kim could have started directly from a template based on established toolkits and workflows,
+while kay-bio accepts configuration in a simpler, standardized format and generative AI assistants
+can guide new users in preparing and debugging their analyses.
+Meanwhile, new RSEs can very easily integrate libkim, kay-bio, and a multitude
+of other tools from the research software corpus into a powerful analysis
+platform that allows researchers to get deeper insights from their research data
+in just a few clicks.
 \end{story}

### 2025-05-13 future_challenges_for_RSEng.tex
Update affiliations
Co-authored-by: Gerasimos Chourdakis <gerasimos.chourdakis@ipvs.uni-stuttgart.de>
Co-authored-by: Jan Philipp Thiele <54978337+jpthiele@users.noreply.github.com>

--- 2025-05-12 future_challenges_for_RSEng.tex
@@
 Michele Martone\authorRef{11},
-Jan Philipp Thiele\authorRef{12},
+Jan Philipp Thiele\authorRef{9}\textsuperscript{,}\authorRef{12},
 Nick del Grosso\authorRef{13}
@@
 \autlabel{2} ZB MED Information Centre for Life Sciences, Cologne, Germany\par
-\autlabel{3} School of Computation, Information and Technology, Technical University of Munich, Garching, Germany;
-             Institute for Parallel and Distributed Systems, University of Stuttgart, Stuttgart, Germany\par
+\autlabel{3} Institute for Parallel and Distributed Systems, University of Stuttgart, Stuttgart, Germany\par
 \autlabel{4} Leibniz University Hannover, Department of Cell Biology and Biophysics, Computational Biology, Germany\par

### 2025-05-13 future_challenges_for_RSEng.tex
Update future_challenges_for_RSEng.tex

@@
 Jan Philipp Thiele\authorRef{9}\textsuperscript{,}\authorRef{12},
-Nick del Grosso\authorRef{13}
+Nicholas A. Del Grosso\authorRef{13}
 } % Authors and references to addresses

### 2025-05-14 future_challenges_for_RSEng.tex
Add affiliations (#21)
Co-authored-by: Gerasimos Chourdakis <gerasimos.chourdakis@ipvs.uni-stuttgart.de>
Co-authored-by: Jan Philipp Thiele <54978337+jpthiele@users.noreply.github.com>
Co-authored-by: Nicholas A. Del Grosso <delgrosso.nick@gmail.com>

--- 2025-05-13 future_challenges_for_RSEng.tex
@@
 %\short{} % Short title of the article (optional)
+\newcommand{\authorRef}[1]{\texorpdfstring{\autref{#1}}{}}
 \author{
-Florian Goth\texorpdfstring{\autref{1}}{},
-Leyla Jael Castro\texorpdfstring{\autref{1}}{},
-Gerasimos Chourdakis\texorpdfstring{\autref{1}}{},
-Simon Christ\texorpdfstring{\autref{1}}{},
-Jeremy Cohen\texorpdfstring{\autref{1}}{},
-Jean-Noël Grad\texorpdfstring{\autref{1}}{},
-Magnus Hagdorn\texorpdfstring{\autref{1}}{},
-Toby Hodges\texorpdfstring{\autref{1}}{},
-Jan Linxweiler\texorpdfstring{\autref{1}}{},
-Frank Löffler\texorpdfstring{\autref{1}}{},
-Michele Martone\texorpdfstring{\autref{1}}{},
-Jan Philipp Thiele\texorpdfstring{\autref{1}}{},
-Nick del Grosso\texorpdfstring{\autref{1}}{}
+Florian Goth\authorRef{1},
+Leyla Jael Castro\authorRef{2},
+Gerasimos Chourdakis\authorRef{3},
+Simon Christ\authorRef{4},
+Jeremy Cohen\authorRef{5},
+Jean-Noël Grad\authorRef{6},
+Magnus Hagdorn\authorRef{7},
+Toby Hodges\authorRef{8},
+Jan Linxweiler\authorRef{9},
+Frank Löffler\authorRef{10},
+Michele Martone\authorRef{11},
+Jan Philipp Thiele\authorRef{9}\textsuperscript{,}\authorRef{12},
+Nicholas A. Del Grosso\authorRef{13}
 } % Authors and references to addresses
-\institute{\autlabel{1} Fantasy University} % Institutes with labels
+\institute{% Institutes with labels
+\autlabel{1} Würzburg-Dresden Cluster of Excellence ct.qmat, University of Würzburg, 97074, Würzburg, Germany\par
+\autlabel{2} ZB MED Information Centre for Life Sciences, Cologne, Germany\par
+\autlabel{3} Institute for Parallel and Distributed Systems, University of Stuttgart, Stuttgart, Germany\par
+\autlabel{4} Leibniz University Hannover, Department of Cell Biology and Biophysics, Computational Biology, Germany\par
+\autlabel{5} Imperial College London, London, UK\par
+\autlabel{6} Institute for Computational Physics, University of Stuttgart, Germany\par
+\autlabel{7} Geschäftsbereich IT, Charité Universitätsmedizin Berlin, Germany\par
+\autlabel{8} The Carpentries, USA\par
+\autlabel{9} Technische Universität Braunschweig, Germany\par
+\autlabel{10} Michael Stifel Center Jena {\&} Friedrich Schiller University Jena, Germany\par
+\autlabel{11} Leibniz Supercomputing Centre, Garching, Germany\par
+\autlabel{12} Weierstrass Institute, Berlin, Germany;
+              Leibniz University Hannover, Institute of Applied Mathematics, Scientific Computing, Hannover, Germany\par
+\autlabel{13} Institute for Experimental Epileptology and Cognition Research, Uniklinikum Bonn, Germany
+}
 \abstract{It's 2025 and we are, admittedly, still working on establishing and growing Research Software Engineering (RSE)

### 2025-05-14 future_challenges_for_RSEng.tex
remove 2_6 and minor fixes

@@
 \institute{% Institutes with labels
-\autlabel{1} Würzburg-Dresden Cluster of Excellence ct.qmat, University of Würzburg, 97074, Würzburg, Germany\par
+\autlabel{1} Institut für theoretische Physik 1, University of Würzburg, 97074, Würzburg, Germany\par
 \autlabel{2} ZB MED Information Centre for Life Sciences, Cologne, Germany\par
@@
 
-\subsection{Education, and changes in the educational system}
-Digital competencies in the general population continue to increase, making jumping to software engineering easier than before.
-
-\begin{whatis}{What do we want to have achieved}{education}
-School curricula have embraced the digital transformation, and have included basics of computer interaction
-in every class level.
-In universities, the first generations of RSE-Master graduates find their way into research departments,
-and accelerate the pace of science.
-\end{whatis}
-
 \subsection{How have academic Institutions transformed by then?}
 \begin{whatis}{What do we want to have achieved}{academic}
-Similar to RSEs that are known by 2035 in university boards, we expect to be the concept of RSE units to be in the minds
+The first generations of RSE-Master graduates find their way into research departments, and accelerate the pace of science.
+Similar to RSEs, that are known by 2035 in university boards, we expect the concept of RSE units to be in the minds
 of academic staff at any level.
@@
 Academia, while claiming to be innovative (cite WissZeitVG here), has organisationally often adopted time-scales for change that are more reminiscient to civil services like pension funds.
-Nevertheless, the presence, even today, of initiatives like the HiddenRef hints at the potential for specialisation of research tasks.
+Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref} \cite{hiddenref} hints at the potential for specialisation of research tasks.
 This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.
-We have alreadyseen some specialisation trends in recent years such as
+We have already seen some specialisation trends in recent years such as
 the HPC-RSE and AI-RSE which are quite recent developments, whose appeal is stimulated by large public and private investments in HPC facilities and AI technologies.
@@
 which can pay off in faster and larger development in the future.
-This is an opportunity that would not possible by short-term funded solo RSEs.
+This is an opportunity that would not be possible by short-term funded solo RSEs.
 
@@
 \end{story}
+\section{What is in this future for RSEs and where to go from here?}
+%FG-NOTE: I think we can use this part to wrap things a bit up, and describe how this future then looks like.
+%GC-NOTE: Shall we add another story box here? Ideas what we write here?
 
-\section{What is in this future for RSEs and where to go from here?}
-FG-NOTE: I think we can use this part to wrap things a bit up, and describe how this future then looks like.
-GC-NOTE: Shall we add another story box here? Ideas what we write here?
-
-Digitalisation will lead to changes in the way societies interact and therefore will translate
-into changes in how academia perceives itself. The differentiation of job profiles in academia,
-that has started with profiles such as the RSE will continue. Societies will be facing new questions, that arise
-from technologies such as generally available AI systems. RSEs will very likely be doing different work,
-than what they are doing today, but their core task of enhancing the scientific output of researchers, other human
-beings, will stay the same.
+Digital competencies in the general population continue to increase, making jumping to software development easier than before.
+But this digitalisation will also lead to changes in the way societies interact and therefore will translate into changes in how academia perceives itself.
+The differentiation of job profiles in academia, that has started with profiles such as the RSE will continue.
+Societies will be facing new questions, that arise from technologies such as generally available AI systems.
+RSEs will very likely be doing different work,
+than what they are doing today, but their core task of enhancing the scientific output of researchers -- other human
+beings -- will stay the same.
 They will still develop software artifacts together with scientists,

### 2025-05-14 future_challenges_for_RSEng.tex
work in some notes from Jan. Write ethics-whatis section

@@
 
-\subsection{Ethics and social consideration}
+\subsection{Ethics and social considerations}
 \begin{whatis}{What do we want to have achieved}{ethics}
-Oh well..., is there really something that we could have achieved here?
-maybe open Science? Data awareness? Data Protection?
+The principles of Good Scientific Practice \cite{dfg_gsp} have been a cornerstone in the conduct of research.
+But their relevance hinges on constant updates in light of how research is performed.
+By 2035 we want to have achieved accepted rules on the use of AI in research tasks and software development.
+Open Science and Open Research are well-established in academia, and supported by respective Open Research Software
+strategies of the political actors. We also hope, that Data Awareness and Data Protection are also still values
+that are protected.
 \end{whatis}
@@
 Sometimes these values are in direct conflict with the aims of their institution. For example, their PI might insist on quickly developing some software without adhering to current best practices.
-A code of ethics and professional conduct for RSEs should be developed by institutions together with RSEs themselves. This can then form the basis for institutional policies and guidelines.
+A code of ethics and professional conduct for RSEs should be developed by institutions together with RSEs themselves.
+This can then form the basis for institutional policies and guidelines.
 
@@
 
-MM-NOTE: (Flo believes it was from Michele) Are we assuming that progress will bring prosperity?
-Will we remain constrained by needing to produce products/artifacts?
-(GC: Where "product/artifacts", maybe also "publications"?)
-
+MM-NOTE:
   - Importance of having personal “red lines” and awareness of how software can be used.
@@
 \begin{whatis}{What do we want to have achieved}{academic}
-The first generations of RSE-Master graduates find their way into research departments, and accelerate the pace of science.
+We expect research software to be a central part of scientific practice on the same level as data and publications.
+This will be supported by the first generations of RSE-Master graduates that find their way into research departments, and accelerate the pace of science.
 Similar to RSEs, that are known by 2035 in university boards, we expect the concept of RSE units to be in the minds
@@
 At universities which have them, we want to see that researchers naturally accept them as part of their used services.
+Additionally researchers will seamlessly interact with infrastructures for using research software.
 \end{whatis}
@@
 But we could imagine new RSE specialisations to emerge in the fields
-of quantum computing, such as the Quantum-RSE, and in federated online platforms.
+of quantum computing, such as the Quantum-RSE, and in federated online platforms for research.
 % European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories
@@
 permanent and well-funded RSE teams.
+Here it can only be strongly hoped, that, with the recognized importance of research software, state-actors
+support the development with a reliable stream of funding.
+This would also require a change of culture in research about the acceptance of research software.\\
 While we hope that the barrier to researchers diminishes, the workshop participants

### 2025-05-14 future_challenges_for_RSEng.tex
iron out remaining Fixmes. spice up stories.

@@
 
-It is already difficult to structure the spectrum of RSE tasks and responsibilities today (FIXME: Willi Hasselbrings classification paper?)
+It is already difficult to structure the spectrum of RSE tasks and responsibilities today given the spectrum of research software \cite{hasselbring2024}
 and it is even more hopeless to structure the aspects that will become relavant for the RSE community in the future.
@@
 \begin{story}{The Definition of RSEs}{defrse}
-A story, about how Kim and Kay started out with a certain job description that now totally
-has changed meaning.
+Kim and Kay chuckled as they looked at their original job descriptions from 2020—"build tools for scientists, manage data, write reusable code."
+Now, their roles were unrecognizable: half AI whisperer, half research ethicist, and part-time diplomat between synthetic and human collaborators.
+They managed not just software, but the personalities of autonomous lab agents and negotiated data-sharing agreements with international AI consortiums.
+Despite the tech upheaval, their deep knowledge of scientific workflows and the messy, human way science actually gets done remained indispensable.
+When labs faltered under algorithmic bias or unclear model provenance, it was Kim and Kay who translated the chaos back into meaning.
+They understood that science didn’t move in straight lines, and that trust—between collaborators, institutions, and even machines—had to be built, not coded.
+Their titles had changed, but their role as bridges between logic and life remained the same.
+In a future of shifting code and constant reinvention, it was their understanding of people that never went out of date.
+
+% Kim and Kay started out as Research Software Engineers.
+% When they began, their work meant cleaning up code, making research reproducible, and helping academics run experiments without crashing HPC clusters.
+Now, they built systems, where AI generated hypotheses, designed experiments, and even debated its own conclusions.
+Their role had shifted from supporting research to shaping how knowledge was created, validated, and understood—an epistemological responsibility as much as a technical one.
+“We used to debug scripts,” Kay said, watching a model suggest a reality-bending theory of space-time.
+Kim nodded: “Now we debug truth.”
 \end{story}
@@
 \begin{whatis}{What do we want to have achieved}{development-rse-community}
-We have more national RSE communities that are interacting with other partners in their nation.
-There are regular international conferences. The installation of RSE Master's programmes led to
-increased numbers of the members of the national societies.
+National RSE communities are increasingly collaborating with local partners, fostering stronger networks across countries.
+Regular international conferences further amplify these connections, promoting the global exchange of ideas.
+The introduction of RSE Master's programs, coupled with a shift in culture towards valuing Research Software, has significantly elevated its profile.
+This growing recognition has led to a surge in membership within national societies, reflecting the increasing importance of RSEng in the research landscape.
 \end{whatis}
@@
 and regional level to support more viable RSE career paths in academia.
-We can expect RSEs to become more easily accessible,
-possibly through centrally-funded long-term positions at research institutions.
+We can expect RSEs to become more easily accessible, possibly through centrally-funded long-term positions at research institutions.
 The community of RSE's will get a significant boost in their acceptance if RSE Masters programmes
 become available at various universities.
-
+With increased numbers the visibility of the RSE community also to industry will become higher.
 \begin{story}{Development of RSE community}{RSEcommunity}
-A story about Kim participating in the first international RSE conference in a country on the southern hemisphere.
+Kay stood proudly at the podium of the first International Research Software Engineering Conference held in the Southern Hemisphere, a historic moment for both her career and the global RSE community.
+Having moved to South Africa, she spoke passionately about the unique challenges researchers in the Global South face—limited access to high-performance computing, unreliable funding, and the persistent digital divide.
+Her keynote highlighted how local innovation thrives despite constraints, often out of necessity rather than choice.
+The audience, a diverse mix of global experts, listened intently as she called for equitable collaborations, not just inclusion as an afterthought.
 \end{story}
@@
 since AI copilots are available 24/7.
-
+Since AI systems are so democratically available and will be used as black boxes,
+the interpretation of their outputs will be far more imortant.
 Currently we are expanding and documenting our work in the open and making our code available,
@@
 \begin{story}{The interaction of RSEs with AI}{RSEs_with_AI}
-A story about Kay co-developing with AI.
+Kim develops advanced scientific software in tandem with an AI agent that codes, debugs, and simulates experiments alongside her.
+The AI proactively suggests optimizations, generates reproducible workflows, and adapts to Kim’s coding style and domain-specific needs.
+With this partnership, Kim accelerates her research output while maintaining full control over scientific direction and ethical oversight.
 \end{story}
@@
 We hope to see that universities use the presence of an RSE Department actively as a means to differentiate themselves from other universities.
-We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department to oversee specific projects[FIXME: cite group paper].
+We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department to oversee specific projects \cite{Kempf2025-draft}.
 We also hope, that specialized clusters form that pool the skills of certain specialised RSEs, with certain RSE units specializing in certain tasks like data management, HPC, or visualization.
@@
 \emph{New challenges}
-If we follow the ideas laid out in (FIXME: cite RSE group paper), then in addition to university wide RSE units,
+If we follow the ideas laid out in \cite{Kempf2025-draft}, then in addition to university wide RSE units,
 there will be embedded RSEs in the departments. Shaping and developing these network bonds will be a task for the future.
-While networking embedded RSEs and RSE units is essential for building robust, sustainable research software infrastructure, but the question arises whether every group truly needs to host and manage its own infrastructure(FIXME: maybe cite this here: https://www.dfg.de/en/research-funding/funding-opportunities/programmes/infrastructure/lis/funding-opportunities/research-software-infrastructures).
+While networking embedded RSEs and RSE units is essential for building robust, sustainable research software infrastructure, but the question arises whether every group truly needs to host and manage its own infrastructure.
 When each team or institution hosts its own systems, this often leads to siloed environments where resources, expertise, and data are isolated, limiting collaboration and creating inefficiencies.
@@
 While local control offers flexibility, it also increases complexity and makes it harder to share innovations or best practices across the wider research community.
-To overcome these challenges, federated infrastructure models are needed, enabling shared access, interoperability, and coordinated development while still allowing for some local autonomy(FIXME: maybe cite this here: https://eurise-network.github.io/).
-Federated approaches break down silos, promote knowledge exchange, and ensure that infrastructure investments benefit the broader ecosystem rather than just isolated groups
+To overcome these challenges, federated infrastructure models are needed, enabling shared access, interoperability, and coordinated development while still allowing for some local autonomy \cite{Jelinek2021}.
+Federated approaches break down silos, promote knowledge exchange, and ensure that infrastructure investments benefit the broader ecosystem rather than just isolated groups.
 
@@
 We would like to see the emergence of more dedicated personel, i.e. other RTPs next to the RSE.
+Also we hope that researchers can build careers at the interface of academia, industry, and policy (\eg in think tanks, NGOs,
+or organisations that do knowledge transfer).
 \end{whatis}
@@
 \begin{story}{New career paths}{New career paths}
-A story, about Ken delving into a new direction as RTP.
+Ken, though trained as a physicist, found his passion not in theory, but in keeping the quantum computers humming.
+He became the go-to expert for troubleshooting entangled qubits and stabilizing delicate cryogenic systems.
+His unique blend of physics knowledge and hands-on problem solving earned him a new title among his peers: the first Quantum RTP.
+In a world where theory meets reality, Ken quietly ensures the future of quantum computing stays on track.
 \end{story}
@@
 %GC-NOTE: Shall we add another story box here? Ideas what we write here?
-
+It is hard to conclude a piece written about the future, but we have tried to list some of the trends that we foresee.
+Nevertheless there are some things that we declared out of scope here.
+Among that is the digital transformation in the wider society.
 Digital competencies in the general population continue to increase, making jumping to software development easier than before.
 But this digitalisation will also lead to changes in the way societies interact and therefore will translate into changes in how academia perceives itself.
+We have also not delved into the interaction with industry much.
+There are also some trends that believe to stay with us.
 The differentiation of job profiles in academia, that has started with profiles such as the RSE will continue.
 Societies will be facing new questions, that arise from technologies such as generally available AI systems.
-RSEs will very likely be doing different work,
+
+We hope that the standard of research software has risen across the board to enable an easy reuse in new research.
+We hope that a cultural change is taking place to recognize research software as the integral part of science that it is
+within broad parts of academia and is supported by political actors.
+To conclude, RSEs will very likely be doing different work
 than what they are doing today, but their core task of enhancing the scientific output of researchers -- other human
@@
 They will still develop software artifacts together with scientists,
-since Better software leads to better research,
+since better software leads to better research,
 better research leads to more better software.

### 2025-05-14 future_challenges_for_RSEng.tex
minor fixes

@@
 Nevertheless, by disseminating the discussed ideas,
-we aim to engage the community in imagining the future.
+we aim to engage the community in creating a future worth living.
 In the following sections, we consider the evolution of the RSE definition,
 the development of RSE education,
+the growth of the RSE communities,
 the impact of automation and generative artificial intelligence in RSE work,
 the emerging ethical considerations,
-and the transformations expected in the research organizations,
+and the transformations expected in the research organisations,
 and if there will be other in academia covered by new RTPs (\emph{Research Technical Professional}s).
@@
 contributed to make research software and its outputs
-more sustainable, robust and maintainable. We see this happening in a number of ways. However,
-we want to ensure that one of the core achievements made by the RSE community in this time
+more sustainable, robust and maintainable.
+We also hope, that Research Software becomes an accepted research artifact on equal footing as data and publications.
+We see this happening in a number of ways.
+However, we want to ensure that one of the core achievements made by the RSE community in this time
 frame is putting the infrastructure and training opportunities in place, alongside the necessary
@@
 \end{whatis}
-% Here we consider how the topics of RSEng will change in the face of the digitalization of society.
+% Here we consider how the topics of RSEng will change in the face of the digitalisation of society.
 % Also here: What will our topics be after version control doesn't need to be taught anymore?
@@
 Kim develops advanced scientific software in tandem with an AI agent that codes, debugs, and simulates experiments alongside her.
-The AI proactively suggests optimizations, generates reproducible workflows, and adapts to Kim’s coding style and domain-specific needs.
+The AI proactively suggests optimisations, generates reproducible workflows, and adapts to Kim’s coding style and domain-specific needs.
 With this partnership, Kim accelerates her research output while maintaining full control over scientific direction and ethical oversight.
@@
 We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department to oversee specific projects \cite{Kempf2025-draft}.
-We also hope, that specialized clusters form that pool the skills of certain specialised RSEs, with certain RSE units specializing in certain tasks like data management, HPC, or visualization.
+We also hope, that specialized clusters form that pool the skills of certain specialised RSEs, with certain RSE units specializing in certain tasks like data management, HPC, or visualisation.
 At universities which have them, we want to see that researchers naturally accept them as part of their used services.
@@
 With longer-term funding and more trust to pick up longer term projects,
-some RSEs can focus on code modernization, cleanup, and modularization,
+some RSEs can focus on code modernisation, cleanup, and modularisation,
 which can pay off in faster and larger development in the future.
@@
 While we hope that the barrier to researchers diminishes, the workshop participants
-still expect some work to do. We hope that the compartmentalization that occurs will help researchers reprioritize their workload.
+still expect some work to do. We hope that the compartmentalisation that occurs will help researchers reprioritize their workload.
 An upcoming challenge will be that, with the emergence of more dedicated RSE units,
@@
 This, however, comes with additional funding challenges,
-as long-term funding is more difficult to acquire in lower levels of an organization.
+as long-term funding is more difficult to acquire in lower levels of an organisation.
 Third-party funding can mitigate this issue in the mid-term.
@@
 It is hard to conclude a piece written about the future, but we have tried to list some of the trends that we foresee.
+The material presented here, is based on a workshop, bt extended with the author's ideas.
 Nevertheless there are some things that we declared out of scope here.

### 2025-05-14 future_challenges_for_RSEng.tex
change What is boxes

@@
 \newcounter{whatiscounter}
-\newtcbtheorem[use counter=whatiscounter, number format=\Alph]{whatis}{What is 2035}{
+\newtcbtheorem[use counter=whatiscounter, number format=\Alph]{whatis}{What do we want to have achieved}{
   enhanced,
@@
 \subsection{Definition and Evolution of RSEs}
-\begin{whatis}{What do we want to have achieved}{definition-evolution-rse}
+\begin{whatis}{}{definition-evolution-rse}
 The term RSE is established in the minds of university boards, as a valuable member in the academic framework.
@@
 \subsection{Development of the RSE community}
-\begin{whatis}{What do we want to have achieved}{development-rse-community}
+\begin{whatis}{}{development-rse-community}
 National RSE communities are increasingly collaborating with local partners, fostering stronger networks across countries.
@@
 \subsection{Impact of AI and automation}
-\begin{whatis}{What do we want to have achieved}{impact-ai-automation}
+\begin{whatis}{}{impact-ai-automation}
 RSEs are educated about AI models, and can help researchers reap their benefits, while
@@
 \subsection{Impact of Complexity and Usability in Research Software}
-\begin{whatis}{What do we want to have achieved}{abstraction}
+\begin{whatis}{}{abstraction}
 With a palette of readily reusable software components,
@@
 \subsection{Ethics and social considerations}
-\begin{whatis}{What do we want to have achieved}{ethics}
+\begin{whatis}{}{ethics}
 The principles of Good Scientific Practice \cite{dfg_gsp} have been a cornerstone in the conduct of research.
@@
 \subsection{How have academic Institutions transformed by then?}
-\begin{whatis}{What do we want to have achieved}{academic}
+\begin{whatis}{}{academic}
 We expect research software to be a central part of scientific practice on the same level as data and publications.
@@
 \subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}
-\begin{whatis}{What do we want to have achieved}{career-paths}
+\begin{whatis}{}{career-paths}
 We would like to see the emergence of more dedicated personel, i.e. other RTPs next to the RSE.

### 2025-05-14 future_challenges_for_RSEng.tex
Proof read up to end of section 2.6.

@@
 
-Research Software Engineering is currently an emerging field,
-with an increasing integration into various aspects of research.
+Research Software Engineering is still a developing field,
+with increasing integration into various aspects of research.
 In this opinion paper, we try to imagine the state of our field
@@
 Nevertheless, by disseminating the discussed ideas,
-we aim to engage the community in creating a future worth living.
-In the following sections, we consider the evolution of the RSE definition,
-the development of RSE education,
-the growth of the RSE communities,
-the impact of automation and generative artificial intelligence in RSE work,
-the emerging ethical considerations,
-and the transformations expected in the research organisations,
-and if there will be other in academia covered by new RTPs (\emph{Research Technical Professional}s).
+we aim to engage the community in creating a future worth living
+and advance the conversation about how Research Software Engineering should, or could, develop.
+In the following sections, we consider a number of areas:
+the evolution of the RSE definition;
+the development of RSE education;
+the growth of the RSE communities;
+the impact of automation and generative artificial intelligence in RSE work;
+the emerging ethical considerations;
+the transformations expected in research organisations;
+and, finally, if there will be other areas in academia covered by new groups of RTPs (\emph{Research Technical Professional}s).
 We conclude the paper with a summary and with our wishes and recommendations.
@@
 Looking ahead over a 10-year timeframe, we hope that the RSE community will have
-contributed to make research software and its outputs
+contributed to making research software and its outputs
 more sustainable, robust and maintainable.
-We also hope, that Research Software becomes an accepted research artifact on equal footing as data and publications.
-We see this happening in a number of ways.
+We also hope that Research Software becomes an accepted research artifact on an equal footing with data and publications.
+We see this happening in a number of ways,
+particularly through movements such as DORA~\cite{DORA} and CoARA~\cite{COARA} which are advocating for a change in the way that research is assessed and outputs recognised.
 However, we want to ensure that one of the core achievements made by the RSE community in this time
@@
 So where do RSEs fit into this picture? Even today we expect to see RSEs providing technical skills and input
-to research activities in a range of different areas, such as
-High-quality, robust architecture and design for research software,
-Seamless integration with research and data infrastructures,
-Development of software for novel and emerging architectures,
-Application of specialist numerical and statistical methods within software,
-Optimisation of codes for use on high-performance computing infrastructure,
-and Green computing - efficient implementations of algorithms/methods and use of hardware
+to research activities in a range of different areas, these include:
+high-quality, robust architecture and design for research software;
+seamless integration with research and data infrastructures;
+development of software for novel and emerging architectures;
+application of specialist numerical and statistical methods within software;
+optimisation of codes for use on high-performance computing infrastructure;
+and green computing - efficient implementations of algorithms/methods and use of hardware
 
 It is already difficult to structure the spectrum of RSE tasks and responsibilities today given the spectrum of research software \cite{hasselbring2024}
-and it is even more hopeless to structure the aspects that will become relavant for the RSE community in the future.
-Nevertheless, we decided to consider
+and it is even more challenging to structure the aspects that will become relevant for the RSE community in the future.
+Nevertheless, we decided to consider:
 \begin{itemize}
-\item The evolution of the term RSE,
-\item Development of the RSE community,
-\item What the impact of AI to the work of RSEs will look like,
-\item the impact of Complexity and Usability in Research Software
-\item Ethics and social considerations,
-\item How education is reshaped by the digital transformation,
-\item Changes to the Academic Institutions,
-\item Impact of alternate career paths for RSEs and other RTPs.
+\item The evolution of the term RSE
+\item Development of the RSE community
+\item What the impact of AI will be to the work of RSEs
+\item The impact of complexity and usability in research software
+\item Ethics and social considerations
+\item How education is reshaped by the digital transformation
+\item Changes to academic institutions
+\item Impact of alternate career paths for RSEs and other RTPs
 \end{itemize}
@@
 \begin{story}{Introduction}{intro}
-In order to give a more explicit setting, imagine a world ten years from now.
+In order to give a more concrete setting, imagine a world ten years from now.
 Not too much to be overwhelmingly far in the future, but distant enough for some changes to occur.
-Over this timeframe Kim\cite{Anzt2021} and Kay\cite{Goth2024}
+Over this timeframe Kim\cite{Anzt2021} and Kay\cite{Goth2024},
+our fictional RSEs who we have used to illustrate specific scenarios in our earlier work,
 were able to ride the waves of their career in Research Software Engineering and are now pondering early retirement on a sunset blessed beach.
@@
 When they look back, what will they see as problems which are now solved?
-Which tasks are persisting, where they are now happy that a younger generation is now taking up the baton and carries forward their work.
-What issues have newly emerged, where they are now just happy to say: “Oh well… My successor takes care of that”.\\
-In order to bring this future to life these two personas will accompany us and we will shed light on certain aspects
+Which tasks are still relevant, where they can be happy that a younger generation is now taking up the baton and carrying forward their work?
+What issues have newly emerged, where they are now just happy to say: “Oh well… my successors will take care of that”.
+In order to bring this future to life, these two personas will accompany us through the rest of this paper and we will shed light on aspects
 of their future in respective small story boxes.
@@
 \begin{whatis}{}{definition-evolution-rse}
-The term RSE is established in the minds of university boards, as a valuable member in the academic framework.
-RSEs are developed using specialized master's program, and their support is valued by researchers.
+The term RSE is established in the minds of university boards and leadership, as a valuable member in the academic framework.
+New RSEs are trained through a specialized master's program, and their support is valued by researchers.
 We still teach researchers basic software skills, but the content of this lecture has changed drastically.
@@
 What will \emph{not} change over that timeframe is the RSE's ability to remain
-flexible and embrace change. Keep an eye open and pick up tools and skills
-as needed as we go along.
+flexible and embrace change, and keep an eye on the wider landscape,
+picking up new tools and skills as needed.
 Communication and collaboration skills will remain essential to engage
@@
 \end{whatis}
-In recent years, we saw the formation of RSE institutes and advocacy groups, such as
+In recent years, we have seen the formation of RSE institutes and advocacy groups, such as
 the UK Software Sustainability Institute\footnote{\url{https://www.software.ac.uk}},
@@
 become available at various universities.
-With increased numbers the visibility of the RSE community also to industry will become higher.
+With increased numbers, the visibility of the RSE community to industry will be enhanced.
 \begin{story}{Development of RSE community}{RSEcommunity}
@@
 With the general availability of AI the interaction with researchers will change at a fundamental level.
-Whereas previously RSEs have been points of knowledge that could translate human language formulated problems into technical descriptions and then provide solutions to it,
-It can be foreseen, that AI drives this low-key interaction to a highly specialized level of knowledge,
+Whereas previously RSEs have been points of knowledge that could translate human language formulated problems into technical descriptions and then provide solutions to them,
+it can be foreseen that AI will drive this low-key interaction to a highly specialized level of knowledge,
 since AI copilots are available 24/7.
@@
 \begin{story}{The interaction of RSEs with AI}{RSEs_with_AI}
-Kim develops advanced scientific software in tandem with an AI agent that codes, debugs, and simulates experiments alongside her.
+Kim develops advanced scientific software in tandem with an AI agent that codes, debugs, and simulates experiments alongside him.
 The AI proactively suggests optimisations, generates reproducible workflows, and adapts to Kim’s coding style and domain-specific needs.
-With this partnership, Kim accelerates her research output while maintaining full control over scientific direction and ethical oversight.
+With this partnership, Kim accelerates his research output while maintaining full control over scientific direction and ethical oversight.
 \end{story}
@@
 and with their skills in managing hierarchical systems of complex software components with clear interfaces,
-RSEs can now build systems of scale and capability not previously possible.
+RSEs can now build reliable, maintainable systems of a scale and capability not previously possible.
 \end{whatis}
@@
 we are now able to develop systems of complexity and capability not previously possible.
-This divide-and-conquer approach allows the scientific community to address questions not previously in grasp.
+This divide-and-conquer approach allows the scientific community to address questions not previously within their grasp.
 When it comes to teaching, it is true that students need to learn how to handle complexity,
@@
 Open Science and Open Research are well-established in academia, and supported by respective Open Research Software
-strategies of the political actors. We also hope, that Data Awareness and Data Protection are also still values
+strategies of the political actors. We also hope that Data Awareness and Data Protection are still values
 that are protected.
@@
 % institutional bonds and ethics
-Sometimes these values are in direct conflict with the aims of their institution. For example, their PI might insist on quickly developing some software without adhering to current best practices.
+Sometimes these values may conflict with the aims of their peers or institution. For example, their PI might insist on quickly developing some software without adhering to current best practices.
 A code of ethics and professional conduct for RSEs should be developed by institutions together with RSEs themselves.
@@
 % society implications
-The technology people working as RSE might have, well directly or indirectly, ethical implications in areas such as social engineering [TODO find source], climate [TODO find source] and more.
+The work being undertaken by RSEs may well have direct, or indirect, ethical implications in areas such as social engineering [TODO find source], climate [TODO find source] and more.
 Another example might be that they are developing code that can also be used for military purposes, such as numerical solvers or computer vision systems.
@@
 \begin{story}{The greater world of Kim and Kay}{ethics}
-10 years have passed since Kim and Kay have started on their journey of being RSEs. Back then,
-at the universities of Eden and Utopia, life was more orderly and harmonious. Kay is working in a project that develops
+10 years have passed since Kim and Kay started on their journey of being RSEs. Back then,
+at the universities of Eden and Utopia, life was more orderly and harmonious. Kay is now working in a project that develops
 software for cheap water monitoring devices in the global south. Kim is currently designing the software service stack
-at the university, and increasingly struggles to find compelling Open Source alternatives that lets them keep their data
-to software that is backed by large corporate entities. So far she still has the support of the university board, since it values
+at the university, and increasingly struggles to find compelling Open Source alternatives that enables them keep complete control of their data
+rather than relying on software and tools that are backed by large corporate entities. So far she still has the support of the university board, since it values
 strategic autonomy.
@@
 Governments around the world are struggling to reposition themselves in this emerging deglobalized world with its new blocks
-and therefore have deprioritized publicly funded research. Kyle decided to go to a defense company since they offer a
+and therefore have deprioritized publicly funded research. Kyle decided to go to an AI hardware company since they offer
 more secure employment in this polarized world.
@@
 This will be supported by the first generations of RSE-Master graduates that find their way into research departments, and accelerate the pace of science.
-Similar to RSEs, that are known by 2035 in university boards, we expect the concept of RSE units to be in the minds
+Similar to RSEs, that are widely known and understood by university boards in 2035, we expect the concept of RSE units to be in the minds
 of academic staff at any level.
@@
 We also hope, that specialized clusters form that pool the skills of certain specialised RSEs, with certain RSE units specializing in certain tasks like data management, HPC, or visualisation.
-At universities which have them, we want to see that researchers naturally accept them as part of their used services.
+At universities which have them, we want to see that researchers naturally accept them as part of their available services.
 Additionally researchers will seamlessly interact with infrastructures for using research software.
@@
 the HPC-RSE and AI-RSE which are quite recent developments, whose appeal is stimulated by large public and private investments in HPC facilities and AI technologies.
-But we could imagine new RSE specialisations to emerge in the fields
+But we could imagine new RSE specialisations emerging in the fields
 of quantum computing, such as the Quantum-RSE, and in federated online platforms for research.
@@
 some RSEs can focus on code modernisation, cleanup, and modularisation,
-which can pay off in faster and larger development in the future.
-This is an opportunity that would not be possible by short-term funded solo RSEs.
+which can pay off in faster and easier development in the future.
+This is an opportunity that would not be possible if relying on short-term funded solo RSEs.
 
@@
 among institutions with successful RSE Teams.
-On a more organisational level, we expect the sharing of models on how how to organise diiferent RSEs, to fit the different organisational constraints.
-After that Organisations will have to come up with models for hiring people.
-While a standardized RSE Master's makes it easier, it will be interesting to see if
-RSE are still often hired by word of mouth within an organisation(``I have this PhD here, before their PostDoc they would be a good fit for the RSE department...'' ) or if the RSE master's degree ensures a hiring process on a more standardised level.
+On a more organisational level, we expect the sharing of models on how to organise different RSEs with different areas of specialist expertise, to fit the different organisational constraints.
+After that, organisations will have to come up with updated models for hiring people.
+While a standardized RSE Masters makes it easier to develop the next generation of practitioners with the necessary technical skills, it will be interesting to see if
+RSE are still often hired by word of mouth within an organisation(``I have this PhD here, before their PostDoc they would be a good fit for the RSE department...'' ) or if the RSE Masters degree ensures a hiring process on a more standardised level.
 
@@
 Kim improves upon the model set forth by the University of Arcadia based on valuable experience. Since the university
-of Avalon has to follow different rules not everything can be straightforwardly adapted. Life would be so much easier,
+of Avalon has to follow different rules, not everything can be straightforwardly adapted. Life would be so much easier
 if at least university regulations would be homogenized, no?
@@
 there will be embedded RSEs in the departments. Shaping and developing these network bonds will be a task for the future.
-While networking embedded RSEs and RSE units is essential for building robust, sustainable research software infrastructure, but the question arises whether every group truly needs to host and manage its own infrastructure.
+While networking embedded RSEs and RSE units is essential for building robust, sustainable research software infrastructure, the question arises whether every group truly needs to host and manage its own infrastructure.
 When each team or institution hosts its own systems, this often leads to siloed environments where resources, expertise, and data are isolated, limiting collaboration and creating inefficiencies.

### 2025-05-15 future_challenges_for_RSEng.tex
Remaining updates from proof read of paper.

@@
 \begin{whatis}{}{career-paths}
-We would like to see the emergence of more dedicated personel, i.e. other RTPs next to the RSE.
+We would like to see the emergence of more dedicated personel, i.e. other (digital) RTPs next to the RSE.
 Also we hope that researchers can build careers at the interface of academia, industry, and policy (\eg in think tanks, NGOs,
@@
 This, however, comes with additional funding challenges,
-as long-term funding is more difficult to acquire in lower levels of an organisation.
+as long-term funding can more difficult to acquire in lower tiers of an of a research organisation (i.e. at departmental or individual research group level).
 Third-party funding can mitigate this issue in the mid-term.
@@
 due to the fact that there are already examples and activities from various universities in this direction
-(examples include the universities of Jena, Heidelberg, Göttingen).
+(examples include the universities of Jena, Heidelberg and Göttingen, as well as several universities in the UK where the concept of dedicated RSE groups began to emerge around 2013).
 We expect funding in central groups to evolve from initially temporary positions to permanent positions,
-as the groups themselves prove themselves valuable.
+as the groups prove themselves to be increasingly valuable and, hence, sustainable.
 Embedded RSEs and central groups will need to network with each other,
@@
 It is hard to conclude a piece written about the future, but we have tried to list some of the trends that we foresee.
-The material presented here, is based on a workshop, bt extended with the author's ideas.
+The material presented here is based on the outputs of a workshop, but extended with the authors' ideas.
 Nevertheless there are some things that we declared out of scope here.
-Among that is the digital transformation in the wider society.
+Among that is the ongoing digital transformation in wider society.
 Digital competencies in the general population continue to increase, making jumping to software development easier than before.
@@
 We have also not delved into the interaction with industry much.
-There are also some trends that believe to stay with us.
+There are also some trends that we believe will stay with us over the coming years.
 The differentiation of job profiles in academia, that has started with profiles such as the RSE will continue.
@@
 
-We hope that the standard of research software has risen across the board to enable an easy reuse in new research.
-We hope that a cultural change is taking place to recognize research software as the integral part of science that it is
-within broad parts of academia and is supported by political actors.
+We hope that the standard of research software will have risen across the board, enabling easier reuse of prior software outputs in new research.
+We also hope that a cultural change is taking place to recognize research software as an integral part of science that it is
+widely recognized and rewarded across academia and is supported by political actors.
 To conclude, RSEs will very likely be doing different work
-than what they are doing today, but their core task of enhancing the scientific output of researchers -- other human
+to what they are doing today, but their core task of enhancing the scientific output of researchers -- other human
 beings -- will stay the same.
@@
 since better software leads to better research,
-better research leads to more better software.
+and better research leads to better software.
 

### 2025-05-16 future_challenges_for_RSEng.tex
update

@@
 Governments around the world are struggling to reposition themselves in this emerging deglobalized world with its new blocks
-and therefore have deprioritized publicly funded research. Kyle decided to go to an AI hardware company since they offer
+and therefore have deprioritized publicly funded research. Kyle decided to go to an cimlausp company since they offer
 more secure employment in this polarized world.
@@
 % AI Systems have been harmed in the creation of this work.
-We thank all the participants of our workshop at deRSE25 in Karlsruhe for their valuable input! 
+We thank all the participants of our workshop at deRSE25 in Karlsruhe for their valuable input!  JC acknowledges support from the UKRI-EPSRC-funded STEP-UP project under grant EP/Y530608/1.
 \end{acknowledge}

### 2025-05-16 future_challenges_for_RSEng.tex
don't do things at night...
fix up missing refs

@@
 and software will become easier to interact with,
-reducing reliance on RSEs to train newcomers on the most basic tasks
-(JN: for example, versioning and archiving?)
-(GC: Higher-level and more intuitive workflows that require less training).
+reducing reliance on RSEs to train newcomers on the most basic tasks.
+This means, that today's workshop content of proper versioning and data management could become obsolete
+by more intuitive tooling, hence basic courses can focus on more higher-level concepts.
 User training sessions might focus on more advanced topics
@@
 the professionalisation%
-\footnote{MM: do we want to decide for UK or US English? Or keep it mixed? Not being a native English speaker I don't know how it reads.}
 of the RSE role.
@@
 % society implications
-The work being undertaken by RSEs may well have direct, or indirect, ethical implications in areas such as social engineering [TODO find source], climate [TODO find source] and more.
+The work being undertaken by RSEs may well have direct, or indirect,
+ethical implications in areas such as social engineering \cite{s2erc,siadati2024},
+climate \cite{Lannelongue2023} and more.
 Another example might be that they are developing code that can also be used for military purposes, such as numerical solvers or computer vision systems.
@@
 A code of ethics and professional conduct for RSEs should be developed by institutions together with RSEs themselves.
-This can then form the basis for institutional policies and guidelines.
-
-MM-NOTE:
-  - Importance of having personal “red lines” and awareness of how software can be used.
+This can then form the basis for institutional policies and guidelines as well as serve as a background for individual developers to consider how software can be used.
 
@@
 Governments around the world are struggling to reposition themselves in this emerging deglobalized world with its new blocks
-and therefore have deprioritized publicly funded research. Kyle decided to go to an cimlausp company since they offer
+and therefore have deprioritized publicly funded research. Kyle decided to go to an defense company since they offer
 more secure employment in this polarized world.
@@
 \end{whatis}
-Academia, while claiming to be innovative (cite WissZeitVG here), has organisationally often adopted time-scales for change that are more reminiscient to civil services like pension funds.
+Academia, while claiming to be innovative \cite{wzvgbayern}, has organisationally often adopted time-scales for change that are more reminiscient to institutions thinking on the time scale of generations.
 Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref} \cite{hiddenref} hints at the potential for specialisation of research tasks.

### 2025-05-16 future_challenges_for_RSEng.tex
Add leyla's keywords

--- 2025-05-16 future_challenges_for_RSEng.tex
@@
 it is worthwhile discussing how RSE will/could change in response to these upcoming challenges in an open discussion.} % Abstract of the article
-\keywords{Keywords go here.} % Keywords for the article
+\keywords{Future RSE, Evolving field, Vision
+} % Keywords for the article
 

### 2025-05-16 future_challenges_for_RSEng.tex
start it

@@
 }
-\abstract{It's 2025 and we are, admittedly, still working on establishing and growing Research Software Engineering (RSE)
+\abstract{
+It is 2025 and we are, admittedly, still working on establishing and growing Research Software Engineering (RSEng)
 as a domain in its own right, with the hope of ensuring that it gets the recognition it deserves.
-Nonetheless, as the use of digital tools becomes ever more pervasive in a researcher's day-to-day tasks, there are already people suggesting that RSE will
-cease to exist as a separate field, and the specialist skills that RSEs provide to support and undertake research will
+Nonetheless, as the use of digital tools becomes ever more pervasive in a researcher's day-to-day tasks, there are already people suggesting that RSEng will
+cease to exist as a separate field, and the specialist skills that Research Software Engineers(RSEs) provide to support and undertake research will
 become a core part of the tooling and skill set of researchers in almost all other domains.
@@
 we also recognise the need to acknowledge that the future is in constant flux and
-it is worthwhile discussing how RSE will/could change in response to these upcoming challenges in an open discussion.} % Abstract of the article
-\keywords{Future RSE, Evolving field, Vision
-} % Keywords for the article
+it is worthwhile discussing how RSEng will/could change in response to these upcoming challenges in an open discussion.
+To that end we held a workshop and structured the resulting discussion around social, technological, and organizational aspects.
+} % Abstract of the article
+\keywords{Keywords go here.} % Keywords for the article
 
@@
 and community in the mid-term future, focusing on future challenges.
+While future projections are common in industry \cite{gartner}, so far there is a projection lacking for the RSEng space.
 We present here the results of a workshop discussion at the
@@
 we aim to engage the community in creating a future worth living
-and advance the conversation about how Research Software Engineering should, or could, develop.
-In the following sections, we consider a number of areas:
-the evolution of the RSE definition;
-the development of RSE education;
-the growth of the RSE communities;
-the impact of automation and generative artificial intelligence in RSE work;
-the emerging ethical considerations;
-the transformations expected in research organisations;
-and, finally, if there will be other areas in academia covered by new groups of RTPs (\emph{Research Technical Professional}s).
+and advance the conversation about how Research Software Engineering should, or could, develop.\\
+It is already difficult to structure the spectrum of RSE tasks and responsibilities today given the spectrum of research software \cite{hasselbring2024}
+and it is even more challenging to structure the aspects that will become relevant for the RSE community in the future.
+Nevertheless, we decided to consider:
+\begin{itemize}
+\item The evolution of the term RSE and its definition
+\item Growth and development of the RSE community
+\item The impact of automation and generative artificial intelligence in RSE work
+\item The impact of complexity and usability in research software
+\item Ethics and social considerations
+\item The transformations expected in academic institutions
+\item Impact of alternate career paths for RSEs and other RTPs(\emph{Research Technical Professional}s)
+\end{itemize}
 We conclude the paper with a summary and with our wishes and recommendations.
@@
 and green computing - efficient implementations of algorithms/methods and use of hardware
-
-It is already difficult to structure the spectrum of RSE tasks and responsibilities today given the spectrum of research software \cite{hasselbring2024}
-and it is even more challenging to structure the aspects that will become relevant for the RSE community in the future.
-Nevertheless, we decided to consider:
-\begin{itemize}
-\item The evolution of the term RSE
-\item Development of the RSE community
-\item What the impact of AI will be to the work of RSEs
-\item The impact of complexity and usability in research software
-\item Ethics and social considerations
-\item How education is reshaped by the digital transformation
-\item Changes to academic institutions
-\item Impact of alternate career paths for RSEs and other RTPs
-\end{itemize}
 

### 2025-05-16 future_challenges_for_RSEng.tex
Update future_challenges_for_RSEng.tex
Co-authored-by: Gerasimos Chourdakis <gerasimos.chourdakis@ipvs.uni-stuttgart.de>

--- 2025-05-16 future_challenges_for_RSEng.tex
@@
 \end{whatis}
-Academia, while claiming to be innovative \cite{wzvgbayern}, has organisationally often adopted time-scales for change that are more reminiscient to institutions thinking on the time scale of generations.
+Academia, while claiming to be innovative \cite{wzvgbayern}, has organisationally often adopted time-scales for change that are more reminiscent to institutions thinking on the time scale of generations.
 Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref} \cite{hiddenref} hints at the potential for specialisation of research tasks.

### 2025-05-16 future_challenges_for_RSEng.tex
finaluise

--- 2025-05-16 future_challenges_for_RSEng.tex
@@
 \newcounter{whatiscounter}
-\newtcbtheorem[use counter=whatiscounter, number format=\Alph]{whatis}{What do we want to have achieved}{
+\newtcbtheorem[use counter=whatiscounter, number format=\Alph]{whatis}{Achieved Goals}{
   enhanced,
@@
 \newcommand{\eg}{e.\,g.}
+\newcommand{\geprislink}[1]{\href{https://gepris.dfg.de/gepris/projekt/#1?language=en}{#1}}
 
@@
 become a core part of the tooling and skill set of researchers in almost all other domains.
-Are we already approaching ``the end of Research Software Engineering''?! While we believe that it is certainly too early to
-already herald the demise of RSE as a separate field,
+Are we already approaching ``the end of Research Software Engineering''?!
+While we believe that it is certainly too early to already herald the demise of RSE as a separate field,
 we also recognise the need to acknowledge that the future is in constant flux and
 it is worthwhile discussing how RSEng will/could change in response to these upcoming challenges in an open discussion.
-To that end we held a workshop and structured the resulting discussion around social, technological, and organizational aspects.
+To figure out these challenges of the future we held a workshop and structured the resulting discussion around social, technological, and organisational aspects.
 } % Abstract of the article
@@
 \section{Aspects of this future}
-%\subsection{What do we want to have achieved by 2035}
+\begin{whatis}{}{whatisexplanation}
+In order to list what has been achieved in a certain aspect within the RSE community of the future of 2035 we use these ``Goal'' boxes.
+\end{whatis}
+
 Looking ahead over a 10-year timeframe, we hope that the RSE community will have
@@
 research activities, this is vital and it will allow RSEs to focus on other more advanced and specialist
-areas. In this context, we do not foresee the demise of RSE as an independent role. However, we
+areas. In this context, we do not foresee the demise of the RSE as an independent role. However, we
 do see the work of RSEs developing and shifting as the research community changes.
@@
 We would, therefore, not expect domain researchers to also become experts in a range of
-advanced RSE topics. 
+advanced RSEng topics. 
 At the same time, we do recognise that the application of core software development best practices
@@
 
-RSE-relevant open educational resources and master's programmes are tracked
+RSEng-relevant open educational resources and master's programmes are tracked
 in the Learning and Teaching RSE database\footnote{\url{https://de-rse.org/learn-and-teach/}}
@@
 widely recognized and rewarded across academia and is supported by political actors.
-To conclude, RSEs will very likely be doing different work
+\section{Conclusion}
+We have tried to elaborate on the state on the future ten years from now.
+It turned out to be a difficult endeavour, since even the full breadth of today's trends is not yet foreseeable.
+We have tried to shed some light on the social, technical and organisational aspects.
+In summary, RSEs will very likely be doing different work
 to what they are doing today, but their core task of enhancing the scientific output of researchers -- other human
@@
 % AI Systems have been harmed in the creation of this work.
-We thank all the participants of our workshop at deRSE25 in Karlsruhe for their valuable input!  JC acknowledges support from the UKRI-EPSRC-funded STEP-UP project under grant EP/Y530608/1.
+We thank all the participants of our workshop at deRSE25 in Karlsruhe for their valuable input!
+FG acknowledges funding from the Deutsche Forschungsgemeinschaft(DFG, German Research Foundation) through the SFB 1170 “Tocotronics”, project Z03 - project number \geprislink{258499086}
+as well as financial support by the Deutsche Forschungsgemeinschaft (DFG, German Research
+Foundation) under Germany’s Excellence Strategy through the Würzburg-Dresden Cluster of Excellence on Complexity and Topology in Quantum Matter – ct.qmat (EXC 2147, project-id \geprislink{390858490}).
+JC acknowledges support from the UKRI-EPSRC-funded STEP-UP project under grant EP/Y530608/1.
 \end{acknowledge}

### 2025-05-16 future_challenges_for_RSEng.tex
typo

--- 2025-05-16 future_challenges_for_RSEng.tex
@@
 and software will become easier to interact with,
-reducing reliance on RSEs to train newcomers on the most basic tasks.
-This means, that today's workshop content of proper versioning and data management could become obsolete
-by more intuitive tooling, hence basic courses can focus on more higher-level concepts.
+reducing reliance on RSEs to train newcomers on the most basic tasks
+(JN: for example, versioning and archiving?)
+(GC: Higher-level and more intuitive workflows that require less training).
 User training sessions might focus on more advanced topics
@@
 the professionalisation%
+\footnote{MM: do we want to decide for UK or US English? Or keep it mixed? Not being a native English speaker I don't know how it reads.}
 of the RSE role.
@@
 % society implications
-The work being undertaken by RSEs may well have direct, or indirect,
-ethical implications in areas such as social engineering \cite{s2erc,siadati2024},
-climate \cite{Lannelongue2023} and more.
+The work being undertaken by RSEs may well have direct, or indirect, ethical implications in areas such as social engineering [TODO find source], climate [TODO find source] and more.
 Another example might be that they are developing code that can also be used for military purposes, such as numerical solvers or computer vision systems.
@@
 A code of ethics and professional conduct for RSEs should be developed by institutions together with RSEs themselves.
-This can then form the basis for institutional policies and guidelines as well as serve as a background for individual developers to consider how software can be used.
+This can then form the basis for institutional policies and guidelines.
+
+MM-NOTE:
+  - Importance of having personal “red lines” and awareness of how software can be used.
 
@@
 \end{whatis}
-Academia, while claiming to be innovative \cite{wzvgbayern}, has organisationally often adopted time-scales for change that are more reminiscent to institutions thinking on the time scale of generations.
+Academia, while claiming to be innovative (cite WissZeitVG here), has organisationally often adopted time-scales for change that are more reminiscient to civil services like pension funds.
 Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref} \cite{hiddenref} hints at the potential for specialisation of research tasks.
@@
 
-On the other hand the existence of people with a certain specilised skillset opens up possiblities for pooling.
+On the other hand the existence of people with a certain specialised skillset opens up possiblities for pooling.
 Universities will benefit from pooling these specialists into RSE units, and offer their services to researchers.

### 2025-05-16 future_challenges_for_RSEng.tex

--- 2025-05-16 future_challenges_for_RSEng.tex
@@
 \end{whatis}
-Academia, while claiming to be innovative (cite WissZeitVG here), has organisationally often adopted time-scales for change that are more reminiscient to civil services like pension funds.
+Academia, while claiming to be innovative \cite{wzvgbayern}, has organisationally often adopted time-scales for change that are more reminiscent to institutions thinking on the time scale of generations.
 Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref} \cite{hiddenref} hints at the potential for specialisation of research tasks.

### 2025-05-17 future_challenges_for_RSEng.tex
unify spelling of Master's programme

--- 2025-05-16 future_challenges_for_RSEng.tex
@@
 The term RSE is established in the minds of university boards and leadership, as a valuable member in the academic framework.
-New RSEs are trained through a specialized master's program, and their support is valued by researchers.
+New RSEs are trained through a specialized Master's programme, and their support is valued by researchers.
 We still teach researchers basic software skills, but the content of this lecture has changed drastically.
@@
 
-RSE-relevant open educational resources and master's programmes are tracked
+RSE-relevant open educational resources and Master's programmes are tracked
 in the Learning and Teaching RSE database\footnote{\url{https://de-rse.org/learn-and-teach/}}
@@
 Regular international conferences further amplify these connections, promoting the global exchange of ideas.
-The introduction of RSE Master's programs, coupled with a shift in culture towards valuing Research Software, has significantly elevated its profile.
+The introduction of RSE Master's programmes, coupled with a shift in culture towards valuing Research Software, has significantly elevated its profile.
 This growing recognition has led to a surge in membership within national societies, reflecting the increasing importance of RSEng in the research landscape.
@@
 We can expect RSEs to become more easily accessible, possibly through centrally-funded long-term positions at research institutions.
-The community of RSE's will get a significant boost in their acceptance if RSE Masters programmes
+The community of RSE's will get a significant boost in their acceptance if RSE Master's programmes
 become available at various universities.
@@
 After that, organisations will have to come up with updated models for hiring people.
-While a standardized RSE Masters makes it easier to develop the next generation of practitioners with the necessary technical skills, it will be interesting to see if
-RSE are still often hired by word of mouth within an organisation(``I have this PhD here, before their PostDoc they would be a good fit for the RSE department...'' ) or if the RSE Masters degree ensures a hiring process on a more standardised level.
+While a standardized RSE Master's degree makes it easier to develop the next generation of practitioners with the necessary technical skills, it will be interesting to see if
+RSE are still often hired by word of mouth within an organisation(``I have this PhD here, before their PostDoc they would be a good fit for the RSE department...'' ) or if the RSE Master's degree ensures a hiring process on a more standardised level.
 

### 2025-05-17 future_challenges_for_RSEng.tex
run spell checker with BE dict

@@
 %\subsection{What do we want to have achieved by 2035}
-Looking ahead over a 10-year timeframe, we hope that the RSE community will have
+Looking ahead over a 10-year time frame, we hope that the RSE community will have
 contributed to making research software and its outputs
 more sustainable, robust and maintainable.
-We also hope that Research Software becomes an accepted research artifact on an equal footing with data and publications.
+We also hope that Research Software becomes an accepted research artefact on an equal footing with data and publications.
 We see this happening in a number of ways,
@@
 Not too much to be overwhelmingly far in the future, but distant enough for some changes to occur.
-Over this timeframe Kim\cite{Anzt2021} and Kay\cite{Goth2024},
+Over this time frame Kim\cite{Anzt2021} and Kay\cite{Goth2024},
 our fictional RSEs who we have used to illustrate specific scenarios in our earlier work,
@@
 
-What will \emph{not} change over that timeframe is the RSE's ability to remain
+What will \emph{not} change over that time frame is the RSE's ability to remain
 flexible and embrace change, and keep an eye on the wider landscape,
@@
 AI is already transforming research processes by enabling rapid analysis of massive datasets, uncovering patterns and insights that would be impossible or prohibitively time-consuming for humans alone to find.
-In research and development, AI accelerates scientific breakthroughs in fields like healthcare, climate science, and materials engineering by automating complex modeling and predictive tasks.
+In research and development, AI accelerates scientific breakthroughs in fields like healthcare, climate science, and materials engineering by automating complex modelling and predictive tasks.
 For software development, AI-powered tools automate code generation, bug detection, and testing, significantly increasing productivity and reducing human error.
@@
 Since AI systems are so democratically available and will be used as black boxes,
-the interpretation of their outputs will be far more imortant.
+the interpretation of their outputs will be far more important.
 Currently we are expanding and documenting our work in the open and making our code available,
@@
 but quickly decided that work in academia with its time limited contracts is not attractive.
-Governments around the world are struggling to reposition themselves in this emerging deglobalized world with its new blocks
-and therefore have deprioritized publicly funded research. Kyle decided to go to an defense company since they offer
+Governments around the world are struggling to reposition themselves in this emerging de-globalised world with its new blocks
+and therefore have de-prioritised publicly funded research. Kyle decided to go to an defence company since they offer
 more secure employment in this polarized world.
@@
 
-On the other hand the existence of people with a certain specialised skillset opens up possiblities for pooling.
+On the other hand the existence of people with a certain specialised skillset opens up possibilities for pooling.
 Universities will benefit from pooling these specialists into RSE units, and offer their services to researchers.
@@
 With longer-term funding and more trust to pick up longer term projects,
-some RSEs can focus on code modernisation, cleanup, and modularisation,
+some RSEs can focus on code modernisation, clean-up, and modularisation,
 which can pay off in faster and easier development in the future.
@@
 While we hope that the barrier to researchers diminishes, the workshop participants
-still expect some work to do. We hope that the compartmentalisation that occurs will help researchers reprioritize their workload.
+still expect some work to do. We hope that the compartmentalisation that occurs will help researchers re-prioritise their workload.
 An upcoming challenge will be that, with the emergence of more dedicated RSE units,
@@
 \begin{whatis}{}{career-paths}
-We would like to see the emergence of more dedicated personel, i.e. other (digital) RTPs next to the RSE.
+We would like to see the emergence of more dedicated personnel, i.e. other (digital) RTPs next to the RSE.
 Also we hope that researchers can build careers at the interface of academia, industry, and policy (\eg in think tanks, NGOs,
@@
 beings -- will stay the same.
-They will still develop software artifacts together with scientists,
+They will still develop software artefacts together with scientists,
 since better software leads to better research,

### 2025-05-17 future_challenges_for_RSEng.tex
reformulation

@@
 Nevertheless, by disseminating the discussed ideas,
-we aim to engage the community in creating a future worth living
-and advance the conversation about how Research Software Engineering should, or could, develop.
+we aim to engage the community in creating a worthwhile future
+and to advance the conversation about how Research Software Engineering should, or could, develop.
 In the following sections, we consider a number of areas:

### 2025-05-17 future_challenges_for_RSEng.tex
split sentence

@@
 advocacy, to ensure that researchers receive core software skills as part of their scientific training.
-With development of software tools and scripted workflows becoming an increasingly important part of everyday
-research activities, this is vital and it will allow RSEs to focus on other more advanced and specialist
+Developing software tools and scripted workflows is becoming an increasingly important part of everyday
+research activities.
+Therefore, it is vital to Improve digital skills of all researchers.
+This will also allow RSEs to focus on other more advanced and specialist
 areas. In this context, we do not foresee the demise of RSE as an independent role. However, we

### 2025-05-17 future_challenges_for_RSEng.tex
swap words around
split sentence

@@
 The specialist skills that RSEs provide take a significant amount of time to develop and maintain.
-We would, therefore, not expect domain researchers to also become experts in a range of
+Therefore, we do not expect domain researchers to also become experts in a range of
 advanced RSE topics. 
@@
 
-It is already difficult to structure the spectrum of RSE tasks and responsibilities today given the spectrum of research software \cite{hasselbring2024}
-and it is even more challenging to structure the aspects that will become relevant for the RSE community in the future.
+It is already difficult to structure the spectrum of RSE tasks and responsibilities today given the spectrum of research software \cite{hasselbring2024}.
+It is even more challenging to structure the aspects that will become relevant for the RSE community in the future.
 Nevertheless, we decided to consider:

### 2025-05-17 future_challenges_for_RSEng.tex
change tense
we want to full programme of lovely sun not only at the end of the day

@@
 In order to give a more concrete setting, imagine a world ten years from now.
-Not too much to be overwhelmingly far in the future, but distant enough for some changes to occur.
+Not too overwhelmingly far ahead in the future, but distant enough for some changes to have occurred.
 Over this time frame Kim\cite{Anzt2021} and Kay\cite{Goth2024},
 our fictional RSEs who we have used to illustrate specific scenarios in our earlier work,
-were able to ride the waves of their career in Research Software Engineering and are now pondering early retirement on a sunset blessed beach.
+were able to ride the waves of their career in Research Software Engineering and are now pondering early retirement on a sun blessed beach.
 %MM-NOTE: the above scenario (early retirement) is contrary to all trends in retirement age development of public employees across the "global north".

### 2025-05-17 future_challenges_for_RSEng.tex
change tense
split sentence

@@
 But the world has moved on.
-When they look back, what will they see as problems which are now solved?
+When they look back, what will they see as problems that have been solved by then?
 Which tasks are still relevant, where they can be happy that a younger generation is now taking up the baton and carrying forward their work?
@@
 This means, that today's workshop content of proper versioning and data management could become obsolete
-by more intuitive tooling, hence basic courses can focus on more higher-level concepts.
+by using more intuitive tools.
+Introductory courses can then focus on more higher-level concepts.
 User training sessions might focus on more advanced topics

### 2025-05-17 future_challenges_for_RSEng.tex
change tense

@@
 \begin{story}{The Definition of RSEs}{defrse}
-Kim and Kay chuckled as they looked at their original job descriptions from 2020—"build tools for scientists, manage data, write reusable code."
-Now, their roles were unrecognizable: half AI whisperer, half research ethicist, and part-time diplomat between synthetic and human collaborators.
-They managed not just software, but the personalities of autonomous lab agents and negotiated data-sharing agreements with international AI consortiums.
-Despite the tech upheaval, their deep knowledge of scientific workflows and the messy, human way science actually gets done remained indispensable.
-When labs faltered under algorithmic bias or unclear model provenance, it was Kim and Kay who translated the chaos back into meaning.
-They understood that science didn’t move in straight lines, and that trust—between collaborators, institutions, and even machines—had to be built, not coded.
-Their titles had changed, but their role as bridges between logic and life remained the same.
-In a future of shifting code and constant reinvention, it was their understanding of people that never went out of date.
+Kim and Kay chuckled when they looked at their original job descriptions from 2020: "build tools for scientists, manage data, write reusable code."
+Now, their roles are unrecognizable: half AI whisperer, half research ethicist, and part-time diplomat negotiating between synthetic and human collaborators.
+They now not only manage software, but also the personalities of autonomous lab agents and negotiate data-sharing agreements with international AI consortia.
+Despite the tech upheaval, their deep knowledge of scientific workflows and the messy, human way science actually gets done remains indispensable.
+When labs faltered under algorithmic bias or unclear model provenance, it was Kim and Kay who transformed the chaos back into meaningful order.
+They understand that science does not move in straight lines, and that trust between collaborators, institutions, and even machines has to be built, not coded.
+Their job titles have changed, but their role as bridges between logic and life remains the same.
+In a future of shifting code and constant reinvention, it is their understanding of people that never went out of date.
 

### 2025-05-17 future_challenges_for_RSEng.tex
change word

@@
 The introduction of RSE Master's programmes, coupled with a shift in culture towards valuing Research Software, has significantly elevated its profile.
-This growing recognition has led to a surge in membership within national societies, reflecting the increasing importance of RSEng in the research landscape.
+This growing recognition has led to a surge in membership of national societies, reflecting the increasing importance of RSEng in the research landscape.
 \end{whatis}

### 2025-05-17 future_challenges_for_RSEng.tex
clarify
rephrase

@@
 and regional level to support more viable RSE career paths in academia.
-We can expect RSEs to become more easily accessible, possibly through centrally-funded long-term positions at research institutions.
+We can expect RSE positions to become more easily accessible, possibly through centrally-funded long-term positions at research institutions.
 The community of RSE's will get a significant boost in their acceptance if RSE Master's programmes
@@
 \begin{whatis}{}{impact-ai-automation}
-RSEs are educated about AI models, and can help researchers reap their benefits, while
+RSEs understand how AI models work, and can help researchers reap their benefits, while
 clearly knowing their limits.

### 2025-05-17 future_challenges_for_RSEng.tex
switch tense

@@
 \begin{story}{Development of RSE community}{RSEcommunity}
-Kay stood proudly at the podium of the first International Research Software Engineering Conference held in the Southern Hemisphere, a historic moment for both her career and the global RSE community.
-Having moved to South Africa, she spoke passionately about the unique challenges researchers in the Global South face—limited access to high-performance computing, unreliable funding, and the persistent digital divide.
-Her keynote highlighted how local innovation thrives despite constraints, often out of necessity rather than choice.
-The audience, a diverse mix of global experts, listened intently as she called for equitable collaborations, not just inclusion as an afterthought.
+Kay stands proudly at the podium of the first International Research Software Engineering Conference held in the Southern Hemisphere, a historic moment for both her career and the global RSE community.
+Having moved to South Africa, she speaks passionately about the unique challenges researchers in the Global South face: limited access to high-performance computing, unreliable funding, and the persistent digital divide.
+Her keynote highlights how local innovation thrives despite constraints, often out of necessity rather than choice.
+The audience, a diverse mix of global experts, listens intently as she calls for equitable collaborations, not just inclusion as an afterthought.
 \end{story}

### 2025-05-17 future_challenges_for_RSEng.tex
fix preposition
capitalise
swap verb

@@
 10 years have passed since Kim and Kay started on their journey of being RSEs. Back then,
-at the universities of Eden and Utopia, life was more orderly and harmonious. Kay is now working in a project that develops
-software for cheap water monitoring devices in the global south. Kim is currently designing the software service stack
+at the universities of Eden and Utopia, life was more orderly and harmonious. Kay is now working on a project that develops
+software for cheap water monitoring devices in the Global South. Kim is currently designing the software service stack
 at the university, and increasingly struggles to find compelling Open Source alternatives that enables them keep complete control of their data
@@
 Governments around the world are struggling to reposition themselves in this emerging de-globalised world with its new blocks
-and therefore have de-prioritised publicly funded research. Kyle decided to go to an defence company since they offer
+and therefore have de-prioritised publicly funded research. Kyle decided to work for a defence company since they offer
 more secure employment in this polarized world.

### 2025-05-17 future_challenges_for_RSEng.tex
state-actors sound a bit too undercover to me. So maybe more specific with who we mean.

@@
 permanent and well-funded RSE teams.
-Here it can only be strongly hoped, that, with the recognized importance of research software, state-actors
+Here it can only be strongly hoped, that, with the recognized importance of research software, research councils and funders
 support the development with a reliable stream of funding.

### 2025-05-17 future_challenges_for_RSEng.tex
change verb, BE spelling

@@
 Avalon which was interested in setting up their own department for RSEs in order to not fall behind the university of Arcadia.
-Kim improves upon the model set forth by the University of Arcadia based on valuable experience. Since the university
+Kim builds upon the model set forth by the University of Arcadia based on his valuable experience. Since the university
 of Avalon has to follow different rules, not everything can be straightforwardly adapted. Life would be so much easier
-if at least university regulations would be homogenized, no?
+if at least university regulations would be homogenised, no?
 \end{story}

### 2025-05-17 future_challenges_for_RSEng.tex
another round of BE spelling

@@
 The term RSE is established in the minds of university boards and leadership, as a valuable member in the academic framework.
-New RSEs are trained through a specialized Master's programme, and their support is valued by researchers.
+New RSEs are trained through a specialised Master's programme, and their support is valued by researchers.
 We still teach researchers basic software skills, but the content of this lecture has changed drastically.
@@
 Whereas previously RSEs have been points of knowledge that could translate human language formulated problems into technical descriptions and then provide solutions to them,
-it can be foreseen that AI will drive this low-key interaction to a highly specialized level of knowledge,
+it can be foreseen that AI will drive this low-key interaction to a highly specialised level of knowledge,
 since AI copilots are available 24/7.
@@
 Nowadays, Kim could have started directly from a template based on established toolkits and workflows,
-while kay-bio accepts configuration in a simpler, standardized format and generative AI assistants
+while kay-bio accepts configuration in a simpler, standardised format and generative AI assistants
 can guide new users in preparing and debugging their analyses.
@@
 and therefore have de-prioritised publicly funded research. Kyle decided to work for a defence company since they offer
-more secure employment in this polarized world.
+more secure employment in this polarised world.
 \end{story}
@@
 We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department to oversee specific projects \cite{Kempf2025-draft}.
-We also hope, that specialized clusters form that pool the skills of certain specialised RSEs, with certain RSE units specializing in certain tasks like data management, HPC, or visualisation.
+We also hope, that specialised clusters form that pool the skills of certain specialised RSEs, with certain RSE units specializing in certain tasks like data management, HPC, or visualisation.
 At universities which have them, we want to see that researchers naturally accept them as part of their available services.
@@
 permanent and well-funded RSE teams.
-Here it can only be strongly hoped, that, with the recognized importance of research software, research councils and funders
+Here it can only be strongly hoped, that, with the recognised importance of research software, research councils and funders
 support the development with a reliable stream of funding.
@@
 After that, organisations will have to come up with updated models for hiring people.
-While a standardized RSE Master's degree makes it easier to develop the next generation of practitioners with the necessary technical skills, it will be interesting to see if
+While a standardised RSE Master's degree makes it easier to develop the next generation of practitioners with the necessary technical skills, it will be interesting to see if
 RSE are still often hired by word of mouth within an organisation(``I have this PhD here, before their PostDoc they would be a good fit for the RSE department...'' ) or if the RSE Master's degree ensures a hiring process on a more standardised level.
@@
 We also hope that a cultural change is taking place to recognize research software as an integral part of science that it is
-widely recognized and rewarded across academia and is supported by political actors.
+widely recognised and rewarded across academia and is supported by political actors.
 To conclude, RSEs will very likely be doing different work

### 2025-05-17 future_challenges_for_RSEng.tex
correct word

@@
 \item Changes to academic institutions
-\item Impact of alternate career paths for RSEs and other RTPs
+\item Impact of alternative career paths for RSEs and other RTPs
 \end{itemize}
@@
 
-\subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}
+\subsection{Impact of Alternative Career Paths for Researchers on the Structure of Research Institutes}
 \begin{whatis}{}{career-paths}

### 2025-05-17 future_challenges_for_RSEng.tex
change word
add missing word

@@
 \begin{whatis}{}{career-paths}
-We would like to see the emergence of more dedicated personnel, i.e. other (digital) RTPs next to the RSE.
+We would like to see the emergence of more dedicated staff, i.e. other (digital) RTPs next to the RSE.
 Also we hope that researchers can build careers at the interface of academia, industry, and policy (\eg in think tanks, NGOs,
@@
 This, however, comes with additional funding challenges,
-as long-term funding can more difficult to acquire in lower tiers of an of a research organisation (i.e. at departmental or individual research group level).
+as long-term funding can be more difficult to acquire in lower tiers of an of a research organisation (i.e. at departmental or individual research group level).
 Third-party funding can mitigate this issue in the mid-term.

### 2025-05-17 future_challenges_for_RSEng.tex
add comma
use noun

@@
 The material presented here is based on the outputs of a workshop, but extended with the authors' ideas.
-Nevertheless there are some things that we declared out of scope here.
+Nevertheless, there are some things that we declared out of scope here.
 Among that is the ongoing digital transformation in wider society.
-Digital competencies in the general population continue to increase, making jumping to software development easier than before.
+Digital competencies in the general population continue to increase, making the jump to software development easier than before.
 But this digitalisation will also lead to changes in the way societies interact and therefore will translate into changes in how academia perceives itself.

### 2025-05-17 future_challenges_for_RSEng.tex

--- 2025-05-17 future_challenges_for_RSEng.tex
@@
 Nevertheless, by disseminating the discussed ideas,
-we aim to engage the community in creating a worthwhile future
-and to advance the conversation about how Research Software Engineering should, or could, develop.
+we aim to engage the community in creating a future worth living
+and advance the conversation about how Research Software Engineering should, or could, develop.
 In the following sections, we consider a number of areas:
@@
 %\subsection{What do we want to have achieved by 2035}
-Looking ahead over a 10-year time frame, we hope that the RSE community will have
+Looking ahead over a 10-year timeframe, we hope that the RSE community will have
 contributed to making research software and its outputs
 more sustainable, robust and maintainable.
-We also hope that Research Software becomes an accepted research artefact on an equal footing with data and publications.
+We also hope that Research Software becomes an accepted research artifact on an equal footing with data and publications.
 We see this happening in a number of ways,
@@
 advocacy, to ensure that researchers receive core software skills as part of their scientific training.
-Developing software tools and scripted workflows is becoming an increasingly important part of everyday
-research activities.
-Therefore, it is vital to Improve digital skills of all researchers.
-This will also allow RSEs to focus on other more advanced and specialist
+With development of software tools and scripted workflows becoming an increasingly important part of everyday
+research activities, this is vital and it will allow RSEs to focus on other more advanced and specialist
 areas. In this context, we do not foresee the demise of RSE as an independent role. However, we
@@
 The specialist skills that RSEs provide take a significant amount of time to develop and maintain.
-Therefore, we do not expect domain researchers to also become experts in a range of
+We would, therefore, not expect domain researchers to also become experts in a range of
 advanced RSE topics. 
@@
 
-It is already difficult to structure the spectrum of RSE tasks and responsibilities today given the spectrum of research software \cite{hasselbring2024}.
-It is even more challenging to structure the aspects that will become relevant for the RSE community in the future.
+It is already difficult to structure the spectrum of RSE tasks and responsibilities today given the spectrum of research software \cite{hasselbring2024}
+and it is even more challenging to structure the aspects that will become relevant for the RSE community in the future.
 Nevertheless, we decided to consider:
@@
 \item Changes to academic institutions
-\item Impact of alternative career paths for RSEs and other RTPs
+\item Impact of alternate career paths for RSEs and other RTPs
 \end{itemize}
@@
 In order to give a more concrete setting, imagine a world ten years from now.
-Not too overwhelmingly far ahead in the future, but distant enough for some changes to have occurred.
-Over this time frame Kim\cite{Anzt2021} and Kay\cite{Goth2024},
+Not too much to be overwhelmingly far in the future, but distant enough for some changes to occur.
+Over this timeframe Kim\cite{Anzt2021} and Kay\cite{Goth2024},
 our fictional RSEs who we have used to illustrate specific scenarios in our earlier work,
-were able to ride the waves of their career in Research Software Engineering and are now pondering early retirement on a sun blessed beach.
+were able to ride the waves of their career in Research Software Engineering and are now pondering early retirement on a sunset blessed beach.
 %MM-NOTE: the above scenario (early retirement) is contrary to all trends in retirement age development of public employees across the "global north".
 But the world has moved on.
-When they look back, what will they see as problems that have been solved by then?
+When they look back, what will they see as problems which are now solved?
 Which tasks are still relevant, where they can be happy that a younger generation is now taking up the baton and carrying forward their work?
@@
 The term RSE is established in the minds of university boards and leadership, as a valuable member in the academic framework.
-New RSEs are trained through a specialised Master's programme, and their support is valued by researchers.
+New RSEs are trained through a specialized master's program, and their support is valued by researchers.
 We still teach researchers basic software skills, but the content of this lecture has changed drastically.
@@
 This means, that today's workshop content of proper versioning and data management could become obsolete
-by using more intuitive tools.
-Introductory courses can then focus on more higher-level concepts.
+by more intuitive tooling, hence basic courses can focus on more higher-level concepts.
 User training sessions might focus on more advanced topics
@@
 
-What will \emph{not} change over that time frame is the RSE's ability to remain
+What will \emph{not} change over that timeframe is the RSE's ability to remain
 flexible and embrace change, and keep an eye on the wider landscape,
@@
 
-RSE-relevant open educational resources and Master's programmes are tracked
+RSE-relevant open educational resources and master's programmes are tracked
 in the Learning and Teaching RSE database\footnote{\url{https://de-rse.org/learn-and-teach/}}
@@
 \begin{story}{The Definition of RSEs}{defrse}
-Kim and Kay chuckled when they looked at their original job descriptions from 2020: "build tools for scientists, manage data, write reusable code."
-Now, their roles are unrecognizable: half AI whisperer, half research ethicist, and part-time diplomat negotiating between synthetic and human collaborators.
-They now not only manage software, but also the personalities of autonomous lab agents and negotiate data-sharing agreements with international AI consortia.
-Despite the tech upheaval, their deep knowledge of scientific workflows and the messy, human way science actually gets done remains indispensable.
-When labs faltered under algorithmic bias or unclear model provenance, it was Kim and Kay who transformed the chaos back into meaningful order.
-They understand that science does not move in straight lines, and that trust between collaborators, institutions, and even machines has to be built, not coded.
-Their job titles have changed, but their role as bridges between logic and life remains the same.
-In a future of shifting code and constant reinvention, it is their understanding of people that never went out of date.
+Kim and Kay chuckled as they looked at their original job descriptions from 2020—"build tools for scientists, manage data, write reusable code."
+Now, their roles were unrecognizable: half AI whisperer, half research ethicist, and part-time diplomat between synthetic and human collaborators.
+They managed not just software, but the personalities of autonomous lab agents and negotiated data-sharing agreements with international AI consortiums.
+Despite the tech upheaval, their deep knowledge of scientific workflows and the messy, human way science actually gets done remained indispensable.
+When labs faltered under algorithmic bias or unclear model provenance, it was Kim and Kay who translated the chaos back into meaning.
+They understood that science didn’t move in straight lines, and that trust—between collaborators, institutions, and even machines—had to be built, not coded.
+Their titles had changed, but their role as bridges between logic and life remained the same.
+In a future of shifting code and constant reinvention, it was their understanding of people that never went out of date.
 
@@
 Regular international conferences further amplify these connections, promoting the global exchange of ideas.
-The introduction of RSE Master's programmes, coupled with a shift in culture towards valuing Research Software, has significantly elevated its profile.
-This growing recognition has led to a surge in membership of national societies, reflecting the increasing importance of RSEng in the research landscape.
+The introduction of RSE Master's programs, coupled with a shift in culture towards valuing Research Software, has significantly elevated its profile.
+This growing recognition has led to a surge in membership within national societies, reflecting the increasing importance of RSEng in the research landscape.
 \end{whatis}
@@
 and regional level to support more viable RSE career paths in academia.
-We can expect RSE positions to become more easily accessible, possibly through centrally-funded long-term positions at research institutions.
-The community of RSE's will get a significant boost in their acceptance if RSE Master's programmes
+We can expect RSEs to become more easily accessible, possibly through centrally-funded long-term positions at research institutions.
+The community of RSE's will get a significant boost in their acceptance if RSE Masters programmes
 become available at various universities.
@@
 \begin{story}{Development of RSE community}{RSEcommunity}
-Kay stands proudly at the podium of the first International Research Software Engineering Conference held in the Southern Hemisphere, a historic moment for both her career and the global RSE community.
-Having moved to South Africa, she speaks passionately about the unique challenges researchers in the Global South face: limited access to high-performance computing, unreliable funding, and the persistent digital divide.
-Her keynote highlights how local innovation thrives despite constraints, often out of necessity rather than choice.
-The audience, a diverse mix of global experts, listens intently as she calls for equitable collaborations, not just inclusion as an afterthought.
+Kay stood proudly at the podium of the first International Research Software Engineering Conference held in the Southern Hemisphere, a historic moment for both her career and the global RSE community.
+Having moved to South Africa, she spoke passionately about the unique challenges researchers in the Global South face—limited access to high-performance computing, unreliable funding, and the persistent digital divide.
+Her keynote highlighted how local innovation thrives despite constraints, often out of necessity rather than choice.
+The audience, a diverse mix of global experts, listened intently as she called for equitable collaborations, not just inclusion as an afterthought.
 \end{story}
@@
 \begin{whatis}{}{impact-ai-automation}
-RSEs understand how AI models work, and can help researchers reap their benefits, while
+RSEs are educated about AI models, and can help researchers reap their benefits, while
 clearly knowing their limits.
@@
 AI is already transforming research processes by enabling rapid analysis of massive datasets, uncovering patterns and insights that would be impossible or prohibitively time-consuming for humans alone to find.
-In research and development, AI accelerates scientific breakthroughs in fields like healthcare, climate science, and materials engineering by automating complex modelling and predictive tasks.
+In research and development, AI accelerates scientific breakthroughs in fields like healthcare, climate science, and materials engineering by automating complex modeling and predictive tasks.
 For software development, AI-powered tools automate code generation, bug detection, and testing, significantly increasing productivity and reducing human error.
@@
 Whereas previously RSEs have been points of knowledge that could translate human language formulated problems into technical descriptions and then provide solutions to them,
-it can be foreseen that AI will drive this low-key interaction to a highly specialised level of knowledge,
+it can be foreseen that AI will drive this low-key interaction to a highly specialized level of knowledge,
 since AI copilots are available 24/7.
 Since AI systems are so democratically available and will be used as black boxes,
-the interpretation of their outputs will be far more important.
+the interpretation of their outputs will be far more imortant.
 Currently we are expanding and documenting our work in the open and making our code available,
@@
 Nowadays, Kim could have started directly from a template based on established toolkits and workflows,
-while kay-bio accepts configuration in a simpler, standardised format and generative AI assistants
+while kay-bio accepts configuration in a simpler, standardized format and generative AI assistants
 can guide new users in preparing and debugging their analyses.
@@
 10 years have passed since Kim and Kay started on their journey of being RSEs. Back then,
-at the universities of Eden and Utopia, life was more orderly and harmonious. Kay is now working on a project that develops
-software for cheap water monitoring devices in the Global South. Kim is currently designing the software service stack
+at the universities of Eden and Utopia, life was more orderly and harmonious. Kay is now working in a project that develops
+software for cheap water monitoring devices in the global south. Kim is currently designing the software service stack
 at the university, and increasingly struggles to find compelling Open Source alternatives that enables them keep complete control of their data
@@
 but quickly decided that work in academia with its time limited contracts is not attractive.
-Governments around the world are struggling to reposition themselves in this emerging de-globalised world with its new blocks
-and therefore have de-prioritised publicly funded research. Kyle decided to work for a defence company since they offer
-more secure employment in this polarised world.
+Governments around the world are struggling to reposition themselves in this emerging deglobalized world with its new blocks
+and therefore have deprioritized publicly funded research. Kyle decided to go to an defense company since they offer
+more secure employment in this polarized world.
 \end{story}
@@
 We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department to oversee specific projects \cite{Kempf2025-draft}.
-We also hope, that specialised clusters form that pool the skills of certain specialised RSEs, with certain RSE units specializing in certain tasks like data management, HPC, or visualisation.
+We also hope, that specialized clusters form that pool the skills of certain specialised RSEs, with certain RSE units specializing in certain tasks like data management, HPC, or visualisation.
 At universities which have them, we want to see that researchers naturally accept them as part of their available services.
@@
 
-On the other hand the existence of people with a certain specialised skillset opens up possibilities for pooling.
+On the other hand the existence of people with a certain specialised skillset opens up possiblities for pooling.
 Universities will benefit from pooling these specialists into RSE units, and offer their services to researchers.
@@
 With longer-term funding and more trust to pick up longer term projects,
-some RSEs can focus on code modernisation, clean-up, and modularisation,
+some RSEs can focus on code modernisation, cleanup, and modularisation,
 which can pay off in faster and easier development in the future.
@@
 permanent and well-funded RSE teams.
-Here it can only be strongly hoped, that, with the recognised importance of research software, research councils and funders
+Here it can only be strongly hoped, that, with the recognized importance of research software, state-actors
 support the development with a reliable stream of funding.
@@
 While we hope that the barrier to researchers diminishes, the workshop participants
-still expect some work to do. We hope that the compartmentalisation that occurs will help researchers re-prioritise their workload.
+still expect some work to do. We hope that the compartmentalisation that occurs will help researchers reprioritize their workload.
 An upcoming challenge will be that, with the emergence of more dedicated RSE units,
@@
 After that, organisations will have to come up with updated models for hiring people.
-While a standardised RSE Master's degree makes it easier to develop the next generation of practitioners with the necessary technical skills, it will be interesting to see if
-RSE are still often hired by word of mouth within an organisation(``I have this PhD here, before their PostDoc they would be a good fit for the RSE department...'' ) or if the RSE Master's degree ensures a hiring process on a more standardised level.
+While a standardized RSE Masters makes it easier to develop the next generation of practitioners with the necessary technical skills, it will be interesting to see if
+RSE are still often hired by word of mouth within an organisation(``I have this PhD here, before their PostDoc they would be a good fit for the RSE department...'' ) or if the RSE Masters degree ensures a hiring process on a more standardised level.
 
@@
 Avalon which was interested in setting up their own department for RSEs in order to not fall behind the university of Arcadia.
-Kim builds upon the model set forth by the University of Arcadia based on his valuable experience. Since the university
+Kim improves upon the model set forth by the University of Arcadia based on valuable experience. Since the university
 of Avalon has to follow different rules, not everything can be straightforwardly adapted. Life would be so much easier
-if at least university regulations would be homogenised, no?
+if at least university regulations would be homogenized, no?
 \end{story}
@@
 
-\subsection{Impact of Alternative Career Paths for Researchers on the Structure of Research Institutes}
+\subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}
 \begin{whatis}{}{career-paths}
-We would like to see the emergence of more dedicated personnel, i.e. other (digital) RTPs next to the RSE.
+We would like to see the emergence of more dedicated personel, i.e. other (digital) RTPs next to the RSE.
 Also we hope that researchers can build careers at the interface of academia, industry, and policy (\eg in think tanks, NGOs,
@@
 We also hope that a cultural change is taking place to recognize research software as an integral part of science that it is
-widely recognised and rewarded across academia and is supported by political actors.
+widely recognized and rewarded across academia and is supported by political actors.
 To conclude, RSEs will very likely be doing different work
@@
 beings -- will stay the same.
-They will still develop software artefacts together with scientists,
+They will still develop software artifacts together with scientists,
 since better software leads to better research,

### 2025-05-17 future_challenges_for_RSEng.tex
Update future_challenges_for_RSEng.tex
Co-authored-by: Magnus Hagdorn <Magnus.Hagdorn@marsupium.org>

@@
 }
-\abstract{It's 2025 and we are, admittedly, still working on establishing and growing Research Software Engineering (RSE)
+\abstract{
+It is 2025 and we are, admittedly, still working on establishing and growing Research Software Engineering (RSEng)
 as a domain in its own right, with the hope of ensuring that it gets the recognition it deserves.
-Nonetheless, as the use of digital tools becomes ever more pervasive in a researcher's day-to-day tasks, there are already people suggesting that RSE will
-cease to exist as a separate field, and the specialist skills that RSEs provide to support and undertake research will
+Nonetheless, as the use of digital tools becomes ever more pervasive in a researcher's day-to-day tasks, there are already people suggesting that RSEng will
+cease to exist as a separate field, and the specialist skills that Research Software Engineers(RSEs) provide to support and undertake research will
 become a core part of the tooling and skill set of researchers in almost all other domains.
-Are we already approaching ``the end of Research Software Engineering''?! While we believe that it is certainly too early to
-already herald the demise of RSE as a separate field,
+Are we already approaching ``the end of Research Software Engineering''?!
+While we believe that it is certainly too early to already herald the demise of RSE as a separate field,
 we also recognise the need to acknowledge that the future is in constant flux and
-it is worthwhile discussing how RSE will/could change in response to these upcoming challenges in an open discussion.} % Abstract of the article
-\keywords{Future RSE, Evolving field, Vision
-} % Keywords for the article
+it is worthwhile discussing how RSEng will/could change in response to these upcoming challenges.
+To figure out these challenges of the future we held a workshop and structured the resulting discussion around social, technological, and organisational aspects.
+} % Abstract of the article
+\keywords{Keywords go here.} % Keywords for the article
 

### 2025-05-17 future_challenges_for_RSEng.tex
Update future_challenges_for_RSEng.tex
Co-authored-by: Magnus Hagdorn <Magnus.Hagdorn@marsupium.org>

@@
 } % Abstract of the article
-\keywords{Keywords go here.} % Keywords for the article
+\keywords{Future RSE, Evolving field, Vision} % Keywords for the article
 
@@
 \item The transformations expected in academic institutions
-\item Impact of alternate career paths for RSEs and other RTPs(\emph{Research Technical Professional}s)
+\item Impact of alternative career paths for RSEs and other RTPs(\emph{Research Technical Professional}s)
 \end{itemize}

### 2025-05-18 future_challenges_for_RSEng.tex
inline items in intro

@@
 and it is even more challenging to structure the aspects that will become relevant for the RSE community in the future.
-Nevertheless, we decided to consider:
-\begin{itemize}
-\item The evolution of the term RSE and its definition
-\item Growth and development of the RSE community
-\item The impact of automation and generative artificial intelligence in RSE work
-\item The impact of complexity and usability in research software
-\item Ethics and social considerations
-\item The transformations expected in academic institutions
-\item Impact of alternative career paths for RSEs and other RTPs(\emph{Research Technical Professional}s)
-\end{itemize}
+Nevertheless, in order to structure the discussion, we decided to consider
+the evolution of the term RSE and its definition,
+the growth and development of the RSE community,
+the impact of automation and generative artificial intelligence in RSE work,
+the impact of complexity and usability in research software,
+ethical and social considerations,
+the expected transformations in academic institutions,
+and the potential impact of alternative career paths for RSEs and other RTPs(\emph{Research Technical Professional}s).
+%
 We conclude the paper with a summary and with our wishes and recommendations.

### 2025-05-18 future_challenges_for_RSEng.tex
work in another comment

@@
 \end{story}
-\section{What is in this future for RSEs and where to go from here?}
+\section{What is in this future for RSEs?}
 %FG-NOTE: I think we can use this part to wrap things a bit up, and describe how this future then looks like.
 %GC-NOTE: Shall we add another story box here? Ideas what we write here?
-It is hard to conclude a piece written about the future, but we have tried to list some of the trends that we foresee.
+In the previous sections, we have tried to list some of the trends that we foresee.
 The material presented here is based on the outputs of a workshop, but extended with the authors' ideas.

### 2025-05-18 future_challenges_for_RSEng.tex
update abstract

@@
 as a domain in its own right, with the hope of ensuring that it gets the recognition it deserves.
-Nonetheless, as the use of digital tools becomes ever more pervasive in a researcher's day-to-day tasks, there are already people suggesting that RSEng will
-cease to exist as a separate field, and the specialist skills that Research Software Engineers(RSEs) provide to support and undertake research will
-become a core part of the tooling and skill set of researchers in almost all other domains.
-Are we already approaching ``the end of Research Software Engineering''?!
-While we believe that it is certainly too early to already herald the demise of RSE as a separate field,
-we also recognise the need to acknowledge that the future is in constant flux and
-it is worthwhile discussing how RSEng will/could change in response to these upcoming challenges.
-To figure out these challenges of the future we held a workshop and structured the resulting discussion around social, technological, and organisational aspects.
+As the field develops, the role and value of Research Software Engineers (RSEs) becomes clearer.
+With the growing importance and scope of software engineering in research,
+researchers who code now begin to fill dedicated, specialized RSE roles crucial for the sustainable development
+of long-term research software projects that can rapidly innovate.
+Looking into the future, participants of the deRSE25 conference discussed how RSEng could evolve in response to the
+rapid changes in the broader social, technological, and organisational landscape.
+This paper presents the main discussion points of the conference participants,
+imagining a future in which RSEs have a well-defined and well-integrated role in research institutions,
+working as engineers of ever more complex research software projects,
+taming a growing range of methods and tools,
+and guiding researchers through technical and ethical considerations
+related to research software.
 } % Abstract of the article

### 2025-05-18 future_challenges_for_RSEng.tex
Mh review (#34)
* unify spelling of Master's programme

* run spell checker with BE dict

* change tense

* state-actors sound a bit too undercover to me.
So maybe more specific with who we mean.

--- 2025-05-18 future_challenges_for_RSEng.tex
@@
 \newcounter{whatiscounter}
-\newtcbtheorem[use counter=whatiscounter, number format=\Alph]{whatis}{Achieved Goals}{
+\newtcbtheorem[use counter=whatiscounter, number format=\Alph]{whatis}{What do we want to have achieved}{
   enhanced,
@@
 \newcommand{\eg}{e.\,g.}
-\newcommand{\geprislink}[1]{\href{https://gepris.dfg.de/gepris/projekt/#1?language=en}{#1}}
 
@@
 }
-\abstract{
-It is 2025 and we are, admittedly, still working on establishing and growing Research Software Engineering (RSEng)
+\abstract{It's 2025 and we are, admittedly, still working on establishing and growing Research Software Engineering (RSE)
 as a domain in its own right, with the hope of ensuring that it gets the recognition it deserves.
-Nonetheless, as the use of digital tools becomes ever more pervasive in a researcher's day-to-day tasks, there are already people suggesting that RSEng will
-cease to exist as a separate field, and the specialist skills that Research Software Engineers(RSEs) provide to support and undertake research will
+Nonetheless, as the use of digital tools becomes ever more pervasive in a researcher's day-to-day tasks, there are already people suggesting that RSE will
+cease to exist as a separate field, and the specialist skills that RSEs provide to support and undertake research will
 become a core part of the tooling and skill set of researchers in almost all other domains.
-Are we already approaching ``the end of Research Software Engineering''?!
-While we believe that it is certainly too early to already herald the demise of RSE as a separate field,
+Are we already approaching ``the end of Research Software Engineering''?! While we believe that it is certainly too early to
+already herald the demise of RSE as a separate field,
 we also recognise the need to acknowledge that the future is in constant flux and
-it is worthwhile discussing how RSEng will/could change in response to these upcoming challenges.
-To figure out these challenges of the future we held a workshop and structured the resulting discussion around social, technological, and organisational aspects.
-} % Abstract of the article
-\keywords{Future RSE, Evolving field, Vision} % Keywords for the article
+it is worthwhile discussing how RSE will/could change in response to these upcoming challenges in an open discussion.} % Abstract of the article
+\keywords{Future RSE, Evolving field, Vision
+} % Keywords for the article
 
@@
 and community in the mid-term future, focusing on future challenges.
-While future projections are common in industry \cite{gartner}, so far there is a projection lacking for the RSEng space.
 We present here the results of a workshop discussion at the
@@
 Nevertheless, by disseminating the discussed ideas,
-we aim to engage the community in creating a future worth living
-and advance the conversation about how Research Software Engineering should, or could, develop.\\
-It is already difficult to structure the spectrum of RSE tasks and responsibilities today given the spectrum of research software \cite{hasselbring2024}
-and it is even more challenging to structure the aspects that will become relevant for the RSE community in the future.
-Nevertheless, in order to structure the discussion, we decided to consider
-the evolution of the term RSE and its definition,
-the growth and development of the RSE community,
-the impact of automation and generative artificial intelligence in RSE work,
-the impact of complexity and usability in research software,
-ethical and social considerations,
-the expected transformations in academic institutions,
-and the potential impact of alternative career paths for RSEs and other RTPs(\emph{Research Technical Professional}s).
-%
+we aim to engage the community in creating a worthwhile future
+and to advance the conversation about how Research Software Engineering should, or could, develop.
+In the following sections, we consider a number of areas:
+the evolution of the RSE definition;
+the development of RSE education;
+the growth of the RSE communities;
+the impact of automation and generative artificial intelligence in RSE work;
+the emerging ethical considerations;
+the transformations expected in research organisations;
+and, finally, if there will be other areas in academia covered by new groups of RTPs (\emph{Research Technical Professional}s).
 We conclude the paper with a summary and with our wishes and recommendations.
@@
 \section{Aspects of this future}
-\begin{whatis}{}{whatisexplanation}
-In order to list what has been achieved in a certain aspect within the RSE community of the future of 2035 we use these ``Goal'' boxes.
-\end{whatis}
-
-Looking ahead over a 10-year timeframe, we hope that the RSE community will have
+%\subsection{What do we want to have achieved by 2035}
+Looking ahead over a 10-year time frame, we hope that the RSE community will have
 contributed to making research software and its outputs
 more sustainable, robust and maintainable.
-We also hope that Research Software becomes an accepted research artifact on an equal footing with data and publications.
+We also hope that Research Software becomes an accepted research artefact on an equal footing with data and publications.
 We see this happening in a number of ways,
@@
 advocacy, to ensure that researchers receive core software skills as part of their scientific training.
-With development of software tools and scripted workflows becoming an increasingly important part of everyday
-research activities, this is vital and it will allow RSEs to focus on other more advanced and specialist
-areas. In this context, we do not foresee the demise of the RSE as an independent role. However, we
+Developing software tools and scripted workflows is becoming an increasingly important part of everyday
+research activities.
+Therefore, it is vital to Improve digital skills of all researchers.
+This will also allow RSEs to focus on other more advanced and specialist
+areas. In this context, we do not foresee the demise of RSE as an independent role. However, we
 do see the work of RSEs developing and shifting as the research community changes.
@@
 The specialist skills that RSEs provide take a significant amount of time to develop and maintain.
-We would, therefore, not expect domain researchers to also become experts in a range of
-advanced RSEng topics. 
+Therefore, we do not expect domain researchers to also become experts in a range of
+advanced RSE topics. 
 At the same time, we do recognise that the application of core software development best practices
@@
 
+It is already difficult to structure the spectrum of RSE tasks and responsibilities today given the spectrum of research software \cite{hasselbring2024}.
+It is even more challenging to structure the aspects that will become relevant for the RSE community in the future.
+Nevertheless, we decided to consider:
+\begin{itemize}
+\item The evolution of the term RSE
+\item Development of the RSE community
+\item What the impact of AI will be to the work of RSEs
+\item The impact of complexity and usability in research software
+\item Ethics and social considerations
+\item How education is reshaped by the digital transformation
+\item Changes to academic institutions
+\item Impact of alternative career paths for RSEs and other RTPs
+\end{itemize}
+
 \begin{story}{Introduction}{intro}
 In order to give a more concrete setting, imagine a world ten years from now.
-Not too much to be overwhelmingly far in the future, but distant enough for some changes to occur.
-Over this timeframe Kim\cite{Anzt2021} and Kay\cite{Goth2024},
+Not too overwhelmingly far ahead in the future, but distant enough for some changes to have occurred.
+Over this time frame Kim\cite{Anzt2021} and Kay\cite{Goth2024},
 our fictional RSEs who we have used to illustrate specific scenarios in our earlier work,
-were able to ride the waves of their career in Research Software Engineering and are now pondering early retirement on a sunset blessed beach.
+were able to ride the waves of their career in Research Software Engineering and are now pondering early retirement on a sun blessed beach.
 %MM-NOTE: the above scenario (early retirement) is contrary to all trends in retirement age development of public employees across the "global north".
 But the world has moved on.
-When they look back, what will they see as problems which are now solved?
+When they look back, what will they see as problems that have been solved by then?
 Which tasks are still relevant, where they can be happy that a younger generation is now taking up the baton and carrying forward their work?
@@
 The term RSE is established in the minds of university boards and leadership, as a valuable member in the academic framework.
-New RSEs are trained through a specialized master's program, and their support is valued by researchers.
+New RSEs are trained through a specialised Master's programme, and their support is valued by researchers.
 We still teach researchers basic software skills, but the content of this lecture has changed drastically.
@@
 
-RSEng-relevant open educational resources and master's programmes are tracked
+RSE-relevant open educational resources and Master's programmes are tracked
 in the Learning and Teaching RSE database\footnote{\url{https://de-rse.org/learn-and-teach/}}
@@
 Regular international conferences further amplify these connections, promoting the global exchange of ideas.
-The introduction of RSE Master's programs, coupled with a shift in culture towards valuing Research Software, has significantly elevated its profile.
-This growing recognition has led to a surge in membership within national societies, reflecting the increasing importance of RSEng in the research landscape.
+The introduction of RSE Master's programmes, coupled with a shift in culture towards valuing Research Software, has significantly elevated its profile.
+This growing recognition has led to a surge in membership of national societies, reflecting the increasing importance of RSEng in the research landscape.
 \end{whatis}
@@
 and regional level to support more viable RSE career paths in academia.
-We can expect RSEs to become more easily accessible, possibly through centrally-funded long-term positions at research institutions.
-The community of RSE's will get a significant boost in their acceptance if RSE Masters programmes
+We can expect RSE positions to become more easily accessible, possibly through centrally-funded long-term positions at research institutions.
+The community of RSE's will get a significant boost in their acceptance if RSE Master's programmes
 become available at various universities.
@@
 Whereas previously RSEs have been points of knowledge that could translate human language formulated problems into technical descriptions and then provide solutions to them,
-it can be foreseen that AI will drive this low-key interaction to a highly specialized level of knowledge,
+it can be foreseen that AI will drive this low-key interaction to a highly specialised level of knowledge,
 since AI copilots are available 24/7.
 Since AI systems are so democratically available and will be used as black boxes,
-the interpretation of their outputs will be far more imortant.
+the interpretation of their outputs will be far more important.
 Currently we are expanding and documenting our work in the open and making our code available,
@@
 but quickly decided that work in academia with its time limited contracts is not attractive.
-Governments around the world are struggling to reposition themselves in this emerging deglobalized world with its new blocks
-and therefore have deprioritized publicly funded research. Kyle decided to go to an defense company since they offer
-more secure employment in this polarized world.
+Governments around the world are struggling to reposition themselves in this emerging de-globalised world with its new blocks
+and therefore have de-prioritised publicly funded research. Kyle decided to work for a defence company since they offer
+more secure employment in this polarised world.
 \end{story}
@@
 permanent and well-funded RSE teams.
-Here it can only be strongly hoped, that, with the recognized importance of research software, state-actors
+Here it can only be strongly hoped, that, with the recognised importance of research software, research councils and funders
 support the development with a reliable stream of funding.
@@
 After that, organisations will have to come up with updated models for hiring people.
-While a standardized RSE Masters makes it easier to develop the next generation of practitioners with the necessary technical skills, it will be interesting to see if
-RSE are still often hired by word of mouth within an organisation(``I have this PhD here, before their PostDoc they would be a good fit for the RSE department...'' ) or if the RSE Masters degree ensures a hiring process on a more standardised level.
+While a standardised RSE Master's degree makes it easier to develop the next generation of practitioners with the necessary technical skills, it will be interesting to see if
+RSE are still often hired by word of mouth within an organisation(``I have this PhD here, before their PostDoc they would be a good fit for the RSE department...'' ) or if the RSE Master's degree ensures a hiring process on a more standardised level.
 
@@
 
-\subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}
+\subsection{Impact of Alternative Career Paths for Researchers on the Structure of Research Institutes}
 \begin{whatis}{}{career-paths}
-We would like to see the emergence of more dedicated personel, i.e. other (digital) RTPs next to the RSE.
+We would like to see the emergence of more dedicated staff, i.e. other (digital) RTPs next to the RSE.
 Also we hope that researchers can build careers at the interface of academia, industry, and policy (\eg in think tanks, NGOs,
@@
 We also hope that a cultural change is taking place to recognize research software as an integral part of science that it is
-widely recognized and rewarded across academia and is supported by political actors.
-\section{Conclusion}
-We have tried to elaborate on the state on the future ten years from now.
-It turned out to be a difficult endeavour, since even the full breadth of today's trends is not yet foreseeable.
-We have tried to shed some light on the social, technical and organisational aspects.
-In summary, RSEs will very likely be doing different work
+widely recognised and rewarded across academia and is supported by political actors.
+To conclude, RSEs will very likely be doing different work
 to what they are doing today, but their core task of enhancing the scientific output of researchers -- other human
 beings -- will stay the same.
-They will still develop software artifacts together with scientists,
+They will still develop software artefacts together with scientists,
 since better software leads to better research,
@@
 % AI Systems have been harmed in the creation of this work.
-We thank all the participants of our workshop at deRSE25 in Karlsruhe for their valuable input!
-FG acknowledges funding from the Deutsche Forschungsgemeinschaft(DFG, German Research Foundation) through the SFB 1170 “Tocotronics”, project Z03 - project number \geprislink{258499086}
-as well as financial support by the Deutsche Forschungsgemeinschaft (DFG, German Research
-Foundation) under Germany’s Excellence Strategy through the Würzburg-Dresden Cluster of Excellence on Complexity and Topology in Quantum Matter – ct.qmat (EXC 2147, project-id \geprislink{390858490}).
-JC acknowledges support from the UKRI-EPSRC-funded STEP-UP project under grant EP/Y530608/1.
+We thank all the participants of our workshop at deRSE25 in Karlsruhe for their valuable input!  JC acknowledges support from the UKRI-EPSRC-funded STEP-UP project under grant EP/Y530608/1.
 \end{acknowledge}

### 2025-05-18 future_challenges_for_RSEng.tex
update ai section

@@
 \end{whatis}
-AI is already transforming research processes by enabling rapid analysis of massive datasets, uncovering patterns and insights that would be impossible or prohibitively time-consuming for humans alone to find.
-In research and development, AI accelerates scientific breakthroughs in fields like healthcare, climate science, and materials engineering by automating complex modelling and predictive tasks.
-For software development, AI-powered tools automate code generation, bug detection, and testing, significantly increasing productivity and reducing human error.
-Developers are shifting from manual coding to orchestrating AI-driven workflows, focusing more on high-level design, integration, and creative problem-solving while AI handles repetitive or routine tasks.
-The integration of AI is democratizing access to advanced technologies and analytics, enabling non-experts to participate in research and software creation, and fostering greater collaboration across disciplines.
-As AI continues to evolve, it will further blur the lines between human and machine roles, making research and software development faster, more efficient, and more innovative than ever before.
-With the general availability of AI the interaction with researchers will change at a fundamental level.
+Generative AI tools, such as GitHub Copilot \cite{Friedmann2021}, are being added to software development tools.
+These AI powered code assistants promise to increase productivity by automatically generating code, and writing test cases and documentation.
+Software developers use them to reduce the number of keystrokes, complete tasks and to recall syntax without having to consult the documentation \cite{Liang2024}.
+Code assistants have the potential to help novice or even non-expert programmers write code \cite{Feldman2024} thus enabling non-experts to participate in research and software creation, and fostering greater collaboration across disciplines.
+However, when these tools are used in practice many users report increased cognitive load and frustration stemming from the difficulty to prompt the system effectively and debugging the generated code \cite{Simkute2025}.
+
+The potential of AI to be a democratising technology can only be realised when the algorithmic divide along with the digital divide is addressed \cite{Yu2020}.
+
+As AI continues to evolve, it will further blur the lines between human and machine roles, making research and software development faster.
+With the general availability of AI, the interaction of RSEs with researchers will change at a fundamental level.
 Whereas previously RSEs have been points of knowledge that could translate human language formulated problems into technical descriptions and then provide solutions to them,
-it can be foreseen that AI will drive this low-key interaction to a highly specialised level of knowledge,
-since AI copilots are available 24/7.
-Since AI systems are so democratically available and will be used as black boxes,
-the interpretation of their outputs will be far more important.
-Currently we are expanding and documenting our work in the open and making our code available,
-therefore there will be more structured material to train GenAI on.
-However, as people get quicker answers from GenAI, the community resources (\eg Stack Overflow, Wikipedia)
-get poorer.
+it can be foreseen that AI will drive this low-key interaction to a highly specialised level of knowledge.
+
+Another risk of the widespread use of AI tools is that community resources such as \eg Stack Overflow and Wikipedia become poorer because there are fewer incentives to maintain them.
 

### 2025-05-18 future_challenges_for_RSEng.tex
Update future_challenges_for_RSEng.tex
Co-authored-by: Gerasimos Chourdakis <gerasimos.chourdakis@ipvs.uni-stuttgart.de>

@@
 Code assistants have the potential to help novice or even non-expert programmers write code \cite{Feldman2024} thus enabling non-experts to participate in research and software creation, and fostering greater collaboration across disciplines.
-However, when these tools are used in practice many users report increased cognitive load and frustration stemming from the difficulty to prompt the system effectively and debugging the generated code \cite{Simkute2025}.
+However, when these tools are used in practice, many users report increased cognitive load and frustration stemming from the difficulty to prompt the system effectively and from debugging the generated code \cite{Simkute2025}.
 

### 2025-05-18 future_challenges_for_RSEng.tex
remove paragraph

--- 2025-05-18 future_challenges_for_RSEng.tex
@@
 }
-\abstract{It's 2025 and we are, admittedly, still working on establishing and growing Research Software Engineering (RSE)
+\abstract{
+It is 2025 and we are, admittedly, still working on establishing and growing Research Software Engineering (RSEng)
 as a domain in its own right, with the hope of ensuring that it gets the recognition it deserves.
-Nonetheless, as the use of digital tools becomes ever more pervasive in a researcher's day-to-day tasks, there are already people suggesting that RSE will
-cease to exist as a separate field, and the specialist skills that RSEs provide to support and undertake research will
-become a core part of the tooling and skill set of researchers in almost all other domains.
-Are we already approaching ``the end of Research Software Engineering''?! While we believe that it is certainly too early to
-already herald the demise of RSE as a separate field,
-we also recognise the need to acknowledge that the future is in constant flux and
-it is worthwhile discussing how RSE will/could change in response to these upcoming challenges in an open discussion.} % Abstract of the article
-\keywords{Future RSE, Evolving field, Vision
-} % Keywords for the article
+As the field develops, the role and value of Research Software Engineers (RSEs) becomes clearer.
+With the growing importance and scope of software engineering in research,
+researchers who code now begin to fill dedicated, specialized RSE roles crucial for the sustainable development
+of long-term research software projects that can rapidly innovate.
+Looking into the future, participants of the deRSE25 conference discussed how RSEng could evolve in response to the
+rapid changes in the broader social, technological, and organisational landscape.
+This paper presents the main discussion points of the conference participants,
+imagining a future in which RSEs have a well-defined and well-integrated role in research institutions,
+working as engineers of ever more complex research software projects,
+taming a growing range of methods and tools,
+and guiding researchers through technical and ethical considerations
+related to research software.
+} % Abstract of the article
+\keywords{Future RSE, Evolving field, Vision} % Keywords for the article
 
@@
 Nevertheless, by disseminating the discussed ideas,
-we aim to engage the community in creating a worthwhile future
+we aim to engage the community in creating a future worth living
 and to advance the conversation about how Research Software Engineering should, or could, develop.
-In the following sections, we consider a number of areas:
-the evolution of the RSE definition;
-the development of RSE education;
-the growth of the RSE communities;
-the impact of automation and generative artificial intelligence in RSE work;
-the emerging ethical considerations;
-the transformations expected in research organisations;
-and, finally, if there will be other areas in academia covered by new groups of RTPs (\emph{Research Technical Professional}s).
+It is already difficult to structure the spectrum of RSE tasks and responsibilities today given the spectrum of research software \cite{hasselbring2024}.
+It is even more challenging to structure the aspects that will become relevant for the RSE community in the future.
+Nevertheless, in order to structure the discussion, we decided to consider
+the evolution of the term RSE and its definition,
+the growth and development of the RSE community,
+the impact of automation and generative artificial intelligence in RSE work,
+the impact of complexity and usability in research software,
+ethical and social considerations,
+the expected transformations in academic institutions,
+and the potential impact of alternative career paths for RSEs and new groups of RTPs(\emph{Research Technical Professional}s).
+%
 We conclude the paper with a summary and with our wishes and recommendations.
@@
 \section{Aspects of this future}
-%\subsection{What do we want to have achieved by 2035}
+\begin{whatis}{}{whatisexplanation}
+In order to list what has been achieved in a certain aspect within the RSE community of the future of 2035 we use these ``Goal'' boxes.
+\end{whatis}
+
 Looking ahead over a 10-year time frame, we hope that the RSE community will have
@@
 research activities.
-Therefore, it is vital to Improve digital skills of all researchers.
+Therefore, it is vital to improve digital skills of all researchers.
 This will also allow RSEs to focus on other more advanced and specialist
-areas. In this context, we do not foresee the demise of RSE as an independent role. However, we
+areas. In this context, we do not foresee the demise of the RSE as an independent role. However, we
 do see the work of RSEs developing and shifting as the research community changes.
@@
 Therefore, we do not expect domain researchers to also become experts in a range of
-advanced RSE topics. 
+advanced RSEng topics.
+
 At the same time, we do recognise that the application of core software development best practices
@@
 and green computing - efficient implementations of algorithms/methods and use of hardware
-
-It is already difficult to structure the spectrum of RSE tasks and responsibilities today given the spectrum of research software \cite{hasselbring2024}.
-It is even more challenging to structure the aspects that will become relevant for the RSE community in the future.
-Nevertheless, we decided to consider:
-\begin{itemize}
-\item The evolution of the term RSE
-\item Development of the RSE community
-\item What the impact of AI will be to the work of RSEs
-\item The impact of complexity and usability in research software
-\item Ethics and social considerations
-\item How education is reshaped by the digital transformation
-\item Changes to academic institutions
-\item Impact of alternative career paths for RSEs and other RTPs
-\end{itemize}
 
@@
 
-RSE-relevant open educational resources and Master's programmes are tracked
+RSEng-relevant open educational resources and Master's programmes are tracked
 in the Learning and Teaching RSE database\footnote{\url{https://de-rse.org/learn-and-teach/}}
@@
 widely recognised and rewarded across academia and is supported by political actors.
-To conclude, RSEs will very likely be doing different work
+\section{Conclusion}
+We have tried to elaborate on the state on the future ten years from now.
+It turned out to be a difficult endeavour, since even the full breadth of today's trends is not yet foreseeable.
+We have tried to shed some light on the social, technical and organisational aspects.
+In summary, RSEs will very likely be doing different work
+
 to what they are doing today, but their core task of enhancing the scientific output of researchers -- other human

### 2025-05-18 future_challenges_for_RSEng.tex

@@
 \end{whatis}
-AI is already transforming research processes by enabling rapid analysis of massive datasets, uncovering patterns and insights that would be impossible or prohibitively time-consuming for humans alone to find.
-In research and development, AI accelerates scientific breakthroughs in fields like healthcare, climate science, and materials engineering by automating complex modelling and predictive tasks.
-For software development, AI-powered tools automate code generation, bug detection, and testing, significantly increasing productivity and reducing human error.
-Developers are shifting from manual coding to orchestrating AI-driven workflows, focusing more on high-level design, integration, and creative problem-solving while AI handles repetitive or routine tasks.
-The integration of AI is democratizing access to advanced technologies and analytics, enabling non-experts to participate in research and software creation, and fostering greater collaboration across disciplines.
-As AI continues to evolve, it will further blur the lines between human and machine roles, making research and software development faster, more efficient, and more innovative than ever before.
-With the general availability of AI the interaction with researchers will change at a fundamental level.
+Generative AI tools, such as GitHub Copilot \cite{Friedmann2021}, are being added to software development tools.
+These AI powered code assistants promise to increase productivity by automatically generating code, and writing test cases and documentation.
+Software developers use them to reduce the number of keystrokes, complete tasks and to recall syntax without having to consult the documentation \cite{Liang2024}.
+Code assistants have the potential to help novice or even non-expert programmers write code \cite{Feldman2024} thus enabling non-experts to participate in research and software creation, and fostering greater collaboration across disciplines.
+However, when these tools are used in practice, many users report increased cognitive load and frustration stemming from the difficulty to prompt the system effectively and from debugging the generated code \cite{Simkute2025}.
+The potential of AI to be a democratising technology can only be realised when the algorithmic divide along with the digital divide is addressed \cite{Yu2020}.
+
+As AI continues to evolve, it will further blur the lines between human and machine roles, making research and software development faster.
+With the general availability of AI, the interaction of RSEs with researchers will change at a fundamental level.
 Whereas previously RSEs have been points of knowledge that could translate human language formulated problems into technical descriptions and then provide solutions to them,
-it can be foreseen that AI will drive this low-key interaction to a highly specialised level of knowledge,
-since AI copilots are available 24/7.
-Since AI systems are so democratically available and will be used as black boxes,
-the interpretation of their outputs will be far more important.
-Currently we are expanding and documenting our work in the open and making our code available,
-therefore there will be more structured material to train GenAI on.
-However, as people get quicker answers from GenAI, the community resources (\eg Stack Overflow, Wikipedia)
-get poorer.
+it can be foreseen that AI will drive this low-key interaction to a highly specialised level of knowledge.
+
+Another risk of the widespread use of AI tools is that community resources such as \eg Stack Overflow and Wikipedia become poorer because there are fewer incentives to maintain them.
 

### 2025-05-18 future_challenges_for_RSEng.tex
add our ideas what using AI tools in future might look like

@@
 
+Another risk of the widespread use of AI tools is that community resources such as \eg Stack Overflow and Wikipedia become poorer because there are fewer incentives to maintain them.
+
+The lessons learned from Human Factors research of the previous decades \cite{Simkute2025} applied to AI tools helped resolve the initial issues outlined above.
+We envision that AI will be transforming research processes by enabling rapid analysis of massive datasets, uncovering patterns and insights that would have been impossible or prohibitively time-consuming for humans alone to find.
+In research and development, AI will accelerate scientific breakthroughs in fields like healthcare, climate science, and materials engineering by automating complex modelling and predictive tasks.
+
 As AI continues to evolve, it will further blur the lines between human and machine roles, making research and software development faster.
@@
 it can be foreseen that AI will drive this low-key interaction to a highly specialised level of knowledge.
-
-Another risk of the widespread use of AI tools is that community resources such as \eg Stack Overflow and Wikipedia become poorer because there are fewer incentives to maintain them.
 

### 2025-05-19 future_challenges_for_RSEng.tex
add some more of the original text
rewrite story. Kyle gone.

@@
 In research and development, AI will accelerate scientific breakthroughs in fields like healthcare, climate science, and materials engineering by automating complex modelling and predictive tasks.
+Developers will be shifting from manual coding to orchestrating AI-driven workflows, focusing more on high-level design, integration, and creative problem-solving while AI handles repetitive or routine tasks.
 
@@
 10 years have passed since Kim and Kay started on their journey of being RSEs. Back then,
-at the universities of Eden and Utopia, life was more orderly and harmonious. Kay is now working on a project that develops
-software for cheap water monitoring devices in the Global South. Kim is currently designing the software service stack
-at the university, and increasingly struggles to find compelling Open Source alternatives that enables them keep complete control of their data
-rather than relying on software and tools that are backed by large corporate entities. So far she still has the support of the university board, since it values
-strategic autonomy.
-Her son Kyle on the other hand started out with a degree in Physics and then took one of the newly created RSE Master's courses,
-but quickly decided that work in academia with its time limited contracts is not attractive.
-Governments around the world are struggling to reposition themselves in this emerging de-globalised world with its new blocks
-and therefore have de-prioritised publicly funded research. Kyle decided to work for a defence company since they offer
-more secure employment in this polarised world.
+at the universities of Eden and Utopia, life was more orderly and harmonious.
+Kim is currently designing the software service stack at the university.
+She increasingly struggles to find compelling Open Source alternatives that enables them to keep complete control of their data
+rather than relying on software and tools that are backed by large, foreign corporate entities.
+So far, she still has the support of the university board, since it values strategic autonomy.
+Kay used to work on a project that developed
+software for cheap water monitoring devices in the Global South, but now funding has dried up.
+Having gone through another fixed term contract he is now fed up
+with work in academia and decides that its time limited contracts are not attractive.
+Now Governments around the world are struggling to reposition themselves in this emerging de-globalised world with its new blocks and therefore have de-prioritised publicly funded research.
+Kay sees the writings on the wall and decides to work for a defence company since they offer more secure employment in this polarised world.
 \end{story}

### 2025-05-19 future_challenges_for_RSEng.tex
one more notion
Update future_challenges_for_RSEng.tex
Co-authored-by: Gerasimos Chourdakis <gerasimos.chourdakis@ipvs.uni-stuttgart.de>

@@
 software for cheap water monitoring devices in the Global South, but now funding has dried up.
-Having gone through another fixed term contract he is now fed up
+Having gone through another fixed term contract, he is now fed up
 with work in academia and decides that its time limited contracts are not attractive.

### 2025-05-19 future_challenges_for_RSEng.tex
update story

--- 2025-05-19 future_challenges_for_RSEng.tex
@@
 \subsection{Impact of AI and automation}
+\label{sec:ai}
 \begin{whatis}{}{impact-ai-automation}
@@
 RSEs need to be aware of these issues and develop their own ``red lines''.
+With the advantages of \autoref{sec:ai} in place, ever-more AI created artifacts will be in the world.
+Therefore we expect a discussion within societies about their acceptable use to take place.
 
@@
 So far, she still has the support of the university board, since it values strategic autonomy.
-Kay used to work on a project that developed
-software for cheap water monitoring devices in the Global South, but now funding has dried up.
-Having gone through another fixed term contract he is now fed up
-with work in academia and decides that its time limited contracts are not attractive.
-Now Governments around the world are struggling to reposition themselves in this emerging de-globalised world with its new blocks and therefore have de-prioritised publicly funded research.
-Kay sees the writings on the wall and decides to work for a defence company since they offer more secure employment in this polarised world.
+Kay had left academia to work in an industrial R\&D department because of a more reliable employment.
+There he gained valuable experience working in an industrial setting.
+However, he was very happy when an opportunity arose to work in a central RSE department with an open-ended contract.
+Apart from his main responsibility of developing and maintaining software for an affordable water monitoring system used in the global south, he now also teaches undergraduates digital literacy.
 \end{story}

### 2025-05-19 future_challenges_for_RSEng.tex
Fix order of authors

@@
 Jeremy Cohen\authorRef{5},
-Jean-Noël Grad\authorRef{6},
-Magnus Hagdorn\authorRef{7},
-Toby Hodges\authorRef{8},
-Jan Linxweiler\authorRef{9},
-Frank Löffler\authorRef{10},
-Michele Martone\authorRef{11},
-Jan Philipp Thiele\authorRef{9}\textsuperscript{,}\authorRef{12},
-Nicholas A. Del Grosso\authorRef{13}
+Nicholas A. Del Grosso\authorRef{6},
+Jean-Noël Grad\authorRef{7},
+Magnus Hagdorn\authorRef{8},
+Toby Hodges\authorRef{9},
+Jan Linxweiler\authorRef{10},
+Frank Löffler\authorRef{11},
+Michele Martone\authorRef{12},\\
+Jan Philipp Thiele\authorRef{10}\textsuperscript{,}\authorRef{13}
 } % Authors and references to addresses
 \institute{% Institutes with labels
-\autlabel{1} Institut für theoretische Physik 1, University of Würzburg, 97074, Würzburg, Germany\par
-\autlabel{2} ZB MED Information Centre for Life Sciences, Cologne, Germany\par
-\autlabel{3} Institute for Parallel and Distributed Systems, University of Stuttgart, Stuttgart, Germany\par
-\autlabel{4} Leibniz University Hannover, Department of Cell Biology and Biophysics, Computational Biology, Germany\par
-\autlabel{5} Imperial College London, London, UK\par
-\autlabel{6} Institute for Computational Physics, University of Stuttgart, Germany\par
-\autlabel{7} Geschäftsbereich IT, Charité Universitätsmedizin Berlin, Germany\par
-\autlabel{8} The Carpentries, USA\par
-\autlabel{9} Technische Universität Braunschweig, Germany\par
-\autlabel{10} Michael Stifel Center Jena {\&} Friedrich Schiller University Jena, Germany\par
-\autlabel{11} Leibniz Supercomputing Centre, Garching, Germany\par
-\autlabel{12} Weierstrass Institute, Berlin, Germany;
-              Leibniz University Hannover, Institute of Applied Mathematics, Scientific Computing, Hannover, Germany\par
-\autlabel{13} Institute for Experimental Epileptology and Cognition Research, Uniklinikum Bonn, Germany
+\autlabel{1} Institut für theoretische Physik 1, University of Würzburg, 97074, Würzburg, Germany\\
+\autlabel{2} ZB MED Information Centre for Life Sciences, Cologne, Germany\\
+\autlabel{3} Institute for Parallel and Distributed Systems, University of Stuttgart, Stuttgart, Germany\\
+\autlabel{4} Leibniz University Hannover, Department of Cell Biology and Biophysics, Computational Biology, Germany\\
+\autlabel{5} Imperial College London, London, UK\\
+\autlabel{6} Institute for Experimental Epileptology and Cognition Research, Uniklinikum Bonn, Germany\\
+\autlabel{7} Institute for Computational Physics, University of Stuttgart, Germany\\
+\autlabel{8} Geschäftsbereich IT, Charité Universitätsmedizin Berlin, Germany\\
+\autlabel{9} The Carpentries, USA\\
+\autlabel{10} Technische Universität Braunschweig, Germany\\
+\autlabel{11} Michael Stifel Center Jena {\&} Friedrich Schiller University Jena, Germany\\
+\autlabel{12} Leibniz Supercomputing Centre, Garching, Germany\\
+\autlabel{13} Weierstrass Institute, Berlin, Germany;
+              Leibniz University Hannover, Institute of Applied Mathematics, Scientific Computing, Hannover, Germany\\
 }

### 2025-05-19 future_challenges_for_RSEng.tex
Language edits

@@
 
-So where do RSEs fit into this picture? Even today we expect to see RSEs providing technical skills and input
-to research activities in a range of different areas, these include:
+So where do RSEs fit into this picture? Even today, we expect to see RSEs providing technical skills and input
+to research activities in a range of different areas, including:
 high-quality, robust architecture and design for research software;
@@
 New RSEs are trained through a specialised Master's programme, and their support is valued by researchers.
-We still teach researchers basic software skills, but the content of this lecture has changed drastically.
+We still teach researchers basic software skills, but the content of this lecture has drastically changed.
 \end{whatis}
@@
 We can expect RSE positions to become more easily accessible, possibly through centrally-funded long-term positions at research institutions.
-The community of RSE's will get a significant boost in their acceptance if RSE Master's programmes
+The community of RSEs will get a significant boost in their acceptance if RSE Master's programmes
 become available at various universities.
@@
 The AI proactively suggests optimisations, generates reproducible workflows, and adapts to Kim’s coding style and domain-specific needs.
-With this partnership, Kim accelerates his research output while maintaining full control over scientific direction and ethical oversight.
+With this assistance, Kim accelerates his research output while maintaining full control over scientific direction and ethical oversight.
 \end{story}
@@
 But their relevance hinges on constant updates in light of how research is performed.
-By 2035 we want to have achieved accepted rules on the use of AI in research tasks and software development.
-Open Science and Open Research are well-established in academia, and supported by respective Open Research Software
+By 2035, we want to have achieved accepted rules on the use of AI in research tasks and software development.
+Open Science and Open Research are by then well-established in academia, and supported by respective Open Research Software
 strategies of the political actors. We also hope that Data Awareness and Data Protection are still values
@@
 Another example might be that they are developing code that can also be used for military purposes, such as numerical solvers or computer vision systems.
-Like engineers, RSEs take the responsibility for the software they develop.
+As engineers, RSEs take the responsibility for the software they develop.
 RSEs need to be aware of these issues and develop their own ``red lines''.
@@
 We expect research software to be a central part of scientific practice on the same level as data and publications.
-This will be supported by the first generations of RSE-Master graduates that find their way into research departments, and accelerate the pace of science.
+This will be supported by the first generations of RSE Master's graduates that find their way into research departments, and accelerate the pace of science.
 Similar to RSEs, that are widely known and understood by university boards in 2035, we expect the concept of RSE units to be in the minds
@@
 We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department to oversee specific projects \cite{Kempf2025-draft}.
-We also hope, that specialised clusters form that pool the skills of certain specialised RSEs, with certain RSE units specializing in certain tasks like data management, HPC, or visualisation.
+We also hope that specialised clusters form, which pool the skills of certain specialised RSEs with certain RSE units specializing in certain tasks like data management, HPC, or visualisation.
 At universities which have them, we want to see that researchers naturally accept them as part of their available services.
@@
 \end{whatis}
-Academia, while claiming to be innovative \cite{wzvgbayern}, has organisationally often adopted time-scales for change that are more reminiscent to institutions thinking on the time scale of generations.
+Academia, while claiming to be innovative \cite{wzvgbayern}, has organisationally often adopted time scales for change that are more reminiscent to institutions thinking on the time scale of generations.
 Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref} \cite{hiddenref} hints at the potential for specialisation of research tasks.
@@
 
-On the other hand the existence of people with a certain specialised skillset opens up possibilities for pooling.
+On the other hand, the existence of people with a certain specialised skill set opens up possibilities for pooling.
 Universities will benefit from pooling these specialists into RSE units, and offer their services to researchers.
@@
 permanent and well-funded RSE teams.
-Here it can only be strongly hoped, that, with the recognised importance of research software, research councils and funders
+Here it can only be strongly hoped that, with the recognised importance of research software, research councils and funders
 support the development with a reliable stream of funding.
@@
 \begin{story}{Changes to the Universities}{changes}
-Kim found work years ago at one of the first RSE units in Arcadia. After a couple of years he moved to the university of
+Kim found a job years ago at one of the first RSE units in Arcadia. After a couple of years, Kim moved to the university of
 Avalon which was interested in setting up their own department for RSEs in order to not fall behind the university of Arcadia.
@@
 \emph{New challenges}
-If we follow the ideas laid out in \cite{Kempf2025-draft}, then in addition to university wide RSE units,
+If we follow the ideas laid out in related community work~\cite{Kempf2025-draft}, then in addition to university wide RSE units,
 there will be embedded RSEs in the departments. Shaping and developing these network bonds will be a task for the future.
@@
 This, however, comes with additional funding challenges,
-as long-term funding can be more difficult to acquire in lower tiers of an of a research organisation (i.e. at departmental or individual research group level).
+as long-term funding can be more difficult to acquire in lower tiers of an of a research organisation (i.e., at departmental or individual research group level).
 Third-party funding can mitigate this issue in the mid-term.
@@
 due to the fact that there are already examples and activities from various universities in this direction
-(examples include the universities of Jena, Heidelberg and Göttingen, as well as several universities in the UK where the concept of dedicated RSE groups began to emerge around 2013).
+(examples include the universities of Jena, Heidelberg, and Göttingen, as well as several universities in the UK where the concept of dedicated RSE groups began to emerge around 2013).
 We expect funding in central groups to evolve from initially temporary positions to permanent positions,
@@
 His unique blend of physics knowledge and hands-on problem solving earned him a new title among his peers: the first Quantum RTP.
-In a world where theory meets reality, Ken quietly ensures the future of quantum computing stays on track.
+In a world where theory meets reality, Ken quietly ensures that the future of quantum computing stays on track.
 \end{story}
@@
 But this digitalisation will also lead to changes in the way societies interact and therefore will translate into changes in how academia perceives itself.
-We have also not delved into the interaction with industry much.
+We have also not sufficiently delved into the interaction with industry.
 There are also some trends that we believe will stay with us over the coming years.
 The differentiation of job profiles in academia, that has started with profiles such as the RSE will continue.
-Societies will be facing new questions, that arise from technologies such as generally available AI systems.
+Societies will be facing new questions, arising from technologies such as generally available AI systems.
 

### 2025-05-19 future_challenges_for_RSEng.tex
Add missing space

@@
 the expected transformations in academic institutions,
-and the potential impact of alternative career paths for RSEs and new groups of RTPs(\emph{Research Technical Professional}s).
+and the potential impact of alternative career paths for RSEs and new groups of RTPs (\emph{Research Technical Professional}s).
 %

### 2025-05-19 future_challenges_for_RSEng.tex
Proof-reading edits by GC (#37)
* Fix order of authors

* Language edits

* Add missing space

* Minor language edits in ethics

@@
 RSEs need to be aware of these issues and develop their own ``red lines''.
-With the advantages of \autoref{sec:ai} in place, ever-more AI created artifacts will be in the world.
-Therefore we expect a discussion within societies about their acceptable use to take place.
+With the advantages of \autoref{sec:ai}, ever-more AI created artifacts will be included in research.
+As a result of this, we expect a discussion within societies about their acceptable use.
 

### 2025-05-19 future_challenges_for_RSEng.tex
Add references to previous work

@@
 with increasing integration into various aspects of research.
+In Germany, a formal call for creating a sustainable environment for research software
+is only a couple of years behind~\cite{Anzt2021},
+while the foundational competencies of a research software engineer have only recently been defined in detail~\cite{Goth2024}.
+Central RSE units are in the process of being imagined~\cite{Kempf2025-draft},
+while an RSE Master's is in the early stages of design. 
 In this opinion paper, we try to imagine the state of our field

### 2025-05-19 future_challenges_for_RSEng.tex
Introduction: Replace reference with footnote, split paragraphs

@@
 while an RSE Master's is in the early stages of design.
+
 In this opinion paper, we try to imagine the state of our field
@@
 We present here the results of a workshop discussion at the
-deRSE25 conference in Karlsruhe\cite{Goth2025EndRSEng}, without
+deRSE25 conference in Karlsruhe\footnote{deRSE25, session ``The End of {RSEng}? Challenges and Risks for {RSEng}'': \url{https://events.hifis.net/event/1741/contributions/14026/}}, without
 claiming to represent the entire deRSE community.
@@
 It is even more challenging to structure the aspects that will become relevant for the RSE community in the future.
-Nevertheless, in order to structure the discussion, we decided to consider
+
+In order to structure the discussion, we decided to consider
 the evolution of the term RSE and its definition,

### 2025-05-19 future_challenges_for_RSEng.tex
Shorten section 2.6
closes #28

--- 2025-05-19 future_challenges_for_RSEng.tex
@@
 \begin{whatis}{}{academic}
-We expect research software to be a central part of scientific practice on the same level as data and publications.
+We expect research software to be a central part of scientific practice, on the same level as data and publications.
 This will be supported by the first generations of RSE Master's graduates that find their way into research departments, and accelerate the pace of science.
-Similar to RSEs, that are widely known and understood by university boards in 2035, we expect the concept of RSE units to be in the minds
+Similar to RSEs, who are widely known and understood by university boards in 2035, we expect the concept of RSE units to be in the minds
 of academic staff at any level.
-We hope to see that universities use the presence of an RSE Department actively as a means to differentiate themselves from other universities.
-We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department to oversee specific projects \cite{Kempf2025-draft}.
+We hope to see that universities actively use the presence of an RSE Department as a means to differentiate themselves from other universities.
+We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects \cite{Kempf2025-draft}.
 We also hope that specialised clusters form, which pool the skills of certain specialised RSEs with certain RSE units specializing in certain tasks like data management, HPC, or visualisation.
-At universities which have them, we want to see that researchers naturally accept them as part of their available services.
-Additionally researchers will seamlessly interact with infrastructures for using research software.
+At universities which get such RSE units, we want to see that researchers naturally accept them as part of their available services.
+Additionally, researchers will seamlessly interact with infrastructures for using research software.
 \end{whatis}
@@
 This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.
-We have already seen some specialisation trends in recent years such as
-the HPC-RSE and AI-RSE which are quite recent developments, whose appeal is stimulated by large public and private investments in HPC facilities and AI technologies.
-But we could imagine new RSE specialisations emerging in the fields
-of quantum computing, such as the Quantum-RSE, and in federated online platforms for research.
+We have already seen some specialisation trends in recent years, such as
+the HPC-RSE and AI-RSE, which are quite recent developments, and whose appeal is stimulated by large public and private investments in HPC facilities and AI technologies.
+We foresee new RSE specialisations emerging, e.g., in the fields
+of quantum computing, such as the Quantum-RSE, or in federated online platforms for research.
 % European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories
@@
 
-On the other hand, the existence of people with a certain specialised skill set opens up possibilities for pooling.
-Universities will benefit from pooling these specialists into RSE units, and offer their services to researchers.
-Hopefully, these RSE-units are endowed with enough trust by the university boards that translates into stable funding.
+At the same time, acquiring people with a certain specialised skill set opens up possibilities for pooling.
+Universities will benefit from pooling these specialists into RSE units and offering their services to researchers in a structured way.
+This compartmentalisation will help researchers innovate in their research, with the support of sustainable software made by or with the help of RSEs.
+These RSE units should be endowed with enough trust by the university boards, translated into stable funding.
 With longer-term funding and more trust to pick up longer term projects,
-some RSEs can focus on code modernisation, clean-up, and modularisation,
-which can pay off in faster and easier development in the future.
-This is an opportunity that would not be possible if relying on short-term funded solo RSEs.
+some RSEs could focus on code modernisation, clean-up, and modularisation,
+which could pay off in faster and easier development in the future.
+This is an opportunity that would not be possible if relying solely on short-term funded, isolated RSEs.
 
-\emph{Which challenges will still persist}\\
-The workshop participants expect that there will still be budgeting and administrative hurdles to creating
-permanent and well-funded RSE teams.
-Here it can only be strongly hoped that, with the recognised importance of research software, research councils and funders
-support the development with a reliable stream of funding.
-This would also require a change of culture in research about the acceptance of research software.\\
-While we hope that the barrier to researchers diminishes, the workshop participants
-still expect some work to do. We hope that the compartmentalisation that occurs will help researchers re-prioritise their workload.
-An upcoming challenge will be that, with the emergence of more dedicated RSE units,
-networking between them will come into the foreground, because there should be a strong need to share best practices
-among institutions with successful RSE Teams.
-On a more organisational level, we expect the sharing of models on how to organise different RSEs with different areas of specialist expertise, to fit the different organisational constraints.
-After that, organisations will have to come up with updated models for hiring people.
-While a standardised RSE Master's degree makes it easier to develop the next generation of practitioners with the necessary technical skills, it will be interesting to see if
-RSE are still often hired by word of mouth within an organisation(``I have this PhD here, before their PostDoc they would be a good fit for the RSE department...'' ) or if the RSE Master's degree ensures a hiring process on a more standardised level.
+Besides all the improvements, the workshop participants still expect some challenges to persist,
+and new challenges to emerge.
+Due to the rise of dedicated RSE units, networking people and teams, federating infrastructures, as well as organizing and sharing knowledge will become important.
+Structuring these new units and joining their administrative support will also help the units to grow.
+A particular example is hiring:
+While a standardised RSE Master's degree will make it easier to develop the next generation of RSEs,
+organizations could develop standardized hiring processes to acquire the skill set that they actually need.
+Budgeting and administrative hurdles will also probably still exist and hinder the formation of permanent and well-funded RSE teams.
+However, we expect the growing (perceived) importance of research software to convince research councils and funders to provide reliable streams of funding.
 
@@
 \end{story}
-
-\emph{New challenges}
-If we follow the ideas laid out in related community work~\cite{Kempf2025-draft}, then in addition to university wide RSE units,
-there will be embedded RSEs in the departments. Shaping and developing these network bonds will be a task for the future.
-While networking embedded RSEs and RSE units is essential for building robust, sustainable research software infrastructure, the question arises whether every group truly needs to host and manage its own infrastructure.
-When each team or institution hosts its own systems, this often leads to siloed environments where resources, expertise, and data are isolated, limiting collaboration and creating inefficiencies.
-Such silos can result in duplicated efforts, higher costs, and challenges in maintaining data integrity and governance, as each group independently manages similar infrastructure and workflows.
-While local control offers flexibility, it also increases complexity and makes it harder to share innovations or best practices across the wider research community.
-To overcome these challenges, federated infrastructure models are needed, enabling shared access, interoperability, and coordinated development while still allowing for some local autonomy \cite{Jelinek2021}.
-Federated approaches break down silos, promote knowledge exchange, and ensure that infrastructure investments benefit the broader ecosystem rather than just isolated groups.
 

### 2025-05-19 future_challenges_for_RSEng.tex
Shorten section 2.6 (#39)
* Shorten section 2.6

Co-authored-by: Florian Goth <florian.goth@uni-wuerzburg.de>

@@
 the HPC-RSE and AI-RSE, which are quite recent developments, and whose appeal is stimulated by large public and private investments in HPC facilities and AI technologies.
-We foresee new RSE specialisations emerging, e.g., in the fields
+We foresee new RSE specialisations emerging, \eg, in the fields
 of quantum computing, such as the Quantum-RSE, or in federated online platforms for research.
@@
 This is an opportunity that would not be possible if relying solely on short-term funded, isolated RSEs.
+We hope that a change of culture in research about the acceptance of research software has happened.
 

### 2025-05-19 future_challenges_for_RSEng.tex
Extend and structure the introduction (#38)
* Add references to previous work

* Introduction: Replace reference with footnote, split paragraphs

@@
 with increasing integration into various aspects of research.
+In Germany, a formal call for creating a sustainable environment for research software
+is only a couple of years behind~\cite{Anzt2021},
+while the foundational competencies of a research software engineer have only recently been defined in detail~\cite{Goth2024}.
+Central RSE units are in the process of being imagined~\cite{Kempf2025-draft},
+while an RSE Master's is in the early stages of design.
+
 In this opinion paper, we try to imagine the state of our field

### 2025-05-19 future_challenges_for_RSEng.tex
Edit section 2.3 to improve pagination

@@
 Generative AI tools, such as GitHub Copilot \cite{Friedmann2021}, are being added to software development tools.
-These AI powered code assistants promise to increase productivity by automatically generating code, and writing test cases and documentation.
-Software developers use them to reduce the number of keystrokes, complete tasks and to recall syntax without having to consult the documentation \cite{Liang2024}.
-Code assistants have the potential to help novice or even non-expert programmers write code \cite{Feldman2024} thus enabling non-experts to participate in research and software creation, and fostering greater collaboration across disciplines.
+These AI-powered code assistants promise increased productivity by automatically generating code, tests, and documentation.
+Software developers use them to reduce the number of keystrokes, draft tasks, and to recall syntax without having to consult the documentation \cite{Liang2024}.
+Code assistants can help novice or even non-expert programmers write code \cite{Feldman2024}, thus enabling non-experts to participate in research and software creation, and fostering greater collaboration across disciplines.
 However, when these tools are used in practice, many users report increased cognitive load and frustration stemming from the difficulty to prompt the system effectively and from debugging the generated code \cite{Simkute2025}.
@@
 
-The lessons learned from Human Factors research of the previous decades \cite{Simkute2025} applied to AI tools helped resolve the initial issues outlined above.
+The lessons learned from Human Factors research of the previous decades \cite{Simkute2025}, applied to AI tools, helped resolve the initial issues outlined above.
 We envision that AI will be transforming research processes by enabling rapid analysis of massive datasets, uncovering patterns and insights that would have been impossible or prohibitively time-consuming for humans alone to find.
@@
 
-As AI continues to evolve, it will further blur the lines between human and machine roles, making research and software development faster.
-With the general availability of AI, the interaction of RSEs with researchers will change at a fundamental level.
+As AI continues to evolve, it will further blur the lines between human and machine roles.
+With the general availability of AI, the interaction of RSEs with researchers will fundamentally change.
 Whereas previously RSEs have been points of knowledge that could translate human language formulated problems into technical descriptions and then provide solutions to them,
-it can be foreseen that AI will drive this low-key interaction to a highly specialised level of knowledge.
+we foresee that AI will drive this low-key interaction to a highly specialised level of knowledge.
 
@@
 
+\vspace{-4ex}
 \subsection{Impact of Complexity and Usability in Research Software}

### 2025-07-18 correspondence/rebuttal-1.md
Add review and rebuttal template

# Rebuttal

Manuscript: "Future challenges for Research Software Engineering"

| Reviewer | Recommendation      | Relevance | Significance | Technical soundness | Presentation |
| ---      | ---                 | ---       | ---          | ---                 | ---          |
| A        | Resubmit for review | very good | good         | very good           | good         |
| B        | Decline submission  | good      | excellent    | fair                | poor         |
| C        | Revisions required  | excellent | very good    | very good           | good         |
| D        | Revisions required  | excellent | very good    | good                | very good    |

## Reviewer A

### Summary

> The authors provide a report and discussion on future trends in Research Software Engineering, mainly focused on Germany, based on a workshop discussion during the deRSE25 conference. They provide a discussion on 7 topics addressing possible trends, goals that will be achieved and stories of relatable personas as well as an outlook about the role and standing of RSEs in the academic system.

### Relevance

> The paper is written as opinion paper, not as research paper. The topic fits perfectly to the special issue. I suppose the audience interest is mainly focused to the German scientific community. The topics provided in the paper are timely and contribute to an ongoing global discussion. The paper provides a perspective and does not state to provide a global analysis.

### Significance

> The paper adds to a discussion about trends in Research Software Engineering. The discussion is an important aspect as reflection for current challenges. The authors state that the topics are not exhaustive and the content of the topic discussions provide some perspective, however do not allow for an in-depth assessment/analysis.

### Technical soundness

> This criterion is difficult to assess as the paper is an opinion paper. There is no transcript of the workshop the authors base their statements on. Thus, an assessment on the soundness is impossible. The authors include relevant topics that will need attention in the near future.

### Quality of presentation

> Especially the changing tenses make the paper a difficult read. The paper is well structured in 4 sections. The second section as main section includes goal statements and persona stories for each topic, respectively. This makes the topics easier to understand and is thus well structured. The formatting and spelling needs some attention. Regarding readability, I provide suggestions below.

### Suggestions

> 1. The authors do not mention Germany as the focus of their discussion, probably because they want to address a broader aspect of RSEng that could apply to other countries or even globally. However, the history, focus and references clearly indicate a German perspective. I suggest either making it clear that the vision and discussion are limited to Germany, or discussing why the focus of the paper is rather nationalistic yet applies to a broader range of locations, and explaining where and why the stated aspects differ.

> 2. The authors switch between the past and present tenses, which makes reading the article difficult. They sometimes address present tools, infrastructure and activities, and sometimes write as if future tools, infrastructure and activities have been made available or developed. For example, the "Achieved Goals" refer to goals achieved by 2035, but the tenses change throughout the boxes. "RSEng-relevant open educational resources..." Some resources are available and some are not; however, the style of writing makes it difficult to tell the difference. I would propose that the authors use tenses to clearly distinguish between past, present and future activities, tools and resources, and not write as if future activities have already been successful. Additionally, the authors write about their hopes, which seem off-topic for a report or opinion paper. I would suggest not writing about hopes at all. Future trend statements should be based on knowledge.

> 3. The authors claim to "present here the results of a workshop discussion at the deRSE25 conference in Karlsruhe". However, they also include their hopes, discussions and own expectations for the future RSE ecosystem (in Germany). This should be mentioned in the introduction and it should be made clear which statements have been put forward during the discussion and which have been added by the authors after the discussion. The authors mention this construct in section 3 ("In the previous sections, we have tried to list some of the trends that we foresee. The material presented here is based on the outputs of a workshop, but extended with the authors’ ideas."), which makes understanding the paper difficult; I would expect such a statement at the beginning. While I do not propose to re-structure the paper (e.g. report first, discussion second), I suggest to review the paper as to which statements have been reported in the discussion and which statement have been brought up by the authors.

> Regarding Chapter 2.3.
> I would propose the authors provide a counter argument/discussion to why AI will not totally make RSEs obsolete. Additionally, the authors paint a quite bright picture of the AI future, however not discussing risks to the scientific processes, e.g. difficulty with recplication, societal trust in science, non-democtratization of science because of access to storage, compute and useful data sets, etc.

> Regarding Chaper 2.6
> "thinking on the time scale of generations". This seems crucial, especially when looking at academia in Germany. Many renowened professors will retire in the next 5-10 years, making way for a younger generation that even may have coded during their career. The authors do not discuss this obvious transformation. Do they no expect at least some RSE professors?

> Minor remarks:
> Author 12 lacks affiliation information

> Keywords could be richer

> "In order to structure the discussion, we decided to consider...". Was the discussion structured as presented or is the paper structured as presented; the sentence implies the former. Is there a rationale why the authors include the following topics? Have there been other topics in discussion that have not been namend in the paper?

> 2. Chapter
> This chapter includes the authors vision of the RSE space in 2035.

> "accepted research artefact on an equal footing with data and publications." If software or data is publicized would these artefacts not also be regarded as publications? I would suggest using "data and text" or "data and text publications" instead.

> "The specialist skills that RSEs provide take a significant amount of time to develop and maintain. Therefore, we do not expect domain researchers to also become experts in a range of advanced RSEng topics." Could the authors name some of these advanced topics here? Or maybe change the position of this paragraph to after the paragraph about the domain researchers.

> Story Box 1:
> "What issues have newly emerged, where they are now just happy to say: .." This seems

> 2.1 Chapter
>
> "might remain unchanged, there might be a shift in the responsibilities of RSEs, in the form of a partial reallocation of the training budget to the other core competencies. " What is the rationale for this statement?

> "Looking at that on a broader scale..." What is meant by "that"? What do the authors look at?

> Chapter 2.2
> "visibility and accelerate the professionalisationof the RSE role" -> missing line spacing

> Story Box 4
> "Kim accelerates his research.." To my understanding, Kim and Kay are female.

> Chapter 2.5
> "As engineers, RSEs take the responsibility for the software they develop. RSEs need to be aware of these issues and develop their own “red lines”." This statement underestimates the complexity of how code/knowledge/algorithms can be reused for military purposes. A scientist would need to know about present military technologies which is unrealistic. I fear the call for responsibility is too much to ask from an individual. Additionally, it creates a huge burden if code or a programme has been reused for military purposes. I would suggest that the authors make clear the boundaries of the responsibility.

> Story Box 6:
> "that are backed by large, foreign corporate entities. So far, she still has the support of the university board, since it values strategic autonomy." This citation shows that the authors seem to write from a German point of view. In the previous sections, international collaboration was highlighted. In a global world, there is no "foreign corporate entity", just collaboration tools from different sources. This is a minor comment here but important for the general remark regarding the perspective given in the paper.

> Chaper 2.7
> "in lower tiers of an of a research organisation", sentence needs attention

> Story Box 8:
> "His unique blend of physics knowledge and hands-on problem solving earned him a new title among his peers: the first Quantum RTP. In a world where theory meets reality, Ken quietly ensures that the future of quantum
> computing stays on track" Is this still a position within the research staff or within the technical staff? The authors seem to imply that HR departements will hire RTPs as research staff. This issue should be discussed.

> Conclusion paragraph. Formatting needs attention.

> Line numbering would make referencing easier

> The reference style in the document looks odd as there is no line spacing between the word and the bracket: ...competencies[WORD]...

\newpage

## Reviewer B

### Summary

> The paper lists a set of expected developments in the future of RSE.

### Relevance

> The paper provides an overview over the challenges and future developments / scenarios
discussed by the conference participants.

### Significance

> The paper overall is very generic and more related to Software Engineering in general than specific to Research Software. Similar papers on Software Engineering issues already exist and provide more details on the actual challenges, expected development and which technologies and approaches may prove relevant. In most cases, this paper is leaving out concrete issues and already existing developments (see e.g. Intellisoft).

### Technical soundness

> The paper is a bit too generic for an in-depth assessment of its technical soundness. Specifically, many of the "core issues" are mentioned but never listed (with exception of "core software skills" - which ever these may be). As noted under significance, some existing developments in Software Engineering have not been properly respected (AI tooling, standardisation etc.) - which all contributed to the simplicity that allows easy generation of software. RSE has specific additional challenges related to soundness, replicability, reproducibility, interpretability etc. which are not addressed in the paper.

### Quality of presentation

> The paper chooses a very interesting approach to present the key challenges / aspects, with a short description of the aspect leading as a mid- to long-term goal and a "futuristic story" as an exemplary scenario. Most of these aspects relate however to communities recognizing the relevance of RSE much more than what the relevance is, and the futuristic scenarios sometimes do not seem to fit the topic of the goal, such as in the context of Ethics / Social Consideration; few of the scenarios provide additional information that would further the understanding of the challenges or goals.

### Suggestions

> I like the idea and the elaboration of the paper, but fail to see relevant contributions in the individual sections, in particular comes to differentiation from Software Engineering. With that respect, related work in Software Engineering already addresses a lot of aspects mentioned in the paper and should be referenced and acknowledged appropriately as it shapes expectations and future work. I'd love to see a better elaboration of the paper, as again it is a relevant and interesting approach.

\newpage

## Reviewer C

### Summary

> A useful overview of outcomes of a valuable workshop, augmented by the author's own opinions, on the future of RSE.

### Relevance

> This paper reflects on a workshop held at de-RSE

### Significance

> This paper addresses many important themes in this field, and makes useful contributions on how they could be advanced.

### Technical soundness

> This is an opinion paper, based on synthesis of the workshop discussions and the authors' knowledge.

### Quality of presentation

> The use of boxes is a great idea, but doesn't quite work. See the detailed notes for ideas on how to improve this.

### Suggestions

> It might be useful to note that this paper on RSE in 2030 exists https://arxiv.org/pdf/2308.07796, although the workshop participants didn’t know about it so it wasn’t incorporated in their thinking.

> Reviewer C also uploaded a file with more feedback

\newpage

## Reviewer D

### Summary

> This paper is the result of a workshop discussion at the deRSE25 (the 5th conference for Research Software Engineering in Germany, in a workshop session titled "The End of RSEng? Challenges and Risks for RSEng"), and includes views and opinions of participants condensed in a well-structured and readable format by the authors.
>
> RSEs (Research Software Engineers) are broadly defined by their core task - "enhancing the scientific output of researchers" - while it is acknowledged that the means by which they currently achieve that end ('RSEng', or 'Research Software Engineering') is likely to evolve towards very different approaches and skillsets in the future, due to the obsoleting and enabling effects of new technologies.
>
> The paper mentions the future introduction of a formalised Masters programme for training RSEs as being an important step in the ensuring the continued professionalisation and acceptance of the RSE role in research institutions and beyond, bringing significant efficiency benefits to the research process. It also reiterates the importance of the recognition of research software as an output of a research-focused career, to be taken into consideration for career progression.
>
> The paper also highlights what roles for RSEs are likely to stay (mostly) the same, vs change: e.g. educating researchers around the creation and use of software will likely still be a feature, although there will be an increasing emphasis on effective and ethical use of AI-based tools in and through the development of software in research. The availability of AI-based tools is anticipated to increase the scope of what RSEs can achieve, while changing the manner and focus of their work.
>
> Despite the ominous phrasing of the source workshop's title ("The End of RSEng?"), the tone of the paper is for the most part optimistic and encouraging of RSEs in and entering the field, and of the value of RSEs within research institutions and beyond.

### Relevance

> Given ECEASST has a focus on software science and technology (including all aspects related to the systematic and rigorous engineering of software and systems), and is intended for the publication of conference proceedings to disseminate current thinking in this space, this paper is highly relevant; it will be of interest to readers at the intersection of software engineering and research in a range of fields.

### Significance

> The views presented in this paper are significant in that they capture recent rapid changes in this landscape (particularly the impact of generative AI), as well as in including timely projections of where this field (and the career trajectories of practitioners) could and should go in the future.

### Technical soundness

> This paper presents the results of a workshop discussion in the form of an opinion piece, but does not appear to give any information about the participants in the workshop, which would have been interesting and might have been a useful framing for this output (in highlighting the areas of expertise and concern of those involved, which would likely have informed and steered any discussions). However, while it would have been of interest to have some details of the participants, it might have presented difficulties around privacy and the flow of the discussion, and may not have added much additional weight to the workshop output, beyond what can be already surmised of the participants based on the title of the conference and the subject of the workshop. The material as presented is still of value without additional information about the participants.

### Quality of presentation

> There are a few cases of less-than-typical use of the English language which are a little distracting, although they do not detract from the message; e.g.:
>
> "In Germany, a formal call for creating a sustainable environment for research software is only a couple of years behind" (behind what?); and
>
> "While future projections are common in industry [Gar25], so far there is a projection lacking for the RSEng space" (slightly stilted and non-standard English in "a projection lacking").
>
> These cases are minor and debatable, and as such do not merit additional editing.

> The use of 'Achieved Goals' boxes to highlight positive future outcomes for the field, and 'Story' boxes to bring the potential changes in the field to a human level with a focus on fictional personas, are nice touches that enhance the readability and impact of this piece.

> Another very minor point: the gender pronouns of the two personas ('Kim' and 'Kay') appear to swap at times; e.g. in Story 3 'Kay' is 'her' and in Story 4 'Kim' is 'him', while in Story 6 'Kay' is 'he' and 'Kim' is 'she')

### Suggestions

> If possible, the inclusion of a small table, figure, and/or a few lines of text summarising the distribution of key features of the workshop participants might be an interesting addition (e.g. level of academic/research education, an indication of their software engineering proficiency/professionalism, their age and/or career stage)... I would be curious as to the level of 'optimism' and expectations for change for the field that participants with different backgrounds might tend to have, although I feel that your paper has merit and interest as it stands, with no such changes needed.

> I am selecting 'Revisions Required' in case you would like to make such an addition (and/or address the other very minor points I made about some cases of slightly non-standard English, and the swapping of genders of your personas), but do not think revisions are strictly necessary; the paper could be published as-is.

### 2025-07-30 future_challenges_for_RSEng.tex
forgot to upload things

--- 2025-05-19 future_challenges_for_RSEng.tex
@@
 \usepackage{color}
+\usepackage{orcidlink}
 \usepackage[dvipsnames]{xcolor}
@@
 \autlabel{11} Michael Stifel Center Jena {\&} Friedrich Schiller University Jena, Germany\\
-\autlabel{12} Leibniz Supercomputing Centre, Garching, Germany\\
+\autlabel{12} orcid: 0000-0003-3239-8554 \orcidlink{0000-0003-3239-8554}\\
 \autlabel{13} Weierstrass Institute, Berlin, Germany;
@@
 and community in the mid-term future, focusing on future challenges.
-While future projections are common in industry \cite{gartner}, so far there is a projection lacking for the RSEng space.
+While future projections are common in industry~\cite{gartner}, so far there is a projection lacking for the RSEng space.
 We present here the results of a workshop discussion at the
@@
 and to advance the conversation about how Research Software Engineering should, or could, develop.
-It is already difficult to structure the spectrum of RSE tasks and responsibilities today given the spectrum of research software \cite{hasselbring2024}.
+It is already difficult to structure the spectrum of RSE tasks and responsibilities today given the spectrum of research software~\cite{hasselbring2024}.
 It is even more challenging to structure the aspects that will become relevant for the RSE community in the future.
@@
 Not too overwhelmingly far ahead in the future, but distant enough for some changes to have occurred.
-Over this time frame Kim\cite{Anzt2021} and Kay\cite{Goth2024},
+Over this time frame Kim~\cite{Anzt2021} and Kay~\cite{Goth2024},
 our fictional RSEs who we have used to illustrate specific scenarios in our earlier work,
@@
 
-While the RSE core competencies\cite{Goth2024} might remain unchanged,
+While the RSE core competencies~\cite{Goth2024} might remain unchanged,
 there might be a shift in the responsibilities of RSEs,
@@
 The Covid-19 pandemic acted as a catalyst for the digital transformation
-of higher education\cite{Bygstad2022}. Students and early-career researchers
+of higher education~\cite{Bygstad2022}. Students and early-career researchers
 are now more familiar with digital learning tools, which lead to a shift
@@
 and to act more as a facilitator of resources,
-and to monitor activities and results over time.''\cite{Bygstad2022}
+and to monitor activities and results over time.''~\cite{Bygstad2022}
 
@@
 \end{whatis}
-Generative AI tools, such as GitHub Copilot \cite{Friedmann2021}, are being added to software development tools.
+Generative AI tools, such as GitHub Copilot~\cite{Friedmann2021}, are being added to software development tools.
 These AI powered code assistants promise to increase productivity by automatically generating code, and writing test cases and documentation.
-Software developers use them to reduce the number of keystrokes, complete tasks and to recall syntax without having to consult the documentation \cite{Liang2024}.
-Code assistants have the potential to help novice or even non-expert programmers write code \cite{Feldman2024} thus enabling non-experts to participate in research and software creation, and fostering greater collaboration across disciplines.
-However, when these tools are used in practice, many users report increased cognitive load and frustration stemming from the difficulty to prompt the system effectively and from debugging the generated code \cite{Simkute2025}.
-The potential of AI to be a democratising technology can only be realised when the algorithmic divide along with the digital divide is addressed \cite{Yu2020}.
+Software developers use them to reduce the number of keystrokes, complete tasks and to recall syntax without having to consult the documentation~\cite{Liang2024}.
+Code assistants have the potential to help novice or even non-expert programmers write code~\cite{Feldman2024} thus enabling non-experts to participate in research and software creation, and fostering greater collaboration across disciplines.
+However, when these tools are used in practice, many users report increased cognitive load and frustration stemming from the difficulty to prompt the system effectively and from debugging the generated code~\cite{Simkute2025}.
+The potential of AI to be a democratising technology can only be realised when the algorithmic divide along with the digital divide is addressed~\cite{Yu2020}.
 
@@
 
-The lessons learned from Human Factors research of the previous decades \cite{Simkute2025} applied to AI tools helped resolve the initial issues outlined above.
+The lessons learned from Human Factors research of the previous decades~\cite{Simkute2025} applied to AI tools helped resolve the initial issues outlined above.
 We envision that AI will be transforming research processes by enabling rapid analysis of massive datasets, uncovering patterns and insights that would have been impossible or prohibitively time-consuming for humans alone to find.
@@
 \begin{whatis}{}{ethics}
-The principles of Good Scientific Practice \cite{dfg_gsp} have been a cornerstone in the conduct of research.
+The principles of Good Scientific Practice~\cite{dfg_gsp} have been a cornerstone in the conduct of research.
 But their relevance hinges on constant updates in light of how research is performed.
@@
 The ethics and values of the Research Software Engineer are grounded in their practice.
-They commit to the health, safety and welfare of the public and act in the interest of society, their employer and their clients \cite{Goth2024}.
+They commit to the health, safety and welfare of the public and act in the interest of society, their employer and their clients~\cite{Goth2024}.
 
@@
 The work being undertaken by RSEs may well have direct, or indirect,
-ethical implications in areas such as social engineering \cite{s2erc,siadati2024},
-climate \cite{Lannelongue2023} and more.
+ethical implications in areas such as social engineering~\cite{s2erc,siadati2024},
+climate~\cite{Lannelongue2023} and more.
 Another example might be that they are developing code that can also be used for military purposes, such as numerical solvers or computer vision systems.
@@
 % data ethics
-Similarly to how data collected by scientists should adhere to the FAIR principles \cite{FAIR}, research software developed by RSEs should adhere to the FAIR4RS principles~\cite{FAIR4RS}.
+Similarly to how data collected by scientists should adhere to the FAIR principles~\cite{FAIR}, research software developed by RSEs should adhere to the FAIR4RS principles~\cite{FAIR4RS}.
 
@@
 We hope to see that universities actively use the presence of an RSE Department as a means to differentiate themselves from other universities.
-We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects \cite{Kempf2025-draft}.
+We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
 We also hope that specialised clusters form, which pool the skills of certain specialised RSEs with certain RSE units specializing in certain tasks like data management, HPC, or visualisation.
@@
 \end{whatis}
-Academia, while claiming to be innovative \cite{wzvgbayern}, has organisationally often adopted time scales for change that are more reminiscent to institutions thinking on the time scale of generations.
-Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref} \cite{hiddenref} hints at the potential for specialisation of research tasks.
+Academia, while claiming to be innovative~\cite{wzvgbayern}, has organisationally often adopted time scales for change that are more reminiscent to institutions thinking on the time scale of generations.
+Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref} hints at the potential for specialisation of research tasks.
 This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.
@@
 RSEs can either be embedded in research groups (belonging to the mid-level university staff),
-or be members of larger RSE units.\cite{Kempf2025-draft}
+or be members of larger RSE units~\cite{Kempf2025-draft}.
 The participants had the opinion that focusing on embedded RSEs would make onboarding new RSEs to existing projects easier.

### 2025-07-30 future_challenges_for_RSEng.tex
Cleanup and bibliography update

@@
 We present here the results of a workshop discussion at the
-deRSE25 conference in Karlsruhe\footnote{deRSE25, session ``The End of {RSEng}? Challenges and Risks for {RSEng}'': \url{https://events.hifis.net/event/1741/contributions/14026/}}, without
-claiming to represent the entire deRSE community.
+deRSE25 conference in Karlsruhe\footnote{deRSE25, session ``The End of {RSEng}?
+Challenges and Risks for {RSEng}'': \url{https://events.hifis.net/event/1741/contributions/14026/}},
+without claiming to represent the entire deRSE community.
 Nevertheless, by disseminating the discussed ideas,
@@
 and to advance the conversation about how Research Software Engineering should, or could, develop.
-It is already difficult to structure the spectrum of RSE tasks and responsibilities today given the spectrum of research software~\cite{hasselbring2024}.
+It is already difficult to structure the spectrum of RSE tasks and responsibilities today given the spectrum of research software~\cite{Hasselbring2024}.
 It is even more challenging to structure the aspects that will become relevant for the RSE community in the future.
@@
 This, however, comes with additional funding challenges,
-as long-term funding can be more difficult to acquire in lower tiers of an of a research organisation (i.e., at departmental or individual research group level).
+as long-term funding can be more difficult to acquire in lower tiers of a research organisation (i.e., at departmental or individual research group level).
 Third-party funding can mitigate this issue in the mid-term.
@@
 We have tried to shed some light on the social, technical and organisational aspects.
-In summary, RSEs will very likely be doing different work
-
-to what they are doing today, but their core task of enhancing the scientific output of researchers -- other human
+In summary, RSEs will very likely be doing different work to what they are doing today,
+but their core task of enhancing the scientific output of researchers -- other human
 beings -- will stay the same.

### 2025-08-01 future_challenges_for_RSEng.tex
Break long lines, sort BibTeX records

@@
 % Your \usepackage commands go here.
+\usepackage[T1]{fontenc}
 \usepackage{fancybox,framed}
@@
 and community in the mid-term future, focusing on future challenges.
-While future projections are common in industry~\cite{gartner}, so far there is a projection lacking for the RSEng space.
+While future projections are common in industry~\cite{Gartner2025},
+so far there is a projection lacking for the RSEng space.
 We present here the results of a workshop discussion at the
@@
 and to advance the conversation about how Research Software Engineering should, or could, develop.
-It is already difficult to structure the spectrum of RSE tasks and responsibilities today given the spectrum of research software~\cite{Hasselbring2024}.
+
+It is already difficult to structure the spectrum of RSE tasks and responsibilities
+today given the spectrum of research software~\cite{Hasselbring2024}.
 It is even more challenging to structure the aspects that will become relevant for the RSE community in the future.
-
 In order to structure the discussion, we decided to consider
@@
 the expected transformations in academic institutions,
-and the potential impact of alternative career paths for RSEs and new groups of RTPs (\emph{Research Technical Professional}s).
-%
+and the potential impact of alternative career paths for RSEs
+and new groups of RTPs (\emph{Research Technical Professional}s).
 We conclude the paper with a summary and with our wishes and recommendations.
@@
 Therefore, it is vital to improve digital skills of all researchers.
-This will also allow RSEs to focus on other more advanced and specialist
-areas. In this context, we do not foresee the demise of the RSE as an independent role. However, we
-do see the work of RSEs developing and shifting as the research community changes.
+This will also allow RSEs to focus on other more advanced and specialist areas.
+In this context, we do not foresee the demise of the RSE as an independent role.
+However, we do see the work of RSEs developing and shifting as the research community changes.
 
@@
 reducing reliance on RSEs to train newcomers on the most basic tasks.
-This means, that today's workshop content of proper versioning and data management could become obsolete
-by using more intuitive tools.
+This means, that today's workshop content of proper versioning
+and data management could become obsolete by using more intuitive tools.
 Introductory courses can then focus on more higher-level concepts.
@@
 \begin{story}{The Definition of RSEs}{defrse}
-Kim and Kay chuckled when they looked at their original job descriptions from 2020: "build tools for scientists, manage data, write reusable code."
+Kim and Kay chuckled when they looked at their original job descriptions from 2020: ``build tools for scientists, manage data, write reusable code.''
 Now, their roles are unrecognizable: half AI whisperer, half research ethicist, and part-time diplomat negotiating between synthetic and human collaborators.
@@
 \begin{whatis}{}{development-rse-community}
-National RSE communities are increasingly collaborating with local partners, fostering stronger networks across countries.
-Regular international conferences further amplify these connections, promoting the global exchange of ideas.
-The introduction of RSE Master's programmes, coupled with a shift in culture towards valuing Research Software, has significantly elevated its profile.
-This growing recognition has led to a surge in membership of national societies, reflecting the increasing importance of RSEng in the research landscape.
+National RSE communities are increasingly collaborating with local partners,
+fostering stronger networks across countries.
+Regular international conferences further amplify these connections,
+promoting the global exchange of ideas. The introduction of RSE Master's programmes,
+coupled with a shift in culture towards valuing Research Software,
+has significantly elevated its profile.
+This growing recognition has led to a surge in membership of national societies,
+reflecting the increasing importance of RSEng in the research landscape.
 \end{whatis}
@@
 We expect these initiatives to help increase the visibility and accelerate
-the professionalisation%
-of the RSE role.
+the professionalisation of the RSE role.
 In addition, a multitude of RSE communities self-organised at the national
 and regional level to support more viable RSE career paths in academia.
-We can expect RSE positions to become more easily accessible, possibly through centrally-funded long-term positions at research institutions.
-The community of RSEs will get a significant boost in their acceptance if RSE Master's programmes
-become available at various universities.
+We can expect RSE positions to become more easily accessible,
+possibly through centrally-funded long-term positions at research institutions.
+The community of RSEs will get a significant boost in their acceptance
+if RSE Master's programmes become available at various universities.
 With increased numbers, the visibility of the RSE community to industry will be enhanced.
 \begin{story}{Development of RSE community}{RSEcommunity}
-Kay stands proudly at the podium of the first International Research Software Engineering Conference held in the Southern Hemisphere, a historic moment for both her career and the global RSE community.
-Having moved to South Africa, she speaks passionately about the unique challenges researchers in the Global South face: limited access to high-performance computing, unreliable funding, and the persistent digital divide.
-Her keynote highlights how local innovation thrives despite constraints, often out of necessity rather than choice.
-The audience, a diverse mix of global experts, listens intently as she calls for equitable collaborations, not just inclusion as an afterthought.
+Kay stands proudly at the podium of the first International Research
+Software Engineering Conference held in the Southern Hemisphere,
+a historic moment for both her career and the global RSE community.
+Having moved to South Africa, she speaks passionately about the unique
+challenges researchers in the Global South face: limited access
+to high-performance computing, unreliable funding,
+and the persistent digital divide.
+Her keynote highlights how local innovation thrives despite constraints,
+often out of necessity rather than choice.
+The audience, a diverse mix of global experts, listens intently as she calls
+for equitable collaborations, not just inclusion as an afterthought.
 \end{story}
@@
 \end{whatis}
-Generative AI tools, such as GitHub Copilot~\cite{Friedmann2021}, are being added to software development tools.
-These AI powered code assistants promise to increase productivity by automatically generating code, and writing test cases and documentation.
-Software developers use them to reduce the number of keystrokes, complete tasks and to recall syntax without having to consult the documentation~\cite{Liang2024}.
-Code assistants have the potential to help novice or even non-expert programmers write code~\cite{Feldman2024} thus enabling non-experts to participate in research and software creation, and fostering greater collaboration across disciplines.
-However, when these tools are used in practice, many users report increased cognitive load and frustration stemming from the difficulty to prompt the system effectively and from debugging the generated code~\cite{Simkute2025}.
-The potential of AI to be a democratising technology can only be realised when the algorithmic divide along with the digital divide is addressed~\cite{Yu2020}.
+Generative AI tools, such as GitHub Copilot~\cite{Friedmann2021},
+are being added to software development tools.
+These AI powered code assistants promise to increase productivity
+by automatically generating code, and writing test cases and documentation.
+Software developers use them to reduce the number of keystrokes,
+complete tasks and to recall syntax without having to consult the documentation~\cite{Liang2024}.
+Code assistants have the potential to help novice or even non-expert programmers
+write code~\cite{Feldman2024} thus enabling non-experts to participate in research
+and software creation, and fostering greater collaboration across disciplines.
+However, when these tools are used in practice, many users report increased cognitive
+load and frustration stemming from the difficulty to prompt the system effectively
+and from debugging the generated code~\cite{Simkute2025}.
+The potential of AI to be a democratising technology can only be realised
+when the algorithmic divide along with the digital divide is addressed~\cite{Yu2020}.
 
-Another risk of the widespread use of AI tools is that community resources such as \eg Stack Overflow and Wikipedia become poorer because there are fewer incentives to maintain them.
+Another risk of the widespread use of AI tools is that community resources
+such as \eg\ Stack Overflow and Wikipedia become poorer because there are fewer incentives to maintain them.
 
-The lessons learned from Human Factors research of the previous decades~\cite{Simkute2025} applied to AI tools helped resolve the initial issues outlined above.
-We envision that AI will be transforming research processes by enabling rapid analysis of massive datasets, uncovering patterns and insights that would have been impossible or prohibitively time-consuming for humans alone to find.
-In research and development, AI will accelerate scientific breakthroughs in fields like healthcare, climate science, and materials engineering by automating complex modelling and predictive tasks.
-Developers will be shifting from manual coding to orchestrating AI-driven workflows, focusing more on high-level design, integration, and creative problem-solving while AI handles repetitive or routine tasks.
+The lessons learned from Human Factors research of the previous decades~\cite{Simkute2025}
+applied to AI tools helped resolve the initial issues outlined above.
+We envision that AI will be transforming research processes by enabling rapid analysis of massive datasets,
+uncovering patterns and insights that would have been impossible or prohibitively time-consuming for humans alone to find.
+In research and development, AI will accelerate scientific breakthroughs in fields like healthcare,
+climate science, and materials engineering by automating complex modelling and predictive tasks.
+Developers will be shifting from manual coding to orchestrating AI-driven workflows,
+focusing more on high-level design, integration, and creative problem-solving while AI handles repetitive or routine tasks.
 
@@
 \begin{story}{The interaction of RSEs with AI}{RSEs_with_AI}
-Kim develops advanced scientific software in tandem with an AI agent that codes, debugs, and simulates experiments alongside him.
-The AI proactively suggests optimisations, generates reproducible workflows, and adapts to Kim’s coding style and domain-specific needs.
+Kim develops advanced scientific software in tandem with an AI agent that codes,
+debugs, and simulates experiments alongside him. The AI proactively suggests optimisations,
+generates reproducible workflows, and adapts to Kim’s coding style and domain-specific needs.
 With this assistance, Kim accelerates his research output while maintaining full control over scientific direction and ethical oversight.
@@
 \begin{story}{The benefits of abstraction}{abstraction}
-Years ago, Kim had to distribute the software library libkim as a source code package with a custom build system and detailed instructions on how to build it on some common systems,
+Years ago, Kim had to distribute the software library libkim as a source code package
+with a custom build system and detailed instructions on how to build it on some common systems,
 investing significant time in developing and maintaining this auxiliary toolkit and documentation.
@@
 % institutional bonds and ethics
-Sometimes these values may conflict with the aims of their peers or institution. For example, their PI might insist on quickly developing some software without adhering to current best practices.
+Sometimes these values may conflict with the aims of their peers or institution.
+For example, their PI might insist on quickly developing some software without adhering to current best practices.
 A code of ethics and professional conduct for RSEs should be developed by institutions together with RSEs themselves.
@@
 The work being undertaken by RSEs may well have direct, or indirect,
-ethical implications in areas such as social engineering~\cite{s2erc,siadati2024},
+ethical implications in areas such as social engineering~\cite{s2erc,Siadati2024},
 climate~\cite{Lannelongue2023} and more.
-Another example might be that they are developing code that can also be used for military purposes, such as numerical solvers or computer vision systems.
+Another example might be that they are developing code that can also be used
+for military purposes, such as numerical solvers or computer vision systems.
 As engineers, RSEs take the responsibility for the software they develop.
@@
 % data ethics
-Similarly to how data collected by scientists should adhere to the FAIR principles~\cite{FAIR}, research software developed by RSEs should adhere to the FAIR4RS principles~\cite{FAIR4RS}.
+Similarly to how data collected by scientists should adhere to the FAIR principles~\cite{FAIR},
+research software developed by RSEs should adhere to the FAIR4RS principles~\cite{FAIR4RS}.
 
@@
 There he gained valuable experience working in an industrial setting.
-However, he was very happy when an opportunity arose to work in a central RSE department with an open-ended contract.
-Apart from his main responsibility of developing and maintaining software for an affordable water monitoring system used in the global south, he now also teaches undergraduates digital literacy.
+However, he was very happy when an opportunity arose to work in a central
+RSE department with an open-ended contract. Apart from his main responsibility
+of developing and maintaining software for an affordable water monitoring system
+used in the global south, he now also teaches undergraduates digital literacy.
 \end{story}
@@
 \end{whatis}
-Academia, while claiming to be innovative~\cite{wzvgbayern}, has organisationally often adopted time scales for change that are more reminiscent to institutions thinking on the time scale of generations.
-Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref} hints at the potential for specialisation of research tasks.
+Academia, while claiming to be innovative~\cite{wzvgbayern},
+has organisationally often adopted time scales for change that are more reminiscent
+to institutions thinking on the time scale of generations.
+Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref}
+hints at the potential for specialisation of research tasks.
 This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.
-We have already seen some specialisation trends in recent years, such as
-the HPC-RSE and AI-RSE, which are quite recent developments, and whose appeal is stimulated by large public and private investments in HPC facilities and AI technologies.
-We foresee new RSE specialisations emerging, \eg, in the fields
-of quantum computing, such as the Quantum-RSE, or in federated online platforms for research.
+We have already seen some specialisation trends in recent years,
+such as the HPC-RSE and AI-RSE, which are quite recent developments,
+and whose appeal is stimulated by large public and private investments in HPC facilities and AI technologies.
+We foresee new RSE specialisations emerging, \eg, in the fields of quantum computing,
+such as the Quantum-RSE, or in federated online platforms for research.
 % European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories
@@
 
-At the same time, acquiring people with a certain specialised skill set opens up possibilities for pooling.
-Universities will benefit from pooling these specialists into RSE units and offering their services to researchers in a structured way.
-This compartmentalisation will help researchers innovate in their research, with the support of sustainable software made by or with the help of RSEs.
+At the same time, acquiring people with a certain specialised skill set
+opens up possibilities for pooling.
+Universities will benefit from pooling these specialists into RSE units
+and offering their services to researchers in a structured way.
+This compartmentalisation will help researchers innovate in their research,
+with the support of sustainable software made by or with the help of RSEs.
 These RSE units should be endowed with enough trust by the university boards, translated into stable funding.
@@
 and new challenges to emerge.
-Due to the rise of dedicated RSE units, networking people and teams, federating infrastructures, as well as organizing and sharing knowledge will become important.
+Due to the rise of dedicated RSE units, networking people and teams,
+federating infrastructures, as well as organizing and sharing knowledge will become important.
 Structuring these new units and joining their administrative support will also help the units to grow.
@@
 Budgeting and administrative hurdles will also probably still exist and hinder the formation of permanent and well-funded RSE teams.
-However, we expect the growing (perceived) importance of research software to convince research councils and funders to provide reliable streams of funding.
+However, we expect the growing (perceived) importance of research software
+to convince research councils and funders to provide reliable streams of funding.
 
@@
 We would like to see the emergence of more dedicated staff, i.e. other (digital) RTPs next to the RSE.
-Also we hope that researchers can build careers at the interface of academia, industry, and policy (\eg in think tanks, NGOs,
+Also we hope that researchers can build careers at the interface of academia, industry, and policy (\eg\ in think tanks, NGOs,
 or organisations that do knowledge transfer).
@@
 This, however, comes with additional funding challenges,
-as long-term funding can be more difficult to acquire in lower tiers of a research organisation (i.e., at departmental or individual research group level).
+as long-term funding can be more difficult to acquire in lower tiers
+of a research organisation (i.e., at departmental or individual research group level).
 Third-party funding can mitigate this issue in the mid-term.
@@
 due to the fact that there are already examples and activities from various universities in this direction
-(examples include the universities of Jena, Heidelberg, and Göttingen, as well as several universities in the UK where the concept of dedicated RSE groups began to emerge around 2013).
+(examples include the universities of Jena, Heidelberg, and Göttingen, as well as several universities
+in the UK where the concept of dedicated RSE groups began to emerge around 2013).
 We expect funding in central groups to evolve from initially temporary positions to permanent positions,
@@
 \begin{story}{New career paths}{New career paths}
-Ken, though trained as a physicist, found his passion not in theory, but in keeping the quantum computers humming.
-He became the go-to expert for troubleshooting entangled qubits and stabilizing delicate cryogenic systems.
-His unique blend of physics knowledge and hands-on problem solving earned him a new title among his peers: the first Quantum RTP.
-In a world where theory meets reality, Ken quietly ensures that the future of quantum computing stays on track.
+Ken, though trained as a physicist, found his passion not in theory,
+but in keeping the quantum computers humming. He became the go-to expert
+for troubleshooting entangled qubits and stabilizing delicate cryogenic systems.
+His unique blend of physics knowledge and hands-on problem solving earned him
+a new title among his peers: the first Quantum RTP.
+In a world where theory meets reality, Ken quietly ensures that the future
+of quantum computing stays on track.
 \end{story}
@@
 Among that is the ongoing digital transformation in wider society.
-Digital competencies in the general population continue to increase, making the jump to software development easier than before.
-But this digitalisation will also lead to changes in the way societies interact and therefore will translate into changes in how academia perceives itself.
+Digital competencies in the general population continue to increase,
+making the jump to software development easier than before.
+But this digitalisation will also lead to changes in the way societies
+interact and therefore will translate into changes in how academia perceives itself.
 We have also not sufficiently delved into the interaction with industry.
@@
 
-We hope that the standard of research software will have risen across the board, enabling easier reuse of prior software outputs in new research.
-We also hope that a cultural change is taking place to recognize research software as an integral part of science that it is
-widely recognised and rewarded across academia and is supported by political actors.
+We hope that the standard of research software will have risen across the board,
+enabling easier reuse of prior software outputs in new research.
+We also hope that a cultural change is taking place to recognize research software
+as an integral part of science that it is widely recognised and rewarded across
+academia and is supported by political actors.
 \section{Conclusion}

### 2025-08-04 future_challenges_for_RSEng.tex
Add perspectives from the SE community

@@
 While future projections are common in industry~\cite{Gartner2025},
-so far there is a projection lacking for the RSEng space.
+so far there is little projection for the RSEng space.
 We present here the results of a workshop discussion at the
@@
 and new groups of RTPs (\emph{Research Technical Professional}s).
+This work supplements other opinion pieces developed in parallel
+in the SE community~\cite{ChueHong2025,Sochat2024Infra,Bencomo2024AEBoK,OliveiraJr2024,Druskat2025},
+which where not known to the workshop participants; remarkably,
+many of the projections made by the participants overlap with predictions made
+by the SE community in ``Research Software Engineering in 2030''~\cite{Katz2023}.
 We conclude the paper with a summary and with our wishes and recommendations.
@@
 In this context, we do not foresee the demise of the RSE as an independent role.
-However, we do see the work of RSEs developing and shifting as the research community changes.
+However, we do see the work of RSEs developing and shifting as the research community changes~\cite{ChueHong2025}.
 
@@
 We can expect RSE positions to become more easily accessible,
-possibly through centrally-funded long-term positions at research institutions.
+possibly through centrally-funded long-term positions at research institutions,
+such as in the bwRSE4HPC service\footnote{\url{https://www.bwrse4hpc.de/68.php}}.
 The community of RSEs will get a significant boost in their acceptance
@@
 and the future of RSE education should aim to prepare students for both roles.
+We note members of the SE community independently reached the same conclusions
+and advocate for the creation of an `Abstraction Engineering Body of Knowledge'
+(AEBoK)\cite{Bencomo2024AEBoK}.
 
@@
 research software developed by RSEs should adhere to the FAIR4RS principles~\cite{FAIR4RS}.
+Members of the SE community have also called for their field to embrace Open Science best practices\cite{OliveiraJr2024,Druskat2025}.
 
@@
 We foresee new RSE specialisations emerging, \eg, in the fields of quantum computing,
-such as the Quantum-RSE, or in federated online platforms for research.
+such as the Quantum-RSE, or in digital research infrastructure.
+We note the SE community has also independently expressed interest in creating an \emph{infrastructure engineering} role~\cite{Sochat2024Infra}
+and funding bodies seem willing to promote such career pathways~\cite{UKRI2025Infra,EuroHPC2024Federated},
 % European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories

### 2025-08-05 correspondence/rebuttal-1.md
Add rebuttal w.r.t. to citation spacing
Add zenodo ref, add text, and write replies.

@@
 
+As suggested also by other referees, we made the workshop notes available at zenodo. A link has been made part of the paper.
+
 > 1.
@@
 report first, discussion second), I suggest to review the paper as to which statements have been reported in the discussion and which statement have been brought up by the authors.
+
+We have put this sentence to the front, an kept it at the end as a reminder.
 
@@
 > The reference style in the document looks odd as there is no line spacing between the word and the bracket: ...competencies[WORD]...
+
+We added non-breaking spaces in front of every citation.
 
@@
 
+Thank you for your suggestion.
+We made the workshop notes available at zenodo and included the link into the paper.
+We definitely agree, that more statisitics about the participants would be valuable, but we didn't measure them....
+We added a respective statement to the introduction.
+
 > I am selecting 'Revisions Required' in case you would like to make such an addition (and/or address the other very minor points I made about some cases of slightly non-standard English, and the swapping of genders of your personas), but do not think revisions are strictly necessary; the paper could be published as-is.

### 2025-08-05 future_challenges_for_RSEng.tex
Add zenodo ref, add text, and write replies.

--- 2025-08-01 future_challenges_for_RSEng.tex
@@
 so far there is a projection lacking for the RSEng space.
-We present here the results of a workshop discussion at the
+We present here the results of a workshop discussion (notes content is archived here \cite{goth2025_WSPad}) at the
 deRSE25 conference in Karlsruhe\footnote{deRSE25, session ``The End of {RSEng}?
@@
 without claiming to represent the entire deRSE community.
+The material presented here is based on the output of the workshop participants, but extended with the authors' ideas.
+Since deRSE25 was co-located with SE25, the annual conference for the Software Engineering community in Germany, the discussion was also open to software engineers,
+but a precise statistic of the workshop participants background was not taken.
+The workshop hosts felt, that the majority of participants were from the RSE community.
 Nevertheless, by disseminating the discussed ideas,
@@
 In the previous sections, we have tried to list some of the trends that we foresee.
-The material presented here is based on the outputs of a workshop, but extended with the authors' ideas.
+The material presented here is based on the outputs of a workshop, and extended with the authors' ideas.
 Nevertheless, there are some things that we declared out of scope here.

### 2025-08-07 future_challenges_for_RSEng.tex

--- 2025-08-01 future_challenges_for_RSEng.tex
@@
 \newcommand{\eg}{e.\,g.}
+
+% PDF accessibility and PDF metadata
+\usepackage{accsupp}
+\newcommand{\authorRef}[1]{\texorpdfstring{\autref{#1}}{}}
+\newcommand{\authorOrcid}[1]{\texorpdfstring{\thinspace\orcidlink{#1}\thinspace}{}}
 \newcommand{\geprislink}[1]{\href{https://gepris.dfg.de/gepris/projekt/#1?language=en}{#1}}
@@
 %\short{} % Short title of the article (optional)
-\newcommand{\authorRef}[1]{\texorpdfstring{\autref{#1}}{}}
 \author{
-Florian Goth\authorRef{1},
-Leyla Jael Castro\authorRef{2},
-Gerasimos Chourdakis\authorRef{3},
-Simon Christ\authorRef{4},
-Jeremy Cohen\authorRef{5},
-Nicholas A. Del Grosso\authorRef{6},
-Jean-Noël Grad\authorRef{7},
-Magnus Hagdorn\authorRef{8},
-Toby Hodges\authorRef{9},
-Jan Linxweiler\authorRef{10},
-Frank Löffler\authorRef{11},
-Michele Martone\authorRef{12},\\
-Jan Philipp Thiele\authorRef{10}\textsuperscript{,}\authorRef{13}
+Florian Goth\authorOrcid{0000-0003-2707-4790}\authorRef{1},
+Leyla Jael Castro\authorOrcid{0000-0003-3986-0510}\authorRef{2},
+Gerasimos Chourdakis\authorOrcid{0000-0002-3977-1385}\authorRef{3},
+\texorpdfstring{\\}{} % line break
+Simon Christ\authorOrcid{0000-0002-5866-1472}\authorRef{4},
+Jeremy Cohen\authorOrcid{0000-0003-4312-2537}\authorRef{5},
+Nicholas A. Del Grosso\thinspace\authorRef{6},
+\texorpdfstring{\\}{} % line break
+Jean-Noël Grad\authorOrcid{0000-0002-5821-4912}\authorRef{7},
+Magnus Hagdorn\authorOrcid{0000-0002-5076-4864}\authorRef{8},
+Toby Hodges\authorOrcid{0000-0003-1766-456X}\authorRef{9},
+Jan Linxweiler\authorOrcid{0000-0002-2755-5087}\authorRef{10},
+\texorpdfstring{\\}{} % line break
+Frank Löffler\authorOrcid{0000-0001-6643-6323}\authorRef{11},
+Michele Martone\authorOrcid{0000-0003-3239-8554}\authorRef{12},
+Jan Philipp Thiele\authorOrcid{0000-0002-8901-6660}\authorRef{10}\textsuperscript{,}\authorRef{13}
 } % Authors and references to addresses
@@
 \autlabel{11} Michael Stifel Center Jena {\&} Friedrich Schiller University Jena, Germany\\
-\autlabel{12} orcid: 0000-0003-3239-8554 \orcidlink{0000-0003-3239-8554}\\
+\autlabel{12} Independent researcher, Germany\\
 \autlabel{13} Weierstrass Institute, Berlin, Germany;

### 2025-08-11 correspondence/rebuttal-1.md
Update correspondence/rebuttal-1.md
Co-authored-by: Gerasimos Chourdakis <gerasimos.chourdakis@ipvs.uni-stuttgart.de>

--- 2025-07-18 correspondence/rebuttal-1.md
@@
 report first, discussion second), I suggest to review the paper as to which statements have been reported in the discussion and which statement have been brought up by the authors.
+
+We have put this sentence to the front, and kept it at the end as a reminder.
 
@@
 
+Thank you for your suggestion.
+We made the workshop notes available on Zenodo and included the link into the paper.
+We definitely agree that more statisitics about the participants would be valuable, but we did not collect any such data.
+We also added a respective statement to the introduction.
+
 > I am selecting 'Revisions Required' in case you would like to make such an addition (and/or address the other very minor points I made about some cases of slightly non-standard English, and the swapping of genders of your personas), but do not think revisions are strictly necessary; the paper could be published as-is.

### 2025-08-11 future_challenges_for_RSEng.tex
Update future_challenges_for_RSEng.tex
Co-authored-by: Gerasimos Chourdakis <gerasimos.chourdakis@ipvs.uni-stuttgart.de>

--- 2025-08-05 future_challenges_for_RSEng.tex
@@
 Since deRSE25 was co-located with SE25, the annual conference for the Software Engineering community in Germany, the discussion was also open to software engineers,
-but a precise statistic of the workshop participants background was not taken.
+but no list or profile of the participants is available.
 The workshop hosts felt, that the majority of participants were from the RSE community.

### 2025-08-11 future_challenges_for_RSEng.tex

--- 2025-08-07 future_challenges_for_RSEng.tex
@@
 so far there is little projection for the RSEng space.
-We present here the results of a workshop discussion at the
+We present here the results of a workshop discussion (notes content is archived here \cite{goth2025_WSPad}) at the
 deRSE25 conference in Karlsruhe\footnote{deRSE25, session ``The End of {RSEng}?
@@
 without claiming to represent the entire deRSE community.
+The material presented here is based on the output of the workshop participants, but extended with the authors' ideas.
+Since deRSE25 was co-located with SE25, the annual conference for the Software Engineering community in Germany, the discussion was also open to software engineers,
+but no list or profile of the participants is available.
+The workshop hosts felt, that the majority of participants were from the RSE community.
 Nevertheless, by disseminating the discussed ideas,

### 2025-08-11 correspondence/rebuttal-1.md
Mention wider risks from AI

--- 2025-07-18 correspondence/rebuttal-1.md
@@
 
+We have extended the AI section to mention further risks from AI.
+While the original manuscript already mentioned issues with AI and a shift of focus in the RSE role,
+we have made these stronger in an additional paragraph.
+
 > Regarding Chaper 2.6
@@
 > Author 12 lacks affiliation information
+
+Author 12 wished to be listed as an independent researcher, which is now reflected in the affiliation. ORCID links have also been added to the other authors.
 

### 2025-08-11 future_challenges_for_RSEng.tex
Mention wider risks from AI

--- 2025-08-07 future_challenges_for_RSEng.tex
@@
 
-Another risk of the widespread use of AI tools is that community resources
-such as \eg\ Stack Overflow and Wikipedia become poorer because there are fewer incentives to maintain them.
+Besides the direct impact in research software and further research output,
+we see a number of risks.
+The workshop participants suggested that the emergence of AI and automation might reduce critical thinking and imagination.
+We address this in \authorRef{sec:complexity}.
+Another risk is that the widespread use of AI tools will make community resources
+such as \eg\ Stack Overflow and Wikipedia (which are used to train AI models) poorer, due to fewer incentives to maintain them.
+Further risks that were later discussed include an increased replication crisis, a de-democratization of research, and the loss of trust in science.
+Due to the non-deterministic nature of generative AI, we expect data generation and analysis to become less reproducible.
+Professional RSEs could safeguard reproducibility by designing, evaluating, and monitoring appropriate workflows.
+Future research will require significant computing and storage resources, as well as useful data and models,
+therefore it will be more driven by institutions able to provide these.
+Finally, an increasing risk is the loss of trust in science,
+in particular due to the automation of the peer-reviewing process~\cite{Naddaf2025}.
+As more and more research will be written by researchers assisted by AI,
+RSEs will spend more of their time in designing and reviewing software written by other humans and AI tools,
+rather than repeatedly crafting similar individual elements.
 
@@
 \subsection{Impact of Complexity and Usability in Research Software}
+\label{sec:complexity}
 \begin{whatis}{}{abstraction}

### 2025-08-11 future_challenges_for_RSEng.tex
Fix LaTeX
Proof-reading

@@
 The workshop participants suggested that the emergence of AI and automation might reduce critical thinking and imagination.
-We address this in \authorRef{sec:complexity}.
+We address this in \autoref{sec:complexity}.
 Another risk is that the widespread use of AI tools will make community resources
 such as \eg\ Stack Overflow and Wikipedia (which are used to train AI models) poorer, due to fewer incentives to maintain them.
-Further risks that were later discussed include an increased replication crisis, a de-democratization of research, and the loss of trust in science.
+Further risks that were later discussed include an intensified replication crisis, a de-democratization of research, and loss of trust in science.
 Due to the non-deterministic nature of generative AI, we expect data generation and analysis to become less reproducible.
@@
 As more and more research will be written by researchers assisted by AI,
-RSEs will spend more of their time in designing and reviewing software written by other humans and AI tools,
-rather than repeatedly crafting similar individual elements.
+RSEs will be freed from repeatedly crafting similar individual elements, but will be spending more of their time in designing and reviewing software written by other humans and AI tools.
 

### 2025-08-11 future_challenges_for_RSEng.tex
Add concluding sentence

@@
 RSEs will be freed from repeatedly crafting similar individual elements, but will be spending more of their time in designing and reviewing software written by other humans and AI tools.
+To address these risks, the RSE of the future could be not only closely overlooking every part of the research software development cycle (ensuring dissemination, archiving, and reproducibility),
+but also contributing to endangered community resources and infrastructure that will keep research accessible to the wider scientific community.
 

### 2025-08-11 future_challenges_for_RSEng.tex
Add pointers to AI4SE literature

@@
 librarians\footnote{\url{https://librarycarpentry.org}},
-and soon HPC practitioners\footnote{\url{https://www.hpc-carpentry.org}}.
+and soon High-Performance Computing (HPC) practitioners\footnote{\url{https://www.hpc-carpentry.org}}.
 
@@
 Generative AI tools, such as GitHub Copilot~\cite{Friedmann2021},
-are being added to software development tools.
-These AI powered code assistants promise to increase productivity
+are being added to software development tools~\cite{Alenezi2025}.
+These AI-powered code assistants promise to increase productivity
 by automatically generating code, and writing test cases and documentation.
@@
 Further risks that were later discussed include an intensified replication crisis, a de-democratization of research, and loss of trust in science.
-Due to the non-deterministic nature of generative AI, we expect data generation and analysis to become less reproducible.
+Due to the non-deterministic nature of generative AI, we expect AI-assisted simulations and and data analysis to become less reproducible.
+This is not an entirely new issue, and similar concerns exist in the HPC community,
+where reproducing simulation data and data analysis workflows presents tough challenges~\cite{Antunes2024}.
 Professional RSEs could safeguard reproducibility by designing, evaluating, and monitoring appropriate workflows.
+When applicable, standardized and version-controlled models should be preferred
+over self-developed models, and self-developed models should be treated like any
+other research output (i.e. publish the trained model, training data and training workflow).
 Future research will require significant computing and storage resources, as well as useful data and models,
@@
 Finally, an increasing risk is the loss of trust in science,
-in particular due to the automation of the peer-reviewing process~\cite{Naddaf2025}.
-As more and more research will be written by researchers assisted by AI,
-RSEs will be freed from repeatedly crafting similar individual elements, but will be spending more of their time in designing and reviewing software written by other humans and AI tools.
-To address these risks, the RSE of the future could be not only closely overlooking every part of the research software development cycle (ensuring dissemination, archiving, and reproducibility),
-but also contributing to endangered community resources and infrastructure that will keep research accessible to the wider scientific community.
+in particular due to the automation of the peer-reviewing process~\cite{Naddaf2025}
+and the malicious use of hidden prompts~\cite{Gibney2025}.
+As researchers increase their reliance on AI assistants,
+RSEs will be freed from repeatedly crafting similar individual elements,
+but will be spending more of their time in designing and reviewing software written by other humans and AI tools.
+To address these risks, the RSE of the future could be not only closely overlooking every part
+of the research software development cycle (ensuring dissemination, archiving, and reproducibility),
+but also contributing to endangered community resources and infrastructure that
+will keep research accessible to the wider scientific community.
 
@@
 With the general availability of AI, the interaction of RSEs with researchers will change at a fundamental level.
-Whereas previously RSEs have been points of knowledge that could translate human language formulated problems into technical descriptions and then provide solutions to them,
+Whereas previously RSEs have been points of knowledge that could translate
+human language formulated problems into technical descriptions and then provide solutions to them,
 it can be foreseen that AI will drive this low-key interaction to a highly specialised level of knowledge.
+RSEs will still act as points of knowledge by assisting users in integrating
+AI tools in their development workflows, understanding the reasoning of AI models,
+and more generally, by improving AI literacy~\cite{Alenezi2025}.
 

### 2025-08-11 future_challenges_for_RSEng.tex
Add references from SE
Co-authored-by: Jeremy Cohen <jcohen02@users.noreply.github.com>

@@
 would expect any researcher who writes code to have in 10 years time. 
-The rapid development of AI means that researchers are likely to have access to tooling that can assist
-them in writing robust, sustainable code, lowering the barrier to developing code, or at least changing the
-required skill set to some extent.
+The rapid development of artificial intelligence (AI) and large language models
+(LLMs) means that researchers are likely to have access to tooling that can assist
+them in writing robust, sustainable code, lowering the barrier to developing code,
+or at least changing the required skill set to some extent.
 They can also be expected to have an understanding of frameworks to support testing, packaging and
@@
 \begin{whatis}{}{impact-ai-automation}
-RSEs understand how AI models work, and can help researchers reap their benefits, while
-clearly knowing their limits.
+RSEs understand how AI models work and how to tune and extend them.
+With these skills they can help researchers reap their benefits,
+while clearly knowing their limits.
 \end{whatis}
@@
 Software developers use them to reduce the number of keystrokes,
-complete tasks and to recall syntax without having to consult the documentation~\cite{Liang2024}.
+focus on higher-level design rather than on idiosyncrasies,
+and recall syntax without having to consult the documentation~\cite{Liang2024}.
 Code assistants have the potential to help novice or even non-expert programmers
@@
 and from debugging the generated code~\cite{Simkute2025}.
+Many factors can influence AI tools adoption~\cite{Russo2024}.
 The potential of AI to be a democratising technology can only be realised
 when the algorithmic divide along with the digital divide is addressed~\cite{Yu2020}.
+We also note members of the SE community have called for improving LLM literacy
+by adapting SE curricula~\cite{Kirova2024}.
 
@@
 We address this in \autoref{sec:complexity}.
-Another risk is that the widespread use of AI tools will make community resources
-such as \eg\ Stack Overflow and Wikipedia (which are used to train AI models) poorer, due to fewer incentives to maintain them.
+Another risk is that the widespread use of AI tools could make community resources
+such as \eg\ Stack Overflow and Wikipedia (which are used to train AI models) poorer
+due to fewer incentives to maintain them, and could compromise LLMs via `model collapse'
+when AI-generated content is added to these websites and becomes part of the training data~\cite{Shumailov2024}.
 Further risks that were later discussed include an intensified replication crisis, a de-democratization of research, and loss of trust in science.

### 2025-08-11 future_challenges_for_RSEng.tex
typo

--- 2025-08-11 future_challenges_for_RSEng.tex
@@
 optimisation of codes for use on high-performance computing infrastructure;
-and green computing - efficient implementations of algorithms/methods and use of hardware
+and green computing - efficient implementations of algorithms/methods and use of hardware.
 
@@
 
-Besides the direct impact in research software and further research output,
-we see a number of risks.
-The workshop participants suggested that the emergence of AI and automation might reduce critical thinking and imagination.
-We address this in \autoref{sec:complexity}.
-Another risk is that the widespread use of AI tools will make community resources
-such as \eg\ Stack Overflow and Wikipedia (which are used to train AI models) poorer, due to fewer incentives to maintain them.
-Further risks that were later discussed include an intensified replication crisis, a de-democratization of research, and loss of trust in science.
-Due to the non-deterministic nature of generative AI, we expect data generation and analysis to become less reproducible.
-Professional RSEs could safeguard reproducibility by designing, evaluating, and monitoring appropriate workflows.
-Future research will require significant computing and storage resources, as well as useful data and models,
-therefore it will be more driven by institutions able to provide these.
-Finally, an increasing risk is the loss of trust in science,
-in particular due to the automation of the peer-reviewing process~\cite{Naddaf2025}.
-As more and more research will be written by researchers assisted by AI,
-RSEs will be freed from repeatedly crafting similar individual elements, but will be spending more of their time in designing and reviewing software written by other humans and AI tools.
-To address these risks, the RSE of the future could be not only closely overlooking every part of the research software development cycle (ensuring dissemination, archiving, and reproducibility),
-but also contributing to endangered community resources and infrastructure that will keep research accessible to the wider scientific community.
+Another risk of the widespread use of AI tools is that community resources
+such as \eg\ Stack Overflow and Wikipedia become poorer because there are fewer incentives to maintain them.
 
@@
 \subsection{Impact of Complexity and Usability in Research Software}
-\label{sec:complexity}
 \begin{whatis}{}{abstraction}

### 2025-08-12 future_challenges_for_RSEng.tex
Add AI surveys
Co-authored-by: Jeremy Cohen <jcohen02@users.noreply.github.com>
Co-authored-by: Magnus Hagdorn <Magnus.Hagdorn@marsupium.org>

--- 2025-08-11 future_challenges_for_RSEng.tex
@@
 These AI-powered code assistants promise to increase productivity
-by automatically generating code, and writing test cases and documentation.
+by automatically generating code, and writing test cases and documentation~\cite{Banh2025}.
 Software developers use them to reduce the number of keystrokes,
@@
 and from debugging the generated code~\cite{Simkute2025}.
-Many factors can influence AI tools adoption~\cite{Russo2024}.
+Many factors can influence the adoption of AI tools~\cite{Russo2024}.
 The potential of AI to be a democratising technology can only be realised
@@
 such as \eg\ Stack Overflow and Wikipedia (which are used to train AI models) poorer
-due to fewer incentives to maintain them, and could compromise LLMs via `model collapse'
-when AI-generated content is added to these websites and becomes part of the training data~\cite{Shumailov2024}.
-Further risks that were later discussed include an intensified replication crisis, a de-democratization of research, and loss of trust in science.
-Due to the non-deterministic nature of generative AI, we expect AI-assisted simulations and and data analysis to become less reproducible.
+due to fewer incentives to maintain them, although a recent survey suggests
+these communities experience less activity from new users and an increased
+average complexity of the questions posted~\cite{Burtch2024}.
+The use of AI assistants in these communities could compromise LLMs via `model collapse'
+when AI-generated content becomes part of the training data~\cite{Shumailov2024}.
+Further risks that were later discussed include an intensified replication crisis,
+a de-democratization of research, and loss of trust in science.
+Due to the non-deterministic nature of generative AI, we expect AI-assisted simulations
+and LLM-assisted data analysis to become less reproducible.
 This is not an entirely new issue, and similar concerns exist in the HPC community,
@@
 Future research will require significant computing and storage resources, as well as useful data and models,
-therefore it will be more driven by institutions able to provide these.
+therefore it will be more likely to be driven by institutions that are able to provide these.
 Finally, an increasing risk is the loss of trust in science,
@@
 RSEs will be freed from repeatedly crafting similar individual elements,
-but will be spending more of their time in designing and reviewing software written by other humans and AI tools.
-To address these risks, the RSE of the future could be not only closely overlooking every part
+but will be spending more of their time designing and reviewing software written by other humans and AI tools.
+To address these risks, the RSE of the future should not only steer every part
 of the research software development cycle (ensuring dissemination, archiving, and reproducibility),
-but also contributing to endangered community resources and infrastructure that
+but also contribute to endangered community resources and infrastructure that
 will keep research accessible to the wider scientific community.
@@
 and more generally, by improving AI literacy~\cite{Alenezi2025}.
+AI-RSE specialists can contribute to the emerging fields of explainable
+AI for SE (XAI4SE)~\cite{Mohammadkhani2023v1} and AI for RSE (AI4RSE)~\cite{Farshidi2025v1}.
 

### 2025-08-13 future_challenges_for_RSEng.tex
Update future_challenges_for_RSEng.tex
Mention AI-RSE specialists in the goal
Remove deprecated paragraph

@@
 \begin{whatis}{}{impact-ai-automation}
-RSEs understand how AI models work and how to tune and extend them.
+RSEs understand how AI models work and AI-RSE specialists know how to tune and extend them.
 With these skills they can help researchers reap their benefits,
@@
 average complexity of the questions posted~\cite{Burtch2024}.
-The use of AI assistants in these communities could compromise LLMs via `model collapse'
+The use of AI assistants in these communities could also compromise LLMs via `model collapse'
 when AI-generated content becomes part of the training data~\cite{Shumailov2024}.
+
 Further risks that were later discussed include an intensified replication crisis,
@@
 will keep research accessible to the wider scientific community.
-
-The lessons learned from Human Factors research of the previous decades~\cite{Simkute2025}
-applied to AI tools helped resolve the initial issues outlined above.
-We envision that AI will be transforming research processes by enabling rapid analysis of massive datasets,
-uncovering patterns and insights that would have been impossible or prohibitively time-consuming for humans alone to find.
-In research and development, AI will accelerate scientific breakthroughs in fields like healthcare,
-climate science, and materials engineering by automating complex modelling and predictive tasks.
-Developers will be shifting from manual coding to orchestrating AI-driven workflows,
-focusing more on high-level design, integration, and creative problem-solving while AI handles repetitive or routine tasks.
 

### 2025-08-14 correspondence/rebuttal-1.md
More SE literature

@@
 
+Thank you for this observation.
+We were unaware of the SE community work on explainable AI for SE (XAI4SE) and AI for RSE (AI4RSE).
+We have extended the section on AI tooling and its challenges using the SE literature on the subject.
+We now mention the issue of reproducibility and draw a parallel with the HPC literature on the subject.
+
 ### Quality of presentation
@@
 I'd love to see a better elaboration of the paper, as again it is a relevant and interesting approach.
+
+This is a fair point.
+We have familiarized ourselves with the SE literature on this subject and have drawn
+parallels between projections made in the SE space and made in the RSE space.
+It was particularly interesting to see the overlap between our manuscript
+and Katz's 2023 paper ``Research Software Engineering in 2030''.
+
+Many participants of the workshop are scientists by training, and therefore have slightly
+different perspectives on the future of RSEng due to the environment they work in.
+They also use a different vocabulary compared to software engineers.
+The addition of SE literature is definitively an improvement for the manuscript.
 

### 2025-08-14 future_challenges_for_RSEng.tex
More SE literature

@@
 Central RSE units are in the process of being imagined~\cite{Kempf2025-draft},
-while an RSE Master's is in the early stages of design.
+while an RSE Master's is in the early stages of design~\cite{Dehne2025-draft}.
 
@@
 and community in the mid-term future, focusing on future challenges.
-While future projections are common in industry~\cite{Gartner2025},
+While future projections are common in software engineering
+(SE)~\cite{Katz2023,Khan2021,Hu2023,Bosch2016a,Boehm2011},
 so far there is little projection for the RSEng space.
@@
 This work supplements other opinion pieces developed in parallel
-in the SE community~\cite{ChueHong2025,Sochat2024Infra,Bencomo2024AEBoK,OliveiraJr2024,Druskat2025},
+in the SE community~\cite{ChueHong2025,Sochat2024Infra,Bencomo2024AEBoK,OliveiraJr2024,Druskat2025,Carleton2022},
 which where not known to the workshop participants; remarkably,
@@
 from full professional RSEs to domain scientists who develop research software.
+We note a similar discussion took place regarding how SE roles had to adapt to
+rapid changes brought by automation, cloud services and new project management
+methodologies~\cite{Meade2019}.
 

### 2025-08-15 future_challenges_for_RSEng.tex
point towards dual use dilemma

--- 2025-08-14 future_challenges_for_RSEng.tex
@@
 climate~\cite{Lannelongue2023} and more.
-Another example might be that they are developing code that can also be used
-for military purposes, such as numerical solvers or computer vision systems.
+Like all areas of research and engineering RSEs are faced with the dual use dilemma~\cite{Bobier2024} where artefacts they develop can be used for military purposes.
 As engineers, RSEs take the responsibility for the software they develop.
+This includes an awareness of ethical issues that may arise.
 RSEs need to be aware of these issues and develop their own ``red lines''.
-With the advantages of \autoref{sec:ai}, ever-more AI created artifacts will be included in research.
+With the advantages of \autoref{sec:ai}, ever-more AI created artefacts will be included in research.
 As a result of this, we expect a discussion within societies about their acceptable use.

### 2025-08-15 future_challenges_for_RSEng.tex

@@
 She increasingly struggles to find compelling Open Source alternatives that enables them to keep complete control of their data
-rather than relying on software and tools that are backed by large, foreign corporate entities.
+rather than relying on software and tools that are backed by large corporate entities.
 So far, she still has the support of the university board, since it values strategic autonomy.

### 2025-08-18 future_challenges_for_RSEng.tex
apply suggestion by @MakisH
Co-authored-by: Gerasimos Chourdakis <gerasimos.chourdakis@ipvs.uni-stuttgart.de>

--- 2025-08-14 future_challenges_for_RSEng.tex
@@
 climate~\cite{Lannelongue2023} and more.
-Another example might be that they are developing code that can also be used
-for military purposes, such as numerical solvers or computer vision systems.
+Like all areas of research and engineering RSEs are faced with the dual use dilemma~\cite{Bobier2024} where artefacts they develop can be used for military purposes.
 As engineers, RSEs take the responsibility for the software they develop.
-RSEs need to be aware of these issues and develop their own ``red lines''.
-With the advantages of \autoref{sec:ai}, ever-more AI created artifacts will be included in research.
+This includes an awareness of ethical issues that may arise.
+RSEs need to be aware of these issues and develop their own ``red lines''
+which need to be supported by organizational and professional guidelines.
+With the advantages of \autoref{sec:ai}, ever-more AI created artefacts will be included in research.
 As a result of this, we expect a discussion within societies about their acceptable use.

### 2025-09-18 future_challenges_for_RSEng.tex
sort out box 1 of issue 52

@@
 more sustainable, robust and maintainable.
-We also hope that Research Software becomes an accepted research artefact on an equal footing with data and publications.
+We also hope that Research Software becomes an accepted research artefact on an equal footing with data publications and traditional text publications.
 We see this happening in a number of ways,
@@
 
-The specialist skills that RSEs provide take a significant amount of time to develop and maintain.
-Therefore, we do not expect domain researchers to also become experts in a range of
-advanced RSEng topics.
-
 At the same time, we do recognise that the application of core software development best practices
 to enhance and address problems with research codebases, that can often make up the bulk of
-current RSE work, will change. As noted above, these are skills that we
-would expect any researcher who writes code to have in 10 years time. 
+current RSE work, will change.
+As noted above, these are skills that we would expect any researcher who writes code to have in 10 years time.
 The rapid development of artificial intelligence (AI) and large language models
@@
 
-So where do RSEs fit into this picture? Even today, we expect to see RSEs providing technical skills and input
-to research activities in a range of different areas, including:
+So where do RSEs fit into this picture?
+Even today, we expect to see RSEs providing technical skills and input to research activities in a range of different areas, including:
 high-quality, robust architecture and design for research software;
@@
 and green computing - efficient implementations of algorithms/methods and use of hardware.
+These specialist skills that RSEs provide take a significant amount of time to develop and maintain,
+which is why we do not expect domain researchers to also become experts in a range of advanced RSEng topics.
 
@@
 Which tasks are still relevant, where they can be happy that a younger generation is now taking up the baton and carrying forward their work?
-What issues have newly emerged, where they are now just happy to say: “Oh well… my successors will take care of that”.
+And what are the issues that have newly emerged during this time?
 In order to bring this future to life, these two personas will accompany us through the rest of this paper and we will shed light on aspects

### 2025-09-18 correspondence/rebuttal-1.md
point towards dual use dilemma
drop foreign
resolve box 2

@@
 
+> "accepted research artefact on an equal footing with data and publications." If software or data is publicized would these artefacts not also be regarded as publications?
+I would suggest using "data and text" or "data and text publications" instead.
+
+We have reworded the text as suggested.
+
+> "The specialist skills that RSEs provide take a significant amount of time to develop and maintain.
+Therefore, we do not expect domain researchers to also become experts in a range of advanced RSEng topics." Could the authors name some of these advanced topics here?
+Or maybe change the position of this paragraph to after the paragraph about the domain researchers.
+
+We have moved this paragraph further down.
+
+> "What issues have newly emerged, where they are now just happy to say: .." This seems
+
+Not clear what was meant here. We have reworded it nevertheless.
+
+> 2.1 Chapter
+> "might remain unchanged, there might be a shift in the responsibilities of RSEs, in the form of a partial reallocation of the training budget to the other core competencies.
+" What is the rationale for this statement?
+
+We reprased it to better reflect that adaptations will be necessary
+
+> "Looking at that on a broader scale..." What is meant by "that"? What do the authors look at?
+
+Clarified that we mean the competencies.
+
+> Chapter 2.2
+> "visibility and accelerate the professionalisation of the RSE role" -> missing line spacing
+
+fixed this.
+
+> Story Box 4
+> "Kim accelerates his research.." To my understanding, Kim and Kay are female.
+
+Reworded it to keep the ambiguity in Gender.
+
 > Regarding Chapter 2.3.
@@
 
+Thank you for the comment. We have changed the sentence to point to the dual use dilemma.
+
 > Story Box 6:
@@
 
-> Chaper 2.7
+Good point, we have dropped foreign as it does not matter where a large cooperation has its head quarters
+
+> Chapter 2.7
 > "in lower tiers of an of a research organisation", sentence needs attention

### 2025-09-18 future_challenges_for_RSEng.tex
resolve box 2

@@
 While the RSE core competencies~\cite{Goth2024} might remain unchanged,
-there might be a shift in the responsibilities of RSEs,
-in the form of a partial reallocation of the training budget
-to the other core competencies.
+there might be a shift in the responsibilities of RSEs and thereby requiring a realignment of the priorities in training.
 Regarding the definition of RSEs, we will still have a spectrum of competencies,
@@
 
-Looking at that on a broader scale, especially the education aspect deserves some attention.\\
+Looking at the competencies on a broader scale, especially the education aspect deserves some attention.\\
 The Covid-19 pandemic acted as a catalyst for the digital transformation
@@
 generates reproducible workflows, and adapts to Kim’s coding style and domain-specific needs.
-With this assistance, Kim accelerates his research output while maintaining full control over scientific direction and ethical oversight.
+With this assistance, Kim's research output is amplified while maintaining full control over the scientific direction and ethical oversight.
 \end{story}

### 2025-09-19 correspondence/rebuttal-1.md
sort out Comment-3

@@
 
+> Chaper 2.7
+> "in lower tiers of an of a research organisation", sentence needs attention
+
+Fixed.
+
 > Minor remarks:
@@
 > This chapter includes the authors vision of the RSE space in 2035.
-
-> "accepted research artefact on an equal footing with data and publications." If software or data is publicized would these artefacts not also be regarded as publications?
-I would suggest using "data and text" or "data and text publications" instead.
-
-> "The specialist skills that RSEs provide take a significant amount of time to develop and maintain.
-Therefore, we do not expect domain researchers to also become experts in a range of advanced RSEng topics." Could the authors name some of these advanced topics here?
-Or maybe change the position of this paragraph to after the paragraph about the domain researchers.
-
-> Story Box 1:
-> "What issues have newly emerged, where they are now just happy to say: .." This seems
-
-> 2.1 Chapter
->
-> "might remain unchanged, there might be a shift in the responsibilities of RSEs, in the form of a partial reallocation of the training budget to the other core competencies.
-" What is the rationale for this statement?
-
-> "Looking at that on a broader scale..." What is meant by "that"? What do the authors look at?
-
-> Chapter 2.2
-> "visibility and accelerate the professionalisationof the RSE role" -> missing line spacing
-
-> Story Box 4
-> "Kim accelerates his research.." To my understanding, Kim and Kay are female.
 
@@
 
-> Chapter 2.7
-> "in lower tiers of an of a research organisation", sentence needs attention
-
 > Story Box 8:
@@
 
+Thanks for pointing this out. We added some discussion that this needs to be resolved by then.
+
 > Conclusion paragraph. Formatting needs attention.
+
+We fixed a typo.
 

### 2025-09-19 future_challenges_for_RSEng.tex
sort out Comment-3

@@
 We would like to see the emergence of more dedicated staff, i.e. other (digital) RTPs next to the RSE.
+Also we expect to see the legal requirements for employing people to be flexible enough to allow for the employment of more RTPs.
 Also we hope that researchers can build careers at the interface of academia, industry, and policy (\eg\ in think tanks, NGOs,
@@
 This, however, comes with additional funding challenges,
-as long-term funding can be more difficult to acquire in lower tiers
-of a research organisation (i.e., at departmental or individual research group level).
+as long-term funding can be more difficult to acquire in lower tiers of a research organisation (i.e., at departmental or individual research group level).
 Third-party funding can mitigate this issue in the mid-term.
@@
 and potentially request and coordinate support from external RSEs.
+Within Germany at least, the distinction whether these position are research staff or technical staff has so far created some discussion.
+The authors hope, that 10 years from now this discussion has resolved towards a solution that enables the employment of RTPs more flexibly,
+and thereby recognizing that the distinction between research staff and technical staff is too restrictive.
 
@@
 \end{story}
+
 \section{What is in this future for RSEs?}
@@
 academia and is supported by political actors.
+
 \section{Conclusion}
-We have tried to elaborate on the state on the future ten years from now.
+We have tried to elaborate on the state of the future ten years from now.
 It turned out to be a difficult endeavour, since even the full breadth of today's trends is not yet foreseeable.

### 2025-09-19 correspondence/rebuttal-1.md
sort out Comment-5

@@
 >
+
+Thanks for pointing this out, we reformulated that sentence.
+
 > "While future projections are common in industry [Gar25], so far there is a projection lacking for the RSEng space" (slightly stilted and non-standard English in "a projection lacking").
 >
+
+Thanks for pointing this out, we clarified these sentences while incorporating changes for other edits.
+
 > These cases are minor and debatable, and as such do not merit additional editing.

### 2025-09-19 future_challenges_for_RSEng.tex
sort out Comment-5

@@
 with increasing integration into various aspects of research.
-In Germany, a formal call for creating a sustainable environment for research software
-is only a couple of years behind~\cite{Anzt2021},
-while the foundational competencies of a research software engineer have only recently been defined in detail~\cite{Goth2024}.
+In Germany, a formal call for creating a sustainable environment for research software has only been raised by the RSE community a couple of years ago~\cite{Anzt2021},
+while the foundational competencies of a research software engineer have just recently been defined in detail~\cite{Goth2024}.
 Central RSE units are in the process of being imagined~\cite{Kempf2025-draft},
@@
 and community in the mid-term future, focusing on future challenges.
-While future projections are common in software engineering
-(SE)~\cite{Katz2023,Khan2021,Hu2023,Bosch2016a,Boehm2011},
-so far there is little projection for the RSEng space.
+While future projections are standard practice in software engineering (SE)~\cite{Katz2023,Khan2021,Hu2023,Bosch2016a,Boehm2011},
+so far future projections are far less common for the RSEng domain.
 We present here the results of a workshop discussion (notes content is archived here \cite{goth2025_WSPad}) at the

### 2025-09-19 correspondence/rebuttal-1.md
sort out Box-6

@@
 
+> Research Software Engineering and Research Software Engineer ae occasionally written in full when the abbreviation could be used
+
+We have used the abbreviation now in all places except where it is part of a name or a paublication. Also we refrained from starting the intro with an abbreviation.
+
+> Why is Research Software capitalised?
+
+Now it is lower-case, except in names or publication titles.
+
+> The titles in section 2 aren’t consistent in how they are capitalised
+
+Thanks for spotting this, now hte titles are properly capitalised.
+
 \newpage

### 2025-09-19 future_challenges_for_RSEng.tex
sort out Box-6

@@
 In Germany, a formal call for creating a sustainable environment for research software has only been raised by the RSE community a couple of years ago~\cite{Anzt2021},
-while the foundational competencies of a research software engineer have just recently been defined in detail~\cite{Goth2024}.
+while the foundational competencies of an RSE have just recently been defined in detail~\cite{Goth2024}.
 Central RSE units are in the process of being imagined~\cite{Kempf2025-draft},
@@
 we aim to engage the community in creating a future worth living
-and to advance the conversation about how Research Software Engineering should, or could, develop.
+and to advance the conversation about how RSEng should, or could, develop.
 
@@
 
-\section{Aspects of this future}
+\section{Aspects of this Future}
 \begin{whatis}{}{whatisexplanation}
@@
 more sustainable, robust and maintainable.
-We also hope that Research Software becomes an accepted research artefact on an equal footing with data publications and traditional text publications.
+We also hope that research software becomes an accepted research artefact on an equal footing with data publications and traditional text publications.
 We see this happening in a number of ways,
@@
 our fictional RSEs who we have used to illustrate specific scenarios in our earlier work,
-were able to ride the waves of their career in Research Software Engineering and are now pondering early retirement on a sun blessed beach.
+were able to ride the waves of their career in RSEng and are now pondering early retirement on a sun blessed beach.
 %MM-NOTE: the above scenario (early retirement) is contrary to all trends in retirement age development of public employees across the "global north".
@@
 
-\subsection{Development of the RSE community}
+\subsection{Development of the RSE Community}
 \begin{whatis}{}{development-rse-community}
@@
 promoting the global exchange of ideas. The introduction of RSE Master's programmes,
-coupled with a shift in culture towards valuing Research Software,
+coupled with a shift in culture towards valuing research software,
 has significantly elevated its profile.
@@
 
-\subsection{Impact of AI and automation}
+\subsection{Impact of AI and Automation}
 \label{sec:ai}
@@
 
-\subsection{Ethics and social considerations}
+\subsection{Ethics and Social Considerations}
 \begin{whatis}{}{ethics}
@@
 By 2035, we want to have achieved accepted rules on the use of AI in research tasks and software development.
-Open Science and Open Research are by then well-established in academia, and supported by respective Open Research Software
+Open Science and Open Research are by then well-established in academia, and supported by respective open research software
 strategies of the political actors. We also hope that Data Awareness and Data Protection are still values
@@
 
-The ethics and values of the Research Software Engineer are grounded in their practice.
+The ethics and values of the RSE are grounded in their practice.
 They commit to the health, safety and welfare of the public and act in the interest of society, their employer and their clients~\cite{Goth2024}.
@@
 
-\subsection{How have academic Institutions transformed by then?}
+\subsection{How have Academic Institutions Transformed by Then?}
 \begin{whatis}{}{academic}
@@
 
-\section{What is in this future for RSEs?}
+\section{What is in this Future for RSEs?}
 %FG-NOTE: I think we can use this part to wrap things a bit up, and describe how this future then looks like.

### 2025-09-19 correspondence/rebuttal-1.md
sort out Comment-8

@@
 
-We have used the abbreviation now in all places except where it is part of a name or a paublication. Also we refrained from starting the intro with an abbreviation.
+We have used the abbreviation now in all places except where it is part of a name or a publication. Also we refrained from starting the intro with an abbreviation.
 
@@
 
-Thanks for spotting this, now hte titles are properly capitalised.
+Thanks for spotting this, now the titles are properly capitalised.
+
+> Page 5
+> In addition, a multitude of RSE communities self-organised at the national and regional level to support more viable RSE career
+> paths in academia. – remove “to”
+
+Used commas and stronger wording to make our intent clearer.
 

### 2025-09-19 correspondence/rebuttal-1.md
sort out Comment-9

@@
 
+> Page 7
+> We acknowledge but would also like to challenge the observations of the participants.
+We believe that “we are standing on the shoulders of giants”: – Up till now in the paper, “we” seems to have meant the research sector at large, but here it is specifically the authors – maybe reword as “The authors acknowledge ….” And “The authors believe …”
+
+Thanks for spotting this. We went throught the paper and thought about the use of "we" in each instance. Now it should refer, to the paper authors exclusively.
+
 \newpage

### 2025-09-19 future_challenges_for_RSEng.tex
sort out Comment-8
sort out Comment-9

@@
 \abstract{
-It is 2025 and we are, admittedly, still working on establishing and growing Research Software Engineering (RSEng)
+It is 2025 and the research software community is, admittedly, still working on establishing and growing Research Software Engineering (RSEng)
 as a domain in its own right, with the hope of ensuring that it gets the recognition it deserves.
@@
 We also hope that research software becomes an accepted research artefact on an equal footing with data publications and traditional text publications.
-We see this happening in a number of ways,
+This is already happening in a number of ways,
 particularly through movements such as DORA~\cite{DORA} and CoARA~\cite{COARA} which are advocating for a change in the way that research is assessed and outputs recognised.
@@
 New RSEs are trained through a specialised Master's programme, and their support is valued by researchers.
-We still teach researchers basic software skills, but the content of this lecture has drastically changed.
+Researchers are still taught basic software skills, but the content of this lecture has drastically changed.
 \end{whatis}
@@
 there might be a shift in the responsibilities of RSEs and thereby requiring a realignment of the priorities in training.
-Regarding the definition of RSEs, we will still have a spectrum of competencies,
+Regarding the definition of RSEs, there will still be a spectrum of competencies,
 from full professional RSEs to domain scientists who develop research software.
@@
 \end{whatis}
-In recent years, we have seen the formation of RSE institutes and advocacy groups, such as
+In recent years, the formation of RSE institutes and advocacy groups can be observed, such as
 the UK Software Sustainability Institute\footnote{\url{https://www.software.ac.uk}},
@@
 the professionalisation of the RSE role.
-In addition, a multitude of RSE communities self-organised at the national
-and regional level to support more viable RSE career paths in academia.
-We can expect RSE positions to become more easily accessible,
+In addition, a multitude of RSE communities, self-organised at the national and regional level,
+argue for more viable RSE career paths in academia.
+It can be expected that RSE positions become more easily accessible,
 possibly through centrally-funded long-term positions at research institutions,
@@
 when the algorithmic divide along with the digital divide is addressed~\cite{Yu2020}.
-We also note members of the SE community have called for improving LLM literacy
+We also note, that members of the SE community have called for improving LLM literacy
 by adapting SE curricula~\cite{Kirova2024}.
@@
 by building hierarchical systems of well-encapsulated, reusable components.
-It is clear that we need both RSEs that can dive deep into single projects of significant complexity,
+It is clear that research will need both RSEs that can dive deep into single projects of significant complexity,
 as well as RSEs that can manage complex systems of almost black-box components,
@@
 But their relevance hinges on constant updates in light of how research is performed.
-By 2035, we want to have achieved accepted rules on the use of AI in research tasks and software development.
+By 2035, accepted rules on the use of AI in research tasks and software development have been achieved.
 Open Science and Open Research are by then well-established in academia, and supported by respective open research software
-strategies of the political actors. We also hope that Data Awareness and Data Protection are still values
-that are protected.
+strategies of the political actors.
+We also hope that Data Awareness and Data Protection are still values that are protected.
 \end{whatis}
@@
 This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.
-We have already seen some specialisation trends in recent years,
+There have already been some specialisation trends observable in recent years,
 such as the HPC-RSE and AI-RSE, which are quite recent developments,

### 2025-09-19 correspondence/rebuttal-1.md
sort out Comment-10

@@
 
+> Page 8
+> By 2035, we want to have achieved accepted rules on the use of AI in research tasks and software development.
+Open Science and Open Research are by then well established in academia, and supported by respective Open Research Software strategies of the political actors.
+We also hope that Data Awareness and Data Protection are still values that are protected.
+– Earlier on the text in these boxes were explained to be what has been achieved by 2035, so this needs rewording.
+And why are some of these terms capitalised?
+
+Thank You for spotting. A stronger wording is now used and capitalisation has been changed. We kept Open Science, Open Research, and Good Scientific Practice since to us they are proper, full nouns.
+
+> Page 9
+> 2.6 How have academic Institutions transformed by then? – this title is quite different to the others of section 2 (which are issues) and would benefit from rewording
+
+We reworded it as a statement.
+
+> As per previous comments, wording like “we expect/hope” should be removed from the Achieved Goals box
+
+Thanks for  spotting this. We increased our boldness and removed every hope...
+
 \newpage

### 2025-09-19 future_challenges_for_RSEng.tex
sort out Comment-10

@@
 strategies of the political actors.
-We also hope that Data Awareness and Data Protection are still values that are protected.
+A strong culture of data awareness and protection has been sustained and is actively maintained across the research community.
 \end{whatis}
@@
 
-\subsection{How have Academic Institutions Transformed by Then?}
+\subsection{Transformations in Academic Institutions}
 \begin{whatis}{}{academic}
-We expect research software to be a central part of scientific practice, on the same level as data and publications.
+Research software is a central part of scientific practice, on the same level as data and publications.
 This will be supported by the first generations of RSE Master's graduates that find their way into research departments, and accelerate the pace of science.
-Similar to RSEs, who are widely known and understood by university boards in 2035, we expect the concept of RSE units to be in the minds
-of academic staff at any level.
-We hope to see that universities actively use the presence of an RSE Department as a means to differentiate themselves from other universities.
-We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
-We also hope that specialised clusters form, which pool the skills of certain specialised RSEs with certain RSE units specializing in certain tasks like data management, HPC, or visualisation.
-At universities which get such RSE units, we want to see that researchers naturally accept them as part of their available services.
+Similar to RSEs, who are widely known and understood by university boards in 2035, academic staff at any level is aware of the concept of RSE units.
+Universities actively use the presence of an RSE Department as a means to differentiate themselves from other universities.
+Some central RSE units will have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
+Specialised clusters have formed, which pool the skills of certain specialised RSEs with certain RSE units specializing in certain tasks like data management, HPC, or visualisation.
+At universities which get such RSE units, it can be observed that researchers naturally accept them as part of their available services.
 Additionally, researchers will seamlessly interact with infrastructures for using research software.

### 2025-09-19 future_challenges_for_RSEng.tex
sort out Comment-11

@@
 Kim builds upon the model set forth by the University of Arcadia based on his valuable experience. Since the university
-of Avalon has to follow different rules, not everything can be straightforwardly adapted. Life would be so much easier
-if at least university regulations would be homogenised, no?
+of Avalon has to follow different legal and administrative rules, not everything can be straightforwardly adapted.
+Life would be so much easier if at least university regulations were homogenised.
 \end{story}
@@
 \begin{whatis}{}{career-paths}
-We would like to see the emergence of more dedicated staff, i.e. other (digital) RTPs next to the RSE.
-Also we expect to see the legal requirements for employing people to be flexible enough to allow for the employment of more RTPs.
-Also we hope that researchers can build careers at the interface of academia, industry, and policy (\eg\ in think tanks, NGOs,
-or organisations that do knowledge transfer).
+As the RSE becomes establishedwe can see the emergence of more dedicated staff, i.e. other (digital) RTPs.
+The legal requirements for employing people have become flexible enough to allow for the employment of more RTPs.
+Work outside of the publication record is more readily accepted and enables researchers to build careers at the interface of academia, industry, and policy
+(\eg\ in think tanks, NGOs, or organisations that do knowledge transfer).
 \end{whatis}
@@
 The material presented here is based on the outputs of a workshop, and extended with the authors' ideas.
-Nevertheless, there are some things that we declared out of scope here.
-Among that is the ongoing digital transformation in wider society.
+Nevertheless, there are some things that we declared out of scope, such as the ongoing digital transformation in wider society.
 Digital competencies in the general population continue to increase,

### 2025-09-19 correspondence/rebuttal-1.md
firt start for generations

--- 2025-09-19 correspondence/rebuttal-1.md
@@
 
-> "accepted research artefact on an equal footing with data and publications." If software or data is publicized would these artefacts not also be regarded as publications?
-I would suggest using "data and text" or "data and text publications" instead.
-
-We have reworded the text as suggested.
-
-> "The specialist skills that RSEs provide take a significant amount of time to develop and maintain.
-Therefore, we do not expect domain researchers to also become experts in a range of advanced RSEng topics." Could the authors name some of these advanced topics here?
-Or maybe change the position of this paragraph to after the paragraph about the domain researchers.
-
-We have moved this paragraph further down.
-
-> "What issues have newly emerged, where they are now just happy to say: .." This seems
-
-Not clear what was meant here. We have reworded it nevertheless.
-
-> 2.1 Chapter
-> "might remain unchanged, there might be a shift in the responsibilities of RSEs, in the form of a partial reallocation of the training budget to the other core competencies.
-" What is the rationale for this statement?
-
-We reprased it to better reflect that adaptations will be necessary
-
-> "Looking at that on a broader scale..." What is meant by "that"? What do the authors look at?
-
-Clarified that we mean the competencies.
-
-> Chapter 2.2
-> "visibility and accelerate the professionalisation of the RSE role" -> missing line spacing
-
-fixed this.
-
-> Story Box 4
-> "Kim accelerates his research.." To my understanding, Kim and Kay are female.
-
-Reworded it to keep the ambiguity in Gender.
-
 > Regarding Chapter 2.3.
@@
 
-> Chaper 2.7
-> "in lower tiers of an of a research organisation", sentence needs attention
-
-Fixed.
+Thanks for bringing this up. We have added that aspect in such a way to not sound too Germany specific.
 
@@
 > This chapter includes the authors vision of the RSE space in 2035.
+
+> "accepted research artefact on an equal footing with data and publications." If software or data is publicized would these artefacts not also be regarded as publications?
+I would suggest using "data and text" or "data and text publications" instead.
+
+> "The specialist skills that RSEs provide take a significant amount of time to develop and maintain.
+Therefore, we do not expect domain researchers to also become experts in a range of advanced RSEng topics." Could the authors name some of these advanced topics here?
+Or maybe change the position of this paragraph to after the paragraph about the domain researchers.
+
+> Story Box 1:
+> "What issues have newly emerged, where they are now just happy to say: .." This seems
+
+> 2.1 Chapter
+>
+> "might remain unchanged, there might be a shift in the responsibilities of RSEs, in the form of a partial reallocation of the training budget to the other core competencies.
+" What is the rationale for this statement?
+
+> "Looking at that on a broader scale..." What is meant by "that"? What do the authors look at?
+
+> Chapter 2.2
+> "visibility and accelerate the professionalisationof the RSE role" -> missing line spacing
+
+> Story Box 4
+> "Kim accelerates his research.." To my understanding, Kim and Kay are female.
 
@@
 
+> Chaper 2.7
+> "in lower tiers of an of a research organisation", sentence needs attention
+
 > Story Box 8:
@@
 
-Thanks for pointing this out. We added some discussion that this needs to be resolved by then.
-
 > Conclusion paragraph. Formatting needs attention.
-
-We fixed a typo.
 
@@
 The addition of SE literature is definitively an improvement for the manuscript.
-
-> Research Software Engineering and Research Software Engineer ae occasionally written in full when the abbreviation could be used
-
-We have used the abbreviation now in all places except where it is part of a name or a publication. Also we refrained from starting the intro with an abbreviation.
-
-> Why is Research Software capitalised?
-
-Now it is lower-case, except in names or publication titles.
-
-> The titles in section 2 aren’t consistent in how they are capitalised
-
-Thanks for spotting this, now the titles are properly capitalised.
-
-> Page 5
-> In addition, a multitude of RSE communities self-organised at the national and regional level to support more viable RSE career
-> paths in academia. – remove “to”
-
-Used commas and stronger wording to make our intent clearer.
 
@@
 >
-
-Thanks for pointing this out, we reformulated that sentence.
-
 > "While future projections are common in industry [Gar25], so far there is a projection lacking for the RSEng space" (slightly stilted and non-standard English in "a projection lacking").
 >
-
-Thanks for pointing this out, we clarified these sentences while incorporating changes for other edits.
-
 > These cases are minor and debatable, and as such do not merit additional editing.

### 2025-09-19 future_challenges_for_RSEng.tex
firt start for generations

--- 2025-09-19 future_challenges_for_RSEng.tex
@@
 \abstract{
-It is 2025 and the research software community is, admittedly, still working on establishing and growing Research Software Engineering (RSEng)
+It is 2025 and we are, admittedly, still working on establishing and growing Research Software Engineering (RSEng)
 as a domain in its own right, with the hope of ensuring that it gets the recognition it deserves.
@@
 with increasing integration into various aspects of research.
-In Germany, a formal call for creating a sustainable environment for research software has only been raised by the RSE community a couple of years ago~\cite{Anzt2021},
-while the foundational competencies of an RSE have just recently been defined in detail~\cite{Goth2024}.
+In Germany, a formal call for creating a sustainable environment for research software
+is only a couple of years behind~\cite{Anzt2021},
+while the foundational competencies of a research software engineer have only recently been defined in detail~\cite{Goth2024}.
 Central RSE units are in the process of being imagined~\cite{Kempf2025-draft},
@@
 and community in the mid-term future, focusing on future challenges.
-While future projections are standard practice in software engineering (SE)~\cite{Katz2023,Khan2021,Hu2023,Bosch2016a,Boehm2011},
-so far future projections are far less common for the RSEng domain.
+While future projections are common in software engineering
+(SE)~\cite{Katz2023,Khan2021,Hu2023,Bosch2016a,Boehm2011},
+so far there is little projection for the RSEng space.
 We present here the results of a workshop discussion (notes content is archived here \cite{goth2025_WSPad}) at the
@@
 we aim to engage the community in creating a future worth living
-and to advance the conversation about how RSEng should, or could, develop.
+and to advance the conversation about how Research Software Engineering should, or could, develop.
 
@@
 
-\section{Aspects of this Future}
+\section{Aspects of this future}
 \begin{whatis}{}{whatisexplanation}
@@
 more sustainable, robust and maintainable.
-We also hope that research software becomes an accepted research artefact on an equal footing with data publications and traditional text publications.
-This is already happening in a number of ways,
+We also hope that Research Software becomes an accepted research artefact on an equal footing with data and publications.
+We see this happening in a number of ways,
 particularly through movements such as DORA~\cite{DORA} and CoARA~\cite{COARA} which are advocating for a change in the way that research is assessed and outputs recognised.
@@
 
+The specialist skills that RSEs provide take a significant amount of time to develop and maintain.
+Therefore, we do not expect domain researchers to also become experts in a range of
+advanced RSEng topics.
+
 At the same time, we do recognise that the application of core software development best practices
 to enhance and address problems with research codebases, that can often make up the bulk of
-current RSE work, will change.
-As noted above, these are skills that we would expect any researcher who writes code to have in 10 years time.
+current RSE work, will change. As noted above, these are skills that we
+would expect any researcher who writes code to have in 10 years time. 
 The rapid development of artificial intelligence (AI) and large language models
@@
 
-So where do RSEs fit into this picture?
-Even today, we expect to see RSEs providing technical skills and input to research activities in a range of different areas, including:
+So where do RSEs fit into this picture? Even today, we expect to see RSEs providing technical skills and input
+to research activities in a range of different areas, including:
 high-quality, robust architecture and design for research software;
@@
 and green computing - efficient implementations of algorithms/methods and use of hardware.
-These specialist skills that RSEs provide take a significant amount of time to develop and maintain,
-which is why we do not expect domain researchers to also become experts in a range of advanced RSEng topics.
 
@@
 our fictional RSEs who we have used to illustrate specific scenarios in our earlier work,
-were able to ride the waves of their career in RSEng and are now pondering early retirement on a sun blessed beach.
+were able to ride the waves of their career in Research Software Engineering and are now pondering early retirement on a sun blessed beach.
 %MM-NOTE: the above scenario (early retirement) is contrary to all trends in retirement age development of public employees across the "global north".
@@
 Which tasks are still relevant, where they can be happy that a younger generation is now taking up the baton and carrying forward their work?
-And what are the issues that have newly emerged during this time?
+What issues have newly emerged, where they are now just happy to say: “Oh well… my successors will take care of that”.
 In order to bring this future to life, these two personas will accompany us through the rest of this paper and we will shed light on aspects
@@
 New RSEs are trained through a specialised Master's programme, and their support is valued by researchers.
-Researchers are still taught basic software skills, but the content of this lecture has drastically changed.
+We still teach researchers basic software skills, but the content of this lecture has drastically changed.
 \end{whatis}
@@
 While the RSE core competencies~\cite{Goth2024} might remain unchanged,
-there might be a shift in the responsibilities of RSEs and thereby requiring a realignment of the priorities in training.
-Regarding the definition of RSEs, there will still be a spectrum of competencies,
+there might be a shift in the responsibilities of RSEs,
+in the form of a partial reallocation of the training budget
+to the other core competencies.
+Regarding the definition of RSEs, we will still have a spectrum of competencies,
 from full professional RSEs to domain scientists who develop research software.
@@
 
-Looking at the competencies on a broader scale, especially the education aspect deserves some attention.\\
+Looking at that on a broader scale, especially the education aspect deserves some attention.\\
 The Covid-19 pandemic acted as a catalyst for the digital transformation
@@
 
-\subsection{Development of the RSE Community}
+\subsection{Development of the RSE community}
 \begin{whatis}{}{development-rse-community}
@@
 promoting the global exchange of ideas. The introduction of RSE Master's programmes,
-coupled with a shift in culture towards valuing research software,
+coupled with a shift in culture towards valuing Research Software,
 has significantly elevated its profile.
@@
 \end{whatis}
-In recent years, the formation of RSE institutes and advocacy groups can be observed, such as
+In recent years, we have seen the formation of RSE institutes and advocacy groups, such as
 the UK Software Sustainability Institute\footnote{\url{https://www.software.ac.uk}},
@@
 the professionalisation of the RSE role.
-In addition, a multitude of RSE communities, self-organised at the national and regional level,
-argue for more viable RSE career paths in academia.
-It can be expected that RSE positions become more easily accessible,
+In addition, a multitude of RSE communities self-organised at the national
+and regional level to support more viable RSE career paths in academia.
+We can expect RSE positions to become more easily accessible,
 possibly through centrally-funded long-term positions at research institutions,
@@
 
-\subsection{Impact of AI and Automation}
+\subsection{Impact of AI and automation}
 \label{sec:ai}
@@
 when the algorithmic divide along with the digital divide is addressed~\cite{Yu2020}.
-We also note, that members of the SE community have called for improving LLM literacy
+We also note members of the SE community have called for improving LLM literacy
 by adapting SE curricula~\cite{Kirova2024}.
@@
 generates reproducible workflows, and adapts to Kim’s coding style and domain-specific needs.
-With this assistance, Kim's research output is amplified while maintaining full control over the scientific direction and ethical oversight.
+With this assistance, Kim accelerates his research output while maintaining full control over scientific direction and ethical oversight.
 \end{story}
@@
 by building hierarchical systems of well-encapsulated, reusable components.
-It is clear that research will need both RSEs that can dive deep into single projects of significant complexity,
+It is clear that we need both RSEs that can dive deep into single projects of significant complexity,
 as well as RSEs that can manage complex systems of almost black-box components,
@@
 
-\subsection{Ethics and Social Considerations}
+\subsection{Ethics and social considerations}
 \begin{whatis}{}{ethics}
@@
 But their relevance hinges on constant updates in light of how research is performed.
-By 2035, accepted rules on the use of AI in research tasks and software development have been achieved.
-Open Science and Open Research are by then well-established in academia, and supported by respective open research software
-strategies of the political actors.
-We also hope that Data Awareness and Data Protection are still values that are protected.
+By 2035, we want to have achieved accepted rules on the use of AI in research tasks and software development.
+Open Science and Open Research are by then well-established in academia, and supported by respective Open Research Software
+strategies of the political actors. We also hope that Data Awareness and Data Protection are still values
+that are protected.
 \end{whatis}
 
-The ethics and values of the RSE are grounded in their practice.
+The ethics and values of the Research Software Engineer are grounded in their practice.
 They commit to the health, safety and welfare of the public and act in the interest of society, their employer and their clients~\cite{Goth2024}.
@@
 
-\subsection{How have Academic Institutions Transformed by Then?}
+\subsection{How have academic Institutions transformed by then?}
 \begin{whatis}{}{academic}
@@
 Additionally, researchers will seamlessly interact with infrastructures for using research software.
+Replacement of aging staff has led to the creation of more RSE professorships.
 \end{whatis}
@@
 has organisationally often adopted time scales for change that are more reminiscent
-to institutions thinking on the time scale of generations.
-Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref}
+to institutions thinking on the time scale of societal generations.
+While sounding negative, this can also be an opportunity,
+if in the time-frame till 2035 a large number of aging, permanent staff is replaced by new people that have observed the
+digital transformation first-hand and have actively participated in it through various grassroots initiatives.\\
+On the other hand, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref}
 hints at the potential for specialisation of research tasks.
 This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.
-There have already been some specialisation trends observable in recent years,
+We have already seen some specialisation trends in recent years,
 such as the HPC-RSE and AI-RSE, which are quite recent developments,
@@
 We would like to see the emergence of more dedicated staff, i.e. other (digital) RTPs next to the RSE.
-Also we expect to see the legal requirements for employing people to be flexible enough to allow for the employment of more RTPs.
 Also we hope that researchers can build careers at the interface of academia, industry, and policy (\eg\ in think tanks, NGOs,
@@
 and potentially request and coordinate support from external RSEs.
-Within Germany at least, the distinction whether these position are research staff or technical staff has so far created some discussion.
-The authors hope, that 10 years from now this discussion has resolved towards a solution that enables the employment of RTPs more flexibly,
-and thereby recognizing that the distinction between research staff and technical staff is too restrictive.
 
@@
 \end{story}
-
-\section{What is in this Future for RSEs?}
+\section{What is in this future for RSEs?}
 %FG-NOTE: I think we can use this part to wrap things a bit up, and describe how this future then looks like.
@@
 academia and is supported by political actors.
-
 \section{Conclusion}
-We have tried to elaborate on the state of the future ten years from now.
+We have tried to elaborate on the state on the future ten years from now.
 It turned out to be a difficult endeavour, since even the full breadth of today's trends is not yet foreseeable.

### 2025-09-19 future_challenges_for_RSEng.tex
Update future_challenges_for_RSEng.tex
Co-authored-by: Jeremy Cohen <jcohen02@users.noreply.github.com>

--- 2025-09-19 future_challenges_for_RSEng.tex
@@
 \end{whatis}
-In recent years, the formation of RSE institutes and advocacy groups can be observed, such as
+In recent years, various RSE-related institutes and advocacy groups have been formed, such as
 the UK Software Sustainability Institute\footnote{\url{https://www.software.ac.uk}},
@@
 when the algorithmic divide along with the digital divide is addressed~\cite{Yu2020}.
-We also note, that members of the SE community have called for improving LLM literacy
+We also note that members of the SE community have called for improving LLM literacy
 by adapting SE curricula~\cite{Kirova2024}.

### 2025-09-19 future_challenges_for_RSEng.tex
Update future_challenges_for_RSEng.tex
Co-authored-by: Jeremy Cohen <jcohen02@users.noreply.github.com>

@@
 This will be supported by the first generations of RSE Master's graduates that find their way into research departments, and accelerate the pace of science.
-Similar to RSEs, who are widely known and understood by university boards in 2035, academic staff at any level is aware of the concept of RSE units.
+Similar to RSEs, who are widely known and understood by university boards in 2035, academic staff members at any level are aware of the concept of RSE units.
 Universities actively use the presence of an RSE Department as a means to differentiate themselves from other universities.

### 2025-09-19 future_challenges_for_RSEng.tex
Update future_challenges_for_RSEng.tex
Co-authored-by: Jeremy Cohen <jcohen02@users.noreply.github.com>

@@
 Specialised clusters have formed, which pool the skills of certain specialised RSEs with certain RSE units specializing in certain tasks like data management, HPC, or visualisation.
-At universities which get such RSE units, it can be observed that researchers naturally accept them as part of their available services.
+At universities which create such RSE units, it can be observed that researchers naturally accept them as part of their available services.
 Additionally, researchers will seamlessly interact with infrastructures for using research software.
@@
 and potentially request and coordinate support from external RSEs.
-Within Germany at least, the distinction whether these position are research staff or technical staff has so far created some discussion.
+Within Germany at least, the distinction around whether these positions are classed as research staff or technical staff has so far created some discussion.
 The authors hope, that 10 years from now this discussion has resolved towards a solution that enables the employment of RTPs more flexibly,

### 2025-09-19 future_challenges_for_RSEng.tex
Update future_challenges_for_RSEng.tex
Co-authored-by: Jeremy Cohen <jcohen02@users.noreply.github.com>

@@
 Within Germany at least, the distinction around whether these positions are classed as research staff or technical staff has so far created some discussion.
-The authors hope, that 10 years from now this discussion has resolved towards a solution that enables the employment of RTPs more flexibly,
+The authors hope that, 10 years from now, this discussion has resolved towards a solution that enables the employment of RTPs more flexibly,
 and thereby recognizing that the distinction between research staff and technical staff is too restrictive.

### 2025-09-19 future_challenges_for_RSEng.tex
Update future_challenges_for_RSEng.tex
Co-authored-by: Jeremy Cohen <jcohen02@users.noreply.github.com>

--- 2025-09-19 future_challenges_for_RSEng.tex
@@
 \end{whatis}
-In recent years, various RSE-related institutes and advocacy groups have been formed, such as
+In recent years, we have seen the formation of RSE institutes and advocacy groups, such as
 the UK Software Sustainability Institute\footnote{\url{https://www.software.ac.uk}},
@@
 when the algorithmic divide along with the digital divide is addressed~\cite{Yu2020}.
-We also note that members of the SE community have called for improving LLM literacy
+We also note members of the SE community have called for improving LLM literacy
 by adapting SE curricula~\cite{Kirova2024}.
@@
 But their relevance hinges on constant updates in light of how research is performed.
-By 2035, accepted rules on the use of AI in research tasks and software development have been achieved.
-Open Science and Open Research are by then well-established in academia, and supported by respective open research software
-strategies of the political actors.
-A strong culture of data awareness and protection has been sustained and is actively maintained across the research community.
+By 2035, we want to have achieved accepted rules on the use of AI in research tasks and software development.
+Open Science and Open Research are by then well-established in academia, and supported by respective Open Research Software
+strategies of the political actors. We also hope that Data Awareness and Data Protection are still values
+that are protected.
 \end{whatis}
 
-The ethics and values of the RSE are grounded in their practice.
+The ethics and values of the Research Software Engineer are grounded in their practice.
 They commit to the health, safety and welfare of the public and act in the interest of society, their employer and their clients~\cite{Goth2024}.
@@
 
-\subsection{Transformations in Academic Institutions}
+\subsection{How have academic Institutions transformed by then?}
 \begin{whatis}{}{academic}
-Research software is a central part of scientific practice, on the same level as data and publications.
+We expect research software to be a central part of scientific practice, on the same level as data and publications.
 This will be supported by the first generations of RSE Master's graduates that find their way into research departments, and accelerate the pace of science.
-Similar to RSEs, who are widely known and understood by university boards in 2035, academic staff members at any level are aware of the concept of RSE units.
-Universities actively use the presence of an RSE Department as a means to differentiate themselves from other universities.
-Some central RSE units will have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
-Specialised clusters have formed, which pool the skills of certain specialised RSEs with certain RSE units specializing in certain tasks like data management, HPC, or visualisation.
-At universities which get such RSE units, it can be observed that researchers naturally accept them as part of their available services.
+Similar to RSEs, who are widely known and understood by university boards in 2035, we expect the concept of RSE units to be in the minds
+of academic staff at any level.
+We hope to see that universities actively use the presence of an RSE Department as a means to differentiate themselves from other universities.
+We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
+We also hope that specialised clusters form, which pool the skills of certain specialised RSEs with certain RSE units specializing in certain tasks like data management, HPC, or visualisation.
+At universities which get such RSE units, we want to see that researchers naturally accept them as part of their available services.
 Additionally, researchers will seamlessly interact with infrastructures for using research software.
+Replacement of staff reaching retirement has led to the creation of more RSE professorships.
 \end{whatis}
@@
 Kim builds upon the model set forth by the University of Arcadia based on his valuable experience. Since the university
-of Avalon has to follow different legal and administrative rules, not everything can be straightforwardly adapted.
-Life would be so much easier if at least university regulations were homogenised.
+of Avalon has to follow different rules, not everything can be straightforwardly adapted. Life would be so much easier
+if at least university regulations would be homogenised, no?
 \end{story}
@@
 \begin{whatis}{}{career-paths}
-As the RSE becomes establishedwe can see the emergence of more dedicated staff, i.e. other (digital) RTPs.
-The legal requirements for employing people have become flexible enough to allow for the employment of more RTPs.
-Work outside of the publication record is more readily accepted and enables researchers to build careers at the interface of academia, industry, and policy
-(\eg\ in think tanks, NGOs, or organisations that do knowledge transfer).
+We would like to see the emergence of more dedicated staff, i.e. other (digital) RTPs next to the RSE.
+Also we hope that researchers can build careers at the interface of academia, industry, and policy (\eg\ in think tanks, NGOs,
+or organisations that do knowledge transfer).
 \end{whatis}
@@
 The material presented here is based on the outputs of a workshop, and extended with the authors' ideas.
-Nevertheless, there are some things that we declared out of scope, such as the ongoing digital transformation in wider society.
+Nevertheless, there are some things that we declared out of scope here.
+Among that is the ongoing digital transformation in wider society.
 Digital competencies in the general population continue to increase,

### 2025-09-19 future_challenges_for_RSEng.tex
Update future_challenges_for_RSEng.tex
Co-authored-by: Jeremy Cohen <jcohen02@users.noreply.github.com>

@@
 While sounding negative, this can also be an opportunity,
-if in the time-frame till 2035 a large number of aging, permanent staff is replaced by new people that have observed the
+if in the time-frame of the next decade, a large number of permanent staff reach retirement and are replaced by a new generation of researchers and technical professionals who have seen the
 digital transformation first-hand and have actively participated in it through various grassroots initiatives.\\

### 2025-09-22 future_challenges_for_RSEng.tex
spell check

--- 2025-09-19 future_challenges_for_RSEng.tex
@@
 With the growing importance and scope of software engineering in research,
-researchers who code now begin to fill dedicated, specialized RSE roles crucial for the sustainable development
+researchers who code now begin to fill dedicated, specialised RSE roles crucial for the sustainable development
 of long-term research software projects that can rapidly innovate.
@@
 Kim and Kay chuckled when they looked at their original job descriptions from 2020: ``build tools for scientists, manage data, write reusable code.''
-Now, their roles are unrecognizable: half AI whisperer, half research ethicist, and part-time diplomat negotiating between synthetic and human collaborators.
+Now, their roles are unrecognisable: half AI whisperer, half research ethicist, and part-time diplomat negotiating between synthetic and human collaborators.
 They now not only manage software, but also the personalities of autonomous lab agents and negotiate data-sharing agreements with international AI consortia.
@@
 Professional RSEs could safeguard reproducibility by designing, evaluating, and monitoring appropriate workflows.
-When applicable, standardized and version-controlled models should be preferred
+When applicable, standardised and version-controlled models should be preferred
 over self-developed models, and self-developed models should be treated like any
@@
 RSEs need to be aware of these issues and develop their own ``red lines''
-which need to be supported by organizational and professional guidelines.
+which need to be supported by organisational and professional guidelines.
 With the advantages of \autoref{sec:ai}, ever-more AI created artefacts will be included in research.
@@
 We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
-We also hope that specialised clusters form, which pool the skills of certain specialised RSEs with certain RSE units specializing in certain tasks like data management, HPC, or visualisation.
+We also hope that specialised clusters form, which pool the skills of certain specialised RSEs with certain RSE units specialising in certain tasks like data management, HPC, or visualisation.
 At universities which get such RSE units, we want to see that researchers naturally accept them as part of their available services.
 Additionally, researchers will seamlessly interact with infrastructures for using research software.
-Replacement of staff reaching retirement has led to the creation of more RSE professorships.
 \end{whatis}
@@
 has organisationally often adopted time scales for change that are more reminiscent
-to institutions thinking on the time scale of societal generations.
-While sounding negative, this can also be an opportunity,
-if in the time-frame till 2035 a large number of aging, permanent staff is replaced by new people that have observed the
-digital transformation first-hand and have actively participated in it through various grassroots initiatives.\\
-On the other hand, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref}
+to institutions thinking on the time scale of generations.
+Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref}
 hints at the potential for specialisation of research tasks.
@@
 Due to the rise of dedicated RSE units, networking people and teams,
-federating infrastructures, as well as organizing and sharing knowledge will become important.
+federating infrastructures, as well as organising and sharing knowledge will become important.
 Structuring these new units and joining their administrative support will also help the units to grow.
@@
 While a standardised RSE Master's degree will make it easier to develop the next generation of RSEs,
-organizations could develop standardized hiring processes to acquire the skill set that they actually need.
+organisations could develop standardised hiring processes to acquire the skill set that they actually need.
 Budgeting and administrative hurdles will also probably still exist and hinder the formation of permanent and well-funded RSE teams.
@@
 but in keeping the quantum computers humming. He became the go-to expert
-for troubleshooting entangled qubits and stabilizing delicate cryogenic systems.
+for troubleshooting entangled qubits and stabilising delicate cryogenic systems.
 His unique blend of physics knowledge and hands-on problem solving earned him
@@
 enabling easier reuse of prior software outputs in new research.
-We also hope that a cultural change is taking place to recognize research software
+We also hope that a cultural change is taking place to recognise research software
 as an integral part of science that it is widely recognised and rewarded across

### 2025-09-22 future_challenges_for_RSEng.tex
change word
"research codebase" sounds a bit strange, especially the plural. I think
just calling it research code sounds better and suffices.

@@
 At the same time, we do recognise that the application of core software development best practices
-to enhance and address problems with research codebases, that can often make up the bulk of
+to enhance and address problems with research code, that can often make up the bulk of
 current RSE work, will change. As noted above, these are skills that we

### 2025-09-22 future_challenges_for_RSEng.tex
replace word
In general I struggle with the word of democratisation of technology. I
think de-democratisation is even worse. The direct opposite is
autocratisation (when we talk about change of political systems). Let's
call it what it really means, democratisation of technologt means making
technology more accessible (especially to non-specialists). In this case
it is making research less accessible.
Update future_challenges_for_RSEng.tex
Co-authored-by: Magnus Hagdorn <Magnus.Hagdorn@marsupium.org>

--- 2025-09-19 future_challenges_for_RSEng.tex
@@
 with increasing integration into various aspects of research.
-In Germany, a formal call for creating a sustainable environment for research software
-is only a couple of years behind~\cite{Anzt2021},
-while the foundational competencies of a research software engineer have only recently been defined in detail~\cite{Goth2024}.
+In Germany, a formal call for creating a sustainable environment for research software has only been raised by the RSE community a couple of years ago~\cite{Anzt2021},
+while the foundational competencies of an RSE have just recently been defined in detail~\cite{Goth2024}.
 Central RSE units are in the process of being imagined~\cite{Kempf2025-draft},
@@
 more sustainable, robust and maintainable.
-We also hope that Research Software becomes an accepted research artefact on an equal footing with data and publications.
-We see this happening in a number of ways,
+We also hope that research software becomes an accepted research artefact on an equal footing with data publications and traditional text publications.
+This is already happening in a number of ways,
 particularly through movements such as DORA~\cite{DORA} and CoARA~\cite{COARA} which are advocating for a change in the way that research is assessed and outputs recognised.
@@
 While the RSE core competencies~\cite{Goth2024} might remain unchanged,
-there might be a shift in the responsibilities of RSEs,
-in the form of a partial reallocation of the training budget
-to the other core competencies.
-Regarding the definition of RSEs, we will still have a spectrum of competencies,
+there might be a shift in the responsibilities of RSEs and thereby requiring a realignment of the priorities in training.
+Regarding the definition of RSEs, there will still be a spectrum of competencies,
 from full professional RSEs to domain scientists who develop research software.
@@
 \end{whatis}
-In recent years, we have seen the formation of RSE institutes and advocacy groups, such as
+In recent years, various RSE-related institutes and advocacy groups have been formed, such as
 the UK Software Sustainability Institute\footnote{\url{https://www.software.ac.uk}},
@@
 when the algorithmic divide along with the digital divide is addressed~\cite{Yu2020}.
-We also note members of the SE community have called for improving LLM literacy
+We also note that members of the SE community have called for improving LLM literacy
 by adapting SE curricula~\cite{Kirova2024}.
@@
 But their relevance hinges on constant updates in light of how research is performed.
-By 2035, we want to have achieved accepted rules on the use of AI in research tasks and software development.
-Open Science and Open Research are by then well-established in academia, and supported by respective Open Research Software
-strategies of the political actors. We also hope that Data Awareness and Data Protection are still values
-that are protected.
+By 2035, accepted rules on the use of AI in research tasks and software development have been achieved.
+Open Science and Open Research are by then well-established in academia, and supported by respective open research software
+strategies of the political actors.
+A strong culture of data awareness and protection has been sustained and is actively maintained across the research community.
 \end{whatis}
 
-The ethics and values of the Research Software Engineer are grounded in their practice.
+The ethics and values of the RSE are grounded in their practice.
 They commit to the health, safety and welfare of the public and act in the interest of society, their employer and their clients~\cite{Goth2024}.
@@
 
-\subsection{How have academic Institutions transformed by then?}
+\subsection{Transformations in Academic Institutions}
 \begin{whatis}{}{academic}
-We expect research software to be a central part of scientific practice, on the same level as data and publications.
+Research software is a central part of scientific practice, on the same level as data and publications.
 This will be supported by the first generations of RSE Master's graduates that find their way into research departments, and accelerate the pace of science.
-Similar to RSEs, who are widely known and understood by university boards in 2035, we expect the concept of RSE units to be in the minds
-of academic staff at any level.
-We hope to see that universities actively use the presence of an RSE Department as a means to differentiate themselves from other universities.
-We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
-We also hope that specialised clusters form, which pool the skills of certain specialised RSEs with certain RSE units specializing in certain tasks like data management, HPC, or visualisation.
-At universities which get such RSE units, we want to see that researchers naturally accept them as part of their available services.
+Similar to RSEs, who are widely known and understood by university boards in 2035, academic staff members at any level are aware of the concept of RSE units.
+Universities actively use the presence of an RSE Department as a means to differentiate themselves from other universities.
+Some central RSE units will have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
+Specialised clusters have formed, which pool the skills of certain specialised RSEs with certain RSE units specializing in certain tasks like data management, HPC, or visualisation.
+At universities which create such RSE units, it can be observed that researchers naturally accept them as part of their available services.
 Additionally, researchers will seamlessly interact with infrastructures for using research software.
-Replacement of staff reaching retirement has led to the creation of more RSE professorships.
 \end{whatis}
@@
 has organisationally often adopted time scales for change that are more reminiscent
-to institutions thinking on the time scale of societal generations.
-While sounding negative, this can also be an opportunity,
-if in the time-frame of the next decade, a large number of permanent staff reach retirement and are replaced by a new generation of researchers and technical professionals who have seen the
-digital transformation first-hand and have actively participated in it through various grassroots initiatives.\\
-On the other hand, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref}
+to institutions thinking on the time scale of generations.
+Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref}
 hints at the potential for specialisation of research tasks.
 This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.
-We have already seen some specialisation trends in recent years,
+Some specialisation trends have already been observed in recent years,
 such as the HPC-RSE and AI-RSE, which are quite recent developments,
@@
 \begin{whatis}{}{career-paths}
-We would like to see the emergence of more dedicated staff, i.e. other (digital) RTPs next to the RSE.
-Also we hope that researchers can build careers at the interface of academia, industry, and policy (\eg\ in think tanks, NGOs,
-or organisations that do knowledge transfer).
+As the RSE becomes establishedwe can see the emergence of more dedicated staff, i.e. other (digital) RTPs.
+The legal requirements for employing people have become flexible enough to allow for the employment of more RTPs.
+Work outside of the publication record is more readily accepted and enables researchers to build careers at the interface of academia, industry, and policy
+(\eg\ in think tanks, NGOs, or organisations that do knowledge transfer).
 \end{whatis}
@@
 and potentially request and coordinate support from external RSEs.
+Within Germany at least, the distinction around whether these positions are classed as research staff or technical staff has so far created some discussion.
+The authors hope that, 10 years from now, this discussion has resolved towards a solution that enables the employment of RTPs more flexibly,
+and thereby recognizing that the distinction between research staff and technical staff is too restrictive.
 
@@
 \end{story}
-\section{What is in this future for RSEs?}
+
+\section{What is in this Future for RSEs?}
 %FG-NOTE: I think we can use this part to wrap things a bit up, and describe how this future then looks like.

### 2025-09-22 future_challenges_for_RSEng.tex
take care of jeremys comment

@@
 \begin{whatis}{}{career-paths}
-As the RSE becomes establishedwe can see the emergence of more dedicated staff, i.e. other (digital) RTPs.
+As the RSE becomes established, we are also seeing the emergence of a wider range of more specialized, technical roles within the research community, i.e. other (digital) RTPs.
 The legal requirements for employing people have become flexible enough to allow for the employment of more RTPs.

### 2025-09-22 future_challenges_for_RSEng.tex
Added some intro text to highlight the German context of the paper.

--- 2025-09-22 future_challenges_for_RSEng.tex
@@
 With the growing importance and scope of software engineering in research,
-researchers who code now begin to fill dedicated, specialised RSE roles crucial for the sustainable development
+researchers who code now begin to fill dedicated, specialized RSE roles crucial for the sustainable development
 of long-term research software projects that can rapidly innovate.
@@
 In Germany, a formal call for creating a sustainable environment for research software
-is only a couple of years behind~\cite{Anzt2021},
+was only published a couple of years ago~\cite{Anzt2021},
 while the foundational competencies of a research software engineer have only recently been defined in detail~\cite{Goth2024}.
@@
 and to advance the conversation about how Research Software Engineering should, or could, develop.
+Since this paper focuses on the outputs of the deRSE25 conference, we present our thoughts and perspectives in a largely German-focused context.
+At the same time, we recognise, the wider international growth of RSEng as a very important development
+for people undertaking software development within the research community.
+Much of our work and our discussions are also building on wider perspectives for the future of RSEng that are currently under
+investigation among RSEng communities in other countries.
+While taking a German-focused view in this paper, we therefore aim to highlight resources and materials from
+the wider international community where we feel this is relevant.
 
@@
 At the same time, we do recognise that the application of core software development best practices
-to enhance and address problems with research code, that can often make up the bulk of
+to enhance and address problems with research codebases, that can often make up the bulk of
 current RSE work, will change. As noted above, these are skills that we
@@
 Kim and Kay chuckled when they looked at their original job descriptions from 2020: ``build tools for scientists, manage data, write reusable code.''
-Now, their roles are unrecognisable: half AI whisperer, half research ethicist, and part-time diplomat negotiating between synthetic and human collaborators.
+Now, their roles are unrecognizable: half AI whisperer, half research ethicist, and part-time diplomat negotiating between synthetic and human collaborators.
 They now not only manage software, but also the personalities of autonomous lab agents and negotiate data-sharing agreements with international AI consortia.
@@
 Professional RSEs could safeguard reproducibility by designing, evaluating, and monitoring appropriate workflows.
-When applicable, standardised and version-controlled models should be preferred
+When applicable, standardized and version-controlled models should be preferred
 over self-developed models, and self-developed models should be treated like any
@@
 RSEs need to be aware of these issues and develop their own ``red lines''
-which need to be supported by organisational and professional guidelines.
+which need to be supported by organizational and professional guidelines.
 With the advantages of \autoref{sec:ai}, ever-more AI created artefacts will be included in research.
@@
 We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
-We also hope that specialised clusters form, which pool the skills of certain specialised RSEs with certain RSE units specialising in certain tasks like data management, HPC, or visualisation.
+We also hope that specialised clusters form, which pool the skills of certain specialised RSEs with certain RSE units specializing in certain tasks like data management, HPC, or visualisation.
 At universities which get such RSE units, we want to see that researchers naturally accept them as part of their available services.
@@
 Due to the rise of dedicated RSE units, networking people and teams,
-federating infrastructures, as well as organising and sharing knowledge will become important.
+federating infrastructures, as well as organizing and sharing knowledge will become important.
 Structuring these new units and joining their administrative support will also help the units to grow.
@@
 While a standardised RSE Master's degree will make it easier to develop the next generation of RSEs,
-organisations could develop standardised hiring processes to acquire the skill set that they actually need.
+organizations could develop standardized hiring processes to acquire the skill set that they actually need.
 Budgeting and administrative hurdles will also probably still exist and hinder the formation of permanent and well-funded RSE teams.
@@
 but in keeping the quantum computers humming. He became the go-to expert
-for troubleshooting entangled qubits and stabilising delicate cryogenic systems.
+for troubleshooting entangled qubits and stabilizing delicate cryogenic systems.
 His unique blend of physics knowledge and hands-on problem solving earned him
@@
 enabling easier reuse of prior software outputs in new research.
-We also hope that a cultural change is taking place to recognise research software
+We also hope that a cultural change is taking place to recognize research software
 as an integral part of science that it is widely recognised and rewarded across

### 2025-09-22 future_challenges_for_RSEng.tex
correct/update my affiliation

@@
 \texorpdfstring{\\}{} % line break
-Frank Löffler\authorOrcid{0000-0001-6643-6323}\authorRef{11},
-Michele Martone\authorOrcid{0000-0003-3239-8554}\authorRef{12},
-Jan Philipp Thiele\authorOrcid{0000-0002-8901-6660}\authorRef{10}\textsuperscript{,}\authorRef{13}
+Frank Löffler\authorOrcid{0000-0001-6643-6323}\authorRef{11}\textsuperscript{,}\authorRef{12},
+Michele Martone\authorOrcid{0000-0003-3239-8554}\authorRef{13},
+Jan Philipp Thiele\authorOrcid{0000-0002-8901-6660}\authorRef{10}\textsuperscript{,}\authorRef{14}
 } % Authors and references to addresses
@@
 \autlabel{10} Technische Universität Braunschweig, Germany\\
-\autlabel{11} Michael Stifel Center Jena {\&} Friedrich Schiller University Jena, Germany\\
-\autlabel{12} Independent researcher, Germany\\
-\autlabel{13} Weierstrass Institute, Berlin, Germany;
+\autlabel{11} Competence Center Digital Research, Friedrich Schiller University Jena, Germany\\
+\autlabel{12} de-RSE e.V.---Society for Research Software in Germany\\
+\autlabel{13} Independent researcher, Germany\\
+\autlabel{14} Weierstrass Institute, Berlin, Germany;
               Leibniz University Hannover, Institute of Applied Mathematics, Scientific Computing, Hannover, Germany\\
@@
 In Germany, a formal call for creating a sustainable environment for research software
-was only published a couple of years ago~\cite{Anzt2021},
+is only a couple of years behind~\cite{Anzt2021},
 while the foundational competencies of a research software engineer have only recently been defined in detail~\cite{Goth2024}.
@@
 and to advance the conversation about how Research Software Engineering should, or could, develop.
-Since this paper focuses on the outputs of the deRSE25 conference, we present our thoughts and perspectives in a largely German-focused context.
-At the same time, we recognise, the wider international growth of RSEng as a very important development
-for people undertaking software development within the research community.
-Much of our work and our discussions are also building on wider perspectives for the future of RSEng that are currently under
-investigation among RSEng communities in other countries.
-While taking a German-focused view in this paper, we therefore aim to highlight resources and materials from
-the wider international community where we feel this is relevant.
 

### 2025-09-22 future_challenges_for_RSEng.tex
update Jan's affiliation matching the one of the unit paper (adding de-RSE)

@@
 Toby Hodges\authorOrcid{0000-0003-1766-456X}\authorRef{9},
-Jan Linxweiler\authorOrcid{0000-0002-2755-5087}\authorRef{10},
+Jan Linxweiler\authorOrcid{0000-0002-2755-5087}\authorRef{10}\textsuperscript{,}\authorRef{11},
 \texorpdfstring{\\}{} % line break
-Frank Löffler\authorOrcid{0000-0001-6643-6323}\authorRef{11}\textsuperscript{,}\authorRef{12},
+Frank Löffler\authorOrcid{0000-0001-6643-6323}\authorRef{12}\textsuperscript{,}\authorRef{11},
 Michele Martone\authorOrcid{0000-0003-3239-8554}\authorRef{13},
@@
 \autlabel{10} Technische Universität Braunschweig, Germany\\
-\autlabel{11} Competence Center Digital Research, Friedrich Schiller University Jena, Germany\\
-\autlabel{12} de-RSE e.V.---Society for Research Software in Germany\\
+\autlabel{11} de-RSE e.V.---Society for Research Software in Germany\\
+\autlabel{12} Competence Center Digital Research, Friedrich Schiller University Jena, Germany\\
 \autlabel{13} Independent researcher, Germany\\

### 2025-09-22 future_challenges_for_RSEng.tex
try better line breaks in affiliation

@@
 Gerasimos Chourdakis\authorOrcid{0000-0002-3977-1385}\authorRef{3},
+Simon Christ\authorOrcid{0000-0002-5866-1472}\authorRef{4},
 \texorpdfstring{\\}{} % line break
-Simon Christ\authorOrcid{0000-0002-5866-1472}\authorRef{4},
 Jeremy Cohen\authorOrcid{0000-0003-4312-2537}\authorRef{5},
 Nicholas A. Del Grosso\thinspace\authorRef{6},
+Jean-Noël Grad\authorOrcid{0000-0002-5821-4912}\authorRef{7},
 \texorpdfstring{\\}{} % line break
-Jean-Noël Grad\authorOrcid{0000-0002-5821-4912}\authorRef{7},
 Magnus Hagdorn\authorOrcid{0000-0002-5076-4864}\authorRef{8},

### 2025-09-22 correspondence/rebuttal-1.md
Update rebuttal notes.

@@
 
+Thank you for highlighting this.
+We have added some text to the second paragraph of the introduction highlighting that with the paper focusing on the outputs of the German RSE conference, the thoughts and perspectives presented are largely German-focused.
+However, we also note that this work has broader relevance and we aim to highlight resources from the wider community where relevant.
+
 > 2.
@@
 Do they no expect at least some RSE professors?
-
-Thanks for bringing this up. We have added that aspect in such a way to not sound too Germany specific.
 

### 2025-09-23 future_challenges_for_RSEng.tex
work out today's idea

--- 2025-09-22 future_challenges_for_RSEng.tex
@@
 \texorpdfstring{\\}{} % line break
-Frank Löffler\authorOrcid{0000-0001-6643-6323}\authorRef{11}\textsuperscript{,}\authorRef{12},
-Michele Martone\authorOrcid{0000-0003-3239-8554}\authorRef{13},
-Jan Philipp Thiele\authorOrcid{0000-0002-8901-6660}\authorRef{10}\textsuperscript{,}\authorRef{14}
+Frank Löffler\authorOrcid{0000-0001-6643-6323}\authorRef{11},
+Michele Martone\authorOrcid{0000-0003-3239-8554}\authorRef{12},
+Jan Philipp Thiele\authorOrcid{0000-0002-8901-6660}\authorRef{10}\textsuperscript{,}\authorRef{13}
 } % Authors and references to addresses
@@
 \autlabel{10} Technische Universität Braunschweig, Germany\\
-\autlabel{11} Competence Center Digital Research, Friedrich Schiller University Jena, Germany\\
-\autlabel{12} de-RSE e.V.---Society for Research Software in Germany\\
-\autlabel{13} Independent researcher, Germany\\
-\autlabel{14} Weierstrass Institute, Berlin, Germany;
+\autlabel{11} Michael Stifel Center Jena {\&} Friedrich Schiller University Jena, Germany\\
+\autlabel{12} Independent researcher, Germany\\
+\autlabel{13} Weierstrass Institute, Berlin, Germany;
               Leibniz University Hannover, Institute of Applied Mathematics, Scientific Computing, Hannover, Germany\\
@@
 
-\subsection{How have academic Institutions transformed by then?}
+\subsection{Transformations in Academic Institutions}
 \begin{whatis}{}{academic}
-We expect research software to be a central part of scientific practice, on the same level as data and publications.
+Research software is a central part of scientific practice, on the same level as data and publications.
 This will be supported by the first generations of RSE Master's graduates that find their way into research departments, and accelerate the pace of science.
-Similar to RSEs, who are widely known and understood by university boards in 2035, we expect the concept of RSE units to be in the minds
-of academic staff at any level.
-We hope to see that universities actively use the presence of an RSE Department as a means to differentiate themselves from other universities.
-We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
-We also hope that specialised clusters form, which pool the skills of certain specialised RSEs with certain RSE units specializing in certain tasks like data management, HPC, or visualisation.
-At universities which get such RSE units, we want to see that researchers naturally accept them as part of their available services.
+Similar to RSEs, who are widely known and understood by university boards in 2035, academic staff members at any level are aware of the concept of RSE units.
+Universities actively use the presence of an RSE Department as a means to differentiate themselves from other universities.
+Some central RSE units will have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
+Specialised clusters have formed, which pool the skills of certain specialised RSEs with certain RSE units specializing in certain tasks like data management, HPC, or visualisation.
+At universities which create such RSE units, it can be observed that researchers naturally accept them as part of their available services.
 Additionally, researchers will seamlessly interact with infrastructures for using research software.
@@
 This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.
-We have already seen some specialisation trends in recent years,
+Some specialisation trends have already been observed in recent years,
 such as the HPC-RSE and AI-RSE, which are quite recent developments,
@@
 \begin{whatis}{}{career-paths}
-We would like to see the emergence of more dedicated staff, i.e. other (digital) RTPs next to the RSE.
-Also we hope that researchers can build careers at the interface of academia, industry, and policy (\eg\ in think tanks, NGOs,
-or organisations that do knowledge transfer).
+As the RSE becomes established, we are also seeing the emergence of a wider range of more specialized, technical roles within the research community, i.e. other (digital) RTPs.
+The legal requirements for employing people have become flexible enough to allow for the employment of more RTPs.
+Work outside of the publication record is more readily accepted and enables researchers to build careers at the interface of academia, industry, and policy
+(\eg\ in think tanks, NGOs, or organisations that do knowledge transfer).
 \end{whatis}
@@
 and potentially request and coordinate support from external RSEs.
+Within Germany at least, the distinction around whether these positions are classed as research staff or technical staff has so far created some discussion.
+The authors hope that, 10 years from now, this discussion has resolved towards a solution that enables the employment of RTPs more flexibly,
+and thereby recognizing that the distinction between research staff and technical staff is too restrictive.
+A similar discussion is currently ongoing in the UK with initiatives like \emph{the Technician's Commitment}~\cite{techniciancommitment}, and \emph{the Hidden Ref}~\cite{hiddenref},
+and finding a resolution with having new job titles, pay grades, and career options according to the level of work that you do.
+The authors' expect these initiatives to spread and find a wider resonance over the world till 2035.
 

### 2025-09-23 future_challenges_for_RSEng.tex
Mli 1 (#65)
* sort out box 1 of issue 52

* resolve box 2

* sort out Comment-3

* sort out Comment-5

* sort out Box-6

* sort out Comment-8

* sort out Comment-9

* sort out Comment-10

* sort out Comment-11


Co-authored-by: Jeremy Cohen <jcohen02@users.noreply.github.com>

Co-authored-by: Magnus Hagdorn <Magnus.Hagdorn@marsupium.org>

* take care of jeremys comment

* work out today's idea

---------

Co-authored-by: Florian Goth <florian.goth@uni-wuerzburg.de>
Co-authored-by: Jeremy Cohen <jcohen02@users.noreply.github.com>
Co-authored-by: Magnus Hagdorn <Magnus.Hagdorn@marsupium.org>

@@
 Gerasimos Chourdakis\authorOrcid{0000-0002-3977-1385}\authorRef{3},
+Simon Christ\authorOrcid{0000-0002-5866-1472}\authorRef{4},
 \texorpdfstring{\\}{} % line break
-Simon Christ\authorOrcid{0000-0002-5866-1472}\authorRef{4},
 Jeremy Cohen\authorOrcid{0000-0003-4312-2537}\authorRef{5},
 Nicholas A. Del Grosso\thinspace\authorRef{6},
+Jean-Noël Grad\authorOrcid{0000-0002-5821-4912}\authorRef{7},
 \texorpdfstring{\\}{} % line break
-Jean-Noël Grad\authorOrcid{0000-0002-5821-4912}\authorRef{7},
 Magnus Hagdorn\authorOrcid{0000-0002-5076-4864}\authorRef{8},
 Toby Hodges\authorOrcid{0000-0003-1766-456X}\authorRef{9},
-Jan Linxweiler\authorOrcid{0000-0002-2755-5087}\authorRef{10},
+Jan Linxweiler\authorOrcid{0000-0002-2755-5087}\authorRef{10}\textsuperscript{,}\authorRef{11},
 \texorpdfstring{\\}{} % line break
-Frank Löffler\authorOrcid{0000-0001-6643-6323}\authorRef{11},
-Michele Martone\authorOrcid{0000-0003-3239-8554}\authorRef{12},
-Jan Philipp Thiele\authorOrcid{0000-0002-8901-6660}\authorRef{10}\textsuperscript{,}\authorRef{13}
+Frank Löffler\authorOrcid{0000-0001-6643-6323}\authorRef{12}\textsuperscript{,}\authorRef{11},
+Michele Martone\authorOrcid{0000-0003-3239-8554}\authorRef{13},
+Jan Philipp Thiele\authorOrcid{0000-0002-8901-6660}\authorRef{10}\textsuperscript{,}\authorRef{14}
 } % Authors and references to addresses
@@
 \autlabel{10} Technische Universität Braunschweig, Germany\\
-\autlabel{11} Michael Stifel Center Jena {\&} Friedrich Schiller University Jena, Germany\\
-\autlabel{12} Independent researcher, Germany\\
-\autlabel{13} Weierstrass Institute, Berlin, Germany;
+\autlabel{11} de-RSE e.V.---Society for Research Software in Germany\\
+\autlabel{12} Competence Center Digital Research, Friedrich Schiller University Jena, Germany\\
+\autlabel{13} Independent researcher, Germany\\
+\autlabel{14} Weierstrass Institute, Berlin, Germany;
               Leibniz University Hannover, Institute of Applied Mathematics, Scientific Computing, Hannover, Germany\\

### 2025-09-23 correspondence/rebuttal-1.md

--- 2025-09-19 correspondence/rebuttal-1.md
@@
 
+> Page 10
+> Life would be so much easier if at least university regulations would be homogenised, no? - reword as Life would be so much easier if at least university regulations were homogenised.
+
+fixed as per comment.
+
+> As per previous comments, wording like “we would like to see” should be removed from the Achieved Goals box
+
+Reworded the box accordingly.
+
+> Page 11
+> Nevertheless, there are some things that we declared out of scope here. Among that is the ongoing digital transformation in wider society. – Reword as one sentence
+
+Reworded the sentence accordingly.
+
+> Page 12
+> In summary, RSEs will very likely be doing different work
+> to what they are doing today - remove this line break
+
+Not sure, how this crept into the final version... It's gone now.
+
 \newpage

### 2025-09-23 future_challenges_for_RSEng.tex

@@
 Further risks that were later discussed include an intensified replication crisis,
-a de-democratization of research, and loss of trust in science.
+making research less accessible, and loss of trust in science.
 Due to the non-deterministic nature of generative AI, we expect AI-assisted simulations

### 2025-09-23 future_challenges_for_RSEng.tex
typos

@@
 At the same time, we do recognise that the application of core software development best practices
-to enhance and address problems with research codebases, that can often make up the bulk of
-current RSE work, will change.
+to enhance and address problems with research codes, that can often make up the bulk of current RSE work, will change.
 As noted above, these are skills that we would expect any researcher who writes code to have in 10 years time.
@@
 Some central RSE units will have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
-Specialised clusters have formed, which pool the skills of certain specialised RSEs with certain RSE units specializing in certain tasks like data management, HPC, or visualisation.
+Specialised clusters have formed, which pool the skills of certain specialised RSEs with certain RSE units specialising in certain tasks like data management, HPC, or visualisation.
 At universities which create such RSE units, it can be observed that researchers naturally accept them as part of their available services.
@@
 \begin{whatis}{}{career-paths}
-As the RSE becomes established, we are also seeing the emergence of a wider range of more specialized, technical roles within the research community, i.e. other (digital) RTPs.
+As the RSE becomes established, we are also seeing the emergence of a wider range of more specialised, technical roles within the research community, i.e. other (digital) RTPs.
 The legal requirements for employing people have become flexible enough to allow for the employment of more RTPs.
@@
 The authors hope that, 10 years from now, this discussion has resolved towards a solution that enables the employment of RTPs more flexibly,
-and thereby recognizing that the distinction between research staff and technical staff is too restrictive.
+and thereby recognising that the distinction between research staff and technical staff is too restrictive.
 A similar discussion is currently ongoing in the UK with initiatives like \emph{the Technician's Commitment}~\cite{techniciancommitment}, and \emph{the Hidden Ref}~\cite{hiddenref},

### 2025-09-23 correspondence/rebuttal-1.md

@@
 
-> Regarding Chaper 2.6
+> Regarding Chapter 2.6
 > "thinking on the time scale of generations".
@@
 
-> Chaper 2.7
+Thanks for bringing this up. We have added that aspect in such a way to not sound too Germany specific.
+
+> Chapter 2.7
 > "in lower tiers of an of a research organisation", sentence needs attention

### 2025-09-23 future_challenges_for_RSEng.tex

@@
 Additionally, researchers will seamlessly interact with infrastructures for using research software.
+Replacement of staff reaching retirement has led to the creation of more RSE professorships.
 \end{whatis}
@@
 has organisationally often adopted time scales for change that are more reminiscent
-to institutions thinking on the time scale of generations.
-Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref}
+to institutions thinking on the time scale of societal generations.
+While sounding negative, this can also be an opportunity,
+if in the time-frame of the next decade, a large number of permanent staff reach retirement and are replaced by a new generation of researchers and technical professionals who have seen the
+digital transformation first-hand and have actively participated in it through various grassroots initiatives.\\
+On the other hand, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref}
 hints at the potential for specialisation of research tasks.

### 2025-09-23 future_challenges_for_RSEng.tex
bring in yesterday's discussions

@@
 to institutions thinking on the time scale of societal generations.
-While sounding negative, this can also be an opportunity,
-if in the time-frame of the next decade, a large number of permanent staff reach retirement and are replaced by a new generation of researchers and technical professionals who have seen the
-digital transformation first-hand and have actively participated in it through various grassroots initiatives.\\
+While sounding negative, this can also be an opportunity that should be used by the RSE community.
+If in the time-frame of the next decade, a large number of permanent staff reach retirement and
+can be replaced by a new generation of researchers and technical professionals who have seen the
+digital transformation first-hand and maybe have actively participated in it through various grassroots initiatives.
+This generational change opens up the opportunity for an active change within the research landscape.
+A simple possibility would be to realign newly appointed professorships with a focus towards software development within their domain.
+A second more progressive step would be to actually appoint new cross-faculty professorships.
 On the other hand, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref}
@@
 We note the SE community has also independently expressed interest in creating an \emph{infrastructure engineering} role~\cite{Sochat2024Infra}
-and funding bodies seem willing to promote such career pathways~\cite{UKRI2025Infra,EuroHPC2024Federated},
+and funding bodies seem willing to promote such career pathways~\cite{UKRI2025Infra,EuroHPC2024Federated}.
+Taking these trends together the generational change should be used as an opportunity to reshape organisational structures to better reflect the work people in research
+are actually doing. The RSE community should take an active role in shaping that discussion in this direction till 2035.
 % European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories

### 2025-09-24 future_challenges_for_RSEng.tex
feat(#44): fix pronouns

--- 2025-09-23 future_challenges_for_RSEng.tex
@@
 Kim develops advanced scientific software in tandem with an AI agent that codes,
-debugs, and simulates experiments alongside him. The AI proactively suggests optimisations,
+debugs, and simulates experiments alongside her. The AI proactively suggests optimisations,
 generates reproducible workflows, and adapts to Kim’s coding style and domain-specific needs.
@@
 Kay had left academia to work in an industrial R\&D department because of a more reliable employment.
-There he gained valuable experience working in an industrial setting.
-However, he was very happy when an opportunity arose to work in a central
-RSE department with an open-ended contract. Apart from his main responsibility
+There she gained valuable experience working in an industrial setting.
+However, she was very happy when an opportunity arose to work in a central
+RSE department with an open-ended contract. Apart from her main responsibility
 of developing and maintaining software for an affordable water monitoring system
-used in the global south, he now also teaches undergraduates digital literacy.
+used in the global south, she now also teaches undergraduates digital literacy.
 \end{story}
@@
 Avalon which was interested in setting up their own department for RSEs in order to not fall behind the university of Arcadia.
-Kim builds upon the model set forth by the University of Arcadia based on his valuable experience. Since the university
+Kim builds upon the model set forth by the University of Arcadia based on her valuable experience. Since the university
 of Avalon has to follow different legal and administrative rules, not everything can be straightforwardly adapted.

### 2025-09-24 future_challenges_for_RSEng.tex
feat(#44): fix tense

@@
 \begin{story}{The Definition of RSEs}{defrse}
-Kim and Kay chuckled when they looked at their original job descriptions from 2020: ``build tools for scientists, manage data, write reusable code.''
+Kim and Kay chuckle when they looked at their original job descriptions from 2020: ``build tools for scientists, manage data, write reusable code.''
 Now, their roles are unrecognisable: half AI whisperer, half research ethicist, and part-time diplomat negotiating between synthetic and human collaborators.
@@
 Despite the tech upheaval, their deep knowledge of scientific workflows and the messy, human way science actually gets done remains indispensable.
-When labs faltered under algorithmic bias or unclear model provenance, it was Kim and Kay who transformed the chaos back into meaningful order.
+When labs falter under algorithmic bias or unclear model provenance, it is Kim and Kay who transform the chaos back into meaningful order.
 They understand that science does not move in straight lines, and that trust between collaborators, institutions, and even machines has to be built, not coded.

### 2025-09-24 future_challenges_for_RSEng.tex
feat(#44): fix tense

@@
 required writing complex code even for simple and common analysis tasks.
-Nowadays, Kim could have started directly from a template based on established toolkits and workflows,
+Nowadays, Kim is able to start directly from a template based on established toolkits and workflows,
 while kay-bio accepts configuration in a simpler, standardised format and generative AI assistants

### 2025-09-24 future_challenges_for_RSEng.tex
split sentence

@@
 Nowadays, Kim is able to start directly from a template based on established toolkits and workflows,
-while kay-bio accepts configuration in a simpler, standardised format and generative AI assistants
-can guide new users in preparing and debugging their analyses.
+while kay-bio accepts configuration in a simpler, standardised format.
+Generative AI assistants can guide new users in preparing and debugging their analyses.
 Meanwhile, new RSEs can very easily integrate libkim, kay-bio, and a multitude

### 2025-09-24 future_challenges_for_RSEng.tex
feat(#44): fix tense
swap words

@@
 so far future projections are far less common for the RSEng domain.
-We present here the results of a workshop discussion (notes content is archived here \cite{goth2025_WSPad}) at the
+Here, we present the results of a workshop discussion (notes content is archived here \cite{goth2025_WSPad}) at the
 deRSE25 conference in Karlsruhe\footnote{deRSE25, session ``The End of {RSEng}?
@@
 Ken, though trained as a physicist, found his passion not in theory,
-but in keeping the quantum computers humming. He became the go-to expert
+but in keeping the quantum computers humming. He is the go-to expert
 for troubleshooting entangled qubits and stabilising delicate cryogenic systems.

### 2025-09-24 future_challenges_for_RSEng.tex
feat(#44): fix tense

@@
 more sustainable, robust and maintainable.
-We also hope that research software becomes an accepted research artefact on an equal footing with data publications and traditional text publications.
+We also hope that research software will have become an accepted research artefact on an equal footing with data publications and traditional text publications.
 This is already happening in a number of ways,

### 2025-09-24 future_challenges_for_RSEng.tex
split long sentence

@@
 such as \eg\ Stack Overflow and Wikipedia (which are used to train AI models) poorer
-due to fewer incentives to maintain them, although a recent survey suggests
-these communities experience less activity from new users and an increased
-average complexity of the questions posted~\cite{Burtch2024}.
+due to fewer incentives to maintain them.
+However, a recent survey suggests these communities experience less activity from new
+users and an increased average complexity of the questions posted~\cite{Burtch2024}.
 The use of AI assistants in these communities could also compromise LLMs via `model collapse'

### 2025-09-24 future_challenges_for_RSEng.tex
split long sentence

@@
 other research output (i.e. publish the trained model, training data and training workflow).
-Future research will require significant computing and storage resources, as well as useful data and models,
-therefore it will be more likely to be driven by institutions that are able to provide these.
+Future research will require significant computing and storage resources, as well as useful data and models.
+Therefore, it will be more likely to be driven by institutions that are able to provide these.
 Finally, an increasing risk is the loss of trust in science,

### 2025-09-24 future_challenges_for_RSEng.tex
split long sentence

@@
 This divide-and-conquer approach allows the scientific community to address questions not previously within their grasp.
-When it comes to teaching, it is true that students need to learn how to handle complexity,
-by being able to develop deep understanding of individual components.
+When it comes to teaching, it is true that students need to learn how to handle complexity.
+They need to develop a deep understanding of the individual components.
 However, it is also important to learn how to manage complexity

### 2025-09-24 future_challenges_for_RSEng.tex
split long sentence

@@
 It is clear that research will need both RSEs that can dive deep into single projects of significant complexity,
-as well as RSEs that can manage complex systems of almost black-box components,
-and the future of RSE education should aim to prepare students for both roles.
+as well as RSEs that can manage complex systems of almost black-box components.
+The future of RSE education should aim to prepare students for both roles.
 We note members of the SE community independently reached the same conclusions

### 2025-09-24 future_challenges_for_RSEng.tex
replace word
drop paragraph as it is already a few lines above

@@
 The principles of Good Scientific Practice~\cite{dfg_gsp} have been a cornerstone in the conduct of research.
-But their relevance hinges on constant updates in light of how research is performed.
+However, their relevance hinges on constant updates in light of how research is performed.
 By 2035, accepted rules on the use of AI in research tasks and software development have been achieved.
@@
 Members of the SE community have also called for their field to embrace Open Science best practices\cite{OliveiraJr2024,Druskat2025}.
-
-A code of ethics and professional conduct for RSEs should be developed by institutions together with RSEs themselves.
-This can then form the basis for institutional policies and guidelines as well as serve as a background for individual developers to consider how software can be used.
 

### 2025-09-24 future_challenges_for_RSEng.tex
feat(#44): fix tense

@@
 Research software is a central part of scientific practice, on the same level as data and publications.
-This will be supported by the first generations of RSE Master's graduates that find their way into research departments, and accelerate the pace of science.
+This is supported by the first generations of RSE Master's graduates that find their way into research departments, and accelerate the pace of science.
 Similar to RSEs, who are widely known and understood by university boards in 2035, academic staff members at any level are aware of the concept of RSE units.
 Universities actively use the presence of an RSE Department as a means to differentiate themselves from other universities.
-Some central RSE units will have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
+Some central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
 Specialised clusters have formed, which pool the skills of certain specialised RSEs with certain RSE units specialising in certain tasks like data management, HPC, or visualisation.
 At universities which create such RSE units, it can be observed that researchers naturally accept them as part of their available services.
-Additionally, researchers will seamlessly interact with infrastructures for using research software.
+Additionally, researchers interact seamlessly with infrastructures for using research software.
 \end{whatis}

### 2025-09-24 future_challenges_for_RSEng.tex
reduce the number of certains

@@
 Some central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
-Specialised clusters have formed, which pool the skills of certain specialised RSEs with certain RSE units specialising in certain tasks like data management, HPC, or visualisation.
+Specialised clusters have formed, which pool the skills of specialised RSEs with RSE units specialising in certain tasks like data management, HPC, or visualisation.
 At universities which create such RSE units, it can be observed that researchers naturally accept them as part of their available services.

### 2025-09-24 future_challenges_for_RSEng.tex
feat(#44): fix tense
fixes(#56): sort out incomplete sentence

--- 2025-09-24 future_challenges_for_RSEng.tex
@@
 so far future projections are far less common for the RSEng domain.
-Here, we present the results of a workshop discussion (notes content is archived here \cite{goth2025_WSPad}) at the
+We present here the results of a workshop discussion (notes content is archived here \cite{goth2025_WSPad}) at the
 deRSE25 conference in Karlsruhe\footnote{deRSE25, session ``The End of {RSEng}?
@@
 more sustainable, robust and maintainable.
-We also hope that research software will have become an accepted research artefact on an equal footing with data publications and traditional text publications.
+We also hope that research software becomes an accepted research artefact on an equal footing with data publications and traditional text publications.
 This is already happening in a number of ways,
@@
 \begin{story}{Introduction}{intro}
-In order to give a more concrete setting, imagine a world ten years from now.
-Not too overwhelmingly far ahead in the future, but distant enough for some changes to have occurred.
+In order to give a more concrete setting, imagine a world ten years from now 
+which is not too overwhelmingly far ahead in the future, but distant enough for some changes to have occurred.
 Over this time frame Kim~\cite{Anzt2021} and Kay~\cite{Goth2024},
@@
 \begin{story}{The Definition of RSEs}{defrse}
-Kim and Kay chuckle when they looked at their original job descriptions from 2020: ``build tools for scientists, manage data, write reusable code.''
+Kim and Kay chuckled when they looked at their original job descriptions from 2020: ``build tools for scientists, manage data, write reusable code.''
 Now, their roles are unrecognisable: half AI whisperer, half research ethicist, and part-time diplomat negotiating between synthetic and human collaborators.
@@
 Despite the tech upheaval, their deep knowledge of scientific workflows and the messy, human way science actually gets done remains indispensable.
-When labs falter under algorithmic bias or unclear model provenance, it is Kim and Kay who transform the chaos back into meaningful order.
+When labs faltered under algorithmic bias or unclear model provenance, it was Kim and Kay who transformed the chaos back into meaningful order.
 They understand that science does not move in straight lines, and that trust between collaborators, institutions, and even machines has to be built, not coded.
@@
 such as \eg\ Stack Overflow and Wikipedia (which are used to train AI models) poorer
-due to fewer incentives to maintain them.
-However, a recent survey suggests these communities experience less activity from new
-users and an increased average complexity of the questions posted~\cite{Burtch2024}.
+due to fewer incentives to maintain them, although a recent survey suggests
+these communities experience less activity from new users and an increased
+average complexity of the questions posted~\cite{Burtch2024}.
 The use of AI assistants in these communities could also compromise LLMs via `model collapse'
@@
 other research output (i.e. publish the trained model, training data and training workflow).
-Future research will require significant computing and storage resources, as well as useful data and models.
-Therefore, it will be more likely to be driven by institutions that are able to provide these.
+Future research will require significant computing and storage resources, as well as useful data and models,
+therefore it will be more likely to be driven by institutions that are able to provide these.
 Finally, an increasing risk is the loss of trust in science,
@@
 Kim develops advanced scientific software in tandem with an AI agent that codes,
-debugs, and simulates experiments alongside her. The AI proactively suggests optimisations,
+debugs, and simulates experiments alongside him. The AI proactively suggests optimisations,
 generates reproducible workflows, and adapts to Kim’s coding style and domain-specific needs.
@@
 This divide-and-conquer approach allows the scientific community to address questions not previously within their grasp.
-When it comes to teaching, it is true that students need to learn how to handle complexity.
-They need to develop a deep understanding of the individual components.
+When it comes to teaching, it is true that students need to learn how to handle complexity,
+by being able to develop deep understanding of individual components.
 However, it is also important to learn how to manage complexity
@@
 It is clear that research will need both RSEs that can dive deep into single projects of significant complexity,
-as well as RSEs that can manage complex systems of almost black-box components.
-The future of RSE education should aim to prepare students for both roles.
+as well as RSEs that can manage complex systems of almost black-box components,
+and the future of RSE education should aim to prepare students for both roles.
 We note members of the SE community independently reached the same conclusions
@@
 required writing complex code even for simple and common analysis tasks.
-Nowadays, Kim is able to start directly from a template based on established toolkits and workflows,
-while kay-bio accepts configuration in a simpler, standardised format.
-Generative AI assistants can guide new users in preparing and debugging their analyses.
+Nowadays, Kim could have started directly from a template based on established toolkits and workflows,
+while kay-bio accepts configuration in a simpler, standardised format and generative AI assistants
+can guide new users in preparing and debugging their analyses.
 Meanwhile, new RSEs can very easily integrate libkim, kay-bio, and a multitude
@@
 The principles of Good Scientific Practice~\cite{dfg_gsp} have been a cornerstone in the conduct of research.
-However, their relevance hinges on constant updates in light of how research is performed.
+But their relevance hinges on constant updates in light of how research is performed.
 By 2035, accepted rules on the use of AI in research tasks and software development have been achieved.
@@
 
+A code of ethics and professional conduct for RSEs should be developed by institutions together with RSEs themselves.
+This can then form the basis for institutional policies and guidelines as well as serve as a background for individual developers to consider how software can be used.
+
 \begin{story}{The greater world of Kim and Kay}{ethics}
@@
 Kay had left academia to work in an industrial R\&D department because of a more reliable employment.
-There she gained valuable experience working in an industrial setting.
-However, she was very happy when an opportunity arose to work in a central
-RSE department with an open-ended contract. Apart from her main responsibility
+There he gained valuable experience working in an industrial setting.
+However, he was very happy when an opportunity arose to work in a central
+RSE department with an open-ended contract. Apart from his main responsibility
 of developing and maintaining software for an affordable water monitoring system
-used in the global south, she now also teaches undergraduates digital literacy.
+used in the global south, he now also teaches undergraduates digital literacy.
 \end{story}
@@
 Avalon which was interested in setting up their own department for RSEs in order to not fall behind the university of Arcadia.
-Kim builds upon the model set forth by the University of Arcadia based on her valuable experience. Since the university
+Kim builds upon the model set forth by the University of Arcadia based on his valuable experience. Since the university
 of Avalon has to follow different legal and administrative rules, not everything can be straightforwardly adapted.
@@
 Ken, though trained as a physicist, found his passion not in theory,
-but in keeping the quantum computers humming. He is the go-to expert
+but in keeping the quantum computers humming. He became the go-to expert
 for troubleshooting entangled qubits and stabilising delicate cryogenic systems.

### 2025-09-24 future_challenges_for_RSEng.tex
fixes(#56): rename goal boxes

@@
 \newcounter{whatiscounter}
-\newtcbtheorem[use counter=whatiscounter, number format=\Alph]{whatis}{Achieved Goals}{
+\newtcbtheorem[use counter=whatiscounter, number format=\Alph]{whatis}{RSEng in 2035}{
   enhanced,
@@
 \begin{whatis}{}{whatisexplanation}
-In order to list what has been achieved in a certain aspect within the RSE community of the future of 2035 we use these ``Goal'' boxes.
+In order to list what has been achieved in a certain aspect within the RSE community of the future of 2035 we use these ``RSEng in 2035'' boxes.
 \end{whatis}

### 2025-09-24 future_challenges_for_RSEng.tex
fixes(#56): swap sentences around to make it clearer that now is 2025

@@
 
+Today, the Carpentries now offer specialised software workshops
+for data scientists\footnote{\url{https://datacarpentry.org}},
+librarians\footnote{\url{https://librarycarpentry.org}},
+and soon High-Performance Computing (HPC) practitioners\footnote{\url{https://www.hpc-carpentry.org}}.
 RSEng-relevant open educational resources and Master's programmes are tracked
@@
 and UK SSI resources hub\footnote{\url{https://www.software.ac.uk/resource-hub}}.
-The Carpentries now offer specialised software workshops
-for data scientists\footnote{\url{https://datacarpentry.org}},
-librarians\footnote{\url{https://librarycarpentry.org}},
-and soon High-Performance Computing (HPC) practitioners\footnote{\url{https://www.hpc-carpentry.org}}.
 

### 2025-09-24 future_challenges_for_RSEng.tex

@@
 required writing complex code even for simple and common analysis tasks.
-Nowadays, Kim could have started directly from a template based on established toolkits and workflows,
-while kay-bio accepts configuration in a simpler, standardised format and generative AI assistants
-can guide new users in preparing and debugging their analyses.
+Nowadays, Kim is able to start directly from a template based on established toolkits and workflows,
+while kay-bio accepts configuration in a simpler, standardised format.
+Generative AI assistants can guide new users in preparing and debugging their analyses.
 Meanwhile, new RSEs can very easily integrate libkim, kay-bio, and a multitude
@@
 
-A code of ethics and professional conduct for RSEs should be developed by institutions together with RSEs themselves.
-This can then form the basis for institutional policies and guidelines as well as serve as a background for individual developers to consider how software can be used.
-
 \begin{story}{The greater world of Kim and Kay}{ethics}
@@
 Research software is a central part of scientific practice, on the same level as data and publications.
-This will be supported by the first generations of RSE Master's graduates that find their way into research departments, and accelerate the pace of science.
+This is supported by the first generations of RSE Master's graduates that find their way into research departments, and accelerate the pace of science.
 Similar to RSEs, who are widely known and understood by university boards in 2035, academic staff members at any level are aware of the concept of RSE units.
 Universities actively use the presence of an RSE Department as a means to differentiate themselves from other universities.
-Some central RSE units will have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
-Specialised clusters have formed, which pool the skills of certain specialised RSEs with certain RSE units specialising in certain tasks like data management, HPC, or visualisation.
+Some central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
+Specialised clusters have formed, which pool the skills of specialised RSEs with RSE units specialising in certain tasks like data management, HPC, or visualisation.
 At universities which create such RSE units, it can be observed that researchers naturally accept them as part of their available services.
-Additionally, researchers will seamlessly interact with infrastructures for using research software.
+Additionally, researchers interact seamlessly with infrastructures for using research software.
 \end{whatis}
@@
 and finding a resolution with having new job titles, pay grades, and career options according to the level of work that you do.
-The authors' expect these initiatives to spread and find a wider resonance over the world till 2035.
+The authors expect these initiatives will have spread and will have found a wider resonance over the world by 2035.
 

### 2025-09-24 future_challenges_for_RSEng.tex
minor fix

@@
 so far future projections are far less common for the RSEng domain.
-Here, we present the results of a workshop discussion (notes content is archived here \cite{goth2025_WSPad}) at the
+Here, we present the results of a workshop discussion (notes are archived here \cite{goth2025_WSPad}) at the
 deRSE25 conference in Karlsruhe\footnote{deRSE25, session ``The End of {RSEng}?

### 2025-09-24 future_challenges_for_RSEng.tex
less hope

@@
 
-Looking ahead over a 10-year time frame, we hope that the RSE community will have
+Looking ahead over a 10-year time frame, we expect that the RSE community will have
 contributed to making research software and its outputs
 more sustainable, robust and maintainable.
-We also hope that research software will have become an accepted research artefact on an equal footing with data publications and traditional text publications.
+We hope that research software will have become an accepted research artefact on an equal footing with data publications and traditional text publications.
 This is already happening in a number of ways,

### 2025-09-24 future_challenges_for_RSEng.tex
be more assertive

@@
 So where do RSEs fit into this picture?
-Even today, we expect to see RSEs providing technical skills and input to research activities in a range of different areas, including:
+Already today, RSEs are providing technical skills and input to research activities in a range of different areas, including:
 high-quality, robust architecture and design for research software;

### 2025-09-24 future_challenges_for_RSEng.tex
make it clearer that the workshop discussions are meant

@@
 % Digitalisation will also feature more heavily in the domain curricula.
-In this discussions the workshop participants found that
+During the discussion the workshop participants found that
 the digital skills of domain scientists will increase,

### 2025-09-24 future_challenges_for_RSEng.tex
reword and remove redundancy

@@
 and with their skills in managing hierarchical systems of complex software components with clear interfaces,
-RSEs can now build reliable, maintainable systems of a scale and capability not previously possible.
+RSEs can now build reliable, maintainable systems of a scale and capability hitherto not possible.
 \end{whatis}
@@
 By reducing the complexity of individual software components,
-we are now able to develop systems of complexity and capability not previously possible.
-This divide-and-conquer approach allows the scientific community to address questions not previously within their grasp.
+we are now able to develop systems of complexity and capability hitherto not possible.
+This divide-and-conquer approach allows the scientific community to address ever larger and more complex problems.
 When it comes to teaching, it is true that students need to learn how to handle complexity.

### 2025-09-24 future_challenges_for_RSEng.tex
insert word
insert comma

@@
 This divide-and-conquer approach allows the scientific community to address ever larger and more complex problems.
-When it comes to teaching, it is true that students need to learn how to handle complexity.
+When it comes to teaching, it is true that students need to learn how to handle this complexity.
 They need to develop a deep understanding of the individual components.
@@
 climate~\cite{Lannelongue2023} and more.
-Like all areas of research and engineering RSEs are faced with the dual use dilemma~\cite{Bobier2024} where artefacts they develop can be used for military purposes.
+Like all areas of research and engineering, RSEs are faced with the dual use dilemma~\cite{Bobier2024} where artefacts they develop can be used for military purposes.
 As engineers, RSEs take the responsibility for the software they develop.

### 2025-09-24 future_challenges_for_RSEng.tex
drop sentence, alternatively we should rewrite this
fix grammar

@@
 This is an opportunity that would not be possible if relying solely on short-term funded, isolated RSEs.
-We hope that a change of culture in research about the acceptance of research software has happened.
 
@@
 Within Germany at least, the distinction around whether these positions are classed as research staff or technical staff has so far created some discussion.
-The authors hope that, 10 years from now, this discussion has resolved towards a solution that enables the employment of RTPs more flexibly,
+The authors hope that, 10 years from now, this discussion has been resolved towards a solution that enables the employment of RTPs more flexibly,
 and thereby recognising that the distinction between research staff and technical staff is too restrictive.

### 2025-09-24 future_challenges_for_RSEng.tex
less hope

@@
 
-We hope that the standard of research software will have risen across the board,
+We anticipate that the standard of research software will have risen across the board,
 enabling easier reuse of prior software outputs in new research.
-We also hope that a cultural change is taking place to recognise research software
+We hope that a cultural change is taking place to recognise research software
 as an integral part of science that it is widely recognised and rewarded across

### 2025-09-25 future_challenges_for_RSEng.tex
Update future_challenges_for_RSEng.tex
Co-authored-by: Gerasimos Chourdakis <gerasimos.chourdakis@ipvs.uni-stuttgart.de>

--- 2025-09-24 future_challenges_for_RSEng.tex
@@
 \newcounter{whatiscounter}
-\newtcbtheorem[use counter=whatiscounter, number format=\Alph]{whatis}{RSEng in 2035}{
+\newtcbtheorem[use counter=whatiscounter, number format=\Alph]{whatis}{Achieved Goals}{
   enhanced,
@@
 so far future projections are far less common for the RSEng domain.
-Here, we present the results of a workshop discussion (notes are archived here \cite{goth2025_WSPad}) at the
+We present here the results of a workshop discussion (notes content is archived here \cite{goth2025_WSPad}) at the
 deRSE25 conference in Karlsruhe\footnote{deRSE25, session ``The End of {RSEng}?
@@
 \begin{whatis}{}{whatisexplanation}
-In order to list what has been achieved in a certain aspect within the RSE community of the future of 2035 we use these ``RSEng in 2035'' boxes.
+In order to list what has been achieved in a certain aspect within the RSE community of the future of 2035 we use these ``Goal'' boxes.
 \end{whatis}
 
-Looking ahead over a 10-year time frame, we expect that the RSE community will have
+Looking ahead over a 10-year time frame, we hope that the RSE community will have
 contributed to making research software and its outputs
 more sustainable, robust and maintainable.
-We hope that research software will have become an accepted research artefact on an equal footing with data publications and traditional text publications.
+We also hope that research software becomes an accepted research artefact on an equal footing with data publications and traditional text publications.
 This is already happening in a number of ways,
@@
 So where do RSEs fit into this picture?
-Already today, RSEs are providing technical skills and input to research activities in a range of different areas, including:
+Even today, we expect to see RSEs providing technical skills and input to research activities in a range of different areas, including:
 high-quality, robust architecture and design for research software;
@@
 \begin{story}{Introduction}{intro}
-In order to give a more concrete setting, imagine a world ten years from now 
-which is not too overwhelmingly far ahead in the future, but distant enough for some changes to have occurred.
+In order to give a more concrete setting, imagine a world ten years from now.
+Not too overwhelmingly far ahead in the future, but distant enough for some changes to have occurred.
 Over this time frame Kim~\cite{Anzt2021} and Kay~\cite{Goth2024},
@@
 % Digitalisation will also feature more heavily in the domain curricula.
-During the discussion the workshop participants found that
+In this discussions the workshop participants found that
 the digital skills of domain scientists will increase,
@@
 
-Today, the Carpentries now offer specialised software workshops
+RSEng-relevant open educational resources and Master's programmes are tracked
+in the Learning and Teaching RSE database\footnote{\url{https://de-rse.org/learn-and-teach/}}
+and UK SSI resources hub\footnote{\url{https://www.software.ac.uk/resource-hub}}.
+The Carpentries now offer specialised software workshops
 for data scientists\footnote{\url{https://datacarpentry.org}},
@@
 and soon High-Performance Computing (HPC) practitioners\footnote{\url{https://www.hpc-carpentry.org}}.
-RSEng-relevant open educational resources and Master's programmes are tracked
-in the Learning and Teaching RSE database\footnote{\url{https://de-rse.org/learn-and-teach/}}
-and UK SSI resources hub\footnote{\url{https://www.software.ac.uk/resource-hub}}.
 
 \begin{story}{The Definition of RSEs}{defrse}
-Kim and Kay chuckle when they looked at their original job descriptions from 2020: ``build tools for scientists, manage data, write reusable code.''
+Kim and Kay chuckled when they looked at their original job descriptions from 2020: ``build tools for scientists, manage data, write reusable code.''
 Now, their roles are unrecognisable: half AI whisperer, half research ethicist, and part-time diplomat negotiating between synthetic and human collaborators.
@@
 and with their skills in managing hierarchical systems of complex software components with clear interfaces,
-RSEs can now build reliable, maintainable systems of a scale and capability hitherto not possible.
+RSEs can now build reliable, maintainable systems of a scale and capability not previously possible.
 \end{whatis}
@@
 By reducing the complexity of individual software components,
-we are now able to develop systems of complexity and capability hitherto not possible.
-This divide-and-conquer approach allows the scientific community to address ever larger and more complex problems.
-When it comes to teaching, it is true that students need to learn how to handle this complexity.
-They need to develop a deep understanding of the individual components.
+we are now able to develop systems of complexity and capability not previously possible.
+This divide-and-conquer approach allows the scientific community to address questions not previously within their grasp.
+When it comes to teaching, it is true that students need to learn how to handle complexity,
+by being able to develop deep understanding of individual components.
 However, it is also important to learn how to manage complexity
@@
 climate~\cite{Lannelongue2023} and more.
-Like all areas of research and engineering, RSEs are faced with the dual use dilemma~\cite{Bobier2024} where artefacts they develop can be used for military purposes.
+Like all areas of research and engineering RSEs are faced with the dual use dilemma~\cite{Bobier2024} where artefacts they develop can be used for military purposes.
 As engineers, RSEs take the responsibility for the software they develop.
@@
 Research software is a central part of scientific practice, on the same level as data and publications.
-This is supported by the first generations of RSE Master's graduates that find their way into research departments, and accelerate the pace of science.
+This will be supported by the first generations of RSE Master's graduates that find their way into research departments, and accelerate the pace of science.
 Similar to RSEs, who are widely known and understood by university boards in 2035, academic staff members at any level are aware of the concept of RSE units.
 Universities actively use the presence of an RSE Department as a means to differentiate themselves from other universities.
-Some central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
-Specialised clusters have formed, which pool the skills of specialised RSEs with RSE units specialising in certain tasks like data management, HPC, or visualisation.
+Some central RSE units will have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
+Specialised clusters have formed, which pool the skills of certain specialised RSEs with certain RSE units specialising in certain tasks like data management, HPC, or visualisation.
 At universities which create such RSE units, it can be observed that researchers naturally accept them as part of their available services.
-Additionally, researchers interact seamlessly with infrastructures for using research software.
+Additionally, researchers will seamlessly interact with infrastructures for using research software.
+Replacement of staff reaching retirement, combined with the intensified research interest in RSE, has led to the creation of more RSE professorships.
 \end{whatis}
@@
 has organisationally often adopted time scales for change that are more reminiscent
-to institutions thinking on the time scale of generations.
-Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref}
+to institutions thinking on the time scale of societal generations.
+While sounding negative, this can also be an opportunity that should be used by the RSE community.
+If in the time-frame of the next decade, a large number of permanent staff reach retirement and
+can be replaced by a new generation of researchers and technical professionals who have seen the
+digital transformation first-hand and maybe have actively participated in it through various grassroots initiatives.
+This generational change opens up the opportunity for an active change within the research landscape.
+A simple possibility would be to realign newly appointed professorships with a focus towards software development within their domain.
+A second more progressive step would be to actually appoint new cross-faculty professorships.
+On the other hand, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref}
 hints at the potential for specialisation of research tasks.
@@
 and finding a resolution with having new job titles, pay grades, and career options according to the level of work that you do.
-The authors expect these initiatives will have spread and will have found a wider resonance over the world by 2035.
+The authors' expect these initiatives to spread and find a wider resonance over the world till 2035.
 

### 2025-09-25 future_challenges_for_RSEng.tex

@@
 with increasing integration into various aspects of research.
-In Germany, a formal call for creating a sustainable environment for research software has only been raised by the RSE community a couple of years ago~\cite{Anzt2021},
-while the foundational competencies of an RSE have just recently been defined in detail~\cite{Goth2024}.
+In Germany, a formal call for creating a sustainable environment for research software
+was only published a couple of years ago~\cite{Anzt2021},
+while the foundational competencies of an RSE have only recently been defined in detail~\cite{Goth2024}.
 Central RSE units are in the process of being imagined~\cite{Kempf2025-draft},
@@
 and to advance the conversation about how RSEng should, or could, develop.
+Since this paper focuses on the outputs of the deRSE25 conference, we present our thoughts and perspectives in a largely German-focused context.
+At the same time, we recognise, the wider international growth of RSEng as a very important development
+for people undertaking software development within the research community.
+Much of our work and our discussions are also building on wider perspectives for the future of RSEng that are currently under
+investigation among RSEng communities in other countries.
+While taking a German-focused view in this paper, we therefore aim to highlight resources and materials from
+the wider international community where we feel this is relevant.
 
@@
 Additionally, researchers will seamlessly interact with infrastructures for using research software.
-Replacement of staff reaching retirement, combined with the intensified research interest in RSE, has led to the creation of more RSE professorships.
 \end{whatis}
@@
 has organisationally often adopted time scales for change that are more reminiscent
-to institutions thinking on the time scale of societal generations.
-While sounding negative, this can also be an opportunity that should be used by the RSE community.
-If in the time-frame of the next decade, a large number of permanent staff reach retirement and
-can be replaced by a new generation of researchers and technical professionals who have seen the
-digital transformation first-hand and maybe have actively participated in it through various grassroots initiatives.
-This generational change opens up the opportunity for an active change within the research landscape.
-A simple possibility would be to realign newly appointed professorships with a focus towards software development within their domain.
-A second more progressive step would be to actually appoint new cross-faculty professorships.
-On the other hand, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref}
+to institutions thinking on the time scale of generations.
+Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref}
 hints at the potential for specialisation of research tasks.
@@
 We note the SE community has also independently expressed interest in creating an \emph{infrastructure engineering} role~\cite{Sochat2024Infra}
-and funding bodies seem willing to promote such career pathways~\cite{UKRI2025Infra,EuroHPC2024Federated}.
-Taking these trends together the generational change should be used as an opportunity to reshape organisational structures to better reflect the work people in research
-are actually doing. The RSE community should take an active role in shaping that discussion in this direction till 2035.
+and funding bodies seem willing to promote such career pathways~\cite{UKRI2025Infra,EuroHPC2024Federated},
 % European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories

### 2025-09-25 future_challenges_for_RSEng.tex
Add german to the workshop participants.

@@
 Gerasimos Chourdakis\authorOrcid{0000-0002-3977-1385}\authorRef{3},
+\texorpdfstring{\\}{} % line break
 Simon Christ\authorOrcid{0000-0002-5866-1472}\authorRef{4},
-\texorpdfstring{\\}{} % line break
 Jeremy Cohen\authorOrcid{0000-0003-4312-2537}\authorRef{5},
 Nicholas A. Del Grosso\thinspace\authorRef{6},
+\texorpdfstring{\\}{} % line break
 Jean-Noël Grad\authorOrcid{0000-0002-5821-4912}\authorRef{7},
-\texorpdfstring{\\}{} % line break
 Magnus Hagdorn\authorOrcid{0000-0002-5076-4864}\authorRef{8},
 Toby Hodges\authorOrcid{0000-0003-1766-456X}\authorRef{9},
-Jan Linxweiler\authorOrcid{0000-0002-2755-5087}\authorRef{10}\textsuperscript{,}\authorRef{11},
+Jan Linxweiler\authorOrcid{0000-0002-2755-5087}\authorRef{10},
 \texorpdfstring{\\}{} % line break
-Frank Löffler\authorOrcid{0000-0001-6643-6323}\authorRef{12}\textsuperscript{,}\authorRef{11},
-Michele Martone\authorOrcid{0000-0003-3239-8554}\authorRef{13},
-Jan Philipp Thiele\authorOrcid{0000-0002-8901-6660}\authorRef{10}\textsuperscript{,}\authorRef{14}
+Frank Löffler\authorOrcid{0000-0001-6643-6323}\authorRef{11},
+Michele Martone\authorOrcid{0000-0003-3239-8554}\authorRef{12},
+Jan Philipp Thiele\authorOrcid{0000-0002-8901-6660}\authorRef{10}\textsuperscript{,}\authorRef{13}
 } % Authors and references to addresses
@@
 \autlabel{10} Technische Universität Braunschweig, Germany\\
-\autlabel{11} de-RSE e.V.---Society for Research Software in Germany\\
-\autlabel{12} Competence Center Digital Research, Friedrich Schiller University Jena, Germany\\
-\autlabel{13} Independent researcher, Germany\\
-\autlabel{14} Weierstrass Institute, Berlin, Germany;
+\autlabel{11} Michael Stifel Center Jena {\&} Friedrich Schiller University Jena, Germany\\
+\autlabel{12} Independent researcher, Germany\\
+\autlabel{13} Weierstrass Institute, Berlin, Germany;
               Leibniz University Hannover, Institute of Applied Mathematics, Scientific Computing, Hannover, Germany\\
@@
 was only published a couple of years ago~\cite{Anzt2021},
-while the foundational competencies of an RSE have only recently been defined in detail~\cite{Goth2024}.
+while the foundational competencies of a research software engineer have only recently been defined in detail~\cite{Goth2024}.
 Central RSE units are in the process of being imagined~\cite{Kempf2025-draft},
@@
 but no list or profile of the participants is available.
-The workshop hosts felt, that the majority of participants were from the RSE community.
+The workshop hosts felt, that the majority of participants were from the German RSE community.
 Nevertheless, by disseminating the discussed ideas,
 we aim to engage the community in creating a future worth living
-and to advance the conversation about how RSEng should, or could, develop.
+and to advance the conversation about how Research Software Engineering should, or could, develop.
 Since this paper focuses on the outputs of the deRSE25 conference, we present our thoughts and perspectives in a largely German-focused context.
@@
 
+The specialist skills that RSEs provide take a significant amount of time to develop and maintain.
+Therefore, we do not expect domain researchers to also become experts in a range of
+advanced RSEng topics.
+
 At the same time, we do recognise that the application of core software development best practices
-to enhance and address problems with research codes, that can often make up the bulk of current RSE work, will change.
-As noted above, these are skills that we would expect any researcher who writes code to have in 10 years time.
+to enhance and address problems with research codebases, that can often make up the bulk of
+current RSE work, will change. As noted above, these are skills that we
+would expect any researcher who writes code to have in 10 years time. 
 The rapid development of artificial intelligence (AI) and large language models
@@
 Further risks that were later discussed include an intensified replication crisis,
-making research less accessible, and loss of trust in science.
+a de-democratization of research, and loss of trust in science.
 Due to the non-deterministic nature of generative AI, we expect AI-assisted simulations
@@
 
-\subsection{Transformations in Academic Institutions}
+\subsection{How have academic Institutions transformed by then?}
 \begin{whatis}{}{academic}
-Research software is a central part of scientific practice, on the same level as data and publications.
+We expect research software to be a central part of scientific practice, on the same level as data and publications.
 This will be supported by the first generations of RSE Master's graduates that find their way into research departments, and accelerate the pace of science.
-Similar to RSEs, who are widely known and understood by university boards in 2035, academic staff members at any level are aware of the concept of RSE units.
-Universities actively use the presence of an RSE Department as a means to differentiate themselves from other universities.
-Some central RSE units will have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
-Specialised clusters have formed, which pool the skills of certain specialised RSEs with certain RSE units specialising in certain tasks like data management, HPC, or visualisation.
-At universities which create such RSE units, it can be observed that researchers naturally accept them as part of their available services.
+Similar to RSEs, who are widely known and understood by university boards in 2035, we expect the concept of RSE units to be in the minds
+of academic staff at any level.
+We hope to see that universities actively use the presence of an RSE Department as a means to differentiate themselves from other universities.
+We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
+We also hope that specialised clusters form, which pool the skills of certain specialised RSEs with certain RSE units specializing in certain tasks like data management, HPC, or visualisation.
+At universities which get such RSE units, we want to see that researchers naturally accept them as part of their available services.
 Additionally, researchers will seamlessly interact with infrastructures for using research software.
@@
 This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.
-Some specialisation trends have already been observed in recent years,
+We have already seen some specialisation trends in recent years,
 such as the HPC-RSE and AI-RSE, which are quite recent developments,
@@
 \begin{whatis}{}{career-paths}
-As the RSE becomes established, we are also seeing the emergence of a wider range of more specialised, technical roles within the research community, i.e. other (digital) RTPs.
-The legal requirements for employing people have become flexible enough to allow for the employment of more RTPs.
-Work outside of the publication record is more readily accepted and enables researchers to build careers at the interface of academia, industry, and policy
-(\eg\ in think tanks, NGOs, or organisations that do knowledge transfer).
+We would like to see the emergence of more dedicated staff, i.e. other (digital) RTPs next to the RSE.
+Also we hope that researchers can build careers at the interface of academia, industry, and policy (\eg\ in think tanks, NGOs,
+or organisations that do knowledge transfer).
 \end{whatis}
@@
 and potentially request and coordinate support from external RSEs.
-Within Germany at least, the distinction around whether these positions are classed as research staff or technical staff has so far created some discussion.
-The authors hope that, 10 years from now, this discussion has resolved towards a solution that enables the employment of RTPs more flexibly,
-and thereby recognising that the distinction between research staff and technical staff is too restrictive.
-A similar discussion is currently ongoing in the UK with initiatives like \emph{the Technician's Commitment}~\cite{techniciancommitment}, and \emph{the Hidden Ref}~\cite{hiddenref},
-and finding a resolution with having new job titles, pay grades, and career options according to the level of work that you do.
-The authors' expect these initiatives to spread and find a wider resonance over the world till 2035.
 
@@
 enabling easier reuse of prior software outputs in new research.
-We also hope that a cultural change is taking place to recognise research software
+We also hope that a cultural change is taking place to recognize research software
 as an integral part of science that it is widely recognised and rewarded across
 academia and is supported by political actors.
-
 \section{Conclusion}
-We have tried to elaborate on the state of the future ten years from now.
+We have tried to elaborate on the state on the future ten years from now.
 It turned out to be a difficult endeavour, since even the full breadth of today's trends is not yet foreseeable.

### 2025-09-25 correspondence/rebuttal-1.md
bring in yesterday's discussions
feat(#44): fix pronouns
Added some intro text to highlight the German context of the paper (#67)
* Added some intro text to highlight the German context of the paper.

* Update rebuttal notes.

* Add german to the workshop participants.

---------

Co-authored-by: Florian Goth <florian.goth@uni-wuerzburg.de>

--- 2025-09-23 correspondence/rebuttal-1.md
@@
 I suggest either making it clear that the vision and discussion are limited to Germany, or discussing why the focus of the paper is rather nationalistic yet applies to a broader range of locations, and explaining where and why the stated aspects differ.
+
+Thank you for highlighting this.
+We have added some text to the second paragraph of the introduction highlighting that with the paper focusing on the outputs of the German RSE conference, the thoughts and perspectives presented are largely German-focused.
+However, we also note that this work has broader relevance and we aim to highlight resources from the wider community where relevant.
 

### 2025-09-25 future_challenges_for_RSEng.tex
Added some intro text to highlight the German context of the paper (#67)
* Added some intro text to highlight the German context of the paper.

* Update rebuttal notes.

* Add german to the workshop participants.

---------

Co-authored-by: Florian Goth <florian.goth@uni-wuerzburg.de>
Update future_challenges_for_RSEng.tex
Co-authored-by: Gerasimos Chourdakis <gerasimos.chourdakis@ipvs.uni-stuttgart.de>

--- 2025-09-25 future_challenges_for_RSEng.tex
@@
 with increasing integration into various aspects of research.
-In Germany, a formal call for creating a sustainable environment for research software
-was only published a couple of years ago~\cite{Anzt2021},
-while the foundational competencies of an RSE have only recently been defined in detail~\cite{Goth2024}.
+In Germany, a formal call for creating a sustainable environment for research software has only been raised by the RSE community a couple of years ago~\cite{Anzt2021},
+while the foundational competencies of an RSE have just recently been defined in detail~\cite{Goth2024}.
 Central RSE units are in the process of being imagined~\cite{Kempf2025-draft},
@@
 and to advance the conversation about how RSEng should, or could, develop.
-Since this paper focuses on the outputs of the deRSE25 conference, we present our thoughts and perspectives in a largely German-focused context.
-At the same time, we recognise, the wider international growth of RSEng as a very important development
-for people undertaking software development within the research community.
-Much of our work and our discussions are also building on wider perspectives for the future of RSEng that are currently under
-investigation among RSEng communities in other countries.
-While taking a German-focused view in this paper, we therefore aim to highlight resources and materials from
-the wider international community where we feel this is relevant.
 
@@
 \begin{story}{The Definition of RSEs}{defrse}
-Kim and Kay chuckled when they looked at their original job descriptions from 2020: ``build tools for scientists, manage data, write reusable code.''
+Kim and Kay chuckle when they look at their original job descriptions from 2020: ``build tools for scientists, manage data, write reusable code.''
 Now, their roles are unrecognisable: half AI whisperer, half research ethicist, and part-time diplomat negotiating between synthetic and human collaborators.

### 2025-09-25 future_challenges_for_RSEng.tex
Update future_challenges_for_RSEng.tex
Co-authored-by: Gerasimos Chourdakis <gerasimos.chourdakis@ipvs.uni-stuttgart.de>
fixing pronouns and tenses (#69)
* feat(#44): fix pronouns

* feat(#44): fix tense

* swap words

* split long sentence

* replace word

* drop paragraph as it is already a few lines above

* reduce the number of certains

* Update future_challenges_for_RSEng.tex

Co-authored-by: Gerasimos Chourdakis <gerasimos.chourdakis@ipvs.uni-stuttgart.de>

@@
 but no list or profile of the participants is available.
-The workshop hosts felt, that the majority of participants were from the RSE community.
+The workshop hosts felt, that the majority of participants were from the German RSE community.
 Nevertheless, by disseminating the discussed ideas,
@@
 Kay had left academia to work in an industrial R\&D department because of a more reliable employment.
-There she gained valuable experience working in an industrial setting.
+There, she gained valuable experience working in an industrial setting.
 However, she was very happy when an opportunity arose to work in a central

### 2025-09-25 future_challenges_for_RSEng.tex

@@
 so far future projections are far less common for the RSEng domain.
-Here, we present the results of a workshop discussion (notes content is archived here \cite{goth2025_WSPad}) at the
+
+Here, we present the results of a workshop discussion (notes are archived here \cite{goth2025_WSPad}) at the
 deRSE25 conference in Karlsruhe\footnote{deRSE25, session ``The End of {RSEng}?
@@
 By reducing the complexity of individual software components,
-we are now able to develop systems of complexity and capability not previously possible.
-This divide-and-conquer approach allows the scientific community to address questions not previously within their grasp.
-When it comes to teaching, it is true that students need to learn how to handle complexity.
+we are now able to develop systems of complexity and capability hitherto not possible.
+This divide-and-conquer approach allows the scientific community to address ever larger and more complex problems.
+When it comes to teaching, it is true that students need to learn how to handle this complexity.
 They need to develop a deep understanding of the individual components.
@@
 Kay had left academia to work in an industrial R\&D department because of a more reliable employment.
-There, she gained valuable experience working in an industrial setting.
+There she gained valuable experience working in an industrial setting.
 However, she was very happy when an opportunity arose to work in a central

### 2025-09-25 future_challenges_for_RSEng.tex
MHagdorn -  yet another read through (#71)
* feat(#44): fix pronouns

* feat(#44): fix tense

* feat(#44): fix tense

* split sentence

* swap words

* split long sentence

* replace word

* drop paragraph as it is already a few lines above

* reduce the number of certains

* fixes(#56): sort out incomplete sentence

* fixes(#56): rename goal boxes

* fixes(#56): swap sentences around to make it clearer that now is 2025

* minor fix

* less hope

* be more assertive

* make it clearer that the workshop discussions are meant

* reword and remove redundancy

* insert word

* insert comma

* drop

--- 2025-09-25 future_challenges_for_RSEng.tex
@@
 but no list or profile of the participants is available.
-The workshop hosts felt, that the majority of participants were from the German RSE community.
+The workshop hosts felt, that the majority of participants were from the RSE community.
 Nevertheless, by disseminating the discussed ideas,
@@
 \begin{story}{The Definition of RSEs}{defrse}
-Kim and Kay chuckle when they look at their original job descriptions from 2020: ``build tools for scientists, manage data, write reusable code.''
+Kim and Kay chuckled when they looked at their original job descriptions from 2020: ``build tools for scientists, manage data, write reusable code.''
 Now, their roles are unrecognisable: half AI whisperer, half research ethicist, and part-time diplomat negotiating between synthetic and human collaborators.
@@
 Kay had left academia to work in an industrial R\&D department because of a more reliable employment.
-There, she gained valuable experience working in an industrial setting.
-However, she was very happy when an opportunity arose to work in a central
-RSE department with an open-ended contract. Apart from her main responsibility
+There he gained valuable experience working in an industrial setting.
+However, he was very happy when an opportunity arose to work in a central
+RSE department with an open-ended contract. Apart from his main responsibility
 of developing and maintaining software for an affordable water monitoring system
-used in the global south, she now also teaches undergraduates digital literacy.
+used in the global south, he now also teaches undergraduates digital literacy.
 \end{story}
@@
 has organisationally often adopted time scales for change that are more reminiscent
-to institutions thinking on the time scale of generations.
-Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref}
+to institutions thinking on the time scale of societal generations.
+The RSE community should plan ahead and use opportunities on that horizon.
+If in the time-frame of the next decade, a large number of permanent staff reach retirement and
+can be replaced by a new generation of researchers and technical professionals who have seen the
+digital transformation first-hand and maybe have actively participated in it through various grassroots initiatives.
+This generational change opens up the opportunity for an active change within the research landscape.
+A simple possibility would be to realign newly appointed professorships with a focus towards software development within their domain.
+A second more progressive step would be to actually appoint new cross-faculty professorships.
+On the other hand, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref}
 hints at the potential for specialisation of research tasks.

### 2025-09-25 future_challenges_for_RSEng.tex
Update future_challenges_for_RSEng.tex
Co-authored-by: Gerasimos Chourdakis <gerasimos.chourdakis@ipvs.uni-stuttgart.de>
address comment

@@
 A simple possibility would be to realign newly appointed professorships with a focus towards software development within their domain.
-A second more progressive step would be to actually appoint new cross-faculty professorships.
-On the other hand, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref}
-hints at the potential for specialisation of research tasks.
+A second, more progressive step would be to actually appoint new cross-faculty professorships.
+Already today the presence, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref} hints at a potential for specialisation of research tasks.
 This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.

### 2025-09-25 future_challenges_for_RSEng.tex
address comment by Magnus and Makish

@@
 to institutions thinking on the time scale of societal generations.
-The RSE community should plan ahead and use opportunities on that horizon.
-If in the time-frame of the next decade, a large number of permanent staff reach retirement and
-can be replaced by a new generation of researchers and technical professionals who have seen the
-digital transformation first-hand and maybe have actively participated in it through various grassroots initiatives.
+During the next decade many positions at universities and research institutions will vacate as the generation of the baby boomers come to retirement age.
+The RSE community should plan ahead and use this opportunity to fill some of these positions with RSEs and other RTPs.
 This generational change opens up the opportunity for an active change within the research landscape.

### 2025-09-25 correspondence/rebuttal-1.md

@@
 
-> Chaper 2.7
+Thanks for bringing this up. We have formulated that aspect towards a call for action without  sounding too Germany specific.
+
+> Chapter 2.7
 > "in lower tiers of an of a research organisation", sentence needs attention
@@
 
+Very true, thanks for pointing this out. We have fixed it. Kim and Kay are sisters.
+
 ### Suggestions

### 2025-09-25 future_challenges_for_RSEng.tex

@@
 so far future projections are far less common for the RSEng domain.
-We present here the results of a workshop discussion (notes content is archived here \cite{goth2025_WSPad}) at the
+
+Here, we present the results of a workshop discussion (notes are archived here \cite{goth2025_WSPad}) at the
 deRSE25 conference in Karlsruhe\footnote{deRSE25, session ``The End of {RSEng}?
@@
 \begin{whatis}{}{whatisexplanation}
-In order to list what has been achieved in a certain aspect within the RSE community of the future of 2035 we use these ``Goal'' boxes.
+In order to list what has been achieved in a certain aspect within the RSE community of the future of 2035 we use these ``RSEng in 2035'' boxes.
 \end{whatis}
 
-Looking ahead over a 10-year time frame, we hope that the RSE community will have
+Looking ahead over a 10-year time frame, we expect that the RSE community will have
 contributed to making research software and its outputs
 more sustainable, robust and maintainable.
-We also hope that research software becomes an accepted research artefact on an equal footing with data publications and traditional text publications.
+We hope that research software will have become an accepted research artefact on an equal footing with data publications and traditional text publications.
 This is already happening in a number of ways,
@@
 and UK SSI resources hub\footnote{\url{https://www.software.ac.uk/resource-hub}}.
-The Carpentries now offer specialised software workshops
-for data scientists\footnote{\url{https://datacarpentry.org}},
-librarians\footnote{\url{https://librarycarpentry.org}},
-and soon High-Performance Computing (HPC) practitioners\footnote{\url{https://www.hpc-carpentry.org}}.
 
 \begin{story}{The Definition of RSEs}{defrse}
-Kim and Kay chuckled when they looked at their original job descriptions from 2020: ``build tools for scientists, manage data, write reusable code.''
+Kim and Kay chuckle when they look at their original job descriptions from 2020: ``build tools for scientists, manage data, write reusable code.''
 Now, their roles are unrecognisable: half AI whisperer, half research ethicist, and part-time diplomat negotiating between synthetic and human collaborators.
@@
 By reducing the complexity of individual software components,
-we are now able to develop systems of complexity and capability not previously possible.
-This divide-and-conquer approach allows the scientific community to address questions not previously within their grasp.
-When it comes to teaching, it is true that students need to learn how to handle complexity,
-by being able to develop deep understanding of individual components.
+we are now able to develop systems of complexity and capability hitherto not possible.
+This divide-and-conquer approach allows the scientific community to address ever larger and more complex problems.
+When it comes to teaching, it is true that students need to learn how to handle this complexity.
+They need to develop a deep understanding of the individual components.
 However, it is also important to learn how to manage complexity
@@
 Research software is a central part of scientific practice, on the same level as data and publications.
-This will be supported by the first generations of RSE Master's graduates that find their way into research departments, and accelerate the pace of science.
+This is supported by the first generations of RSE Master's graduates that find their way into research departments, and accelerate the pace of science.
 Similar to RSEs, who are widely known and understood by university boards in 2035, academic staff members at any level are aware of the concept of RSE units.
 Universities actively use the presence of an RSE Department as a means to differentiate themselves from other universities.
-Some central RSE units will have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
-Specialised clusters have formed, which pool the skills of certain specialised RSEs with certain RSE units specialising in certain tasks like data management, HPC, or visualisation.
+Some central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
+Specialised clusters have formed, which pool the skills of specialised RSEs with RSE units specialising in certain tasks like data management, HPC, or visualisation.
 At universities which create such RSE units, it can be observed that researchers naturally accept them as part of their available services.
-Additionally, researchers will seamlessly interact with infrastructures for using research software.
+Additionally, researchers interact seamlessly with infrastructures for using research software.
 Replacement of staff reaching retirement, combined with the intensified research interest in RSE, has led to the creation of more RSE professorships.
@@
 This is an opportunity that would not be possible if relying solely on short-term funded, isolated RSEs.
-We hope that a change of culture in research about the acceptance of research software has happened.
+We hope that research institutions will have recognised the importance of research software and the need to support the role of RSEs.
 

### 2025-09-26 future_challenges_for_RSEng.tex
a try at keywords

@@
 } % Abstract of the article
-\keywords{Future RSE, Evolving field, Vision} % Keywords for the article
+\keywords{Future RSEng, trends in RSEng, Educational Advancements, emergent academic jobs, academic reorganisation} % Keywords for the article
 
@@
 Additionally, researchers interact seamlessly with infrastructures for using research software.
-Replacement of staff reaching retirement, combined with the intensified research interest in RSE, has led to the creation of more RSE professorships.
 \end{whatis}
@@
 has organisationally often adopted time scales for change that are more reminiscent
-to institutions thinking on the time scale of societal generations.
-During the next decade many positions at universities and research institutions will vacate as the generation of the baby boomers come to retirement age.
-The RSE community should plan ahead and use this opportunity to fill some of these positions with RSEs and other RTPs.
-This generational change opens up the opportunity for an active change within the research landscape.
-A simple possibility would be to realign newly appointed professorships with a focus towards software development within their domain.
-A second, more progressive step would be to actually appoint new cross-faculty professorships.
-Already today the presence, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref} hints at a potential for specialisation of research tasks.
+to institutions thinking on the time scale of generations.
+Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref}
+hints at the potential for specialisation of research tasks.
 This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.

### 2025-09-26 future_challenges_for_RSEng.tex
write something how the workshop was held and where the questions are from

@@
 } % Abstract of the article
-\keywords{Future RSEng, trends in RSEng, Educational Advancements, emergent academic jobs, academic reorganisation} % Keywords for the article
+\keywords{Future RSE, Evolving field, Vision} % Keywords for the article
 
@@
 The material presented here is based on the output of the workshop participants, but extended with the authors' ideas.
-Since deRSE25 was co-located with SE25, the annual conference for the Software Engineering community in Germany, the discussion was also open to software engineers,
-but no list or profile of the participants is available.
+Since deRSE25 was co-located with SE25, the annual conference for the Software Engineering community in Germany, the discussion was also open to software engineers, but no list or profile of the participants is available.
 The workshop hosts felt, that the majority of participants were from the German RSE community.
+
+The workshop was held as an open discussion of 45 minutes between the participants.
+As can be seen from the workshop material~\cite{goth2025_WSPad}, we prepared some slides to steer the discussion
+towards the timeframe of 2035.
+For a start we prepared questions that we openly asked to the audience,
+while also allowing the participants to meander towards questions of their interest.
+The prepared questions stem from internal discussions among the workshop hosts.
+It started from discussions whether the recently defined competencies of RSEs~\cite{Goth2024}
+are just snapshots of our time, or whether they can constitute something that stands the test of time.
+Then we asked ourselves how the community of RSEs would interact with some emerging trends.
+
 Nevertheless, by disseminating the discussed ideas,
@@
 It is already difficult to structure the spectrum of RSE tasks and responsibilities
-today given the spectrum of research software~\cite{Hasselbring2024}.
+today, given the spectrum of research software~\cite{Hasselbring2024}.
 It is even more challenging to structure the aspects that will become relevant for the RSE community in the future.
-In order to structure the discussion, we decided to consider
+In order to structure the workshop discussion for this publication, we decided to consider
 the evolution of the term RSE and its definition,
@@
 and new groups of RTPs (\emph{Research Technical Professional}s).
+This roughly mirrors the questions asked to the audience, where the entire workshop content is condensed into these topics.
+
 This work supplements other opinion pieces developed in parallel

### 2025-09-26 correspondence/rebuttal-1.md
rebuttal
write something how the workshop was held and where the questions are from
A bit of tidy up in the rebuttal for reviewer B vs. C

@@
 
-Thanks for bringing this up. We have formulated that aspect towards a call for action without  sounding too Germany specific.
-
 > Chapter 2.7
@@
 
+\newpage
+
+## Reviewer C
+
+### Summary
+
+> A useful overview of outcomes of a valuable workshop, augmented by the author's own opinions, on the future of RSE.
+
+### Relevance
+
+> This paper reflects on a workshop held at de-RSE
+
+### Significance
+
+> This paper addresses many important themes in this field, and makes useful contributions on how they could be advanced.
+
+### Technical soundness
+
+> This is an opinion paper, based on synthesis of the workshop discussions and the authors' knowledge.
+
+### Quality of presentation
+
+> The use of boxes is a great idea, but doesn't quite work. See the detailed notes for ideas on how to improve this.
+
+### Suggestions
+
+> It might be useful to note that this paper on RSE in 2030 exists https://arxiv.org/pdf/2308.07796, although the workshop participants didn’t know about it so it wasn’t incorporated in their thinking.
+
+> Reviewer C also uploaded a file with more feedback
+
+> Page 2
+> In order to list what has been achieved in a certain aspect within the RSE community of the future of 2035 we use these “Goal” boxes.
+– The use of the term “Achieved Goals” is a bit confusing.
+I understand it’s meant that these would have been achieved by 2035, but then the text under these boxes talks about 2035 as a future state.
+I’d suggest rewording these boxes to “2035 Goals’ or similar.
+> Page 3
+> Story 1: In order to give a more concrete setting, imagine a world ten years from now.
+Not too overwhelmingly far ahead in the future, but distant enough for some changes to have occurred.
+– These are both incomplete sentences, ie it's a group of words that doesn't express a complete thought and is missing a necessary element, like a subject or a verb.
+> Page 4
+> RSEng-relevant open educational resources and Master’s programmes are tracked in the Learning and Teaching RSE database2 and UK SSI resources hub3.
+The Carpentries now offer specialised software workshops for data scientists4, librarians5, and soon HPC practitioners6.
+– This seems to be about the future state, but it’s not clear what “now” means here (probably 2035?) – for example, a few paragraphs earlier there is wording “This means, that today’s workshop content of proper versioning and data management could become obsolete by using more intuitive tools.” – where today means 2025.
+Can it be made clearer throughout the paper when the text is talking about 2025 versus 2035?
+
 > Page 5
@@
 Not sure, how this crept into the final version... It's gone now.
-
-\newpage
-
-## Reviewer C
-
-### Summary
-
-> A useful overview of outcomes of a valuable workshop, augmented by the author's own opinions, on the future of RSE.
-
-### Relevance
-
-> This paper reflects on a workshop held at de-RSE
-
-### Significance
-
-> This paper addresses many important themes in this field, and makes useful contributions on how they could be advanced.
-
-### Technical soundness
-
-> This is an opinion paper, based on synthesis of the workshop discussions and the authors' knowledge.
-
-### Quality of presentation
-
-> The use of boxes is a great idea, but doesn't quite work. See the detailed notes for ideas on how to improve this.
-
-### Suggestions
-
-> It might be useful to note that this paper on RSE in 2030 exists https://arxiv.org/pdf/2308.07796, although the workshop participants didn’t know about it so it wasn’t incorporated in their thinking.
-
-> Reviewer C also uploaded a file with more feedback
 

### 2025-09-26 correspondence/rebuttal-1.md
provide answers to some reviewer feedback

@@
 Future trend statements should be based on knowledge.
+
+Thanks for spotting. We have used stronger language in the revision. We have also passed over the document to make the tenses homogeneous.
 
@@
 I’d suggest rewording these boxes to “2035 Goals’ or similar.
+
+We have reworded them to RSEng in 2035.
+
 > Page 3
@@
 – These are both incomplete sentences, ie it's a group of words that doesn't express a complete thought and is missing a necessary element, like a subject or a verb.
+
+We have reworded this sentence in order to provide more natural british english.
+
 > Page 4
@@
 Can it be made clearer throughout the paper when the text is talking about 2025 versus 2035?
+
+Since the inconsistent use tof times was mentioned by multiple reviewers, we went over the paper multiple times to be more consistent with regard to the use of tenses.
 

### 2025-09-26 future_challenges_for_RSEng.tex
fix typo
provide answers to some reviewer feedback

--- 2025-09-26 future_challenges_for_RSEng.tex
@@
 \begin{story}{Introduction}{intro}
-In order to give a more concrete setting, imagine a world ten years from now 
-which is not too overwhelmingly far ahead in the future, but distant enough for some changes to have occurred.
+To provide a more concrete setting, imagine a world ten years from now --
+not too overwhelmingly far into the future, but distant enough for some changes to have occurred.
 Over this time frame Kim~\cite{Anzt2021} and Kay~\cite{Goth2024},
@@
 At universities which create such RSE units, it can be observed that researchers naturally accept them as part of their available services.
-Additionally, researchers interact seamlessly with infrastructures for using research software.
+Additionally7, researchers interact seamlessly with infrastructures for using research software.
 \end{whatis}

### 2025-09-26 future_challenges_for_RSEng.tex

@@
 At universities which create such RSE units, it can be observed that researchers naturally accept them as part of their available services.
-Additionally7, researchers interact seamlessly with infrastructures for using research software.
+Additionally, researchers interact seamlessly with infrastructures for using research software.
+Replacement of staff reaching retirement, combined with the intensified research interest in RSE, has led to the creation of more RSE professorships.
 \end{whatis}
@@
 has organisationally often adopted time scales for change that are more reminiscent
-to institutions thinking on the time scale of generations.
-Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref}
-hints at the potential for specialisation of research tasks.
+to institutions thinking on the time scale of societal generations.
+During the next decade many positions at universities and research institutions will vacate as the generation of the baby boomers come to retirement age.
+The RSE community should plan ahead and use this opportunity to fill some of these positions with RSEs and other RTPs.
+This generational change opens up the opportunity for an active change within the research landscape.
+A simple possibility would be to realign newly appointed professorships with a focus towards software development within their domain.
+A second, more progressive step would be to actually appoint new cross-faculty professorships.
+Already today the presence, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref} hints at a potential for specialisation of research tasks.
 This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.

### 2025-09-26 future_challenges_for_RSEng.tex
typo

--- 2025-09-26 future_challenges_for_RSEng.tex
@@
 At universities which create such RSE units, it can be observed that researchers naturally accept them as part of their available services.
-Additionally7, researchers interact seamlessly with infrastructures for using research software.
+Additionally, researchers interact seamlessly with infrastructures for using research software.
 \end{whatis}

### 2025-09-26 future_challenges_for_RSEng.tex

--- 2025-09-26 future_challenges_for_RSEng.tex
@@
 The material presented here is based on the output of the workshop participants, but extended with the authors' ideas.
-Since deRSE25 was co-located with SE25, the annual conference for the Software Engineering community in Germany, the discussion was also open to software engineers,
-but no list or profile of the participants is available.
+Since deRSE25 was co-located with SE25, the annual conference for the Software Engineering community in Germany, the discussion was also open to software engineers, but no list or profile of the participants is available.
 The workshop hosts felt, that the majority of participants were from the German RSE community.
+
+The workshop was held as an open discussion of 45 minutes between the participants.
+As can be seen from the workshop material~\cite{goth2025_WSPad}, we prepared some slides to steer the discussion
+towards the time frame of 2035.
+For a start we prepared questions that we openly asked to the audience,
+while also allowing the participants to meander towards questions of their interest.
+The prepared questions stem from internal discussions among the workshop hosts.
+It started from discussions whether the recently defined competencies of RSEs~\cite{Goth2024}
+are just snapshots of our time, or whether they can constitute something that stands the test of time.
+Then we asked ourselves how the community of RSEs would interact with some emerging trends.
+
 Nevertheless, by disseminating the discussed ideas,

### 2025-09-26 future_challenges_for_RSEng.tex
rewording

--- 2025-09-26 future_challenges_for_RSEng.tex
@@
 
-At the same time, acquiring people with a certain specialised skill set
-opens up possibilities for pooling.
+While every faculty could hire the specialists they require, academia would be better
+served if they get pooled into dedicated units in order to make them available for the entire university.
 Universities will benefit from pooling these specialists into RSE units

### 2025-09-26 future_challenges_for_RSEng.tex
jngrads feedback

@@
 
-While every faculty could hire the specialists they require, academia would be better
-served if they get pooled into dedicated units in order to make them available for the entire university.
-Universities will benefit from pooling these specialists into RSE units
+While every faculty could hire the specialists they require,
+Universities benefit more from pooling these specialists into RSE units
 and offering their services to researchers in a structured way.

### 2025-09-28 future_challenges_for_RSEng.tex
Update future_challenges_for_RSEng.tex
Co-authored-by: Gerasimos Chourdakis <gerasimos.chourdakis@ipvs.uni-stuttgart.de>

--- 2025-09-26 future_challenges_for_RSEng.tex
@@
 } % Abstract of the article
-\keywords{Future RSE, Evolving field, Vision} % Keywords for the article
+\keywords{Future RSEng, trends in RSEng, educational advancements, emergent academic jobs, academic reorganisation} % Keywords for the article
 
@@
 
-While every faculty could hire the specialists they require, academia would be better
-served if they get pooled into dedicated units in order to make them available for the entire university.
+At the same time, acquiring people with a certain specialised skill set
+opens up possibilities for pooling.
 Universities will benefit from pooling these specialists into RSE units

### 2025-09-28 future_challenges_for_RSEng.tex
fixes
Update future_challenges_for_RSEng.tex
Co-authored-by: Magnus Hagdorn <Magnus.Hagdorn@marsupium.org>

@@
 } % Abstract of the article
-\keywords{Future RSEng, trends in RSEng, educational advancements, emergent academic jobs, academic reorganisation} % Keywords for the article
+\keywords{Future RSE, Evolving field, Vision} % Keywords for the article
 
@@
 Additionally, researchers interact seamlessly with infrastructures for using research software.
+Replacement of staff reaching retirement, combined with the intensified research interest in RSE, has led to the creation of more RSE professorships.
 \end{whatis}
@@
 has organisationally often adopted time scales for change that are more reminiscent
-to institutions thinking on the time scale of generations.
-Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref}
-hints at the potential for specialisation of research tasks.
+to institutions thinking on the time scale of societal generations.
+During the next decade many positions at universities and research institutions will be vacated as the generation of the baby boomers come to retirement age.
+The RSE community should plan ahead and use this opportunity to fill some of these positions with RSEs and other RTPs.
+This generational change opens up the opportunity for an active change within the research landscape.
+A simple possibility would be to realign newly appointed professorships with a focus towards software development within their domain.
+A second, more progressive step would be to actually appoint new cross-faculty professorships.
+Already today the presence, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref} hints at a potential for specialisation of research tasks.
 This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.

### 2025-09-28 future_challenges_for_RSEng.tex
Update future_challenges_for_RSEng.tex
Co-authored-by: Magnus Hagdorn <Magnus.Hagdorn@marsupium.org>
Update future_challenges_for_RSEng.tex
Co-authored-by: Gerasimos Chourdakis <gerasimos.chourdakis@ipvs.uni-stuttgart.de>

--- 2025-09-28 future_challenges_for_RSEng.tex
@@
 The material presented here is based on the output of the workshop participants, but extended with the authors' ideas.
-Since deRSE25 was co-located with SE25, the annual conference for the Software Engineering community in Germany, the discussion was also open to software engineers,
-but no list or profile of the participants is available.
+Since deRSE25 was co-located with SE25, the annual conference for the Software Engineering community in Germany, the discussion was also open to software engineers, but no list or profile of the participants is available.
 The workshop hosts felt, that the majority of participants were from the German RSE community.
+
+The workshop was held as an open discussion of 45 minutes between the participants.
+As can be seen from the workshop material~\cite{goth2025_WSPad}, we prepared some slides to steer the discussion
+towards the time frame of 2035.
+For a start we prepared questions that we openly asked to the audience,
+while also allowing the participants to meander towards questions of their interest.
+The prepared questions stem from internal discussions among the workshop hosts.
+It started from discussions whether the recently defined competencies of RSEs~\cite{Goth2024}
+are just snapshots of our time, or whether they can constitute something that stands the test of time.
+Then we asked ourselves how the community of RSEs would interact with some emerging trends.
+The collected material was further discussed among the authors in follow-up meetings, structuring and refining the ideas, adding new ideas as described in each of the following sections.
+
 Nevertheless, by disseminating the discussed ideas,

### 2025-09-28 future_challenges_for_RSEng.tex
a try at keywords (#73)
* a try at keywords

* rebuttal


Co-authored-by: Florian Goth <florian.goth@uni-wuerzburg.de>
Co-authored-by: Gerasimos Chourdakis <gerasimos.chourdakis@ipvs.uni-stuttgart.de>

--- 2025-09-28 future_challenges_for_RSEng.tex
@@
 } % Abstract of the article
-\keywords{Future RSEng, trends in RSEng, educational advancements, emergent academic jobs, academic reorganisation} % Keywords for the article
+\keywords{trends in RSEng, educational advancements, emergent academic jobs, academic reorganisation} % Keywords for the article
 
@@
 
-At the same time, acquiring people with a certain specialised skill set
-opens up possibilities for pooling.
-Universities will benefit from pooling these specialists into RSE units
+While every faculty could hire the specialists they require,
+Universities benefit more from pooling these specialists into RSE units
 and offering their services to researchers in a structured way.

### 2025-09-28 future_challenges_for_RSEng.tex
first start for generations (#66)
* firt start for generations

Co-authored-by: Jeremy Cohen <jcohen02@users.noreply.github.com>

* bring in yesterday's discussions

Co-authored-by: Gerasimos Chourdakis <gerasimos.chourdakis@ipvs.uni-stuttgart.de>

Co-authored-by: Magnus Hagdorn <Magnus.Hagdorn@marsupium.org>

* address comment

* address comment by Magnus and Makish

* Update future_challenges_for_RSEng.tex

Co-authored-by: Magnus Hagdorn <Magnus.Hagdorn@marsupium.org>

---------

Co-authored-by: Florian Goth <florian.goth@uni-wuerzburg.de>
Co-authored-by: Jeremy Cohen <jcohen02@users.noreply

--- 2025-09-28 future_challenges_for_RSEng.tex
@@
 } % Abstract of the article
-\keywords{Future RSE, Evolving field, Vision} % Keywords for the article
+\keywords{trends in RSEng, educational advancements, emergent academic jobs, academic reorganisation} % Keywords for the article
 
@@
 Taking these trends together the generational change should be used as an opportunity to reshape organisational structures to better reflect the work people in research
-are actually doing. The RSE community should take an active role in shaping that discussion in this direction till 2035.
+are actually doing. The RSE community should take an active role in shaping this discussion.
 % European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories

### 2025-09-28 correspondence/rebuttal.md

--- 2025-09-26 correspondence/rebuttal-1.md
@@
 
+Thanks for bringing this up. We have formulated that aspect towards a call for action without sounding too Germany specific.
+
 > Chapter 2.7
@@
 
+Thank You, we added some more keywords.
+
 > "In order to structure the discussion, we decided to consider...".
@@
 Have there been other topics in discussion that have not been namend in the paper?
+
+Thanks for bringing this up. We have clarified this, that the workshop was an open discussion with some prepared questions.
+We also explain how the questions have been rooted in some of the discussions of the teachingRSE project.
 

### 2025-09-28 correspondence/rebuttal.md
Add rebuttal build step (#76)
* Add rebuttal build step

* CI: Install pandoc

* Debugging CI

* bugfix

* final fixes to the rebuttal

---------

Co-authored-by: Florian Goth <florian.goth@uni-wuerzburg.de>

@@
 
-Reworded it to keep the ambiguity in Gender.
+Reworded it since Kim and Kay are sisters.
 
@@
 
-Since the inconsistent use tof times was mentioned by multiple reviewers, we went over the paper multiple times to be more consistent with regard to the use of tenses.
+Since the inconsistent use of times was mentioned by multiple reviewers, we went over the paper multiple times to be more consistent with regard to the use of tenses.
 
